### 2021-05-10 data/PrelimAnalysis/PreliminaryAnalysis.Rmd
update

---
title: Weather anomalies associated with wildfires in the Great Plains
subtitle: Preliminary analysis
date: '`r Sys.Date()`'
author: DAM
output:
  html_document:
    keep_md: false
    toc: true
    toc_float: true
    theme: united
    highlight: tango
---

```{r setup, echo=FALSE, warning=FALSE, message = FALSE, results='hide'}
knitr::opts_chunk$set(message = FALSE, warning=FALSE, 
               echo=FALSE, eval=TRUE) #, fig.path='figures/', dev=c('png', 'pdf'))

pacman::p_load(tidyverse, sf, lubridate, grid)
options(dplyr.summarise.inform = FALSE)
source('https://raw.githubusercontent.com/devanmcg/rangeR/master/R/CustomGGplotThemes.R')
```

```{r data_loading}
wd =  "C:/Users/devan.mcgranahan/GithubProjects/GreatPlainsFire"
# GIS layers
  load(paste0(wd, './gis/Robjects/us_sf.Rdata'))
  load(paste0(wd, '/gis/Robjects/gp_l3.Rdata'))
  load(paste0(wd, '/gis/Robjects/GP_WF.Rdata'))
  load("C:/Users/devan.mcgranahan/GithubProjects/FireWeather/mapping/data/region_states.Rdata")

# weather data 
  load(paste0(wd, '/data/HistoricalWeather/NWGP_top_20.Rdata'))
  load(paste0(wd, '/data/PrelimAnalysis/FireWeekMeans.Rdata'))
  load(paste0(wd, '/data/PrelimAnalysis/FireEventWx.Rdata'))
```

# Main take-aways

In a preliminary anaylsis of 20 wildfires in the Northern Great Plains, several weather variables showed substantial departures (*anomalies*) from the historical range of these values (*normal*) for the week in which the fires occurred:

* Vapor Pressure Deficit averaged 2.4 standard deviations above normal
* Potential Evapotranspiration averaged 2.3 standard deviations above normal
* Relative humidity averaged 1.7 standard deviations below normal 
* NFDRS products BI and ERC averaged 1.8 and 2.0 above normal, respectively

# Location of fires

The preliminary analysis is limited to the 20 largest fires in the MTBS database from the Northwestern Great Plains EPA Level III ecoregion:

```{r incident_map}
fires <- 
  NWGP_top_20 %>%
  group_by(incident) %>%
  slice(1) %>%
  st_as_sf(coords = c("lon", "lat"), 
           crs = 4326) %>%
    st_transform(26915)

inset_gg <-
  ggplot() + theme_map(14) +
    geom_sf(data = us_sf, 
            fill = "lightgrey") + 
    geom_sf(data = gp_l3,
            fill = "darkblue", color = NA) + 
    geom_sf(data = gp_l3 %>%
                    filter(US_L3NAME == "Northwestern Great Plains"), 
            fill = "lightblue")

main_gg <-
  ggplot() + theme_map(14) +
    geom_sf(data = gp_l3 %>%
                    filter(US_L3NAME == "Northwestern Great Plains"), 
            fill = "lightblue") + 
    geom_sf(data = region_states, fill = NA) +
    geom_sf(data = fires, 
            pch=21, 
            size = 4, stroke = 1.25,
            fill = "darkblue", 
            color = 'white')

grid.newpage() 
main <- viewport(width = 1, height = 1, x = 0.5, y = 0.5) 
inset <- viewport(width = 0.5, height = 0.5, x = 0.2, y = 0.25)
print(main_gg, vp = main)
print(inset_gg, vp = inset)
# dev.off() 
```

# Day-of conditions relative to 'normal'

## Methods

1. Determine centroid of each perimeter
2. Define a "weather week" as three days prior to ignition and three days after
3. Fetch historical weather data for the week, 1980-2020, from gridMET data
4. Calculate mean for each variable for each annual weather week 
5. Plot day-of conditions relative to distribution of historical values (1980-2020)

## Results

In most of the 20 fires, several variables had consistent relationships relative to 'normal' values:

* Vapor Pressure Deficit (VPD) appeared the highest on the day of ignition relative to median values
* Aside from 5 fires with apparently no Burn Index data, BI tended to be higher than normal on days of ignition

```{r incident_histories, fig.height = 12}
ggplot() + theme_bw(16) +
  # geom_violin(data = FireWeekMeans %>%
  #                     filter(param == "vpd"),  
  #             aes(x = incident, 
  #                 y = Mean), 
  #             draw_quantiles = c(0.15, 0.50, 0.85), 
  #             fill = "lightblue", alpha = 0.5) +
  geom_boxplot(data = FireWeekMeans, 
               aes(x = incident,
                   y = Mean),
               outlier.shape = NA, 
               fill = "lightblue", alpha = 0.5) +
  geom_point(data = FireEventWx, 
             aes(x = incident, 
                 y = value), 
             pch = 21, fill = "blue", 
             col = "white", stroke = 1.1, size = 4) +
  labs(y = '') +
  coord_flip() +
  facet_wrap(~param, scales = 'free') + 
  theme(panel.grid.major.y = element_blank(), 
        axis.text.y = element_blank())
```

Variables include

| gridMET code     | Weather variable                        |
|:-----------------|:----------------------------------------| 
| `burn_index`     | Fire behavior (NFRDS)                   |
| `energy_release` | Energy Release Component (NFRDS)        |
| `pet_grass`      | Potential Evapotranspiration            |
| `prcp`           | Daily rainfall                          |
| `rhmin`          | Lowest RH for the day                   |
| `shum`           | Daily mean specific humidity            |
| `srad`           | Daily mean surface shortwave radiation  |
| `tmax`           | Daily maximum air temperature (dry bulb)|
| `vpd`            | Vapor Pressure Deficit                  |
| `wind_vel`       | Daily mean wind speed                   |

# Trends in anomalies

Here I've calculated the anomaly for each variable as

$$ \frac{'Day~of'~value - median}{standard~deviation} $$

This graph summarizes departures of day-of-ignition values from median values, with departure expressed as standard deviations away from the median value.  
The data combine the median and 95% confidence interval of departures for all 20 fires.

```{r incident_anomalies}
FireWeekMeans %>%
  group_by(incident, param) %>%
  summarize(Median = median(Mean, na.rm = TRUE), 
            SD = sd(Mean, na.rm = TRUE)) %>%
  ungroup() %>%
  full_join(select(FireEventWx, date, incident, param, value)) %>%
  mutate(value = case_when(
    param == "burn_index" & value <= 0.001 ~ NA_real_, 
    TRUE ~ value
  )) %>%
  mutate(anomaly = (value - Median) / SD ) %>%
  group_by(param) %>%
  summarize(lwr = quantile(anomaly, probs = 0.025, na.rm = TRUE), 
            median = quantile(anomaly, 0.50, na.rm = TRUE), 
            upr = quantile(anomaly, 0.975, na.rm = TRUE)) %>%
  ggplot() + theme_bw(14) + 
  geom_vline(xintercept = 0) + 
  geom_errorbarh(aes(y = param,
                     xmin = lwr, 
                     xmax = upr), 
                 color = "darkblue", 
                 height = 0.25, size = 1) +
  geom_point(aes(x = median, 
                 y = param), 
             pch = 21, size = 4, stroke = 1.1,
             col = "darkblue", fill = 'lightblue') + 
  labs(y = "Weather parameter", 
       x = "Standard deviations from median") + 
  theme(panel.grid.major.y = element_blank())
```

This can be interpreted as the median vapor pressure deficit for these 20 fires was 2.5 standard deviations above normal for the week in which the fire occurred. 
The variable with the next greatest anomaly was PET. 
RH had less of an anomaly than might be expected. 
Of the two NFRDS products included here, BI had a consistently high anomaly and ERC was not signficantly different than normal.

# Next steps

* Expand geographic scope to entire Great Plains region (Dark blue in map)
* Statistical models to identify variable or combination of variables with most consistent anomalies
* Identify seasonal patterns in anomalies 
* Identify spatial and temporal changes in the seasonality, frequency, magnitude, and duration of anomalies

### 2021-05-10 data/PrelimAnalysis/SecondaryAnalysis.Rmd
update

--- 2021-05-10 data/PrelimAnalysis/PreliminaryAnalysis.Rmd
@@
 title: Weather anomalies associated with wildfires in the Great Plains
-subtitle: Preliminary analysis
+subtitle: Preliminary analysis with all events in MTBS database
 date: '`r Sys.Date()`'
@@
 
-pacman::p_load(tidyverse, sf, lubridate, grid)
+pacman::p_load(tidyverse, sf, lubridate, grid, wesanderson)
 options(dplyr.summarise.inform = FALSE)
@@
 # GIS layers
-  load(paste0(wd, './gis/Robjects/us_sf.Rdata'))
-  load(paste0(wd, '/gis/Robjects/gp_l3.Rdata'))
-  load(paste0(wd, '/gis/Robjects/GP_WF.Rdata'))
-  load("C:/Users/devan.mcgranahan/GithubProjects/FireWeather/mapping/data/region_states.Rdata")
+  load(paste0(wd, '/gis/Robjects/mapping.Rdata'))
 
-# weather data 
-  load(paste0(wd, '/data/HistoricalWeather/NWGP_top_20.Rdata'))
-  load(paste0(wd, '/data/PrelimAnalysis/FireWeekMeans.Rdata'))
-  load(paste0(wd, '/data/PrelimAnalysis/FireEventWx.Rdata'))
+# Historical weather data for events
+  load(paste0(wd, '/data/HistoricalWeather/HistWx.Rdata'))
+  load(paste0(wd, '/data/HistoricalWeather/DayOfWx.Rdata'))
 ```
@@
 
-```{r incident_map}
-fires <- 
-  NWGP_top_20 %>%
-  group_by(incident) %>%
-  slice(1) %>%
-  st_as_sf(coords = c("lon", "lat"), 
-           crs = 4326) %>%
-    st_transform(26915)
-
-inset_gg <-
+```{r map}
 ggplot() + theme_map(14) +
-    geom_sf(data = us_sf, 
-            fill = "lightgrey") + 
-    geom_sf(data = gp_l3,
-            fill = "darkblue", color = NA) + 
-    geom_sf(data = gp_l3 %>%
-                    filter(US_L3NAME == "Northwestern Great Plains"), 
-            fill = "lightblue")
-
-main_gg <-
-  ggplot() + theme_map(14) +
-    geom_sf(data = gp_l3 %>%
-                    filter(US_L3NAME == "Northwestern Great Plains"), 
-            fill = "lightblue") + 
-    geom_sf(data = region_states, fill = NA) +
-    geom_sf(data = fires, 
-            pch=21, 
-            size = 4, stroke = 1.25,
-            fill = "darkblue", 
-            color = 'white')
-
-grid.newpage() 
-main <- viewport(width = 1, height = 1, x = 0.5, y = 0.5) 
-inset <- viewport(width = 0.5, height = 0.5, x = 0.2, y = 0.25)
-print(main_gg, vp = main)
-print(inset_gg, vp = inset)
-# dev.off() 
+  geom_sf(data = mapping$us, 
+          fill = "grey90") + 
+  geom_sf(data = mapping$gp_er, 
+          aes(fill = L2))  + 
+  geom_sf(data = mapping$gp_st, 
+          fill = NA, 
+          color = 'grey50') +
+  scale_fill_manual(name = 'Level 2 ecoregion',
+                    values = wes_palette("Zissou1")) 
 ```

### 2021-05-10 gg_radar.R

@@ -1,0 +1,279 @@
+gg_radar <- function (plot.data, 
+                      base.size = NULL, 
+                      font.radar = "sans", 
+                      values.radar = c("0%", "50%", "100%"), 
+                      axis.labels = colnames(plot.data)[-1], 
+                      grid.min = 0, 
+                      grid.mid = 0.5, 
+                      grid.max = 1, 
+                      centre.y = grid.min - ((1/9) *  (grid.max - grid.min)), 
+                      plot.extent.x.sf = 1, 
+                      plot.extent.y.sf = 1.2, 
+                      x.centre.range = 0.02 * (grid.max - centre.y), 
+                      label.centre.y = FALSE, 
+                      grid.line.width = 0.5, 
+                      gridline.min.linetype = "longdash", 
+                      gridline.mid.linetype = "longdash", 
+                      gridline.max.linetype = "longdash", 
+                      gridline.min.colour = "grey", 
+                      gridline.mid.colour = "#007A87", 
+                      gridline.max.colour = "grey", 
+                      grid.label.size = NULL, 
+                      gridline.label.offset = -0.1 * (grid.max - centre.y), 
+                      label.gridline.min = TRUE, 
+                      label.gridline.mid = TRUE, 
+                      label.gridline.max = TRUE, 
+                      axis.label.offset = 1.15, 
+                      axis.label.size = NULL, 
+                      axis.line.colour = "grey", 
+                      group.line.width = 1.5, 
+                      group.point.size = 6, 
+                      group.colours = NULL, 
+                      background.circle.colour = "#D7D6D1", 
+                      background.circle.transparency = 0.2, 
+                      plot.legend = if (nrow(plot.data) > 1) TRUE else FALSE, 
+                      legend.title = "", 
+                      plot.title = "", 
+                      legend.text.size = NULL, 
+                      legend.direction = NULL,
+                      legend.position = NULL) 
+{
+  
+  funcCircleCoords <- function(center = c(0, 0), r = 1, npoints = 100) {
+                                tt <- seq(0, 2 * pi, length.out = npoints)
+                                xx <- center[1] + r * cos(tt)
+                                yy <- center[2] + r * sin(tt)
+                                return(data.frame(x = xx, y = yy))
+                              }
+                              
+  CalculateAxisPath <- function(var.names, min, max) {
+                        # var.names <- c("v1","v2","v3","v4","v5")
+                        n.vars <- length(var.names) # number of vars (axes) required
[…]

### 2021-05-10 script/FireFrequencyRadarMap.R

@@ -1,0 +1,151 @@
+pacman::p_load(tidyverse, lubridate, wesanderson, grid)
+pacman::p_load_current_gh("ricardo-bion/ggradar")
+
+load('./data/HistoricalWeather/DayOfWx.Rdata')
+
+pal40 <- wes_palette("Zissou1", 40, type = "continuous") 
+
+events_count <- 
+DayOfWx %>% 
+  mutate(month = month(date, label = T), 
+         decade = floor_date(date, "10 year"), 
+         decade = paste0(format(decade, '%Y'), ' - ', format(decade+years(10), '%Y'))) %>%
+  group_by(L2, decade, month) %>%
+  summarize(events = n() ) %>%
+  ungroup() 
+  
+sc_rp <-
+  events_count %>%
+    ungroup() %>%
+    filter(L2 == "South Central Semi-Arid Prairies") %>%
+    select(-L2) %>% 
+    pivot_wider(names_from = 'month', 
+                values_from = 'events', 
+                values_fill = 0, 
+                names_sort = TRUE) %>%
+  gg_radar(
+    axis.label.size = 4,
+    grid.label.size = 3,
+    legend.text.size = 8,
+    values.radar = c(0, 1250, 2500),
+    grid.min = 0, grid.mid = 1250, grid.max = 2500,
+    # Polygons
+      group.line.width = 0.4, 
+      group.point.size = 1.5,
+      group.colours = wes_palette("FantasticFox1")[c(3,2,1,5)],
+    # Background and grid lines
+    background.circle.colour = "white",
+    gridline.mid.colour = "grey",
+    plot.legend = FALSE, 
+    plot.title = 'South Central Semi-Arid Prairies') +
+  theme(plot.title = element_text(size = 8, hjust = 0.5))
+  
+tp_rp <- 
+  events_count %>%
+    ungroup() %>%
+    filter(L2 == "Temperate Prairies") %>%
+    select(-L2) %>% 
+    pivot_wider(names_from = 'month', 
+                values_from = 'events', 
+                values_fill = 0, 
+                names_sort = TRUE) %>%
+    gg_radar(
+      axis.label.size = 4,
+      grid.label.size = 3,
+      legend.text.size = 8,
+      values.radar = c(0, 250, 500),
+      grid.min = 0, grid.mid = 250, grid.max = 500,
+      # Polygons
+        group.line.width = 0.4, 
+        group.point.size = 1.5,
[…]

### 2021-05-10 script/GettingElevationDataNG.R

@@ -1,0 +1,39 @@
+pacman::p_load(tidyverse, sf)
+load('./albersEAC.Rdata')
+
+load("C:/Users/devan.mcgranahan/GithubProjects/GreatPlainsFire/gis/Robjects/gpng_perims3k_allots.Rdata")
+
+# NPS shapefiles 
+
+gpng_sf <- gpng_perims3k_allots %>%
+                st_transform(albersEAC)
+
+pacman::p_load(foreach, doParallel)
+cores=detectCores()
+
+{
+  cl <- makeCluster(cores[1]) 
+  registerDoParallel(cl)
+  begin = Sys.time()
+  
+    foreach(p=1:length(gp_nps_sf$unit), 
+            .combine=bind_rows) %dopar% {
+      pacman::p_load(tidyverse, sf)
+      p_sf <-  
+        gp_nps_sf %>%
+        slice(p) 
+      p_sf %>%
+        as_Spatial()  %>%
+        elevatr::get_elev_raster(z = 14, 
+                                 clip = 'bbox', 
+                                 verbose = F, 
+                                 override_size_check = T) %>%
+        raster::writeRaster(paste0('S:/DevanMcG/GIS/SpatialData/GreatPlains/DEM/', 
+                                   str_replace_all(p_sf$unit, " ", "_"), '.tif'))
+            }
+    beepr::beep() 
+    stopCluster(cl)
+    Sys.time() - begin
+}   
+
+

### 2021-05-10 script/GettingSoilsDataNG.R

@@ -1,0 +1,116 @@
+# pacman::p_load(RODBC, aqp, soilDB)
+
+pacman::p_load(tidyverse, sf)
+
+buffs <- gpng_perims3k %>%
+          st_transform(4326)
+
+pacman::p_load(foreach, doParallel)
+cores=detectCores()
+
+{
+cl <- makeCluster(cores[1]) 
+registerDoParallel(cl)
+begin = Sys.time()
+
+AllFireBuffs <-
+  foreach(p=1:length(buffs$FIRENAME), 
+          .combine=bind_rows, 
+          .packages=c('tidyverse', 'sf')) %dopar% {
+
+# Get just one feature from shp
+  p_sf <-  
+    buffs %>%
+      slice(p) 
+# Fetch map units for the feature
+  mukey_sf <-
+    p_sf %>%
+      as_Spatial() %>%
+       soilDB::SDA_spatialQuery(
+                        what = 'geom', 
+                        db = "SSURGO", 
+                        geomIntersection = TRUE) %>%
+        st_as_sf()
+# Define bbox to query from shp
+  aoi <- 
+    p_sf %>%
+     as_Spatial() %>%
+      rgeos::writeWKT()
+
+# Create SQL query strings...
+# ...one for ecological sites:
+es_q = paste0(
+  "SELECT 
+    component.mukey, component.cokey, component.comppct_r, component.compname, 
+    coecoclass.ecoclassname 
+  FROM component 
+    INNER JOIN coecoclass ON component.cokey = coecoclass.cokey
+  WHERE mukey IN (
+    SELECT * from SDA_Get_Mukey_from_intersection_with_WktWgs84('", aoi, "'))")
+#... another for soil texture:
+tex_q = paste0(
+  "SELECT
+    muname, mu.mukey, -- attributes from table 'mapunit'
+    comppct_r, compname, c.cokey, -- attributes from table 'component'
+    ch.chkey, -- attributes from table 'chorizon'
+    sandtotal_r, silttotal_r, claytotal_r --total sand, silt and clay fractions from table 'chorizon'
+  FROM sacatalog sac
+    INNER JOIN legend l ON l.areasymbol = sac.areasymbol AND l.areatypename = 'Non-MLRA Soil Survey Area'
+    INNER JOIN mapunit mu ON mu.lkey = l.lkey
[…]

### 2021-05-10 script/GrazingAllotmentsNG.R

@@ -1,0 +1,85 @@
+pacman::p_load(tidyverse, sf)
+source('https://raw.githubusercontent.com/devanmcg/rangeR/master/R/CustomGGplotThemes.R')
+
+s_gis = 'S:/DevanMcG/GIS/SpatialData/US/USFS'
+
+# Get a boundary for the Great Plains (EPA L3)
+  GP_L3 <- 
+    read_sf('S:/DevanMcG/GIS/SpatialData/US/EPAecoregions/L3', 
+            'us_eco_l3_state_boundaries') %>%
+    filter(NA_L1NAME == 'GREAT PLAINS') %>%
+    st_transform(4269)
+  GP <- st_union(GP_L3)
+
+# Filter national-level data down to the Great Plains
+
+  gp_ng <- read_sf(paste0(s_gis, '/NationalGrasslands'), 
+                    'S_USA.NationalGrassland') %>%
+                st_intersection(GP)
+  
+  gp_perims <- read_sf(paste0(s_gis, '/FinalFirePerimeter'), 
+                       'S_USA.FinalFirePerimeter') %>%
+                  filter(FIREYEAR >= 1985) %>%
+                  mutate(valid = st_is_valid(.)) %>%
+                  filter(valid != FALSE) %>%
+                  select(-valid) %>%
+                  st_intersection(GP)
+  
+  gpng_perims <- gp_perims %>%
+                    st_intersection(gp_ng %>% select(GRASSLANDN)) %>%
+                    filter(TOTALACRES >= 247.12) %>% # > 100 ha
+                    select(GRASSLANDN, FIRENAME, DISCOVERYD, 
+                           SIZECLASS, STATCAUSE, COMMENTS, OWNERAGENC)
+  
+  gpng_perims3k <- gpng_perims %>%
+                    st_transform(26915) %>%
+                      st_buffer(3200) %>%
+                    st_transform(4269) %>%
+                    st_intersection(gp_ng)
+  
+  gpng_allots <- read_sf(paste0(s_gis, '/RangeAllotments'), 
+                    'S_USA.Allotment')  %>%
+                    st_intersection(gp_ng %>% select(GRASSLANDN)) %>%
+                  select(GRASSLANDN, ALLOTMENT_:ALLOTMEN_2)
+
+  gpng_perims3k_allots <- gpng_perims3k %>%
+                            st_intersection(gpng_allots)
+  # save(gpng_perims3k_allots, file = './gis/Robjects/gpng_perims3k_allots.Rdata')
+  gpng_allots <- 
[…]

### 2021-05-10 script/GreatPlainsMapping.R

@@ -1,0 +1,123 @@
+epa <- read_sf('S:/DevanMcG/GIS/SpatialData/US/EPAecoregions/L3', 
+               'us_eco_l3_state_boundaries')
+aea <- st_crs(epa)
+
+load('C:/Users/devan.mcgranahan/GithubProjects/GreatPlainsFire/gis/Robjects/us_sf.Rdata')
+us_sf <- us_sf %>%
+          st_transform(aea) %>%
+            st_simplify()
+
+gp_states <- 
+  epa %>%
+  as_tibble() %>%
+  select(NA_L1NAME, STATE_NAME) %>%
+  filter(NA_L1NAME == "GREAT PLAINS") %>%
+  filter(STATE_NAME != "Wisconsin")
+
+gp_sf <- epa %>%
+          filter(STATE_NAME %in% gp_states$STATE_NAME) %>%
+          select(STATE_NAME) %>%
+          mutate(area = st_area(.)) %>%
+          group_by(STATE_NAME) %>%
+          summarize(area = sum(area)) %>%
+          ungroup() %>%
+          select(-area) 
+
+gp_l2 <- epa %>% 
+          filter(NA_L1NAME == "GREAT PLAINS") %>%
+            mutate(area = st_area(.)) %>%
+            group_by(NA_L2NAME) %>%
+            summarize(area = sum(area)) %>%
+            select(-area) %>%
+            rename(L2 = NA_L2NAME) %>%
+            mutate(L2 = str_to_title(L2))
+
+gp_l3 <- epa %>% 
+          filter(NA_L1NAME == "GREAT PLAINS") %>%
+          select(NA_L2NAME, US_L3NAME) %>%
+          rename(L2 = NA_L2NAME, 
+                 L3 = US_L3NAME) %>%
+          mutate(L2 = str_to_title(L2))
+
+gp_l2_5 <- epa %>% 
+  filter(NA_L1NAME == "GREAT PLAINS") %>%
+  select(NA_L2NAME, US_L3NAME) %>%
+  rename(L2 = NA_L2NAME, 
+         L3 = US_L3NAME) %>%
+  mutate(L2 = str_to_title(L2)) %>%
+  mutate(L2 = case_when(
+    L2 %in% c("Tamaulipas-Texas Semiarid Plain", 
+              'Texas-Louisiana Coastal Plain') ~ "Texas Southern Plains", 
+    TRUE ~ L2) ) %>%
+  mutate(L2_5 = case_when(
[…]

### 2021-05-10 script/HistoricalWeatherSummaries.R

@@ -1,0 +1,83 @@
+pacman::p_load(tidyverse, sf, multidplyr, wesanderson)
+
+# Load historical weather data
+  load('./data/HistoricalWeather/RawHistWx.Rdata')
+  load('./data/HistoricalWeather/RawDayOfWx.Rdata')
+
+# Load fire spatial information
+  gis_fp = 'C:/Users/devan.mcgranahan/GithubProjects/GreatPlainsFire/gis'
+  load(file=paste0(gis_fp, '/Robjects/GP_WF.Rdata'))
+
+# Add EPA ecoregion columns to historical weather data
+  HistWx <-
+    GP_WF  %>%
+      select(Event_ID, Incid_Name, NA_L2NAME, US_L3NAME) %>%
+      unite(event, c(Event_ID, Incid_Name), sep = '_') %>%
+      rename(L2 = NA_L2NAME, 
+             L3 = US_L3NAME) %>%
+      mutate(L2 = str_to_title(L2)) %>%
+      as_tibble() %>%
+      select(-geometry) %>%
+      full_join(HistWx)
+  #save(HistWx, file = './data/HistoricalWeather/HistWx.Rdata')
+  
+  DayOfWx <-
+    GP_WF  %>%
+    select(Event_ID, Incid_Name, NA_L2NAME, US_L3NAME) %>%
+    unite(event, c(Event_ID, Incid_Name), sep = '_') %>%
+    rename(L2 = NA_L2NAME, 
+           L3 = US_L3NAME) %>%
+    mutate(L2 = str_to_title(L2)) %>%
+    as_tibble() %>%
+    select(-geometry) %>%
+    full_join(DayOfWx)
+  #save(DayOfWx, file = './data/HistoricalWeather/DayOfWx.Rdata')
+
+# Summarize event-level data to L2 ecoregion
+  
+  L2_AnnualMeans <-
+    HistWx %>%
+    filter(!is.na(param)) %>%
+      group_by(L2, year, param) %>%
+      summarize(Mean = mean(TenDayMean, na.rm = TRUE)) %>%
+      ungroup() 
+  
+  L2_DayOfMeans <-
+    DayOfWx %>%
+    filter(!is.na(param)) %>%
+    group_by(L2, param) %>%
+    summarize(Mean = mean(DayOfValue, na.rm = TRUE), 
+              SD = sd(DayOfValue, na.rm = TRUE),
[…]

### 2021-05-10 script/LoadGridMetData_foreach.R

@@ -1,0 +1,75 @@
+pacman::p_load(tidyverse, sf, climateR)
+
+gis_fp = 'C:/Users/devan.mcgranahan/GithubProjects/GreatPlainsFire/gis'
+
+load(file=paste0(gis_fp, '/Robjects/GP_WF.Rdata'))
+
+# Identify variables to fetch
+elements = c(
+  'precipitation_amount', 
+  'daily_minimum_relative_humidity',
+  'daily_mean_specific_humidity',
+  'daily_mean_shortwave_radiation_at_surface',
+  'daily_maximum_temperature',
+  'daily_mean_wind_speed',
+  'daily_mean_burning_index_g',
+  'daily_mean_energy_release_component-g',
+  #'daily_mean_palmer_drought_severity_index',
+  'daily_mean_reference_evapotranspiration_grass',
+  'daily_mean_vapor_pressure_deficit')
+
+params <- filter(param_meta$gridmet, description %in% elements )$common.name
+
+fires <- 
+  GP_WF %>%
+    st_centroid() %>% 
+      st_transform(4326) %>%
+  select(Event_ID, Incid_Name, Ig_Date) %>%
+  mutate(start = format(Ig_Date - lubridate::days(5), '%m-%d'), 
+         end = format(Ig_Date + lubridate::days(5), '%m-%d'), 
+         date = format(Ig_Date, '%Y-%m-%d')) %>%
+  unite("event", c(Event_ID, Incid_Name), sep = '_') 
+
+# Download 40 years of historical weather data from each fire location
+yr_range = 1980:2020
+
+pacman::p_load(foreach, doParallel)
+cores=detectCores()
+{
+cl <- makeCluster(cores[1]) 
+registerDoParallel(cl)
+
+begin = Sys.time()
+HistWx <- 
+  foreach(f=1:length(fires$event[1:2]), 
+          .combine = rbind, 
+          .errorhandling = 'remove', 
+          .inorder = FALSE) %:% 
+  
+  foreach(y=1:length(yr_range[1:10]), 
+          .combine = 'rbind', 
+          .errorhandling = 'remove', 
+          .inorder = FALSE, 
+          .packages = c('tidyverse', 'sf', 'climateR')) %dopar% {
+        fire = slice(fires, f)
+        climateR::getGridMET(
+          AOI = fire, 
+          param = params,
+          startDate = paste0(yr_range[y],'-', 
+                             as_tibble(fire)['start']), 
+          endDate = paste0(yr_range[y],'-', 
[…]

### 2021-05-10 script/LoadTrimSpatialData.R

@@ -3,17 +3,19 @@ pacman::p_load(tidyverse, sf)
 gis_fp = 'C:/Users/devan.mcgranahan/GithubProjects/GreatPlainsFire/gis'
 
 GP_L3 <- 
-  read_sf(paste0(gis_fp, '/L3ecoregions'), 'us_eco_l3') %>%
+  read_sf('S:/DevanMcG/GIS/SpatialData/US/EPAecoregions/L3', 
+          'us_eco_l3_state_boundaries') %>%
     filter(NA_L1NAME == 'GREAT PLAINS')
 # save(GP_L3, file=paste0(gis_fp, '/Robjects/GP_L3.Rdata'))
 
 albersEAC = st_crs(GP_L3)
 
-GP_WF <- 
-  read_sf(paste0(gis_fp, '/mtbs'), 'mtbs_perims_DD') %>%
+GP_fires <- 
+  read_sf('S:/DevanMcG/GIS/SpatialData/US/mtbs', 
+          'mtbs_perims_DD') %>%
       st_transform(albersEAC) %>%
         st_intersection(GP_L3) %>%
-          filter(Incid_Type == "Wildfire")
+          filter(Incid_Type %in% c('Wildfire', 'Prescribed Fire', 'Complex'))
 
 # save(GP_WF, file=paste0(gis_fp, '/Robjects/GP_WF.Rdata'))
 

### 2021-05-10 script/NationalGrasslandsFires.R

@@ -1,0 +1,70 @@
+pacman::p_load(tidyverse, sf)
+
+source('https://raw.githubusercontent.com/devanmcg/rangeR/master/R/CustomGGplotThemes.R')
+
+gis_fp = 'C:/Users/devan.mcgranahan/GithubProjects/GreatPlainsFire/gis'
+
+load(file=paste0(gis_fp, '/Robjects/GP_WF.Rdata'))
+
+gp_fires <-
+  GP_fires %>%
+    st_transform(26915) 
+
+gp_fires_2k <-
+  GP_fires %>% 
+    st_buffer(2000) 
+
+  # st_write(gp_fires, paste0(gis_fp, '/BufferedPerimeters/GP_UTM15.shp'), append = FALSE)
+  # st_write(gp_fires_2k, paste0(gis_fp, '/BufferedPerimeters/BufferedGP_UTM15.shp'), append = FALSE)
+
+
+ggplot() + theme_map() + 
+  geom_sf(data = gp_wf_2k %>%
+                  slice(1:5),
+          aes(fill = Incid_Type),
+          fill = 'darkblue') %>%
+  geom_sf(data = GP_WF %>%
+            slice(1:5) %>%
+            st_transform(26915),
+          aes(fill = Incid_Type),
+          fill = 'lightblue')
+
+NG_fires <-
+  read_sf('S:/DevanMcG/GIS/SpatialData/US/USFS/NationalGrasslands', 
+          'S_USA.NationalGrassland') %>% 
+  st_transform(26915) %>%
+  st_intersection(gp_fires)
+
+# st_write(NG_fires, paste0(gis_fp, '/BufferedPerimeters/NGfires_UTM15.shp'), append = FALSE)
+
+# While this starts the buffer process, 
+# a better (dissolved) buffer was created in QGIS: 
+# S:/DevanMcG/GIS/SpatialData/GreatPlains/NGfireBuffer_UTM15.shp
+NG_fires_2k <-
+  read_sf('S:/DevanMcG/GIS/SpatialData/US/USFS/NationalGrasslands', 
+          'S_USA.NationalGrassland') %>% 
+    st_transform(26915) %>%
+    st_intersection(NG_fires) %>% 
+  st_buffer(2000) 
+
+NG_fires_2k %>% 
+  st_transform(4326) %>%
[…]

### 2021-05-10 script/RAPfetcher.sh

@@ -1,0 +1,1 @@
+gdal_translate -co compress=lzw -co tiled=yes -co bigtiff=yes /vsicurl/http://rangeland.ntsg.umt.edu/data/rap/rap-vegetation-cover/v2/vegetation-cover-v2-2000.tif -projwin -106.67 48.43 -102.19 45.1 S:\DevanMcG\GIS\SpatialData\RAP\MonDak2000.tif

### 2021-05-11 script/LoadGridMetData.R
update point historical data

@@ -22,7 +22,7 @@ params <- filter(param_meta$gridmet, description %in% elements )$common.name
 
 fires_LongLat <- 
   GP_WF %>%
-    slice(1:5) %>%
+    # slice(1:5) %>%
       st_centroid() %>% 
       st_transform(4326)
 

### 2021-05-11 script/LoadGridMetDataHistoricalGrid.R

@@ -1,0 +1,54 @@
+pacman::p_load(tidyverse, sf, climateR)
+
+load('./gis/Robjects/GP_Wx_SamplePoints.Rdata')
+
+points <- 
+  GP_Wx_SamplePoints %>%
+    st_transform(4326)
+
+# Identify variables to fetch
+elements = c(
+  'precipitation_amount', 
+  'daily_minimum_relative_humidity',
+  'daily_mean_specific_humidity',
+  'daily_mean_shortwave_radiation_at_surface',
+  'daily_maximum_temperature',
+  'daily_mean_wind_speed',
+  'daily_mean_burning_index_g',
+  'daily_mean_energy_release_component-g',
+  'daily_mean_reference_evapotranspiration_grass',
+  'daily_mean_vapor_pressure_deficit')
+
+params <- filter(param_meta$gridmet, description %in% elements )$common.name
+
+# Download 40 years of historical weather data for each sample point
+  fast.download <- edit(fast.download) 
+# Non=parallel (getGridMET's fast.download modified to use all six cores)
+{
+  begin = Sys.time()
+  PointHistWx <- tibble() 
+    for(p in 1:length(points$geometry))  {
+              pt = slice(points, p)
+              getGridMET(
+                AOI = pt, 
+                param = params,
+                startDate = '1980-01-01', 
+                endDate = '2020-12-31') %>% 
+                select(-source, -lat, -lon) %>%
+                pivot_longer(names_to = "param", 
+                             values_to = "value", 
+                             -date) %>%
+                mutate(date = format(date, '%Y-%m')) %>%
+                group_by(date, param) %>%
+                summarize(MonthMean = mean(value)) %>%
+                ungroup()  %>%
+                mutate(point = p) %>%
+                bind_rows(PointHistWx) -> PointHistWx
+              closeAllConnections()
+            } 
+
+  Sys.time() - begin 
+}
+  
+ # save(PointHistWx, file = './data/HistoricalWeather/PointHistWx.Rdata') 
+ 

### 2023-06-24 SpringSummerSeverity/script/GetGridMET.R
start NAPC talk

@@ -1,0 +1,65 @@
+pacman::p_load(tidyverse, sf)
+pacman::p_load_gh("devanmcg/climateR") 
+# pacman::p_load_gh('fickse/mesowest')
+
+# Identify variables to fetch
+  elements = c(
+    'precipitation_amount', 
+    'daily_minimum_relative_humidity',
+    'daily_maximum_temperature',
+    'daily_mean_wind_speed',
+    'daily_mean_palmer_drought_severity_index',
+    'daily_mean_reference_evapotranspiration_grass',
+    'dead_fuel_moisture_100hr', 
+    'daily_mean_burning_index_g',
+    'daily_mean_energy_release_component-g',
+    'daily_mean_vapor_pressure_deficit')
+
+params <- filter(param_meta$gridmet, description %in% elements )$common.name
+
+
+
+pacman::p_load(foreach, doSNOW)
+
+{
+  cgrec_pt <- read_sf('./gis/boundaries/CGREC_bbox_4326.shp') %>% st_centroid()
+  yr_range = 1980:2020
+  mo_range = c('04', '05', '06', '07', '08', '09', '10')
+  
+  cores=parallel::detectCores()
+  cl <- makeCluster(cores) 
+  registerDoSNOW(cl )
+  
+  begin = Sys.time()
+  HistWx <- 
+    foreach(y=1:length(yr_range), 
+            .combine = bind_rows, 
+            .errorhandling = 'remove', 
+            .inorder = FALSE) %:% 
+    
+    foreach(m=1:length(mo_range), 
+            .combine = bind_rows, 
+            .errorhandling = 'remove', 
+            .inorder = FALSE, 
+            .packages = c('tidyverse', 'sf', 'climateR')) %dopar% {
+              getGridMET(
+                AOI = cgrec_pt, 
+                param = params,
+                startDate = paste0(yr_range[y],'-', 
+                                   mo_range[m], '-', 
+                                   '01'), 
+                endDate = paste0(yr_range[y],'-', 
+                                 mo_range[m], '-', 
+                                 '30') )  %>% 
+                select(-source, -lat, -lon) %>%
+                mutate(hdwi = wind_vel * vpd) %>%
+                pivot_longer(names_to = "param", 
+                             values_to = "value", 
+                             -date) 
+              # closeAllConnections()
+            }
+  stopCluster(cl)
+  Sys.time() - begin 
+} 
+
+save(HistWx, file = './data/HistWx.Rdata')

### 2023-06-24 SpringSummerSeverity/script/PlottingFireWx.R

@@ -1,0 +1,23 @@
+
+
+load('./data/HistWx.Rdata')
+
+HistWx %>%
+  separate(date, c('year', 'month', 'day'), sep = "-", remove = F) %>%
+  mutate(season = case_when(
+    month %in% c('04', '05') ~ 'spring',
+    month %in% c('08') ~ 'summer'), 
+    day = format(date, '%b-%d')) %>%
+  filter(!is.na(season),
+         !is.na(value), 
+         param %in% c("burn_index", "fmoist_100", "energy_release", "palmer",        
+                      "pet_grass", "hdwi" )) %>%
+  group_by(param, season, day) %>%
+  summarize(mean = mean(value, na.rm = TRUE), 
+            SEM = sd(value, na.rm = TRUE) / sqrt(n()), 
+            .groups = 'drop') %>%
+  ggplot() + theme_bw() + 
+    geom_smooth(aes(x = day, 
+                  y = mean, 
+                  group = 1) ) +
+  facet_grid(param ~ season, scales = 'free')

### 2023-06-24 SpringSummerSeverity/script/boundaries.R

@@ -1,0 +1,18 @@
+pacman::p_load(tidyverse, sf)
+
+fp = './gis/boundaries'
+
+# WGS bbox for EarthExplorer
+  read_sf('S:/DevanMcG/GIS/SpatialData/NorthDakota/raw/boundaries/AllPastureBoundaries', 
+                           'AllPastureBoundaries') %>%
+    st_union() %>%
+    st_bbox() %>%
+    st_as_sfc() %>%
+    st_transform(4326) %>%
+    write_sf(paste0(fp, '/CGREC_bbox_4326.shp'), 
+             append = FALSE)
+# Get CGREC pastures and patches
+  read_sf('S:/DevanMcG/GIS/SpatialData/PBGunits/CGREC/patches', 'patches') %>%
+    filter(Unit %in% c('Bob', 'Barker')) %>%
+    write_sf(paste0(fp, '/CGREC_pbg_26913.shp'), 
+             append = FALSE)

### 2023-06-24 RxFireReview/SoilHeatingReview.Rmd

---
title: Soil Heating effects of prescribed fire in US Great Plains
author: "Devan Allen McGranahan"
output:
  word_document:
    reference_docx: arial_template.docx
csl: SpringerNature.csl
bibliography: SoilheatingBib.bib
---

```{r setup, include=FALSE, eval = TRUE}
knitr::opts_chunk$set(echo = FALSE, eval = FALSE)
# pacman::p_load(tidyverse, pander)
```

Soil is a natural system comprised of minerals, organic matter, air, and water. 
Soil development begins with the weathering of local geologic parent material and is continually influenced by climate and organisms. 
Soil is not just the natural medium for plant growth---it is the habitat for organisms that regulate the delivery of multiple ecosystem services, and a substantial repository for elements and minerals important to global processes, such as carbon and nitrogen.

Many soil properties can be affected by wildland fire, either directly or indirectly. 
Variability in these affects comes largely from the intensity and duration of heat exposure as well as the environmental conditions before and after fire [@mcgranahan2021; @pereira2019]. 
Direct effects are those produced by exposure to heat generated by combustion---e.g., chemical conversion of organic nitrogen to mineral ammonium, protein denaturation. 
Indirect effects are responses that follow environmental alterations attributable to fire---e.g., biomass removal, increased insolation, and second-order vegetation dynamics.

## General patterns of soil heating

There is little evidence that prescribed fire has substantial impact on biotic or abiotic soil properties in the Great Plains. 
The overall lack of effect is primarily due to the fact that soil heating in grassland ecosystems is relatively benign, due to the low residence time of flame fronts in fuelbeds dominated by fine fuels, especially under the moderate fire weather conditions typical of prescribed fire in the region. 
Soil itself resists heating---the majority of heat released from surface combustion rises, and organic matter and air effectively insulate against energy from moving downward. 
Research from Europe indicates that only under conditions of very high soil moisture and prolonged heat input will substantial heating penetrate more than 3-5 cm into the soil profile and subsequently affect soil properties [@giovannini1997; @valette1994; @barreiro2020]. 
The limited data on soil heating during prescribed fire in the Great Plains provides evidence of this drastic attenuation (Table 1).

Table 1: Selection of studies reporting soil surface and/or sub-surface temperatures from prescribed fire in Great Plains grassland communities. 
All reported data collected from thermocouples connected to digital dataloggers.
Sensor placement refers to where the hot end of thermocouple was placed; negative values indicate depth below soil surface. 
* Values derived via soil heating model.

Study	         | Prairie type (Location)       |Sensor placement | Temperature ($^\circ$C)
---------------|-------------------------------|-----------------|------------------------
@archibold2003 | Fescue prairie (Saskatchewan) | −1 cm	         | 22
@archibold1998 | Mixed grass (Saskatchewan)    | Soil surface    | 252
		           |                               | −5 cm	         | 12
@zopfi2020     | Mixed grass (central North Dakota) | Soil surface | 150
@zopfi2020     | Mixed grass (sw North Dakota) | Soil surface | 120
@mcgranahan2022| Mixed grass (sw North Dakota)    | Soil surface    | 275
               |                               | −2 cm *         | 100 *
               |                               | −5 cm *         | 50 *
@ohrtman2015   | Tallgrass (South Dakota)      | Soil surface	   | 384
@anderson2016  | Tallgrass (Texas)             | −5 cm           | 28

Many fire effects on soils are modulated by the fire regime, particularly the frequency and intensity of fire. 
While infrequent fire can increase nitrogen availability and increase nutrient cycling rates, annual burning increased soil respiration in tallgrass prairie [@johnson2001]. 
Indeed, annual burning might generally elevate the net export of necessary nutrients above what deposition and mineralization can replace [@brye2006]. 
Conversely, low-intensity fires in *Poa pratensis*-dominated mixed grass prairie in North Dakota had no appreciable effect on soil properties [@gerhard2022]. 
The conclusion here was that soil properties are resilient to heating, but given that soil surface temperatures in the studied burned units averaged just 150$^\circ$C [@zopfi2020], it is just as likely that the soil strata measured received very little heating.

## Specific belowground responses to soil heating

Prescribed fire tends to increase soil surface temperature and decrease soil moisture, but the magnitude of reported effects are generally small. 
In western Oklahoma, soil temperatures were 1-3$^\circ$C higher after burns, but soil moisture on these sandy soils did not vary among treatments [@vermeire2005]. 
In eastern Montana, burned plots were 0.5$^\circ$C warmer and 1% drier than unburned plots [@vermeire2011]. 
In southern Saskatchewan, soil moisture did not vary signficantly between burned and unburned sites while in only one instance was a 5$^\circ$C increase in soil temperature significantly greater than in unburned controls [@yang2013]. 
But drier soil might not be a universal response to fire, especially in grazed areas: @spiess2021 reported *higher* soil moisture in burned patches in which grazers maintained low-stature vegetation, suggesting reducing the amount of leaf tissue capable of moving soil moisture into the atmosphere might conserve water in the soil. 
This hypothesis, however, has yet to be formally tested at the plant level, although altered water flux has been documented in burned prairie [@fischer2012].

Plant-available, or *mineralized*, nitrogen (N) has been widely reported to increase in response to fire. 
Although much of this research has been conducted beyond the Great Plains [@mcgranahan2021], work from our region corroborates these trends. 
In one of few studies to directly measure mineral N and soil heating, @mcgranahan2022 found that ammonium levels increased immediately after fall fire in North Dakota, and nitrate levels were elevated 7 mo after fire, into the next growing season (Fig. 1).
This same pattern was reported in Colorado shortgrass steppe [@augustine2014].
Similarly, @reinhart2016 also reported increases in soil N following fire in eastern Montana.

Soil microbes have generally been described as resilient to fire---in the event that limited reductions in microbial biomass occur immediately after fire, abundance typically recovers within weeks or months [@mcgranahan2022]. 
But fire-driven reductions are not universally observed [@reinhart2016] and microbial community composition is typically unaffected by fire [@dangi2010; @spiess2021; @mcgranahan2022].

Most reported fire effects on soil properties are limited to upper soil layers. 
For example, in mesquite-prairie rangeland in Texas, @dai2006 found fire regimes that included summer fires increased soil organic C and total N in the top 20 cm, but had no effect on either variable lower in the profile. 
Thus, the role of fire in long-term, ecosystem-level soil nutrient dynamics, such as carbon sequestration, in the Great Plains likely lies mostly in indirect effects related to plant community composition---e.g., herbaceous vs. woody vegetation---than direct effects of fire. 
In some systems, *pyrogenic* charcoal byproducts of wildland fire known as *PyC* or *PyOM* have been shown to decompose more slowly than similar amounts of carbon or organic matter unaltered by heating, suggesting a recalcitrance that might contribute to carbon sequestration [@quilliam2013]. 
However, experimental work on grass tissue, specifically, reports similar rates of microbial activity [@hilscher2009; @hilscher2011], suggesting carbonization does little to enhance the carbon storage capacity of the dominant vegetation type in the Great Plains.

# References

### 2023-06-25 SpringSummerSeverity/talk/NAPC2023.tex
work on talk

\documentclass[11pt]{beamer}

\usetheme[progressbar=frametitle]{metropolis}
\usepackage{appendixnumberbeamer}
\usepackage{FiraSans}
\usepackage{graphicx} 
\usepackage{booktabs}
\usepackage[scale=2]{ccicons}
\usepackage{adjustbox}
\usepackage{soul}
\usepackage{caption}
	\captionsetup[figure]{	labelformat=empty, 
					font=scriptsize,
					labelfont=scriptsize, 
					justification=raggedleft}

\usepackage{pgfplots}
\usepgfplotslibrary{dateplot}

\usepackage{xspace}
\newcommand{\themename}{\textbf{\textsc{metropolis}}\xspace}

\title{Barriers to a summer fire regime in northern prairies }
\subtitle{Ecological, physical, and social}
\date{}
\author{Devan Allen McGranahan } 
\institute{\emph{Research Rangeland Management Scientist\textemdash Ecologist} \\
		     USDA Agricultural Research Service \\ 
	    	     Miles City, Montana}

\newcommand\blfootnotetext[1]{%
  \begingroup
  \renewcommand\thefootnote{}\footnote{#1}%
  \addtocounter{footnote}{-1}%
  \endgroup
}

\begin{document}

\maketitle

\begin{frame}{The Western fire regime concept } 
	\begin{center}
		\begin{figure}
			\includegraphics[width=1\linewidth]{figs/FireRegimeNAPC} 
		\end{figure}
	\end{center}
\end{frame}

\begin{frame}{The Western fire regime concept}
	\begin{center}
		\begin{figure}
			\includegraphics[width=0.9\linewidth]{figs/FireRegimeModulatorsNAPC} 
		\end{figure}
	\end{center}
\end{frame}

\begin{frame}{Rx fire in the Northern Great Plains}
	\begin{center}
		\begin{figure}
			\includegraphics[width=1\linewidth]{figs/region_map-1.pdf} 
		\end{figure}
	\end{center}
\end{frame}

\begin{frame}{Rx fire in the Northern Great Plains}
	\begin{center}
		\begin{figure}
			\includegraphics[width=1\linewidth]{figs/PBGmaps} 
		\end{figure}
	\end{center}
\end{frame}

\begin{frame}{Burns visible from space}
	\begin{center}
		\begin{figure}
			\includegraphics[width=1\linewidth]{figs/BobSWBurnedOutline2} 
		\end{figure}
	\end{center}
\end{frame}

\begin{frame}{Spotty success in completing summer burns}

	\begin{center}
		\begin{figure}
			\includegraphics[width=0.9\linewidth]{figs/BarkerBurns-1} 
			\caption{2019 burn map for southern study block}
		\end{figure}
	\end{center}
\end{frame}

\section{A tale of two fires}

\begin{frame}{What a difference a few days makes! } 
	\begin{columns}
		\begin{column}{0.5\textwidth}
				\begin{center}
				\begin{figure}
					%\caption{5 May 2018}
					\includegraphics[width=0.9\linewidth]{figs/GoodBurn} 
				\end{figure}
			\end{center}
		\end{column}
		\begin{column}{0.5\textwidth}  
			\begin{center}
				\begin{figure}
					%\caption{16 May 2018}
					\includegraphics[width=1\linewidth]{figs/StreeterFire_17}  
				\end{figure}
			\end{center}
		\end{column}
	\end{columns}
\centering
\adjustbox{max height=\dimexpr\textheight-5.5cm\relax,
	max width=1\textwidth}{
	\begin{tabular}{ccc}
		\toprule
		5 May 2018 & \textbf{Date} & 16 May 2018 \\
		Weather & \textbf{Wind} & \\
		Fuels & \textbf{RH} & Yes \\ \bottomrule
	\end{tabular}
}
\end{frame}

\begin{frame}{Barriers to summer Rx fire } 
	All opportunities and limitations fit within fire regime concept
	\begin{itemize}
		\item Biophysical 
		\item[] \emph{mostly, \alert{too wet}}
		\begin{itemize}
			\item High live moisture fuel content (\emph{photosynthesis})
			\item High dead moisture fuel content (\emph{humidity})
			\item Poor convection/smoke dispersal (\emph{humidity})
		\end{itemize}
	\item[]
		\item Social 
		\item[] \emph{mostly, \alert{too dry}}
		\begin{itemize}
			\item Local burn restrictions
			\item Control issues
			\end{itemize}
	\end{itemize}

\end{frame}

\section{Biophysical barriers}

\section{Social barriers}

\begin{frame}{County-level burn restrictions}
	\begin{center}
		\begin{figure}
			\includegraphics[width=1\linewidth]{figs/FireDeclarations.png} 
		\end{figure}
	\end{center}
\end{frame}

\begin{frame}{County-level burn restrictions}
	\begin{center}
		\begin{figure}
			\includegraphics[width=1\linewidth]{figs/StutsmanBurnBan} 
		\end{figure}
	\end{center}
\end{frame}

\begin{frame}{Tying burn restrictions to real-time conditions}
	\begin{center}
		\begin{figure}
			\includegraphics[width=1\linewidth]{figs/StutsmanOrdinance} 
		\end{figure}
	\end{center}
\end{frame}

\begin{frame}{Fire, grazing, and the pre-European landscape}
\alert{Pyric herbivory}
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
	\item Recent burns have high-quality forage
	\item[]
	\item Grazers focus on burned areas
	\item[]
	\item Vegetation grows up in other areas...
		\begin{itemize}
			\item providing habitat
			\item fueling future fires
		\end{itemize}
	\end{itemize}
\end{column}
\begin{column}{0.55\textwidth}  
\begin{center}
\begin{figure}
 \includegraphics[width=1\linewidth]{figs/ghd}

 \end{figure}
\end{center}
\end{column}
\end{columns}
\end{frame}

\begin{frame}{Disturbance-driven vs. inherent heterogeneity}
\begin{center}
\begin{figure}
 \includegraphics[width=1\linewidth]{figs/StreeterFire_17} 
 \end{figure}
\end{center}
\end{frame}

\begin{frame}{Testing PBG 2.0 in the Northern Great Plains } 
	\begin{columns}
		\begin{column}{0.6\textwidth}
			\begin{itemize}
				\item \alert{Streeter (Central Grasslands)}
				\begin{itemize}
					\item \textbf{Compares grazing systems}
					\item 4 pastures x 3 systems; cattle
					\item \emph{Poa pratensis}-invaded mixed-grass prairie
					\item 46 cm (18") annual rainfall
				\end{itemize}
				\item \alert{Hettinger}
				\begin{itemize}
					\item \textbf{Compares cattle \& sheep}
					\item 3 x cattle, 3 x sheep PBG
					\item Mostly non-native, cool-season grasses
					\item post-CRP pasture stands
					\item 40 cm (16") annual rainfall
				\end{itemize}
			\end{itemize}
		\end{column}
		\begin{column}{0.45\textwidth}  
			\begin{center}
				\begin{figure}
					\includegraphics[width=1\linewidth]{figs/BobPasture}  
				\end{figure}
			\end{center}
		\end{column}
	\end{columns}
\end{frame}

\begin{frame}{Mopping up} 
\alert{Prescribed fire has direct benefits to livestock production}
\begin{columns}
\begin{column}{0.55\textwidth}
\begin{itemize}
	\item Livestock focus grazing in recently-burned patches
	\item[]
	\item Burned areas have better forage
	\item[]
	\item Fire has positive or neutral effects on soil
	\item[]
	\item Patch-burning enhances conservation outcomes, too
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}  
\begin{center}
\begin{figure}
\includegraphics[width=1\linewidth]{figs/mop_up} 
 \end{figure}
\end{center}
\end{column}
\end{columns}
\end{frame}

\begin{frame}{Thank you!}
Any questions before you all run away?
\begin{center}
\begin{figure}
 \includegraphics[width=1\linewidth]{figs/fire_cows_close}

 \end{figure}
\end{center}
\end{frame}

\end{document}

### 2023-06-25 SpringSummerSeverity/script/NorthDakotaMapping.R

@@ -1,0 +1,50 @@
+pacman::p_load(tidyverse, sf, geonames, 
+               grid, gridExtra)
+source('https://raw.githubusercontent.com/devanmcg/rangeR/master/R/CustomGGplotThemes.R')
+
+load('C:/Users/devan.mcgranahan/GithubProjects/SpatialFireBehavior/paper/figures/mapping.Rdata')
+# Make maps
+
+load('./paper/figures/mapping.Rdata')
+
+region_map <-
+  ggplot() + theme_map(14)  +
+  geom_sf(data = mapping$us, 
+          fill = 'lightgrey', 
+          color = NA)+ 
+    geom_sf(data = mapping$gp_l1, 
+            fill = 'lightblue') +
+    geom_sf(data = mapping$nd,
+            fill = 'darkblue', 
+            color = "darkgrey") +
+    geom_sf(data = mapping$us, 
+            fill = NA, 
+            color = "white")
+
+  
+state_map <-
+  ggplot() + theme_map(14)  +
+    geom_sf(data = mapping$nd,
+            fill = 'NA', 
+            color = "darkgrey") + 
+    geom_sf(data = mapping$nd_l3,
+            aes(fill = L3), 
+            show.legend = FALSE) +
+    geom_sf_label(data = mapping$nd_l3,
+                  aes(label = L3)) +
+    geom_sf(data = mapping$pts, 
+            aes(shape = feature), 
+            size = 4,
+            show.legend = FALSE) +
+    scale_shape_manual(values = c(16, 16, 17)) +
+    geom_sf_text(data = mapping$pts,
+                 aes(label = name), 
+                 fontface = c("bold", 'bold', 'italic', 'italic'),
+                 nudge_y = 20000,
+                 nudge_x = c(-35000, 0, 0, -15000),
+            show.legend = FALSE)
+
+v1<-viewport(width = 1, height = 1, x = 0.5, y = 0.5) #plot area for the main map
+v2<-viewport(width = 0.45, height = 0.45, x = 0.71, y = 0.75) #plot area for the inset map
+print(state_map,vp=v1) 
+print(region_map,vp=v2)

### 2023-06-25 SpringSummerSeverity/talk/FigureGenerator.Rmd

---
title: Figures for NAPC talk
output:
  html_document:
    keep_md: false
    theme: united
    highlight: tango
---

```{r setup, echo=FALSE, warning=FALSE, message = FALSE, results='hide'}
knitr::opts_chunk$set(message = FALSE, warning=FALSE, 
                      echo=FALSE, eval=TRUE, fig.path='C:/Users/devan.mcgranahan/GithubProjects/GreatPlainsFire/SpringSummerSeverity/talk/figs/', dev=c('png', 'pdf')) 
# Packages 
  pacman::p_load(tidyverse, sf, grid, gridExtra)
  pacman::p_load_gh('devanmcg/wesanderson')
# Data 
  # regional map data
    load('C:/Users/devan.mcgranahan/GithubProjects/SpatialFireBehavior/paper/figures/mapping.Rdata')
  # study data
    cgrec_gpkg = 'C:/Users/devan.mcgranahan/GithubProjects/GreatPlainsFire/SpringSummerSeverity/gis/boundaries/CGREC_PBG_26914.gpkg'
```

```{r region_map, fig.width=8, eval = FALSE}
mapping$nd_l3 <- mapping$nd_l3 %>% st_simplify(dTolerance = 1000)
mapping$gp_l1 <- mapping$gp_l1 %>% st_simplify(dTolerance = 10000)

region_map <-
  ggplot() + theme_void(24)  +
  geom_sf(data = mapping$us, 
          fill = 'lightgrey', 
          color = NA)+ 
    geom_sf(data = mapping$gp_l1, 
            fill = wes_palettes$DeadLiveLight[1]) +
    geom_sf(data = mapping$nd,
            fill = wes_palettes$DeadLiveLight[2], 
            color = "darkgrey") +
    geom_sf(data = mapping$us, 
            fill = NA, 
            color = "white")

state_map <-
  ggplot() + theme_void(30)  +
    geom_sf(data = mapping$nd,
            fill = 'NA', 
            color = "darkgrey") + 
    geom_sf(data = filter(mapping$nd_l3,
                          L3 != 'Northwestern Great Plains'), 
            fill = wes_palettes$DeadLiveLight[1], 
            show.legend = FALSE) +
    geom_sf_label(data = filter(mapping$nd_l3,
                                L3 != 'Northwestern Great Plains'),
                  aes(label = L3), 
                  size = 6) +
    geom_sf(data = filter(mapping$pts, 
                          name != "Hettinger REC"), 
            aes(shape = feature), 
            size = 4,
            show.legend = FALSE) +
    scale_shape_manual(values = c(16, 16, 17)) +
    geom_sf_text(data = filter(mapping$pts, 
                               name != "Hettinger REC"), 
                 aes(label = name), 
                  size = 6, 
                 fontface = c("bold", 'bold', 'italic'),
                 nudge_y = c(20000, 20000, -20000), 
                 nudge_x = c(-35000, 0, -15000),
            show.legend = FALSE)

v1<-viewport(width = 1, height = 1, x = 0.5, y = 0.5) #plot area for the main map
v2<-viewport(width = 0.45, height = 0.45, x = 0.71, y = 0.75) #plot area for the inset map
print(state_map,vp=v1) 
print(region_map,vp=v2)
```

```{r BarkerBurns}
  # Fire mapping

 # st_layers(cgrec_gpkg )

  pastures <- st_read(cgrec_gpkg, 'Pastures') 
  patches <- st_read(cgrec_gpkg, 'PasturePatches')

  fires <- st_read(cgrec_gpkg, 'FirePerimeters') %>%
              mutate(Year = as.factor(Year))

    ggplot() + theme_void() +
      geom_sf(data = filter(fires, unit == 'Barker'), 
      aes(fill = Year, 
          alpha = Season)) +
      geom_sf(data = filter(patches, Unit == 'Barker'), 
              fill = NA, 
              color = 'black', 
              lty = 2, 
              size = 1) + 
      geom_sf(data = filter(pastures, Unit == 'Barker'), 
              fill = NA, 
              color = 'black', 
              lty = 1, 
              size = 2) +
      geom_sf_text(data = filter(fires, unit == 'Barker'), 
                   aes(label = Year), 
                   nudge_y = 50) +
      geom_sf_text(data = filter(fires, unit == 'Barker'), 
                   aes(label = substr(Season, 1, 3)), 
                   nudge_y = -50) +
      scale_alpha_manual(values = c(0.25, 0.8)) + 
      theme(legend.position = 'none')
```

### 2023-06-26 SpringSummerSeverity/talk/NAPC2023.tex
finish talk

@@
 \usepackage{FiraSans}
+\usepackage{textcomp} %% put this in your preamble
 \usepackage{graphicx} 
@@
 
+\begin{frame}{For every season: Burn, burn, burn...}
+		\vspace{-3em}	
+	\begin{columns}
+
+		\begin{column}{0.5\textwidth}
+			\begin{itemize} 
+				\item Conventional R\textsubscript{x} fire conducted during primarily dormant season
+				\item[]
+				\item Increased interest in burning during non-dormant season
+				\item[]
+				\begin{itemize}
+					\item Awareness of pre-colonial fire regimes
+					\item Diversify management
+				\end{itemize}
+
+			\end{itemize}
+
+		\end{column}
+		\begin{column}{0.5\textwidth}  
+			\begin{center}
+				\begin{figure}
+					\includegraphics[width=1\linewidth]{figs/StreeterRxFire}
+
+				\end{figure}
+			\end{center}
+		\end{column}
+	\end{columns}
+
+	%\alert{As with most components of fire regime, \textbf{seasonality is a construct}}
+\end{frame}
+
+\begin{frame}{For every season: Burn, burn, burn...}
+
+	\begin{columns}
+		\begin{column}{0.5\textwidth}
+			\begin{itemize} 
+				\item Conventional R\textsubscript{x} fire conducted during primarily dormant season
+				\item[]
+				\item Increased interest in burning during non-dormant season
+				\item[]
+				\begin{itemize}
+					\item Awareness of pre-colonial fire regimes
+					\item Diversify management
+				\end{itemize}
+
+			\end{itemize}
+
+		\end{column}
+		\begin{column}{0.5\textwidth}  
+			\begin{center}
+				\begin{figure}
+					\includegraphics[width=1\linewidth]{figs/StreeterRxFire}
+
+				\end{figure}
+			\end{center}
+		\end{column}
+	\end{columns}
+\alert{As with most fire regime parameters, \textbf{seasonality is a construct}}
+\end{frame}
+
 \begin{frame}{The Western fire regime concept } 
@@
 
-\begin{frame}{The Western fire regime concept}
+\begin{frame}{Controls over fire behavior and effects}
 	\begin{center}
@@
 
-\begin{frame}{Rx fire in the Northern Great Plains}
+\begin{frame}{Case study: R\textsubscript{x} fire in the Northern Great Plains}
 	\begin{center}
@@
 
-\begin{frame}{Rx fire in the Northern Great Plains}
+\begin{frame}{R\textsubscript{x} fire in the Northern Great Plains}
 	\begin{center}
@@
 
+\begin{frame}{Remotely-sensed burn severity products}
+	\begin{center}
+		\begin{figure}
+			\includegraphics[width=1\linewidth]{figs/severity_gg-1} 
+			\caption{An example from Dunn Ranch Prairie (McGranahan et al. 2022 \emph{Land})}
+		\end{figure}
+	\end{center}
+\end{frame}
+
 \begin{frame}{Spotty success in completing summer burns}
@@
 
+\begin{frame}{Barriers to summer R\textsubscript{x} fire } 
+	All opportunities and limitations fit within fire regime concept
+	\begin{itemize}
+		\item Biophysical 
+		\item[] \emph{mostly, \alert{too wet}}
+		\begin{itemize}
+			\item High live moisture fuel content (\emph{photosynthesis})
+			\item High dead moisture fuel content (\emph{humidity})
+			\item Poor convection/smoke dispersal (\emph{humidity})
+		\end{itemize}
+		\item[]
+		\item Social 
+		\item[] \emph{mostly, \alert{too dry}}
+		\begin{itemize}
+			\item Local burn restrictions
+			\item Control issues
+		\end{itemize}
+	\end{itemize}
+
+\end{frame}
+
 \section{A tale of two fires}
 
-\begin{frame}{What a difference a few days makes! } 
+\begin{frame}{What a difference a few days make! } 
 	\begin{columns}
@@
 		5 May 2018 & \textbf{Date} & 16 May 2018 \\
-		Weather & \textbf{Wind} & \\
-		Fuels & \textbf{RH} & Yes \\ \bottomrule
+		79 & \textbf{Air temp (F)} & 82 \\
+		5.9 & \textbf{Wind (m s\textsuperscript{-1})} & 2.4 \\
+		22 & \textbf{min RH (\%)} & 24 \\
+		37 & \textbf{Dew Point (F)} & 46 \\\bottomrule
 	\end{tabular}
@@
 
-\begin{frame}{Barriers to summer Rx fire } 
-	All opportunities and limitations fit within fire regime concept
+\section{Biophysical barriers}
+
+\begin{frame}{The terrible, humid, green, no-good day}
+
+	\begin{columns}
+		\begin{column}{0.6\textwidth}
+	\alert{13 August 2018}
 	\begin{itemize} 
-		\item Biophysical 
-		\item[] \emph{mostly, \alert{too wet}}
+		\item Forecast called for $<~40\%$ RH by noon, but...
 		\begin{itemize}
-			\item High live moisture fuel content (\emph{photosynthesis})
-			\item High dead moisture fuel content (\emph{humidity})
-			\item Poor convection/smoke dispersal (\emph{humidity})
+			\item 54\% at 1200
+			\item 50\% at 1500
 		\end{itemize}
 	\item[]
-		\item Social 
-		\item[] \emph{mostly, \alert{too dry}}
+		\item Modeled historical data (gridMET) says RH reached 28\%
+	\item[]
+		\item Local factors constrain ignition and spread
 				\begin{itemize}
-			\item Local burn restrictions
-			\item Control issues
-			\end{itemize}
+			\item Transpiration by green plants
+			\item Humid air resists heating, lift
 		\end{itemize}
 
+	\end{itemize}
+
+		\end{column}
+		\begin{column}{0.4\textwidth}  
+			\begin{center}
+				\begin{figure}
+					\includegraphics[width=0.9\linewidth]{figs/TorreGreenFuel}
+
+				\end{figure}
+			\end{center}
+		\end{column}
+	\end{columns}
 \end{frame}
-
-\section{Biophysical barriers}
 
@@
 
-\begin{frame}{Fire, grazing, and the pre-European landscape}
-\alert{Pyric herbivory}
-\begin{columns}
-\begin{column}{0.5\textwidth}
-\begin{itemize}
-	\item Recent burns have high-quality forage
-	\item[]
-	\item Grazers focus on burned areas
-	\item[]
-	\item Vegetation grows up in other areas...
-		\begin{itemize}
-			\item providing habitat
-			\item fueling future fires
-		\end{itemize}
-	\end{itemize}
-\end{column}
-\begin{column}{0.55\textwidth}  
+\begin{frame}{Fire regime management}
+	Difficult to satisfy multiple socially-constructed parameters (e.g. \emph{force patterns})
 	\begin{center}
 		\begin{figure}
- \includegraphics[width=1\linewidth]{figs/ghd}
-
- \end{figure}
-\end{center}
-\end{column}
-\end{columns}
-\end{frame}
-
-\begin{frame}{Disturbance-driven vs. inherent heterogeneity}
-\begin{center}
-\begin{figure}
- \includegraphics[width=1\linewidth]{figs/StreeterFire_17} 
+			\includegraphics[width=1\linewidth]{figs/venn} 
 		\end{figure}
@@
 
-\begin{frame}{Testing PBG 2.0 in the Northern Great Plains } 
-	\begin{columns}
-		\begin{column}{0.6\textwidth}
-			\begin{itemize}
-				\item \alert{Streeter (Central Grasslands)}
-				\begin{itemize}
-					\item \textbf{Compares grazing systems}
-					\item 4 pastures x 3 systems; cattle
-					\item \emph{Poa pratensis}-invaded mixed-grass prairie
-					\item 46 cm (18") annual rainfall
-				\end{itemize}
-				\item \alert{Hettinger}
-				\begin{itemize}
-					\item \textbf{Compares cattle \& sheep}
-					\item 3 x cattle, 3 x sheep PBG
-					\item Mostly non-native, cool-season grasses
-					\item post-CRP pasture stands
-					\item 40 cm (16") annual rainfall
-				\end{itemize}
-			\end{itemize}
-		\end{column}
-		\begin{column}{0.45\textwidth}  
+\begin{frame}{Fire regime management}
+	\vspace{-3em}
+	Better to tweak controllable components towards desired objectives (e.g. \emph{support processes})
 	\begin{center}
 		\begin{figure}
-					\includegraphics[width=1\linewidth]{figs/BobPasture}  
+			\includegraphics[width=1\linewidth]{figs/levers} 
 		\end{figure}
 	\end{center}
-		\end{column}
-	\end{columns}
+\end{frame}
+
+\begin{frame}{Fire regime management}
+	Better to tweak controllable components towards desired objectives (e.g. \emph{support processes})
+	\begin{center}
+		\begin{figure}
+			\includegraphics[width=1\linewidth]{figs/levers} 
+		\end{figure}
+	\end{center}
+\alert{Great Plains are in a \textbf{fire deficit} and management is best focused on \textbf{burning new acres}}
 \end{frame}
@@
 \begin{frame}{Mopping up}
-\alert{Prescribed fire has direct benefits to livestock production}
+
 \begin{columns}
@@
 \begin{itemize}
-	\item Livestock focus grazing in recently-burned patches
+	\item Fight the deficit: Burn what you can, when you can, but try to add new acres
 	\item[]
-	\item Burned areas have better forage
+		\item Focus on fine-scale levers to accomplish objectives regardless of season
 	\item[]
-	\item Fire has positive or neutral effects on soil
-	\item[]
-	\item Patch-burning enhances conservation outcomes, too
+	\item Consider emphasis on burn completeness: \emph{defend desired refugia}
+
 \end{itemize}

### 2023-10-10 SpringSummerSeverity/script/GetGridMET.R
surface update

@@ -52,7 +52,9 @@ pacman::p_load(foreach, doSNOW)
                                  mo_range[m], '-', 
                                  '30') )  %>% 
                 select(-source, -lat, -lon) %>%
-                mutate(hdwi = wind_vel * vpd) %>%
+                mutate(hdwi = wind_vel * vpd, 
+                       tmax = tmax = 275.15, 
+                       ) %>%
                 pivot_longer(names_to = "param", 
                              values_to = "value", 
                              -date) 

### 2023-10-10 SpringSummerSeverity/script/PlottingFireWx.R

@@ -1,9 +1,49 @@
-
+pacman::p_load(tidyverse, sf)
 
 load('./data/HistWx.Rdata')
 
+# Example days 
+
+HistWx %>%
+  filter(as.character(date) %in% c('2018-05-05', '2018-05-16', '2018-08-13')) %>%
+  pivot_wider(names_from = param, 
+              values_from = value) %>% View() 
+
+ndawn <- 
+read_csv('./data/StreeterStudyPeriodNDAWN.csv')
+
+ndawn %>%
+  select(-`Station Name`, -Latitude, -Longitude, -Elevation) %>%
+unite(col = 'date', c(Year, Month, Day), sep = '-') %>%
+filter(date %in% c('2018-5-5', '2018-5-16', '2018-8-13')) %>%
+  pivot_wider(names_from = param, 
+              values_from = value) %>% View() 
+
+ndawn %>%
+  select(-`Station Name`, -Latitude, -Longitude, -Elevation) %>%
+  unite(col = 'date', c(Year, Month, Day), sep = '-') %>%
+  filter(date %in% c('2018-5-5', '2018-5-16'),  
+         Hour %in% c(1500)) 
+
+ndawn %>%
+  select(-`Station Name`, -Latitude, -Longitude, -Elevation) %>%
+  unite(col = 'date', c(Year, Month, Day), sep = '-') %>%
+  filter(date %in% c('2018-5-5', '2018-5-16'),  
+         Hour %in% c(1500)) 
+
+
+ndawn %>%
+  select(-`Station Name`, -Latitude, -Longitude, -Elevation) %>%
+  unite(col = 'date', c(Year, Month, Day), sep = '-') %>%
+  filter(date %in% c('2018-5-5', '2018-5-16')) %>%  
+  group_by(date) %>%
+    summarise(RH = min(RH) )
+
+
 HistWx %>%
   separate(date, c('year', 'month', 'day'), sep = "-", remove = F) %>%
+  filter(year == '2018') %>%
+  arrange(desc(as.numeric(month)) ) 
   mutate(season = case_when(
     month %in% c('04', '05') ~ 'spring',
     month %in% c('08') ~ 'summer'), 

### 2023-10-10 SpringSummerSeverity/script/boundaries.R

@@ -16,3 +16,37 @@ fp = './gis/boundaries'
     filter(Unit %in% c('Bob', 'Barker')) %>%
     write_sf(paste0(fp, '/CGREC_pbg_26913.shp'), 
              append = FALSE)
+  
+  # Fire mapping 
+
+  cgrec_gpkg = './gis/boundaries/CGREC_PBG_26914.gpkg'
+  st_layers(cgrec_gpkg ) 
+  
+  pastures <- st_read(cgrec_gpkg, 'Pastures') 
+  patches <- st_read(cgrec_gpkg, 'PasturePatches') 
+
+  fires <- st_read(cgrec_gpkg, 'FirePerimeters') %>%
+              mutate(Year = as.factor(Year))
+
+    ggplot() + theme_void() +
+      geom_sf(data = filter(fires, unit == 'Barker'), 
+      aes(fill = Year, 
+          alpha = Season)) +
+      geom_sf(data = filter(patches, Unit == 'Barker'), 
+              fill = NA, 
+              color = 'black', 
+              lty = 2, 
+              size = 1) + 
+      geom_sf(data = filter(pastures, Unit == 'Barker'), 
+              fill = NA, 
+              color = 'black', 
+              lty = 1, 
+              size = 2) +
+      geom_sf_text(data = filter(fires, unit == 'Barker'), 
+                   aes(label = Year), 
+                   nudge_y = 50) +
+      geom_sf_text(data = filter(fires, unit == 'Barker'), 
+                   aes(label = substr(Season, 1, 3)), 
+                   nudge_y = -50) +
+      scale_alpha_manual(values = c(0.25, 0.8)) + 
+      theme(legend.position = 'none')

### 2024-10-29 SpringSummerSeverity/writeup.Rmd
updates

---
title: Spring vs. summer fire regimes in central North Dakota
author: DA McGranahan & JP Angerer??
output: 
  html_document:
    highlight: tango
---

```{r setup, echo=FALSE, warning=FALSE, message = FALSE}
knitr::opts_chunk$set(message = FALSE, warning=FALSE, 
                      echo=FALSE, eval=TRUE)
pacman::p_load(tidyverse)
# load data 
  load('./data/NoFireIndices.Rdata')
  load('./data/BurnIndices.Rdata')
# some necessary things
  seasons <- 
    tibble(season = c('Spring', 'Summer'), 
           start = c(as.Date('2024-04-25'), as.Date('2024-08-07')), 
           end = c(as.Date('2024-05-25'), as.Date('2024-08-27'))) 
```

# In a nutshell

## Background

* From 2017-2020, a project was conducted to restore spatially-patchy Rx fire in Central ND
* Two burning regimes:
  - 4 x 40-acre, full-patch burns each spring 
  - 4 x 20-acre, half-patch burns, first half in spring, second half in summer
* Mixed success in restoring fire: 
  - All spring burns were completed in each year 
  - Summer burns only completed in 2/4 years *but not for lack of trying*

## Questions

* Are there biophysical limitations on summer fires? 
  - *e.g., maybe it can just be too green and/or too wet and/or too humid to burn*
* Can trends in vegetation and weather explain why summer burns were successful in some years and not in others? 
  - *If so, maybe one can predict whether a summer burn is worth trying or not, or predicting summer wildfire risk*
* Potential implications for intentions behind growing season burns vs. pre-European summer fire regimes?
  - *Folks recognize pre-European fire regimes included considerable summer fire, and talk about greater variability in burn severity with growing-season fire as a biodiversity-friendly reason to burnin the summer. But what if pre-European summer burns were actually quite intense and burned severely because they actually only occurred under drought conditions?*

## Objectives

* Explore patterns in greenness and humidity to determine effects on fire effectiveness
* Identify anomalies and ranges of conditions conducive to prescribed fire

# Remotely-sensed data

Pulled burn severity (NBR) and several vegetation indices from Sentinel-2 imagery for 2017-2020 growing seasons. 
Data sampled from regular gridded points within burn units

Vegetation indices:

* NDVI (Normalized Difference Vegetation Index)---Measure of vegetation greenness
* NDMI (Normalized Difference Moisture Index)---Measure of moisture content of vegetation
* MSI (Moisture Stress Index)---Another measure of moisture content sometimes used in fire risk analysis. Higher = drier

## Vegetation data

These data are from four *unburned* pastures in an adjacent continuous grazing treatment to remove any confounding effects of prescribed burns.

### Seasonal trends

Summer fires were successful in 2017 and 2020. 
Vegetation in both years was less green and generally more dry.

```{r veg_trends, fig.cap="Blue shaded areas span the first and last dates of completed burns in the spring and summer.", fig.height = 8, fig.width = 10}
NoFireIndices %>%
  pivot_longer(names_to = 'index', 
               values_to = 'value', 
               cols = c(ndvi:ndmi)) %>%
  group_by(ImageDate, pasture, index) %>%
    summarise(Mean = mean(value), 
              SEM = sd(value)/sqrt(n()), 
              .groups = 'drop') %>%
  mutate(ImageDate = as.Date(ImageDate, '%Y-%m-%d'), 
         Year = lubridate::year(ImageDate), 
         Year = as.factor(Year), 
         SampleDate = format(ImageDate, '%m-%d'), 
         SampleDate = as.Date(SampleDate, '%m-%d'), 
         index = toupper(index), 
          index = fct_rev(index)) %>%
  filter(Year != 2016) %>%
ggplot(aes(x = SampleDate, 
           color = Year, 
           fill = Year, 
           group = interaction(pasture, Year))) + theme_bw(14) +
  geom_rect(data= seasons, inherit.aes = FALSE,
              aes(xmin=start, xmax=end, ymin=-Inf, ymax=+Inf), 
              fill='lightblue', alpha=0.5) +
  geom_smooth(aes(y = Mean, group = Year),
              se = T,
              level = 0.99) +
  geom_errorbar(aes(ymin = Mean - SEM, 
                    ymax = Mean + SEM),
                alpha = 0.5) +
  geom_line(aes(y = Mean),
             alpha = 0.5) +
  labs(x = 'Image date', y = 'Index value') + 
  facet_wrap(~index, ncol=1, scale = 'free_y') +
  scale_x_date( date_breaks = 'month', 
                date_labels = '%b-%d', 
                minor_breaks = NULL) 
```

### Seasons summarized

* Greenness:
  - NDVI during all spring burns generally pretty low
  - Successful summer burn years higher greenness than any spring...
  - ... still slightly but distinctly lower than unsuccessful years

* Moisture content: 
  - High variability in spring seasons across years
  - Successful summer seasons drier than in unsuccessful years

```{r veg_by_season, fig.cap="Data summarized by year from shaded blue ranges above.", fig.height = 4, fig.width = 10}
# Get average values for burn seasons 
  NoFireIndices %>%
    pivot_longer(names_to = 'index', 
                 values_to = 'value', 
                 cols = c(ndvi:ndmi)) %>%
    group_by(ImageDate, pasture, index) %>%
    summarise(Median = median(value) , 
              .groups = 'drop') %>%
    mutate(ImageDate = as.Date(ImageDate, '%Y-%m-%d'), 
           Year = lubridate::year(ImageDate), 
           Year = as.factor(Year), 
           SampleDate = format(ImageDate, '%m-%d'), 
           SampleDate = as.Date(SampleDate, '%m-%d'), 
           season = case_when(
             between(SampleDate, seasons$start[1], seasons$end[1]) ~ 'Spring',
             between(SampleDate, seasons$start[2], seasons$end[2]) ~ 'Summer', 
             TRUE ~ NA ) ) %>%
    filter(!is.na(season)) %>%
    group_by(season, Year, index) %>%
    summarize(SeasonMean = mean(Median), 
              SeasonSEM = sd(Median)/sqrt(n()), 
              .groups = 'drop') %>%
    mutate(status = case_when(
      season == 'Summer' & Year %in% c('2018', '2019') ~ "Not completed", 
      TRUE ~ "Completed"), 
          index = toupper(index), 
          index = fct_rev(index)) %>%
    ggplot(aes(x = season, y = SeasonMean, color = Year)) + theme_bw(14) +
    geom_errorbar(aes(ymin = SeasonMean - SeasonSEM, 
                      ymax = SeasonMean + SeasonSEM), 
                  width = 0.25, lwd = 0.95, 
                  position = position_dodge(width = 0.33), 
                  key_glyph = "point")  +
    geom_point(aes(pch = interaction(status, Year)), 
               position = position_dodge(width = 0.33), 
                size = 4, stroke = 2, show.legend = F) +
    labs(x = 'Burn season', 
         y = 'Index value', 
         caption = 'X = Year/season in which burns were not completed') + 
    scale_shape_manual(values = c(16, 16, 4, 16, 4, 16)) + 
    facet_wrap(~index, ncol = 3, scale = "free") +
    theme(axis.text.x = element_text(color = 'black')) +
    guides(color=guide_legend(override.aes = list(pch=15, size =4))) 
```

## Burn severity

Overall it appears greener fuels consistently result in less-complete burns (although 2018 was weird).

Trends between fuel moisture content and severity were less consistent. 
Again 2018 behaved strangely---fuel in 4 burn units in 1 block of pastures was extremely dry but still didn't burn with particularly high severity; less-dry fuel in the other block tended to burn slightly more severely. 
The dry block was burned a couple weeks earlier than the other block.

```{r sev_veg, fig.cap="Vegetation indices pulled from closest pre-burn image, same as pre-burn NBR.", fig.height = 3.8, fig.width = 10}
BurnIndices  %>%
  pivot_longer(names_to = 'name', 
               values_to = 'index', 
               cols = c(ndvi, msi, ndmi)) %>%
  group_by(Year, Season, fire, name) %>%
  summarise_at(.vars = vars(dNBR, index),
               .funs = c(Mean="mean", 
                         SEM = function(x) {sd(x)/sqrt(n()) } )) %>%
  ungroup() %>%
  mutate(name = toupper(name), 
         name = fct_rev(name)) %>% 
  filter(Season == "Spring") %>%
  ggplot(aes(x = index_Mean, y = dNBR_Mean, color = Year)) + theme_bw(14) +
  geom_smooth(aes(x = index_Mean, y = dNBR_Mean), 
              method = 'lm', 
              color = 'black',
              se = T, alpha = 0.2) +
  geom_errorbar(aes(ymin = dNBR_Mean - dNBR_SEM, 
                    ymax = dNBR_Mean + dNBR_SEM)) +
  geom_errorbarh(aes(xmin = index_Mean - index_SEM, 
                     xmax = index_Mean + index_SEM)) +
  geom_point() + 
  labs(x = 'Pre-burn index value', 
       y = 'Burn severity (dNBR)') + 
  geom_smooth(method = 'lm', se = F) +
  facet_wrap(~name, scales = 'free')
```

```{r, eval = FALSE}
full_join(by = 'fire', 
fires %>%
  as_tibble() %>%
  filter(Year == '2018') %>%
  select(fire, PreBurn),

BurnIndices  %>%
  pivot_longer(names_to = 'name', 
               values_to = 'index', 
               cols = c(ndvi, msi, ndmi)) %>%
  group_by(Year, Season, fire, name) %>%
  summarise_at(.vars = vars(dNBR, index),
               .funs = c(Mean="mean", 
                         SEM = function(x) {sd(x)/sqrt(n()) } )) %>%
  ungroup() %>%
  filter(Year == '2018') %>%
  select(fire, name, dNBR_Mean, index_Mean) %>%
  pivot_wider(names_from = name, 
              values_from = index_Mean) ) %>%
  rename(dNBR = dNBR_Mean) %>%
  arrange(desc(msi)) 
```

# Weather data

*Pending*

### 2024-10-29 SpringSummerSeverity/script/ChuggingImagery.R

@@ -9,98 +9,88 @@ patches <- st_read(cgrec_gpkg, 'PasturePatches')
 NoFirePts <- st_read(cgrec_gpkg, 'SamplePoints') %>%
               filter(location == 'Refuge')
 
-# Get NDVI data for unburned pastures
-  ndvi_dir = 'S:/DevanMcG/Projects/SpringSummerSeverity/sentinel/NDVI'
-  ndvi_images <- list.files(ndvi_dir)
+# Get veg data for unburned pastures
+  imagery_dir = 'S:/DevanMcG/Projects/SpringSummerSeverity/sentinel/MultiBand'
+  images <- list.files(imagery_dir)
   
   { 
     begin = Sys.time() 
     pacman::p_load(foreach, doSNOW)
     cores = parallel::detectCores()
-    cl <- makeCluster(cores -1 , methods = F, useXDR = F)
+    cl <- makeCluster(cores, methods = F, useXDR = F)
     registerDoSNOW(cl)
-    NoFireNDVI <-
-      foreach(i=1:length(ndvi_images), 
+    NoFireIndices <-
+      foreach(i=1:length(images), 
               .combine = 'bind_rows',
               .errorhandling = 'remove', 
               .packages=c('tidyverse', 'sf')) %dopar% {
-        image = ndvi_images[i]
-        image_path = paste0(ndvi_dir, '/', image)
+        image = images[i]
+        image_path = paste0(imagery_dir, '/', image)
         ras <- terra::rast(image_path)
-        names(ras) <- 'ndvi'
-        ndvi_float = (ras-10000)/10000
-        terra::extract( ndvi_float,
+        names(ras) <- c('nbr', 'ndvi', 'msavi', 'msi', 'ndmi') 
+        ras <- ras[[c('ndvi', 'msi','ndmi')]]
+        float = (ras-10000)/10000
+        terra::extract(float,
           NoFirePts %>%
             select(pasture, sample) %>%
             terra::vect() , 
           FUN = mean, 
           bind = TRUE)  %>%
         st_as_sf() %>%
+          as_tibble() %>%
         mutate(ImageDate = substr(image, 1, 10)) %>%
-        as_tibble() %>%
-        select(ImageDate, pasture, sample, ndvi) 
+        select(ImageDate, pasture, sample, ndvi:ndmi) 
             }
     stopCluster(cl)
   Sys.time() - begin 
   }
-
-NoFireNDVI %>%
[…]

### 2024-10-29 SpringSummerSeverity/script/rgee_landsat_refuge.R

@@ -1,0 +1,180 @@
+pacman::p_load(tidyverse, sf, rgee, rgeeExtra)
+#ee_Initialize(drive = TRUE)
+ee$Initialize(project='ee-devanmcg')
+rgeeExtra::extra_Initialize()
+
+# CGREC scene days to fetch from GEE
+
+  passes <- 
+    readxl::read_xlsx('./data/BurnData.xlsx', 'SentinelDays') %>%
+    unite('PassDate', sep = '-') %>%
+    mutate(PassDate = as.Date(PassDate, '%Y-%m-%d'), 
+           PassDate = as.character(PassDate)) 
+  
+  passes <- as.vector(unique(passes))[[1]]
+
+# CGREC boundaries 
+  cgrec_gpkg = './gis/boundaries/CGREC_PBG_26914.gpkg'
+  st_layers(cgrec_gpkg ) 
+  
+  pastures <- st_read(cgrec_gpkg, 'Pastures') %>%
+                mutate(Unit = recode(Unit, 'Bob' = 'North', 'Barker' = 'South'))
+  patches <- st_read(cgrec_gpkg, 'PasturePatches') %>%
+                mutate(Unit = recode(Unit, 'Bob' = 'North', 'Barker' = 'South'))
+  fires <- st_read(cgrec_gpkg, 'FirePerimeters') %>%
+            mutate(Year = as.factor(Year), 
+                   unit = recode(unit, 'Bob' = 'North', 'Barker' = 'South'))
+  # Create GEE geometries
+    CGREC_PBG <-
+      pastures %>%
+        st_union() %>%
+        st_transform(4326) %>% 
+        sf_as_ee()
+    CGREC_NorthPBG <-
+      pastures %>%
+        filter(Unit == 'North') %>%
+        st_union() %>%
+        st_transform(4326) %>% 
+        sf_as_ee()
+    CGREC_SouthPBG <-
+      pastures %>%
+        filter(Unit == 'South') %>%
+        st_union() %>%
+        st_transform(4326) %>%
+        sf_as_ee()
+    refuge_ee <- st_read(cgrec_gpkg, 'SamplePoints') %>%
+      filter(location == 'Refuge')  %>%
+      st_transform(4326) %>%
+      st_union() %>%
+      st_bbox() %>%
+      st_as_sfc(., crs=4326) %>%
+      sf_as_ee()
+
+# Turn on GEE connection
+  #ee_Initialize(drive = TRUE)
[…]

### 2024-11-25 SpringSummerSeverity/paper/FigureGenerator.Rmd
SpringSummerSeverity paper draft

---
title: Evaluating an attempt to restore summer fire in the Northern Great Plains
author: DA McGranahan & JP Angerer??
output: 
  html_document:
    highlight: tango
---

```{r setup, echo=FALSE, warning=FALSE, message = FALSE}
knitr::opts_chunk$set(message = FALSE, warning=FALSE, 
                      echo=FALSE, eval=TRUE, fig.path = '', dev='png' )
pacman::p_load(tidyverse, cowplot, patchwork, wesanderson)
  BuGnPal = c( "#67A9CF" ,"#3690C0" ,"#02818A", "#016450")
  FFgrad = wes_palette("FantasticFox1")[c(3,2,4,5)]
# load data 
  load('../data/NoFireIndices.Rdata')
  load('../data/BurnIndices.Rdata')
  load('../data/NoFireTrends.Rdata')
  load('../data/WxData.Rdata')
# some necessary things
  seasons <- 
    tibble(season = c('Spring', 'Summer'), 
           start = c(as.Date('2024-04-25'), as.Date('2024-08-07')), 
           end = c(as.Date('2024-05-25'), as.Date('2024-08-27'))) 
```

```{r}
LandsatSums <-
  NoFireTrends  %>%
    separate(ImageDate, into=c('year', 'month', 'day'), sep = '-') %>%
    mutate(season = ifelse(month=='04', 'Spring', 'Summer')) %>%
    pivot_longer(names_to = 'index', 
                 values_to = 'value', 
                 cols = c(ndvi:ndmi)) %>%
    group_by(year, season, pasture, index) %>%
    summarize(Value = mean(value), 
              .groups = 'drop') 
```

# In a nutshell

## Background

* From 2017-2020, a project was conducted to restore spatially-patchy Rx fire in Central ND
* Two burning regimes:
  - 4 x 40-acre, full-patch burns each spring 
  - 4 x 20-acre, half-patch burns, first half in spring, second half in summer
* Mixed success in restoring fire: 
  - All spring burns were completed in each year 
  - Summer burns only completed in 2/4 years *but not for lack of trying*

## Questions

* Are there biophysical limitations on summer fires? 
  - *e.g., maybe it can just be too green and/or too wet and/or too humid to burn*
* Can trends in vegetation and weather explain why summer burns were successful in some years and not in others? 
  - *If so, maybe one can predict whether a summer burn is worth trying or not, or predicting summer wildfire risk*
* Potential implications for intentions behind growing season burns vs. pre-European summer fire regimes?
  - *Folks recognize pre-European fire regimes included considerable summer fire, and talk about greater variability in burn severity with growing-season fire as a biodiversity-friendly reason to burnin the summer. But what if pre-European summer burns were actually quite intense and burned severely because they actually only occurred under drought conditions?*

## Objectives

* Explore patterns in fuel greenness and fire weather to determine effects on fire effectiveness
* Identify anomalies and ranges of conditions conducive to prescribed fire

## Data

Pulled burn severity (NBR) and NDVI from Sentinel-2 imagery for 2017-2020 growing seasons.
Also pulled historical (1990-2022) NDVI from Landsat for long-term trends. 
NDVI = Normalized Difference Vegetation Index; Measure of vegetation greenness.
Data sampled from regular gridded points within burn units.

Also retrieved historical hourly weather data, 1990-2022, from local mesonet station.

# Results

## Vegetation data

```{r ndvi_gg, fig.cap="Sentinel imagery summarized for the study years. A: NDVI data from four adjacent unburned pastures avoids any confounding effects of prescribed burns. X denotes seasons in which burns were not completed. B: Burn severity (dNBR) plotted against fuelbed greenness (NDVI) retrieved from the pre-burn Sentinel-2 image used for pre-fire NBR. Points represent mean for individual prescribed fires with associated standard error; circles and solid trendline indicate spring burns while broken trendline and triangles denote summer burns.", fig.height = 7, fig.width = 5}
# Get average values for burn seasons 
p1 <-
  NoFireIndices %>%
    group_by(ImageDate, pasture) %>%
    summarise(NDVI = median(ndvi) , 
              .groups = 'drop') %>%
    mutate(ImageDate = as.Date(ImageDate, '%Y-%m-%d'), 
           Year = lubridate::year(ImageDate), 
           Year = as.factor(Year), 
           SampleDate = format(ImageDate, '%m-%d'), 
           SampleDate = as.Date(SampleDate, '%m-%d'), 
           season = case_when(
             between(SampleDate, seasons$start[1], seasons$end[1]) ~ 'Spring',
             between(SampleDate, seasons$start[2], seasons$end[2]) ~ 'Summer', 
             TRUE ~ NA ) ) %>%
    filter(!is.na(season)) %>%
    group_by(season, Year) %>%
    summarize(SeasonMean = mean(NDVI), 
              SeasonSEM = sd(NDVI)/sqrt(n()), 
              .groups = 'drop') %>%
    mutate(status = case_when(
      season == 'Summer' & Year %in% c('2018', '2019') ~ "Not completed", 
      TRUE ~ "Completed") ) %>%  
    ggplot(aes(x = season, color = Year)) + theme_bw(14) +
    geom_errorbar(aes(ymin = SeasonMean - SeasonSEM, 
                      ymax = SeasonMean + SeasonSEM), 
                  width = 0.25, lwd = 0.95, 
                  alpha = 0.5,
                  position = position_dodge(width = 0.33), 
                  key_glyph = "point")  +
    geom_point(aes( y = SeasonMean, 
                    pch = interaction(status, Year)), 
               position = position_dodge(width = 0.33), 
                size = 4, stroke = 2, show.legend = F) +
    labs(x = 'Burn season', 
         y = 'NDVI', 
         ) + 
    scale_shape_manual(values = c(16, 16, 4, 16, 4, 16)) + 
   scale_color_manual(values = FFgrad) +  
    theme(axis.text.x = element_text(color = 'black')    ) +
    guides(color=guide_legend(override.aes = list(pch=15, alpha = 1, size =4))) 
p2 <-
BurnIndices  %>%
  group_by(Year, Season, fire) %>%
  summarise_at(.vars = vars(dNBR, ndvi),
               .funs = c(Mean="mean", 
                         SEM = function(x) {sd(x)/sqrt(n()) } )) %>%
  ungroup() %>%
  ggplot(aes(x = ndvi_Mean, y = dNBR_Mean, color = Year)) + theme_bw(14) +
  geom_smooth(data = . %>%
                filter(Season == 'Spring'), 
              aes(x = ndvi_Mean, y = dNBR_Mean, lty=Season), 
              method = 'lm', 
              color = 'black',
              se = T, alpha = 0.2) +
    geom_smooth(data = . %>%
                filter(Season == 'Summer'), 
              aes(x = ndvi_Mean, y = dNBR_Mean), 
              method = 'lm', 
              lty = 2,
              color = 'black',
              se = F, alpha = 0.2) +
  geom_errorbar(aes(ymin = dNBR_Mean - dNBR_SEM, 
                    ymax = dNBR_Mean + dNBR_SEM)) +
  geom_errorbarh(aes(xmin = ndvi_Mean - ndvi_SEM, 
                     xmax = ndvi_Mean + ndvi_SEM)) +
  geom_point(aes(fill = Year, 
                 shape = Season), 
             color = 'black',
             size = 3) + 
  scale_shape_manual(values=c(21, 24)) +
   scale_fill_manual(values = FFgrad) +
   scale_color_manual(values = FFgrad) +  
  labs(x = 'Pre-burn NDVI', 
       y = 'Burn severity (dNBR)') +
  theme(  legend.position = 'none', 
          legend.position.inside = c(0.8, 0.85))

plot_grid(p1, p2,  labels = "AUTO", 
          ncol = 1, rel_heights = c(0.8, 1))
```

```{r green_trends, fig.cap="Study season data (points) plotted within the context of 42-year historical trends (boxplots). X indicates seasons prescribed fires were not completed. A: NDVI data retrieved from Landsat imagery. B: Annual growing season rainfall accummulated by two specific dates associated with the spring and summer burn seasons, respectively.", fig.width = 5, fig.height = 7}
p3 <-
LandsatSums %>%
                            filter(year %in% c(2017:2020), 
                                   index == 'ndvi') %>%
                            group_by(year, season) %>%
                            summarize(Mean = mean(Value),
                                      SEM = sd(Value) / sqrt(n()),
                                      .groups = 'drop') %>%
                            mutate(status = case_when(
                              season == 'Summer' & 
                              year %in% c('2018', '2019') ~ "Not completed", 
                              TRUE ~ "Completed") )%>% 
  ggplot(aes(x = season)) + theme_bw(14) + 
      geom_boxplot(data = LandsatSums %>%
                     filter(index == 'ndvi')  ,
                 aes( y = Value),
                 outliers = F,
                 staplewidth = 0.25) +
  # geom_errorbar(aes( ymin = Mean - SEM,
  #                    ymax = Mean + SEM,
  #                    color = year),
  #               width = 0.5,
  #               position = position_dodge(width = 0.25)) +
         geom_point(#data = , 
        aes(y = Mean, 
                color = year, shape = status), 
             position = position_dodge(width = 0.25), 
             size = 4, stroke = 3, show.legend = T) + 
    labs(x = 'Burn season', y = 'NDVI') + 
       scale_color_manual("Year", values = FFgrad) + 
  scale_shape_manual(values = c(16, 4), guide = 'none') +

  theme(axis.text.x = element_text(color = 'black'), 
        panel.grid.major.x = element_blank() ) +
    guides(color=guide_legend(override.aes = list(pch=15, size =4)))
p4 <- 
    WxData$Rainfall %>% 
    filter(month(date) %in% c(04:09)) %>%
    separate(date, into = c('Year', 'Month', 'Day'),sep = '-') %>%
    group_by(Year) %>%
    mutate(CumPrecip = cumsum(Rainfall), 
           Month = recode(Month, '05' ="May", '08'='Aug'), 
           Month = paste0(Month, ' ', Day)) %>%
    filter(Month %in% c('May 15', 'Aug 01' )) %>%
    mutate(Month = fct_rev(Month)) %>%
    ggplot() + theme_bw(14) + 
      geom_boxplot(aes(x = Month, y = CumPrecip),
                   outliers = F,
                   staplewidth = 0.25) +
    geom_point(data = . %>%
                 filter(as.numeric(Year) %in% c(2017:2020)) %>%
                  mutate(status = case_when(
                    Month == 'Aug 01' & 
                      Year %in% c('2018', '2019') ~ "Not completed", 
                    TRUE ~ "Completed") ), 
            aes(x = Month, y = CumPrecip,
                color = Year, shape = status), 
            position = position_dodge(width = 0.25), 
            size = 4, stroke = 3, show.legend = T) + 
  labs(x = 'Date', y = 'Cumulative Rainfall (mm)') + 
  scale_shape_manual(values = c(16, 4), guide = 'none') +
  scale_color_manual("Year", values = FFgrad) + 
  theme(axis.text.x = element_text(color = 'black'), 
        panel.grid.major.x = element_blank() ) +
  guides(color=guide_legend(override.aes = list(pch=15, size =4)))

p3 / p4 +
  patchwork::plot_layout(
    guides = "collect",
    nrow = 2
  ) +
  patchwork::plot_annotation(
    tag_levels = "A"
  ) &
  theme(legend.position = "right", legend.direction = "vertical")
```

## Weather data

```{r weather, fig.cap="Four fire weather variables plotted by burn day values (open triangles), study season means (open circles and X for seasons without completed burns), and 42-year historical trends (boxplots).", fig.height = 9}
RainyDays <- 
   WxData$Rainfall %>%
      filter( ! is.na(season), Rainfall > 0.1) 
# Seasonal trends over time   
  WxData$HourlyWx %>%
    filter(! date %in% RainyDays$date, # remove rainy days with no chance of burning
           ! is.na(season)) %>%
    group_by(Year, date, season) %>%
    summarise_at(.vars = vars(c(DewPoint, RelHum, WindSpeed,VPD)),
                 .funs = c("mean")) %>%
    ungroup() %>%
    pivot_longer(names_to = 'variable', 
                 values_to = 'value', 
                 cols = DewPoint:VPD) %>%
    group_by(Year, season, variable) %>%
    summarise(Value = mean(value), 
              .groups = 'drop') %>%
    mutate(variable = recode(variable, 
                             'DewPoint' = 'Dew Point (°C)', 
                             'RelHum' = 'Relative Humidity (%)', 
                             'VPD' = 'Vapor Pressure Deficit', 
                             'WindSpeed' = 'Wind speed (m/s)') , 
           Year = as.character(Year), 
           season = fct_rev(season)) %>%
    ggplot() + theme_bw(16) + 
      geom_boxplot( aes(x=season, 
                       y = Value),
                   outliers = F, size = 0.75,
                   staplewidth = 0.25) +
      geom_point(data = . %>%
                   filter(Year %in% c(2017:2020)) %>%
                   mutate(status = case_when(
                     season == 'Summer' & Year %in% c('2018', '2019') ~ "Not completed", 
                     TRUE ~ "Completed") ) , 
                 aes(x = season, y = Value, 
                     color = Year, shape = status), 
                 position = position_nudge(x = -.2), 
                  fill = 'grey90', 
                 size = 3, stroke = 2, show.legend = T) + 
    geom_point(data = WxData$BurnDayWx , 
               aes(x = season, y = value, color =Year), 
               position = position_nudge(x = +.2), 
               shape = 24, fill = 'grey90', 
               size = 3, stroke = 2,
                show.legend = F) + 
      labs(x = 'Burn season', y = 'Weather variable value') + 
      facet_wrap(~variable, scales = 'free_x', ncol = 1)  +
      scale_shape_manual(values = c(21, 4), guide = 'none') +
    scale_color_manual(values = FFgrad) +  
      theme(axis.text.y = element_text(color = 'black', angle = 45, hjust = 0.5), 
            panel.grid.major.y = element_blank() , 
            strip.text = element_text(face='bold'), 
            strip.background = element_rect(fill = '#A1D99B')) +
    coord_flip() + 
      guides(color=guide_legend(override.aes = list(pch=15, size =4)))
```

# Supplemenary Info

## Seasonal trends

Summer fires were successful in 2017 and 2020. 
Vegetation in both years was less green and rainfall lower.

```{r ndvi_trend, eval = TRUE, fig.cap="Green shaded areas span the first and last dates of completed burns in the spring and summer.", fig.height = 4, fig.width = 8}
NoFireIndices %>%
  group_by(ImageDate, pasture) %>%
    summarise(Mean = mean(ndvi), 
              SEM = sd(ndvi)/sqrt(n()), 
              .groups = 'drop') %>%
  mutate(ImageDate = as.Date(ImageDate, '%Y-%m-%d'), 
         Year = lubridate::year(ImageDate), 
         Year = as.factor(Year), 
         SampleDate = format(ImageDate, '%m-%d'), 
         SampleDate = as.Date(SampleDate, '%m-%d') ) %>%
  filter(Year != 2016) %>%
ggplot(aes(x = SampleDate, 
           color = Year, 
           fill = Year, 
           group = interaction(pasture, Year))) + theme_bw(14) +
  geom_rect(data= seasons, inherit.aes = FALSE,
              aes(xmin=start, xmax=end, ymin=-Inf, ymax=+Inf), 
              fill='#A1D99B', alpha=0.5) +
  geom_smooth(aes(y = Mean, group = Year),
              se = T,
              level = 0.99) +
  geom_errorbar(aes(ymin = Mean - SEM, 
                    ymax = Mean + SEM),
                alpha = 0.5) +
  geom_line(aes(y = Mean),
             alpha = 0.5) +
  labs(x = 'Image date', y = 'NDVI') + 
    scale_fill_manual(values = FFgrad) + 
  scale_color_manual(values = FFgrad) + 
  scale_x_date( date_breaks = 'month', 
                date_labels = '%b-%d', 
                minor_breaks = NULL) 
```

```{r precip_trend, eval = TRUE, fig.cap="Thick black line shows smoothed 40-yr trend. Green shaded areas span the first and last dates of completed burns in the spring and summer.", fig.height = 4, fig.width = 8}
  WxData$Rainfall %>% 
    filter(month(date) %in% c(04:09)) %>%
    separate(date, into = c('Year', 'Month', 'Day'),sep = '-') %>%
    group_by(Year) %>%
    mutate(CumPrecip = cumsum(Rainfall)) %>%
    ggplot(aes(x = day)) + theme_bw(14) +
    geom_rect(data= seasons, inherit.aes = FALSE,
              aes(xmin=start, xmax=end, ymin=-Inf, ymax=+Inf), 
              fill='#A1D99B', alpha=0.5) +
    geom_line(aes(y = CumPrecip, group = Year),
              color = 'grey40', 
              alpha = 0.5) +
    geom_smooth(aes(y = CumPrecip),
                se = F,
                color = 'black',
                size = 1,
                level = 0.99) +
    geom_line(data = . %>%
                filter(as.numeric(Year) %in% c(2017:2020)) , 
              aes(y = CumPrecip, 
                  color = Year,
                  group = Year),
              size = 2) +
    labs(x = 'Day of year', y = 'Cumulative rainfall (mm)') + 
    scale_fill_manual(values = FFgrad) + 
    scale_color_manual(values = FFgrad) + 
    scale_x_date( date_breaks = 'month', 
                  date_labels = '%b-%d', 
                  minor_breaks = NULL)
```

```{r veg_trends, eval = FALSE, fig.cap="Blue shaded areas span the first and last dates of completed burns in the spring and summer.", fig.height = 8, fig.width = 10}
NoFireIndices %>%
  pivot_longer(names_to = 'index', 
               values_to = 'value', 
               cols = c(ndvi:ndmi)) %>%
  group_by(ImageDate, pasture, index) %>%
    summarise(Mean = mean(value), 
              SEM = sd(value)/sqrt(n()), 
              .groups = 'drop') %>%
  mutate(ImageDate = as.Date(ImageDate, '%Y-%m-%d'), 
         Year = lubridate::year(ImageDate), 
         Year = as.factor(Year), 
         SampleDate = format(ImageDate, '%m-%d'), 
         SampleDate = as.Date(SampleDate, '%m-%d'), 
         index = toupper(index), 
          index = fct_rev(index)) %>%
  filter(Year != 2016) %>%
ggplot(aes(x = SampleDate, 
           color = Year, 
           fill = Year, 
           group = interaction(pasture, Year))) + theme_bw(14) +
  geom_rect(data= seasons, inherit.aes = FALSE,
              aes(xmin=start, xmax=end, ymin=-Inf, ymax=+Inf), 
              fill='#A1D99B', alpha=0.5) +
  geom_smooth(aes(y = Mean, group = Year),
              se = T,
              level = 0.99) +
  geom_errorbar(aes(ymin = Mean - SEM, 
                    ymax = Mean + SEM),
                alpha = 0.5) +
  geom_line(aes(y = Mean),
             alpha = 0.5) +
  labs(x = 'Image date', y = 'Index value') + 
  facet_wrap(~index, ncol=1, scale = 'free_y') +
    scale_fill_manual(values = FFgrad) + 
  scale_color_manual(values = FFgrad) + 
  scale_x_date( date_breaks = 'month', 
                date_labels = '%b-%d', 
                minor_breaks = NULL) 
```

```{r veg_by_season, eval = FALSE, fig.cap="Data summarized by year from shaded blue ranges above.", fig.height = 4, fig.width = 10}
# Get average values for burn seasons 
  NoFireIndices %>%
    pivot_longer(names_to = 'index', 
                 values_to = 'value', 
                 cols = c(ndvi:ndmi)) %>%
    group_by(ImageDate, pasture, index) %>%
    summarise(Median = median(value) , 
              .groups = 'drop') %>%
    mutate(ImageDate = as.Date(ImageDate, '%Y-%m-%d'), 
           Year = lubridate::year(ImageDate), 
           Year = as.factor(Year), 
           SampleDate = format(ImageDate, '%m-%d'), 
           SampleDate = as.Date(SampleDate, '%m-%d'), 
           season = case_when(
             between(SampleDate, seasons$start[1], seasons$end[1]) ~ 'Spring',
             between(SampleDate, seasons$start[2], seasons$end[2]) ~ 'Summer', 
             TRUE ~ NA ) ) %>%
    filter(!is.na(season)) %>%
    group_by(season, Year, index) %>%
    summarize(SeasonMean = mean(Median), 
              SeasonSEM = sd(Median)/sqrt(n()), 
              .groups = 'drop') %>%
    mutate(status = case_when(
      season == 'Summer' & Year %in% c('2018', '2019') ~ "Not completed", 
      TRUE ~ "Completed"), 
          index = toupper(index), 
          index = fct_rev(index)) %>%
    ggplot(aes(x = season, y = SeasonMean, color = Year)) + theme_bw(14) +
    geom_errorbar(aes(ymin = SeasonMean - SeasonSEM, 
                      ymax = SeasonMean + SeasonSEM), 
                  width = 0.25, lwd = 0.95, 
                  position = position_dodge(width = 0.33), 
                  key_glyph = "point")  +
    geom_point(aes(pch = interaction(status, Year)), 
               position = position_dodge(width = 0.33), 
                size = 4, stroke = 2, show.legend = F) +
    labs(x = 'Burn season', 
         y = 'Index value', 
         caption = 'X = Year/season in which burns were not completed') + 
    scale_shape_manual(values = c(16, 16, 4, 16, 4, 16)) + 
    facet_wrap(~index, ncol = 3, scale = "free") +
    theme(axis.text.x = element_text(color = 'black')) +
    guides(color=guide_legend(override.aes = list(pch=15, size =4))) 
```

```{r sev_veg, eval = FALSE, fig.cap="Vegetation indices pulled from closest pre-burn image, same as pre-burn NBR.", fig.height = 3.8, fig.width = 10}
BurnIndices  %>%
  pivot_longer(names_to = 'name', 
               values_to = 'index', 
               cols = c(ndvi, msi, ndmi)) %>%
  group_by(Year, Season, fire, name) %>%
  summarise_at(.vars = vars(dNBR, index),
               .funs = c(Mean="mean", 
                         SEM = function(x) {sd(x)/sqrt(n()) } )) %>%
  ungroup() %>%
  mutate(name = toupper(name), 
         name = fct_rev(name)) %>% 
  filter(Season == "Spring") %>%
  ggplot(aes(x = index_Mean, y = dNBR_Mean, color = Year)) + theme_bw(14) +
  geom_smooth(aes(x = index_Mean, y = dNBR_Mean), 
              method = 'lm', 
              color = 'black',
              se = T, alpha = 0.2) +
  geom_errorbar(aes(ymin = dNBR_Mean - dNBR_SEM, 
                    ymax = dNBR_Mean + dNBR_SEM)) +
  geom_errorbarh(aes(xmin = index_Mean - index_SEM, 
                     xmax = index_Mean + index_SEM)) +
  geom_point() + 
  labs(x = 'Pre-burn index value', 
       y = 'Burn severity (dNBR)') + 
  geom_smooth(method = 'lm', se = F) +
  facet_wrap(~name, scales = 'free')
```

```{r, eval = FALSE}
full_join(by = 'fire', 
fires %>%
  as_tibble() %>%
  filter(Year == '2018') %>%
  select(fire, PreBurn),

BurnIndices  %>%
  pivot_longer(names_to = 'name', 
               values_to = 'index', 
               cols = c(ndvi, msi, ndmi)) %>%
  group_by(Year, Season, fire, name) %>%
  summarise_at(.vars = vars(dNBR, index),
               .funs = c(Mean="mean", 
                         SEM = function(x) {sd(x)/sqrt(n()) } )) %>%
  ungroup() %>%
  filter(Year == '2018') %>%
  select(fire, name, dNBR_Mean, index_Mean) %>%
  pivot_wider(names_from = name, 
              values_from = index_Mean) ) %>%
  rename(dNBR = dNBR_Mean) %>%
  arrange(desc(msi)) 
```

## Historical trends

```{r eval = FALSE}
LandsatSums %>%
  filter(year %in% c(2017:2020)) %>%
  group_by(year, season, index) %>%
  summarize(Mean = mean(Value),
            SEM = sd(Value)/sqrt(n()), 
            .groups = 'drop') %>%
  mutate(status = case_when(
    season == 'Summer' & year %in% c('2018', '2019') ~ "Not completed", 
    TRUE ~ "Completed"), 
    index = toupper(index), 
    index = fct_rev(index) ) %>%
    ggplot(aes(x=season)) + theme_bw(14) + 
    geom_boxplot(data = LandsatSums %>% mutate(index = toupper(index), 
                                               index = fct_rev(index)),
                         aes( y = Value),
                         outliers = F, 
                 staplewidth = 0.25) +
    geom_point(aes(y = Mean, color = year, shape = status), 
               position = position_dodge(width = 0.25), 
               size = 3, stroke = 2, show.legend = F) + 
    facet_wrap(~index, scales = 'free_y')  +
  scale_shape_manual(values = c(16, 4)) +
  theme(axis.text.x = element_text(color = 'black'), 
        panel.grid.major.x = element_blank() )
```

```{r }
StudySums <- 
  LandsatSums %>%
    filter(year %in% c(2017:2020)) %>%
    group_by(year, season, index) %>%
    summarize(Mean = mean(Value),
              SEM = sd(Value)/sqrt(n()), 
              .groups = 'drop') %>%
    mutate(status = case_when(
      season == 'Summer' & year %in% c('2018', '2019') ~ "Not completed", 
      TRUE ~ "Completed"), 
      index = toupper(index), 
      index = fct_rev(index) )

HistoricalSpectralData <- 
  bind_rows(
  LandsatSums %>%
  filter(year %in% c(1990:1999))%>% 
  mutate(range = '1990-1999', 
         index = toupper(index), 
         index = fct_rev(index)), 
  LandsatSums %>%
    filter(year %in% c(2012:2022))%>% 
    mutate(range = '2012-2022', 
           index = toupper(index), 
           index = fct_rev(index)), 
  LandsatSums %>%
    mutate(range = '1990-2022', 
           index = toupper(index), 
           index = fct_rev(index)) )
```

```{r historical_ndvi, eval = FALSE, fig.width = 7, fig.height =3}
  ggplot() + theme_bw(14) + 
  geom_boxplot(data = HistoricalSpectralData  %>% filter(index == 'NDVI'),
               aes(x=season, 
                   y = Value),
               outliers = F, 
               staplewidth = 0.25) +
  # geom_errorbar(aes( ymin = Mean - SEM, 
  #                    ymax = Mean + SEM, 
  #                    color = year), 
  #               width = 0.5, 
  #               position = position_dodge(width = 0.25)) + 
  geom_point(data = StudySums %>% filter(index == 'NDVI'), 
            aes(x = season, y = Mean, 
                color = year, shape = status), 
             position = position_dodge(width = 0.25), 
             size = 3, stroke = 2, show.legend = T) + 
    labs(x = 'Burn season', y = 'NDVI') + 
  facet_wrap(~range)  +
  scale_shape_manual(values = c(16, 4), guide = 'none') +
    scale_color_manual(values = FFgrad) +
  theme(axis.text.x = element_text(color = 'black'), 
        panel.grid.major.x = element_blank() ) +
    guides(color=guide_legend(override.aes = list(pch=15, size =4)))
```

```{r historical_phenology, eval = TRUE, fig.width = 8, fig.height =11.5}
WxData$HourlyWx %>%
  filter(! date %in% RainyDays$date, # remove rainy days with no chance of burning
         ! is.na(season)) %>%
  group_by(Year, date, season) %>%
  summarise_at(.vars = vars(c(DewPoint, RelHum, WindSpeed,VPD)),
               .funs = c("mean")) %>%
  ungroup() %>%
  pivot_longer(names_to = 'variable', 
               values_to = 'value', 
               cols = DewPoint:VPD) %>%
  group_by(Year, season, variable) %>%
  summarise(Value = mean(value), 
            .groups = 'drop') %>%
  mutate(Year = as.character(Year), 
         #season = fct_rev(season), 
         period = case_when(
           Year %in% c(1990:1999) ~  '1990-1999', 
           Year %in% c(2012:2022) ~  '2012-2022', 
           TRUE ~ NA
         )) %>%
  filter(!is.na(period)) %>%
  bind_rows(WxData$Rainfall %>% 
              filter(month(date) %in% c(04:09)) %>%
              separate(date, into = c('Year', 'Month', 'Day'),sep = '-') %>%
              group_by(Year) %>%
              mutate(CumPrecip = cumsum(Rainfall), 
                     Month = recode(Month, '05' ="May", '08'='Aug'), 
                     season = paste0(Month, ' ', Day)) %>%
              ungroup() %>%
              filter(season %in% c('May 15', 'Aug 01' )) %>%
              mutate(season = ifelse(season == 'May 15', 'Spring', 'Summer'), 
                     period = case_when(
                       Year %in% c(1990:1999) ~  '1990-1999', 
                       Year %in% c(2012:2022) ~  '2012-2022', 
                       TRUE ~ NA
                     ) )  %>%
              filter(!is.na(period)) %>%
              select(Year, season, CumPrecip, period) %>%
              pivot_longer(names_to = 'variable', 
                           values_to = 'Value', 
                           cols = CumPrecip))  %>%
  bind_rows(HistoricalSpectralData  %>% 
              select(-pasture) %>%
              filter(index == 'NDVI', 
                     range != '1990-2022') %>%
              rename(Year = year, period = range, variable = index)
            ) %>%
  mutate( variable = factor(variable, levels = c('NDVI', 'CumPrecip', 'DewPoint', 
                                                 'RelHum', 'VPD', 'WindSpeed')), 
          variable = recode(variable, 
                           'NDVI' = "Fuelbed\ngreenness\n(NDVI)", 
                           'CumPrecip' = 'Cumulative\nrainfall\n(mm)',
                           'DewPoint' = 'Dew\nPoint\n(°C)', 
                           'RelHum' = 'Relative\nHumidity\n(%)', 
                           'VPD' = 'Vapor\nPressure\nDeficit', 
                           'WindSpeed' = 'Wind\nspeed\n(m/s)'), 
          period = recode(period, '1990-1999' = '1990-99', '2012-2022' = '2012-22')) %>%
  ggplot() + theme_bw(16) + 
  geom_hline(data = . %>%
               filter(as.numeric(Year) %in% c(2017:2020)) %>%
               mutate(status = case_when(
                 season == 'Summer' & 
                   Year %in% c('2018', '2019') ~ "Not completed", 
                 TRUE ~ "Completed") ) %>%
               group_by(variable, Year, season, status) %>%
               slice(1) , 
             aes(yintercept = Value,
                 color = Year, lty = status), 
             lwd = 1 ) + 
  geom_boxplot(aes(x=period, 
                   y = Value),
               outliers = F, 
               staplewidth = 0.25) +
  labs(x = 'Historical period', y = ' ') + 
  facet_grid(variable~season, scales = "free_y")  +
  scale_linetype_manual("Burn status", values = c(1,2)) +
  scale_color_manual("Study year", values = FFgrad) +
  theme(axis.text.x = element_text(color = 'black'), 
        panel.grid.major.x = element_blank() , 
        strip.text.y = element_text(angle = 0), 
        strip.background = element_rect(fill = '#A1D99B')) 
```

```{r eval = FALSE}
HistoricalSpectralData  %>% 
  filter(index == 'NDVI', 
         ! is.na(Value), 
         range != '1990-2022') %>%
  group_by(season, range) %>%
  summarise(boxplot= list(setNames(boxplot.stats(Value)$stats,                                      c('bottom_whisker','bottom_hinge','median','top_hinge','top_whisker'))), 
              .groups = 'drop') %>%
    unnest_wider(boxplot) %>%
  mutate(across(c(bottom_whisker:top_whisker), ~ round(., 2)) ) %>%
  pander::pander()
  #   summarise(across(Value, 
  #                    list(min=min, Q1=~quantile(., probs = 0.25),
  #                         median=median, Q3=~quantile(., probs = 0.75),
  #                         max=max),  
  #                    .names = "{.fn}"), 
  #             .groups = 'drop') %>%
  # mutate(across(c(min:max), ~ round(., 2)) ) %>%
  # pander::pander()
```

### 2024-11-25 SpringSummerSeverity/paper/SummerFireRestoration.Rmd
SpringSummerSeverity paper draft

---
title: Evaluating an attempt to restore summer fire in the Northern Great Plains
author:
- Devan Allen McGranahan* and Jay P. Angerer
- Livestock and Range Research Laboratory, USDA Agricultural Research Service, Miles City, Montana, USA 
- "*Correspondence: Devan.McGranahan@usda.gov"
output: # html_document
  word_document:
    reference_docx: re_template.docx
    keep_md: yes
csl: restoration-ecology.csl
bibliography: 
  - SpringSummerFire.bib 
always_allow_html: true
---

```{r setup, echo=FALSE, warning=FALSE, message = FALSE, results='hide'}
knitr::opts_chunk$set(message = FALSE, warning=FALSE, 
                      echo=FALSE, eval=TRUE)
pacman::p_load(tidyverse, knitr, kableExtra)
# Some caption tools 
  t_nums = captioner::captioner(prefix = "Table")
  f_nums = captioner::captioner(prefix = "Fig.")
  t_ref = function(x) { str_extract(t_nums(x), "[^:]*")   }
  f_ref = function(x) { str_extract(f_nums(x), "[^:]*")   }
```

# Abstract

There is growing interest in diversifying human-managed fire regimes. 
In many North American grasslands, late growing season burns re-introduce fire to periods most prone to lightning-driven fire prior to wildfire suppression policies. 
We report here on restoring summer fire in central North Dakota, USA, from a research project in which summer fire was only successfully achieved in two out of four years. 
We use remotely-sensed imagery and local weather data to assess whether fuel or weather conditions limited burning in the summer, and to compare fire environmental conditions and subsequent burn severity across prescribed burns conducted in the spring and summer. 
Finally, we review historical data to determine if conditions have changed in either the spring or summer burn seasons over 42 years. 
We found that burn severity declined with fuelbed greenness but was independent of burn season---summer burns could effect as high of severity as spring burns despite having greener fuelbeds.
What little phenological change seems to have occurred at the study location---slightly greener fuelbeds and slightly lower relative humidity in spring---likely offset each other to some degree. 
Overall, we found little evidence that successful summer burns were anomalies, and conclude that it is reasonable for managers to incorporate late growing season fire into prescribed fire programs with the caveat that some summers will simply be too wet and/or too green to burn.

**Keywords:** Burn severity; Phenology; Seasonality of fire; Pyrodiversity

# Introduction

Fire is essential to the maintenance of grasslands and other open ecosystems, and has been described as a key component of restoring pre-colonial ecological conditions [@bond2021; @fuhlendorf2017]. 
The nature of fire in a given ecosystem is described as a regime. 
While determining the vegetation layer(s) that carry fire and the typical frequency and intensity of burning are relatively straightforward, seasonality and ignition source are much more difficult to reconstruct [@mcgranahan2021]. 
As such, the appropriate season(s) in which to burn is among the most debated aspects of fire restoration projects.

Burning in different seasons can produce different effects. 
While combustion in the wildland fire environment is theoretically possible whenever a competent ignition source is introduced to a sufficient amount of dry vegetation, climate often determines whether these conditions occur simultaneously. 
Lightning is the primary source of natural ignitions responsible for most fire starts prior to the emergence of hominids several hundred thousand years ago [@pyne1994]. 
As humans evolved and their technological capacity increased, so too did the influence of their ignitions.
In regions with extant cultural fire use, the seasonality of fire is often dictated by localized and traditional land-use goals [@mcgranahan2022]. 
In most Western countries, however, cultural burning has been replaced by systematic, command-and-control fire management in which most unplanned ignitions are suppressed and intentional burning occurs under strict prescriptions [e.g., @steen-adams2017].

Restoring summer fire is of particular interest among managers in the North American Great Plains, where the influence of Indigenous ignitions was extensive and interacted with climate [@boyd2002; @roos2018]. 
Prescribed fire in the Great Plains has conventionally occurred while vegetation is dormant, although @engle2001 argued prairie was resilient to fire in any season and challenged managers to consider summer burns. 
Although subsequent research has shown benefits of summer fire for diversity and productivity [@ansley2007a; @decker2019], a survey of grassland managers found that summer remains the least frequent season for prescribed fire [@harmon-threatt2016]. 
Meanwhile, wide variability in responses to seasonal fire have been reported within and across plant functional groups, often driven by disproportionately large responses by few species [@steuter1987; @mndela2023; @biondini1989; @sparks1998]. 
Considerable variability in responses to summer fire has been attributed to the fact that higher humidity and fuel moisture during the growing season often reduce fire intensity and effect lower severity [@steuter1987; @tangney2022].

The Northern Great Plains might have had one of the most pronounced differentials in fire seasonality between lightning-caused and anthropogenic ignitions. 
Fire was frequent in the region for at least 400 years prior to European colonization [@umbanhowar1996].
Contemporary accounts of fires set by Native Americans and early Europeans beginning in 1673 identified a bimodal pattern of human-set fires in the spring and fall [@higgins1986]. 
But for the last 80 years, at least, lightning-caused fires occurred predominately in late July and August [@higgins1984; @mcgranahan2024a].

In this paper we report on mixed success in conducting summer prescribed fire in the Northern Great Plains. 
As researchers attempted to restore spatially-patchy prescribed fire to grazed mixed-grass prairie, two patterns of fire were introduced: one in which an entire one-quarter patch was burned each spring, and another in which half of the patch---one-eighth of the pasture---was burned in the spring, followed by the second half-patch scheduled for a late summer burn, in an attempt to provide a late-season flush of high-quality forage [e.g., @allred2011; @nichols2021] and determine whether a greater number of patches would increase spatial heterogeneity  [e.g., @mcgranahan2018b].
While all spring burns were completed as planned, summer fires were only successful in two out of four years. 
Here we present seasonal and long-term remotely-sensed burn severity and fuel data alongside weather information in considering whether successful summer burns are likely to be the exception or the norm based on historical patterns.

# Methods

## Study location

The prescribed fire restoration project was established at the North Dakota State University Central Grasslands Research Extension Center, near Streeter, in south-central North Dakota, USA (99°25′W, 42°42′N). 
Vegetation at the Center is typical northern mixed-grass prairie with substantial abundance of the non-native C~3~ grasses *Poa pratensis* and *Bromus inermis*. 
In 2017, a research project began in which 12, 65-ha pastures were assigned to one of each of three management strategies (N = 4), none of which involved internal fences: continuous grazing, in which livestock had unhindered access to pastures with no additional disturbance; spring-only patch burning, in which a 16-ha patch was burned each spring and livestock allowed unhindered access; and spring-summer patch burning, in which half of a 16-ha patch was burned in the spring, and the second half scheduled to be burned in the summer. 
All pastures were stocked with approximately 22 Angus *Bos taurus* cow-calf pairs from May-October; see @duquette2022a for a complete description of the study location and experimental design.

A total of 40 prescribed fires were conducted over four years, 2017--2020. 
All 32 planned spring burns were completed, between 25 April and 25 May. 
Only eight of 16 planned summer burns were completed: 4 in 2017, and 4 in 2020, between 13-25 August. 
Burn units were surrounded with plowed mineral firebreaks and ignitions varied between backing and head fires, and interior strip ignitions, with handheld drip torches.

## Data

Two types of data are presented here: remotely-sensed imagery from earth observation satellites, and weather data logged by the Center's mesonet weather station. 
We used remotely-sensed data from two separate sources. 
For the study period alone, 2017--2020, we used the European Space Agency's Sentinel-2 mission, which provides hyperspectral imagery at 10 $\times$ 10 m resolution at 5-10 day intervals. 
For historical context, 1990--2022, we used Landsat resources from the US Geological Survey, which provides hyperspectral imagery at 30 $\times$ 30 m resolution at 7-14 day intervals; Sentinel-2 only began in 2016, while the first Landsat mission launched in the mid 1980s.

Remotely-sensed data consisted of two indices: the Normalized Burn Ratio (NBR), which is used to estimate burn severity, and Normalized Difference Vegetation Index (NDVI), used to estimate greenness of vegetation. 
Sentinel-2 imagery was retrieved from the online Copernicus Browser using a custom EvalScript to return values for each index as a separate band in a geo-referenced raster file; each scene was inspected to ensure cloud-free imagery over pasture locations beginning with the first snow-free scene through to the final snow-free scene each year. 
To calculate burn severity, we subtracted the first post-fire image from the most-recent pre-fire image for each burn unit, generating a burn severity index known as differenced Normalized Burn Ratio (dNBR). 
Estimates of fuel greenness were made with the NDVI from the same most-recent pre-burn image.

Landsat imagery was processed with Google Earth Engine using a custom script to merge missions over time, mask clouds, and calculate NDVI for the burn seasons defined above, 1990--2022. 
See Supplemental Information for scripts used for both Copernicus and Google Earth Engine.

Remotely-sensed values were extracted from raster images at sample points using the *terra* package for the **R** statistical environment [@hijmans2022; @rcoreteam2024a].
Data from the patch-burned pastures were retrieved from gridded points assigned with 30 m spacing never less than 10 m from the edges of burn units, with ponds and ephemeral water bodies within burn units excluded. 
Data from unburned control pastures were retrieved from points used in forage quality assessments by @wanchuk2024.

Weather data---which included air temperature, relative humidity, dew point, windspeed, and vapor pressure deficit---were retrieved from the North Dakota Agricultural Weather Network's Streeter station [@ndawn2023]. 
To characterize day-of fire weather for successful burn days, we calculated mean values for operational burn periods (1000--1700) from hourly data for days recorded as having at least one completed fire. 
To characterize seasonal fire weather, we used daily precipitation totals to eliminate days with more than 25 mm precipitation---days no fire would be conducted under any other circumstance---and calculated mean values for the operational period for each spring (25 April--25 May) and summer (7--27 August) burn season. 
Seasonal values were determined for both the study period (spring and summer burn seasons, 2017--2020) and the historical range (1 April- 31 October, 1990--2023).

To examine potential changes in conditions over time, all data related to the fire environment---NDVI, cumulative rainfall, and average hourly fire weather data for the operational period---were summarized for each burn season in the first and final decades of the dataset, 1990--1999 and 2012--2022, respectively.

# Results

Grassland fuelbeds were less green during each of the spring fire seasons (Sentinel-2 NDVI 0.3-0.4) than in the summer (NDVI > 0.5), with the two seasons in which summer burns could not be completed being the most green (NDVI > 0.6; `r f_ref("sentinel")`A). 
Burn severity ranged widely within and across seasons and years (`r f_ref("sentinel")`B). 
Among the 32 spring burns, there was a general decline in burn severity with increasing fuelbed greenness. 
The low number of successful summer burns precluded a linear assessment, but despite having categorically higher greenness, summer burning effected the same range of variability in burn severity as in the spring (`r f_ref("sentinel")`B).

Fuelbed greenness during successful burn seasons was close to the typical range over the 42-year historical period (`r f_ref("GreenRain")`A), while cumulative rainfall was at or below the typical historical range (`r f_ref("GreenRain")`B). 
Conversely, for years in which summer burns were not completed, both summer fuelbed greenness and accumulated rainfall by 1 August were atypically high relative to their historical ranges (`r f_ref("GreenRain")`).
<!-- Note that spring NDVI derived from Landsat (`r f_ref("GreenRain")`A) was slightly higher than the same data from Sentinel-2 (`r f_ref("sentinel")`A); the reason for the difference is unclear but might relate to the lower temporal resolution of Landsat imagery skewing the spring average with later, greener values and fewer early, snow-free images. -->
<!-- Notably, summer values appear quite consistent between Landsat and Sentinel-2.  -->

Observed fire weather varied widely within and across the four study years, and generally spanned the entire range of historical values, although wind speeds on the day of prescribed burns never exceeded the historical median in either season, and only one burn day exceeded the historical median relative humidity (`r f_ref("weather")`). 
Throughout the 42-year dataset, summer burn seasons generally had higher moisture, greater vapor pressure deficient, and lower wind speed than spring burn seasons.

Few substantial changes in the fire environment appear to have occurred since the 1990s in either spring or summer burn seasons (`r f_ref("historical")`). 
The two most notable changes---increased spring fuelbed greenness and lower spring dewpoint and relative humidity---might cancel each other out, with drier air potentially mitigating the impact of higher-moisture fine fuels. 
In two instances---spring windspeed and summer rainfall---the more apparent temporal trend was for greater variability around the typical range, rather than substantial shifts in typical values.

# Discussion

We used remotely-sensed imagery and weather data to inform whether, in the four years summer burns were attempted in a research project in the Northern Great Plains, the two successful years, or the two unsuccessful years, were more likely outcomes based on typical fuel and weather conditions. 
Comparing seasonal values to historical ranges suggests that unsuccessful burn seasons were the anomalies. 
As such, successfully completing prescribed fire in the late growing season in the Northern Great Plains is a reasonable management objective. 
However, prairie fuelbeds might simply be too green to carry fire in some years, likely those with above-normal rainfall over the course of the growing season.

The data provide little indication that fire weather posed a barrier to successful summer burning. 
Overall, burn season fire weather was highly variable across study years, and burn day weather spanned the gamut (the exception was wind speed, which was below average in all study years). 
For example, for the four years summer burns were planned, one successful year and one unsuccessful year were at the high end of typical relative humidity, and the other successful and unsuccessful years were at the low end of typical relative humidity. 
In a study of these and similar grasslands that included a subset of the spring fires reported here, higher wind speed and greater vapor pressure deficit contributed to higher rates of spread, a measure of fire intensity [@mcgranahan2023a]. 
But overall, the statistical analysis provided little explanation for the substantial variability in fire behavior.

<!-- Experience does suggest fire weather conditions can limit specific prescribed fire operations.  -->
<!-- For example, a burn was planned around a favorable weather forecast on August 13, 2018, and a full crew was on-site that morning.  -->
<!-- But relative humidity did not fall as predicted in the spot weather forecast, remaining above 50\% until 1500 (the burn was called off by 1400 after repeated test fires failed to spread).  -->
<!-- Certainly, actual weather defies forecasts in all seasons.  -->
<!-- But in the context of summer prescribed fire, when fuels are more green and windspeed generally lower (`r f_ref("historical")`), stubborn relative humidity is more of a barrier in the absence of mitigating factors.  -->

Although trends towards lower relative humidity might mitigate trends towards greener fuelbeds in the spring, cumulative seasonal rainfall, fuelbed greenness, and relative humidity likely all increase together in the summer, when plants are photosynthetically active. 
Conversely, our historical comparisons suggest the spring burn season is perhaps most susceptible to global changes, although complex interactions between drivers of global change introduce substantial uncertainty. 
In addition to continental-scale increases in air temperature and variability in precipitation [@kukal2016; @pendergrass2017], fire managers in the Northern Great Plains must also deal with local changes. 
For example, native grasslands of the Northern Plains---including the pastures considered here---have been invaded with exotic, cool-season grasses that green up early, change fuel structure, and reduce rate of spread [@gasch2020; @yurkonis2019].

Furthermore, regional atmospheric modification might also modulate the effectiveness of prescribed fire even if local conditions are amenable for burning.
Put simply, fire intensity generally increases with atmospheric instability, which facilitates rapid convection of smoke away from the reaction zone and draws in fresh oxygen for combustion [@potter2012]. 
Instability, meanwhile, is modulated by soil moisture and land use: arid areas tend to have lower soil moisture and greater convective potential, but greater atmospheric moisture can increase stability [@chapman2021; @hiestand2024]. 
Evidence from the US Corn Belt suggests transpiration from highly-productive crop fields contributes to atmospheric moisture content [@hill2019], and the western spread of the Corn Belt into the Great Plains is well-established [@wright2013]. 
But the potential interactions and impacts of regional climate and land-use changes on fire management have not been sufficiently explored.

Research into opportunities to conduct prescribed fire might consider phenological changes in the fire environment, rather than conventional attempts to assess changes in the availability of "weather windows"---the set of conditions under which prescribed fire might possibly be conducted [e.g., @weir2011]. 
While some research has used the concept of the weather window to identify seasonal shifts in  potentially favorable burn conditions---e.g., @yurkonis2019 showed a decrease in early spring burn days and an increase in late summer burn days for central North Dakota---the focus on prescriptions generally masks interactions between mitigating and exacerbating factors, fails to account for fuelbed dynamics, and is agnostic to antecedent weather and fuel conditioning. 
We suggest instead an approach that uses the type of data presented here to classify the seasonality of fuels and fire weather to investigate temporal changes in the phenology of the fire environment as a whole.

<!-- Even including fuel load did little to improve explanatory power---while fuel load had a positive linear relationship with observed flame temperatures in a subset of these fires, it was not related to rate of spread or surface heating [@mcgranahan2023a]. -->

<!-- On account of ecological and management histories, fuel loads across burn units were similar, with internal variability attributable to topoedaphic effects (slope, soil texture) and year-to-year differences in precipitation.  -->

\newpage

```{r sentinel, fig.cap = sent_cap}
sent_cap <- f_nums(name = "sentinel", 
                   caption = "\\label{sentinel}Sentinel imagery summarized for the study years. A: Normalized Difference Vegetation Index (NDVI) from four adjacent unburned pastures avoids any confounding effects of prescribed burns. X denotes seasons in which burns were not completed. B: Burn severity (differenced Normalized Burn Ratio, dNBR) plotted against fuelbed greenness (NDVI) retrieved from the pre-burn Sentinel-2 image used for pre-fire NBR. Points represent mean for individual prescribed fires with associated standard error; circles and solid trendline indicate spring burns while broken trendline and triangles denote summer burns.")
include_graphics('ndvi_gg-1.png')
```

\newpage

```{r GreenRain, fig.cap = GreenRain_cap}
GreenRain_cap <- f_nums(name = "GreenRain", 
                   caption = "\\label{GreenRain}Study season data (points) plotted within the context of 42-year historical trends (boxplots). X indicates seasons prescribed fires were not completed. A: Normalized Difference Vegetation Index (NDVI) data retrieved from Landsat imagery. B: Annual growing season rainfall accummulated by two specific dates associated with the spring and summer burn seasons, respectively.")
include_graphics('green_trends-1.png')
```

\newpage

```{r weather, fig.cap = weather_cap}
weather_cap <- f_nums(name = "weather", 
                   caption = "\\label{weather}Four fire weather variables plotted by burn day values (open triangles), study season means (open circles and X for seasons without completed burns), and 42-year historical trends (boxplots). ")
include_graphics('weather-1.png')
```

\newpage

```{r historical, fig.cap = hist_cap}
hist_cap <- f_nums(name = "historical", 
                   caption = "\\label{historical}Temporal trends in the phenology of the fire environment. Boxplots represent data from the first decade of the dataset used here (1990-1999) to the final decade (2012-2022), by burn season. Horizontal lines represent average conditions for the four study years (2017-2020).")
include_graphics('historical_phenology-1.png')
```

\newpage

# References

### 2024-11-25 SpringSummerSeverity/writeup.Rmd
SpringSummerSeverity paper draft

@@
 ---
-title: Spring vs. summer fire regimes in central North Dakota
+title: Evaluating an attempt to restore summer fire in the Northern Great Plains
 author: DA McGranahan & JP Angerer??
@@
                       echo=FALSE, eval=TRUE)
-pacman::p_load(tidyverse)
+pacman::p_load(tidyverse, cowplot, patchwork, wesanderson)
+  BuGnPal = c( "#67A9CF" ,"#3690C0" ,"#02818A", "#016450")
+  FFgrad = wes_palette("FantasticFox1")[c(3,2,4,5)]
 # load data 
@@
   load('./data/BurnIndices.Rdata')
+  load('./data/NoFireTrends.Rdata')
+  load('./data/WxData.Rdata')
 # some necessary things
@@
            end = c(as.Date('2024-05-25'), as.Date('2024-08-27'))) 
+```
+
+```{r}
+LandsatSums <-
+  NoFireTrends  %>%
+    separate(ImageDate, into=c('year', 'month', 'day'), sep = '-') %>%
+    mutate(season = ifelse(month=='04', 'Spring', 'Summer')) %>%
+    pivot_longer(names_to = 'index', 
+                 values_to = 'value', 
+                 cols = c(ndvi:ndmi)) %>%
+    group_by(year, season, pasture, index) %>%
+    summarize(Value = mean(value), 
+              .groups = 'drop') 
 ```
@@
 
-* Explore patterns in greenness and humidity to determine effects on fire effectiveness
+* Explore patterns in fuel greenness and fire weather to determine effects on fire effectiveness
 * Identify anomalies and ranges of conditions conducive to prescribed fire
 
-# Remotely-sensed data
+## Data
 
-Pulled burn severity (NBR) and several vegetation indices from Sentinel-2 imagery for 2017-2020 growing seasons. 
-Data sampled from regular gridded points within burn units
+Pulled burn severity (NBR) and NDVI from Sentinel-2 imagery for 2017-2020 growing seasons.
+Also pulled historical (1990-2022) NDVI from Landsat for long-term trends. 
+NDVI = Normalized Difference Vegetation Index; Measure of vegetation greenness.
+Data sampled from regular gridded points within burn units.
 
-Vegetation indices:
+Also retrieved historical hourly weather data, 1990-2022, from local mesonet station.
 
-* NDVI (Normalized Difference Vegetation Index)---Measure of vegetation greenness
-* NDMI (Normalized Difference Moisture Index)---Measure of moisture content of vegetation
-* MSI (Moisture Stress Index)---Another measure of moisture content sometimes used in fire risk analysis. Higher = drier
+# Results
 
@@
 
-These data are from four *unburned* pastures in an adjacent continuous grazing treatment to remove any confounding effects of prescribed burns.
+```{r ndvi_gg, fig.cap="Sentinel imagery summarized for the study years.
+A: NDVI data from four adjacent unburned pastures avoids any confounding effects of prescribed burns.
+X denotes seasons in which burns were not completed.
+B: Burn severity (dNBR) plotted against fuelbed greenness (NDVI) retrieved from the pre-burn Sentinel-2 image used for pre-fire NBR.
+Points represent mean for individual prescribed fires with associated standard error; circles and solid trendline indicate spring burns while broken trendline and triangles denote summer burns.", fig.height = 7, fig.width = 5}
+# Get average values for burn seasons 
+p1 <-
+  NoFireIndices %>%
+    group_by(ImageDate, pasture) %>%
+    summarise(NDVI = median(ndvi) , 
+              .groups = 'drop') %>%
+    mutate(ImageDate = as.Date(ImageDate, '%Y-%m-%d'), 
+           Year = lubridate::year(ImageDate), 
+           Year = as.factor(Year), 
+           SampleDate = format(ImageDate, '%m-%d'), 
+           SampleDate = as.Date(SampleDate, '%m-%d'), 
+           season = case_when(
+             between(SampleDate, seasons$start[1], seasons$end[1]) ~ 'Spring',
+             between(SampleDate, seasons$start[2], seasons$end[2]) ~ 'Summer', 
+             TRUE ~ NA ) ) %>%
+    filter(!is.na(season)) %>%
+    group_by(season, Year) %>%
+    summarize(SeasonMean = mean(NDVI), 
+              SeasonSEM = sd(NDVI)/sqrt(n()), 
+              .groups = 'drop') %>%
+    mutate(status = case_when(
+      season == 'Summer' & Year %in% c('2018', '2019') ~ "Not completed", 
+      TRUE ~ "Completed") ) %>%  
+    ggplot(aes(x = season, color = Year)) + theme_bw(14) +
+    geom_errorbar(aes(ymin = SeasonMean - SeasonSEM, 
+                      ymax = SeasonMean + SeasonSEM), 
+                  width = 0.25, lwd = 0.95, 
+                  alpha = 0.5,
+                  position = position_dodge(width = 0.33), 
+                  key_glyph = "point")  +
+    geom_point(aes( y = SeasonMean, 
+                    pch = interaction(status, Year)), 
+               position = position_dodge(width = 0.33), 
+                size = 4, stroke = 2, show.legend = F) +
+    labs(x = 'Burn season', 
+         y = 'NDVI', 
+         ) + 
+    scale_shape_manual(values = c(16, 16, 4, 16, 4, 16)) + 
+   scale_color_manual(values = FFgrad) +  
+    theme(axis.text.x = element_text(color = 'black')    ) +
+    guides(color=guide_legend(override.aes = list(pch=15, alpha = 1, size =4))) 
+p2 <-
+BurnIndices  %>%
+  group_by(Year, Season, fire) %>%
+  summarise_at(.vars = vars(dNBR, ndvi),
+               .funs = c(Mean="mean", 
+                         SEM = function(x) {sd(x)/sqrt(n()) } )) %>%
+  ungroup() %>%
+  ggplot(aes(x = ndvi_Mean, y = dNBR_Mean, color = Year)) + theme_bw(14) +
+  geom_smooth(data = . %>%
+                filter(Season == 'Spring'), 
+              aes(x = ndvi_Mean, y = dNBR_Mean, lty=Season), 
+              method = 'lm', 
+              color = 'black',
+              se = T, alpha = 0.2) +
+    geom_smooth(data = . %>%
+                filter(Season == 'Summer'), 
+              aes(x = ndvi_Mean, y = dNBR_Mean), 
+              method = 'lm', 
+              lty = 2,
+              color = 'black',
+              se = F, alpha = 0.2) +
+  geom_errorbar(aes(ymin = dNBR_Mean - dNBR_SEM, 
+                    ymax = dNBR_Mean + dNBR_SEM)) +
+  geom_errorbarh(aes(xmin = ndvi_Mean - ndvi_SEM, 
+                     xmax = ndvi_Mean + ndvi_SEM)) +
+  geom_point(aes(fill = Year, 
+                 shape = Season), 
+             color = 'black',
+             size = 3) + 
+  scale_shape_manual(values=c(21, 24)) +
+   scale_fill_manual(values = FFgrad) +
+   scale_color_manual(values = FFgrad) +  
+  labs(x = 'Pre-burn NDVI', 
+       y = 'Burn severity (dNBR)') +
+  theme(  legend.position = 'none', 
+          legend.position.inside = c(0.8, 0.85))
 
-### Seasonal trends
+plot_grid(p1, p2,  labels = "AUTO", 
+          ncol = 1, rel_heights = c(0.8, 1))
+```
+
+```{r fig.cap="Study season data (points) plotted within the context of 42-year historical trends (boxplots).
+X indicates seasons prescribed fires were not completed.
+A: NDVI data retrieved from Landsat imagery.
+B: Annual growing season rainfall accummulated by two specific dates associated with the spring and summer burn seasons, respectively.", fig.width = 5, fig.height = 7}
+p3 <-
+LandsatSums %>%
+                            filter(year %in% c(2017:2020), 
+                                   index == 'ndvi') %>%
+                            group_by(year, season) %>%
+                            summarize(Mean = mean(Value),
+                                      SEM = sd(Value) / sqrt(n()),
+                                      .groups = 'drop') %>%
+                            mutate(status = case_when(
+                              season == 'Summer' & 
+                              year %in% c('2018', '2019') ~ "Not completed", 
+                              TRUE ~ "Completed") )%>% 
+  ggplot(aes(x = season)) + theme_bw(14) + 
+      geom_boxplot(data = LandsatSums %>%
+                     filter(index == 'ndvi')  ,
+                 aes( y = Value),
+                 outliers = F,
+                 staplewidth = 0.25) +
+  # geom_errorbar(aes( ymin = Mean - SEM,
+  #                    ymax = Mean + SEM,
+  #                    color = year),
+  #               width = 0.5,
+  #               position = position_dodge(width = 0.25)) +
+         geom_point(#data = , 
+        aes(y = Mean, 
+                color = year, shape = status), 
+             position = position_dodge(width = 0.25), 
+             size = 4, stroke = 3, show.legend = T) + 
+    labs(x = 'Burn season', y = 'NDVI') + 
+       scale_color_manual("Year", values = FFgrad) + 
+  scale_shape_manual(values = c(16, 4), guide = 'none') +
+
+  theme(axis.text.x = element_text(color = 'black'), 
+        panel.grid.major.x = element_blank() ) +
+    guides(color=guide_legend(override.aes = list(pch=15, size =4)))
+p4 <- 
+    WxData$Rainfall %>% 
+    filter(month(date) %in% c(04:09)) %>%
+    separate(date, into = c('Year', 'Month', 'Day'),sep = '-') %>%
+    group_by(Year) %>%
+    mutate(CumPrecip = cumsum(Rainfall), 
+           Month = recode(Month, '05' ="May", '08'='Aug'), 
+           Month = paste0(Month, ' ', Day)) %>%
+    filter(Month %in% c('May 15', 'Aug 01' )) %>%
+    mutate(Month = fct_rev(Month)) %>%
+    ggplot() + theme_bw(14) + 
+      geom_boxplot(aes(x = Month, y = CumPrecip),
+                   outliers = F,
+                   staplewidth = 0.25) +
+    geom_point(data = . %>%
+                 filter(as.numeric(Year) %in% c(2017:2020)) %>%
+                  mutate(status = case_when(
+                    Month == 'Aug 01' & 
+                      Year %in% c('2018', '2019') ~ "Not completed", 
+                    TRUE ~ "Completed") ), 
+            aes(x = Month, y = CumPrecip,
+                color = Year, shape = status), 
+            position = position_dodge(width = 0.25), 
+            size = 4, stroke = 3, show.legend = T) + 
+  labs(x = 'Date', y = 'Cumulative Rainfall (mm)') + 
+  scale_shape_manual(values = c(16, 4), guide = 'none') +
+  scale_color_manual("Year", values = FFgrad) + 
+  theme(axis.text.x = element_text(color = 'black'), 
+        panel.grid.major.x = element_blank() ) +
+  guides(color=guide_legend(override.aes = list(pch=15, size =4)))
+
+p3 / p4 +
+  patchwork::plot_layout(
+    guides = "collect",
+    nrow = 2
+  ) +
+  patchwork::plot_annotation(
+    tag_levels = "A"
+  ) &
+  theme(legend.position = "right", legend.direction = "vertical")
+```
+
+## Weather data
+
+```{r, fig.cap="Four fire weather variables plotted by burn day values (open triangles), study season means (open circles and X for seasons without completed burns), and 42-year historical trends (boxplots).", fig.height = 9}
+RainyDays <- 
+   WxData$Rainfall %>%
+      filter( ! is.na(season), Rainfall > 0.1) 
+# Seasonal trends over time   
+  WxData$HourlyWx %>%
+    filter(! date %in% RainyDays$date, # remove rainy days with no chance of burning
+           ! is.na(season)) %>%
+    group_by(Year, date, season) %>%
+    summarise_at(.vars = vars(c(DewPoint, RelHum, WindSpeed,VPD)),
+                 .funs = c("mean")) %>%
+    ungroup() %>%
+    pivot_longer(names_to = 'variable', 
+                 values_to = 'value', 
+                 cols = DewPoint:VPD) %>%
+    group_by(Year, season, variable) %>%
+    summarise(Value = mean(value), 
+              .groups = 'drop') %>%
+    mutate(variable = recode(variable, 
+                             'DewPoint' = 'Dew Point (°C)', 
+                             'RelHum' = 'Relative Humidity (%)', 
+                             'VPD' = 'Vapor Pressure Deficit', 
+                             'WindSpeed' = 'Wind speed (m/s)') , 
+           Year = as.character(Year), 
+           season = fct_rev(season)) %>%
+    ggplot() + theme_bw(16) + 
+      geom_boxplot( aes(x=season, 
+                       y = Value),
+                   outliers = F, size = 0.75,
+                   staplewidth = 0.25) +
+      geom_point(data = . %>%
+                   filter(Year %in% c(2017:2020)) %>%
+                   mutate(status = case_when(
+                     season == 'Summer' & Year %in% c('2018', '2019') ~ "Not completed", 
+                     TRUE ~ "Completed") ) , 
+                 aes(x = season, y = Value, 
+                     color = Year, shape = status), 
+                 position = position_nudge(x = -.2), 
+                  fill = 'grey90', 
+                 size = 3, stroke = 2, show.legend = T) + 
+    geom_point(data = WxData$BurnDayWx , 
+               aes(x = season, y = value, color =Year), 
+               position = position_nudge(x = +.2), 
+               shape = 24, fill = 'grey90', 
+               size = 3, stroke = 2,
+                show.legend = F) + 
+      labs(x = 'Burn season', y = 'Weather variable value') + 
+      facet_wrap(~variable, scales = 'free_x', ncol = 1)  +
+      scale_shape_manual(values = c(21, 4), guide = 'none') +
+    scale_color_manual(values = FFgrad) +  
+      theme(axis.text.y = element_text(color = 'black', angle = 45, hjust = 0.5), 
+            panel.grid.major.y = element_blank() , 
+            strip.text = element_text(face='bold'), 
+            strip.background = element_rect(fill = '#A1D99B')) +
+    coord_flip() + 
+      guides(color=guide_legend(override.aes = list(pch=15, size =4)))
+```
+
+# Supplemenary Info
+
+## Seasonal trends
 
 Summer fires were successful in 2017 and 2020. 
-Vegetation in both years was less green and generally more dry.
+Vegetation in both years was less green and rainfall lower.
 
-```{r veg_trends, fig.cap="Blue shaded areas span the first and last dates of completed burns in the spring and summer.", fig.height = 8, fig.width = 10}
+```{r ndvi_trend, eval = TRUE, fig.cap="Green shaded areas span the first and last dates of completed burns in the spring and summer.", fig.height = 4, fig.width = 8}
+NoFireIndices %>%
+  group_by(ImageDate, pasture) %>%
+    summarise(Mean = mean(ndvi), 
+              SEM = sd(ndvi)/sqrt(n()), 
+              .groups = 'drop') %>%
+  mutate(ImageDate = as.Date(ImageDate, '%Y-%m-%d'), 
+         Year = lubridate::year(ImageDate), 
+         Year = as.factor(Year), 
+         SampleDate = format(ImageDate, '%m-%d'), 
+         SampleDate = as.Date(SampleDate, '%m-%d') ) %>%
+  filter(Year != 2016) %>%
+ggplot(aes(x = SampleDate, 
+           color = Year, 
+           fill = Year, 
+           group = interaction(pasture, Year))) + theme_bw(14) +
+  geom_rect(data= seasons, inherit.aes = FALSE,
+              aes(xmin=start, xmax=end, ymin=-Inf, ymax=+Inf), 
+              fill='#A1D99B', alpha=0.5) +
+  geom_smooth(aes(y = Mean, group = Year),
+              se = T,
+              level = 0.99) +
+  geom_errorbar(aes(ymin = Mean - SEM, 
+                    ymax = Mean + SEM),
+                alpha = 0.5) +
+  geom_line(aes(y = Mean),
+             alpha = 0.5) +
+  labs(x = 'Image date', y = 'NDVI') + 
+    scale_fill_manual(values = FFgrad) + 
+  scale_color_manual(values = FFgrad) + 
+  scale_x_date( date_breaks = 'month', 
+                date_labels = '%b-%d', 
+                minor_breaks = NULL) 
+```
+
+```{r precip_trend, eval = TRUE, fig.cap="Thick black line shows smoothed 40-yr trend.
+Green shaded areas span the first and last dates of completed burns in the spring and summer.", fig.height = 4, fig.width = 8}
+  WxData$Rainfall %>% 
+    filter(month(date) %in% c(04:09)) %>%
+    separate(date, into = c('Year', 'Month', 'Day'),sep = '-') %>%
+    group_by(Year) %>%
+    mutate(CumPrecip = cumsum(Rainfall)) %>%
+    ggplot(aes(x = day)) + theme_bw(14) +
+    geom_rect(data= seasons, inherit.aes = FALSE,
+              aes(xmin=start, xmax=end, ymin=-Inf, ymax=+Inf), 
+              fill='#A1D99B', alpha=0.5) +
+    geom_line(aes(y = CumPrecip, group = Year),
+              color = 'grey40', 
+              alpha = 0.5) +
+    geom_smooth(aes(y = CumPrecip),
+                se = F,
+                color = 'black',
+                size = 1,
+                level = 0.99) +
+    geom_line(data = . %>%
+                filter(as.numeric(Year) %in% c(2017:2020)) , 
+              aes(y = CumPrecip, 
+                  color = Year,
+                  group = Year),
+              size = 2) +
+    labs(x = 'Day of year', y = 'Cumulative rainfall (mm)') + 
+    scale_fill_manual(values = FFgrad) + 
+    scale_color_manual(values = FFgrad) + 
+    scale_x_date( date_breaks = 'month', 
+                  date_labels = '%b-%d', 
+                  minor_breaks = NULL)
+```
+
+```{r veg_trends, eval = FALSE, fig.cap="Blue shaded areas span the first and last dates of completed burns in the spring and summer.", fig.height = 8, fig.width = 10}
 NoFireIndices %>%
@@
               aes(xmin=start, xmax=end, ymin=-Inf, ymax=+Inf), 
-              fill='lightblue', alpha=0.5) +
+              fill='#A1D99B', alpha=0.5) +
   geom_smooth(aes(y = Mean, group = Year),
@@
   facet_wrap(~index, ncol=1, scale = 'free_y') +
+    scale_fill_manual(values = FFgrad) + 
+  scale_color_manual(values = FFgrad) + 
   scale_x_date( date_breaks = 'month', 
@@
 
-### Seasons summarized
-
-* Greenness:
-  - NDVI during all spring burns generally pretty low
-  - Successful summer burn years higher greenness than any spring...
-  - ... still slightly but distinctly lower than unsuccessful years
-
-* Moisture content: 
-  - High variability in spring seasons across years
-  - Successful summer seasons drier than in unsuccessful years
-
-```{r veg_by_season, fig.cap="Data summarized by year from shaded blue ranges above.", fig.height = 4, fig.width = 10}
+```{r veg_by_season, eval = FALSE, fig.cap="Data summarized by year from shaded blue ranges above.", fig.height = 4, fig.width = 10}
 # Get average values for burn seasons 
@@
 
-## Burn severity
-
-Overall it appears greener fuels consistently result in less-complete burns (although 2018 was weird).
-
-Trends between fuel moisture content and severity were less consistent. 
-Again 2018 behaved strangely---fuel in 4 burn units in 1 block of pastures was extremely dry but still didn't burn with particularly high severity; less-dry fuel in the other block tended to burn slightly more severely.
-The dry block was burned a couple weeks earlier than the other block.
-
-```{r sev_veg, fig.cap="Vegetation indices pulled from closest pre-burn image, same as pre-burn NBR.", fig.height = 3.8, fig.width = 10}
+```{r sev_veg, eval = FALSE, fig.cap="Vegetation indices pulled from closest pre-burn image, same as pre-burn NBR.", fig.height = 3.8, fig.width = 10}
 BurnIndices  %>%
@@
 
-# Weather data
+## Historical trends
 
-*Pending*
+```{r eval = FALSE}
+LandsatSums %>%
+  filter(year %in% c(2017:2020)) %>%
+  group_by(year, season, index) %>%
+  summarize(Mean = mean(Value),
+            SEM = sd(Value)/sqrt(n()), 
+            .groups = 'drop') %>%
+  mutate(status = case_when(
+    season == 'Summer' & year %in% c('2018', '2019') ~ "Not completed", 
+    TRUE ~ "Completed"), 
+    index = toupper(index), 
+    index = fct_rev(index) ) %>%
+    ggplot(aes(x=season)) + theme_bw(14) + 
+    geom_boxplot(data = LandsatSums %>% mutate(index = toupper(index), 
+                                               index = fct_rev(index)),
+                         aes( y = Value),
+                         outliers = F, 
+                 staplewidth = 0.25) +
+    geom_point(aes(y = Mean, color = year, shape = status), 
+               position = position_dodge(width = 0.25), 
+               size = 3, stroke = 2, show.legend = F) + 
+    facet_wrap(~index, scales = 'free_y')  +
+  scale_shape_manual(values = c(16, 4)) +
+  theme(axis.text.x = element_text(color = 'black'), 
+        panel.grid.major.x = element_blank() )
+```
+
+```{r, fig.width = 7, fig.height =3}
+StudySums <- 
+  LandsatSums %>%
+    filter(year %in% c(2017:2020)) %>%
+    group_by(year, season, index) %>%
+    summarize(Mean = mean(Value),
+              SEM = sd(Value)/sqrt(n()), 
+              .groups = 'drop') %>%
+    mutate(status = case_when(
+      season == 'Summer' & year %in% c('2018', '2019') ~ "Not completed", 
+      TRUE ~ "Completed"), 
+      index = toupper(index), 
+      index = fct_rev(index) )
+
+HistoricalData <- 
+  bind_rows(
+  LandsatSums %>%
+  filter(year %in% c(1990:1999))%>% 
+  mutate(range = '1990-1999', 
+         index = toupper(index), 
+         index = fct_rev(index)), 
+  LandsatSums %>%
+    filter(year %in% c(2012:2022))%>% 
+    mutate(range = '2012-2022', 
+           index = toupper(index), 
+           index = fct_rev(index)), 
+  LandsatSums %>%
+    mutate(range = '1990-2022', 
+           index = toupper(index), 
+           index = fct_rev(index)) )
+
+  ggplot() + theme_bw(14) + 
+  geom_boxplot(data = HistoricalData  %>% filter(index == 'NDVI'),
+               aes(x=season, 
+                   y = Value),
+               outliers = F, 
+               staplewidth = 0.25) +
+  # geom_errorbar(aes( ymin = Mean - SEM, 
+  #                    ymax = Mean + SEM, 
+  #                    color = year), 
+  #               width = 0.5, 
+  #               position = position_dodge(width = 0.25)) + 
+  geom_point(data = StudySums %>% filter(index == 'NDVI'), 
+            aes(x = season, y = Mean, 
+                color = year, shape = status), 
+             position = position_dodge(width = 0.25), 
+             size = 3, stroke = 2, show.legend = T) + 
+    labs(x = 'Burn season', y = 'NDVI') + 
+  facet_wrap(~range, scales = 'free_y')  +
+  scale_shape_manual(values = c(16, 4), guide = 'none') +
+    scale_color_manual(values = FFgrad) +
+  theme(axis.text.x = element_text(color = 'black'), 
+        panel.grid.major.x = element_blank() ) +
+    guides(color=guide_legend(override.aes = list(pch=15, size =4)))
+```

### 2024-11-25 SpringSummerSeverity/script/ChuggingImagery.R

@@ -9,6 +9,13 @@ patches <- st_read(cgrec_gpkg, 'PasturePatches')
 NoFirePts <- st_read(cgrec_gpkg, 'SamplePoints') %>%
               filter(location == 'Refuge')
 
+NoFirePts %>%
+  st_union() %>%
+  st_bbox() %>%
+  st_as_sfc(., crs=4326) %>%
+  ggplot() + theme_void() +
+    geom_sf() 
+
 # Get veg data for unburned pastures
   imagery_dir = 'S:/DevanMcG/Projects/SpringSummerSeverity/sentinel/MultiBand'
   images <- list.files(imagery_dir)
@@ -202,3 +209,40 @@ fires %>%
     select(fire)
   
   fires %>% filter(fire %in% NegSev$fire)
+  
+# Get Landsat data for unburned pastures
+  imagery_dir = 'S:/DevanMcG/Projects/SpringSummerSeverity/landsat/LandsatSeasonal'
+  images <- list.files(imagery_dir)
+  
+  { 
+    begin = Sys.time() 
+    pacman::p_load(foreach, doSNOW)
+    cores = parallel::detectCores()
+    cl <- makeCluster(cores, methods = F, useXDR = F)
+    registerDoSNOW(cl)
+    NoFireTrends <-
+      foreach(i=1:length(images), 
+              .combine = 'bind_rows',
+              .errorhandling = 'remove', 
+              .packages=c('tidyverse', 'sf')) %dopar% {
+                image = images[i]
+                image_path = paste0(imagery_dir, '/', image)
+                ras <- terra::rast(image_path)
+                ras <- ras[[c('ndvi', 'msi','ndmi')]]
+                 terra::extract(ras,
+                               NoFirePts %>%
+                                 select(pasture, sample) %>%
+                                 st_transform(4326) %>%
+                                 terra::vect() , 
+                               FUN = mean, 
+                               bind = TRUE)  %>%
+                  st_as_sf() %>%
+                  as_tibble() %>%
+                  mutate(ImageDate = tools::file_path_sans_ext(image))  %>%
+                  select(ImageDate, pasture, sample, ndvi:ndmi) 
+              }
+    stopCluster(cl)
+    Sys.time() - begin 
+  }
+  # save(NoFireTrends, file = './data/NoFireTrends.Rdata')
+  

### 2024-11-25 SpringSummerSeverity/script/ChuggingWeather.R

@@ -1,0 +1,68 @@
+pacman::p_load(tidyverse, readxl, RColorBrewer)
+
+seasons <- tibble(season = c('spring', 'summer'), 
+                 start = c('04-20', '07-20'), 
+                 end = c('06-10', '09-10') ) %>%
+          mutate(start = as.Date(start, '%m-%d'), 
+                 end = as.Date(end, '%m-%d'))
+
+BurnDays <- 
+  read_xlsx('./data/BurnData.xlsx', 'BurnDays') %>%
+  filter(is.na(certainty)) %>%
+    mutate(date = paste(year, date), 
+           date = as.Date(date, '%Y %B %d')) %>%
+    select(date) %>%
+  distinct() 
+
+WxData <- lst() 
+# Get daily data 
+  DailyWx <- 
+    read_xlsx('./data/CGREC_weather.xlsx', 'daily')
+# Identify rainy days
+  WxData$Rainfall <-
+    DailyWx %>%
+    select(Year, Month, Day, Rainfall) %>% 
+    unite( c(Month, Day), col = 'day', sep = '-', remove = F) %>%
+    mutate(day = as.Date(day, '%m-%d')  , 
+           season = case_when(
+             between(day, seasons$start[1], seasons$end[1]) ~'Spring', 
+             between(day, seasons$start[2], seasons$end[2]) ~'Summer',
+             TRUE ~ NA      
+           ) ) %>% 
+     unite( c(Year, Month, Day), col = 'date', sep = '-')
+
+
+# Hourly data 
+  WxData$HourlyWx <- 
+    read_xlsx('./data/CGREC_weather.xlsx', 'hourly') %>%
+    filter(between(Hour, 1000, 1700)) %>%
+    unite( c(Year, Month, Day), col = 'date', sep = '-', remove = F) %>%
+    unite( c(Month, Day), col = 'day', sep = '-', remove = F) %>%
+    mutate(day = as.Date(day, '%m-%d'), 
+           season = case_when(
+             between(day, seasons$start[1], seasons$end[1]) ~'Spring', 
+             between(day, seasons$start[2], seasons$end[2]) ~'Summer',
+             TRUE ~ NA      
+           ) ) %>%
+    mutate(e = 6.11 * (10 ^ ( (7.5 * DewPoint)/ (237.3 + DewPoint) ) ), 
+           es = 6.11 * (10 ^ ( (7.5 * AirTemp)/ (237.3 + AirTemp) ) ), 
+           VPD = es - e) 
+# Get burn day weather
[…]

### 2024-11-25 SpringSummerSeverity/script/rgee_landsat_refuge.R

@@ -49,13 +49,21 @@ rgeeExtra::extra_Initialize()
       st_bbox() %>%
       st_as_sfc(., crs=4326) %>%
       sf_as_ee()
+    
+    st_read(cgrec_gpkg, 'SamplePoints') %>%
+      filter(location == 'Refuge')  %>%
+      st_transform(4326) %>%
+      st_union() %>%
+      st_bbox() %>%
+      st_as_sfc(., crs=4326) %>%
+      write_sf('./data/AOI/refuge.shp')
 
 # Turn on GEE connection
   #ee_Initialize(drive = TRUE)
   ee$Initialize(project='ee-devanmcg')
   extra_Initialize()
   
-  years = 1990:2023
+  years = 2019:2023
   seasons = tibble(season = c('spring', 'summer'), 
                    start = c('04-20', '07-20'), 
                    end = c('06-10', '09-10') ) 
@@ -68,38 +76,101 @@ rgeeExtra::extra_Initialize()
   for(s in 1:2) {
     StartDate = paste0(years[y], '-', slice(seasons, s)$start )
     EndDate = paste0(years[y], '-', slice(seasons, s)$end )
-    collection <- case_when(
+    ImageCol <- case_when(
       years[y] %>% between(collections$start[1], collections$end[1]) ~ collections$ee_tag[1], 
       years[y] %>% between(collections$start[2], collections$end[2]) ~ collections$ee_tag[2],
       years[y] %>% between(collections$start[3], collections$end[3]) ~ collections$ee_tag[3]
     )
-  spectra_ee_image <- 
-    ee$ImageCollection(collection)$
-    filterBounds(refuge_ee) %>%
-    ee$ImageCollection$filterDate(StartDate, EndDate) %>%
-    ee$ImageCollection$Extra_preprocess() %>%
-    ee$Image$Extra_maskClouds(prob = 75,buffer = 300,cdi = -0.5) %>%
-    ee$Image$Extra_spectralIndex(c("NDVI",'NDMI',"MSI")) %>%
-    ee$ImageCollection$toBands() 
-    ImageDate <- spectra_ee_image$bandNames()$getInfo()[1] %>% str_sub(13,20)
-    FileDate = paste(str_sub(ImageDate, 1,4), str_sub(ImageDate, 5,6), str_sub(ImageDate, 7,8), sep = '_')
-    names(spectra_ee_image) <- c("NDVI",'NDMI',"MSI")
-
-   ee_as_rast(
[…]

### 2024-11-25 SpringSummerSeverity/script/rgee_sandbox.R

@@ -1,6 +1,6 @@
-pacman::p_load(tidyverse, sf, rgee)
+pacman::p_load(tidyverse, sf, rgee, rgeeExtra)
 
-ee_Initialize(drive = TRUE)
+
 
 lem <- read_sf('C:/Users/devan.mcgranahan/USDA/Rangeland responses to fire - Fire x herbivory/Lemonade Fire/mapping/SpatialData/LemonadePerimeter', 
                'LemonadePerimeter') %>%
@@ -24,9 +24,9 @@ ee$Image$Extra_spectralIndex(c("NDVI", "NDREI", 'NDMI', 'MSAVI', 'SAVI'))
   
   burns_ee <- ee$ImageCollection("COPERNICUS/S2_SR")$
                 filterBounds(lem_box_ee) %>%
-                ee$ImageCollection$filterDate('2022-09-01', '2022-09-30') %>% 
-                ee$ImageCollection$Extra_preprocess() %>%
-                ee$Image$Extra_spectralIndex(c("NBR", "NBR2", 'CSI')) %>% 
+                ee$ImageCollection$Extra_closest("2018-09-22",  2, "week") %>% 
+                #ee$ImageCollection$Extra_preprocess() %>%
+                #ee$Image$Extra_spectralIndex(c("NBR", "NBR2", 'CSI')) %>% 
                 ee$ImageCollection$toBands() 
 names(burns_ee) <- c("NBR", "NBR2", 'CSI')
 burns_ee$bandNames()$getInfo()

### 2024-12-11 SpringSummerSeverity/paper/AllScriptTogetherSI.Rmd
draft SpringSummerSeverity

---
title: Evaluating an attempt to restore summer fire in the Northern Great Plains
author:
- Devan Allen McGranahan and Jay P. Angerer
subtitle:  'Supplementary Information'
header-includes:
  - \usepackage{amsmath}
  - \renewcommand{\familydefault}{\sfdefault}
  - \usepackage[colorlinks=true, citecolor=blue, urlcolor=blue,linkcolor=blue, pdfborder={0 0 0}]{hyperref}
  - \urlstyle{same}
  - \usepackage{xcolor} 
  - \usepackage{booktabs}
  - \raggedright
  - \usepackage{listings}
output:
  pdf_document:
    highlight: tango
    keep_tex: yes
    toc: false
---

This document provides script for fetching remotely-sensed imagery and processing results in the \textsf{R} statistical environment.

# Remote sensing

## Sentinel-2

This \texttt{EvalScript} can be used in the [Copernicus browser](https://browser.dataspace.copernicus.eu/) as a Custom script to create a Custom visualization that can be exported as a 16-bit \texttt{.tiff} for raster analysis. 
The user must select the area of interest and imagery dates and is responsible for assessing cloudiness.

\begin{lstlisting}[language=Java]

//VERSION=3
function setup() {
  return {
    input: ["B04", "B08", "B8A", "B11", "B12"],
    output: { bands: 2, sampleType: "UINT16" }
  };
}

function evaluatePixel(sample) {
  let nbr = index(sample.B08, sample.B12);
  let ndvi = index(sample.B08, sample.B04);
  // apply offset for UINT16 
  return [10000 * nbr + 10000, 
          10000 * ndvi + 10000]; 
}
\end{lstlisting}

\clearpage
## Landsat

This script for [Google Earth Engine](https://code.earthengine.google.com/) exports a composited NDVI \texttt{.tiff} file to the user's Google Drive for the area of interest defined as \texttt{geometry} for an eight week period beginning with each date given in \texttt{listDate}. 
The script combines imagery from Landsat missions 5, 7, \& 8 and handles cloud masking.

\begin{lstlisting}[language=Java]
// Geometry
var geometry: Polygon, 4 vertices
  type: Polygon
  coordinates: List (1 element)
    0: List (5 elements)
      0: [-99.51276195869765,46.71741699210676]
      1: [-99.42521465645156,46.71741699210676]
      2: [-99.42521465645156,46.778346255312734]
      3: [-99.51276195869765,46.778346255312734]
      4: [-99.51276195869765,46.71741699210676]
  geodesic: false

// Main script

var batch = require('users/fitoprincipe/geetools:batch');

///****variables that need to be changed by user
//****ADD google drive directory name that you want to download files to
var gdrivedir = 'landsat'

var listDate= ["1992-04-20", "1992-07-20", "1993-04-20", "1993-07-20", 
               "1994-04-20", "1994-07-20", "1995-04-20", "1995-07-20", 
               "1996-04-20", "1996-07-20", "1997-04-20", "1997-07-20", 
               "1998-04-20", "1998-07-20", "1999-04-20", "1999-07-20", 
               "2000-04-20", "2000-07-20", "2001-04-20", "2001-07-20", 
               "2002-04-20", "2002-07-20", "2003-04-20", "2003-07-20",
               "2004-04-20", "2004-07-20", "2005-04-20", "2005-07-20", 
               "2006-04-20", "2006-07-20", "2007-04-20", "2007-07-20", 
               "2008-04-20", "2008-07-20", "2009-04-20", "2009-07-20",
               "2010-04-20", "2010-07-20", "2011-04-20", "2011-07-20", 
               "2012-04-20", "2012-07-20", "2013-04-20", "2013-07-20", 
               "2014-04-20", "2014-07-20", "2015-04-20", "2015-07-20",
               "2016-04-20", "2016-07-20", "2017-04-20", "2017-07-20", 
               "2018-04-20", "2018-07-20", "2019-04-20", "2019-07-20", 
               "2020-04-20", "2020-07-20", "2021-04-20", "2021-07-20",
               "2022-04-20", "2022-07-20"]

listDate.forEach(function (listDate) {
  //yearRanges.forEach(function (yearRange) {
    exportTimeseries(listDate)
  })
//})
function exportTimeseries(listDate) {

  // Defines a base date/time for the following examples.
  var startDate = ee.Date(listDate);
  print(startDate, 'The start date/time');
  print(startDate.format('YYYYMMdd'))
  var sdate = startDate.format('YYYYMMdd')
  var endDate = startDate.advance(8, 'week')
  print(endDate, 'The end date/time');
  var edate = endDate.format('YYYYMMdd')

  print(endDate.format('YYYYMMdd'))

  //from https://developers.google.com/earth-engine/tutorials/
  //                community/extract-raster-values-for-points
  //function to mask cloud and shadow pixels
  function fmask(img) {
    var cloudShadowBitMask = 1 << 4;
    var cloudsBitMask = 1 << 3;
    var qa = img.select('QA_PIXEL');
    var mask = qa.bitwiseAnd(cloudShadowBitMask).eq(0)
      .and(qa.bitwiseAnd(cloudsBitMask).eq(0));
    return img.updateMask(mask);
  }

  // Selects and renames bands of interest for Landsat OLI.
  function renameOli(img) {
    return img.select(
      ['SR_B2', 'SR_B3', 'SR_B4', 'SR_B5', 'SR_B6', 'SR_B7'],
      ['Blue', 'Green', 'Red', 'NIR', 'SWIR1', 'SWIR2']);
  }

  // Selects and renames bands of interest for TM/ETM+.
  function renameEtm(img) {
    return img.select(
      ['SR_B1', 'SR_B2', 'SR_B3', 'SR_B4', 'SR_B5', 'SR_B7'],
      ['Blue', 'Green', 'Red', 'NIR', 'SWIR1', 'SWIR2']);
  }

  // Prepares (cloud masks and renames) OLI images.
  function prepOli(img) {
    img = fmask(img);
    //img = lcfmask(img)
    img = renameOli(img);
    return img;
  }

  // Prepares (cloud masks and renames) TM/ETM+ images.
  function prepEtm(img) {
    img = fmask(img);
    //img = lcfmask(img)
    img = renameEtm(img);
    return img;
  }

  // Apply scaling factors
  function applyScaleFactors(image) {
    var opticalBands = image.select('SR_B.').multiply(0.0000275).add(-0.2);
    var thermalBands = image.select('ST_B.*').multiply(0.00341802).add(149.0);
    return image.addBands(opticalBands, null, true)
                .addBands(thermalBands, null, true);
  }

  // Get surface reflectance collections for Landsat
  // maps scaling factors and cloud masking functions

  var oliCol = ee.ImageCollection('LANDSAT/LC08/C02/T1_L2')
              .filter(ee.Filter.bounds(geometry))
              .map(applyScaleFactors)
              .map(prepOli);

  var etmCol = ee.ImageCollection('LANDSAT/LE07/C02/T1_L2')
                .filter(ee.Filter.bounds(geometry))
                .map(applyScaleFactors)
                .map(prepEtm);

  var tmCol = ee.ImageCollection('LANDSAT/LT05/C02/T1_L2')
              .filter(ee.Filter.bounds(geometry))
              .map(applyScaleFactors)
              .map(prepEtm) ;

  var landsatCol = oliCol.merge(etmCol).merge(tmCol)
                   .filterDate(startDate, endDate) ;

  // Calculate NDVI 
  var addIndices = function(image) {
     var ndvi = image.normalizedDifference(['NIR', 'Red'])
                .rename('ndvi').float();
  return image.addBands([ndvi]);
  };

  var indices = landsatCol.map(addIndices) 
                .select('ndvi');
  // Create composite of images from date range
    var composite = indices.mean() ;

    var fileName = ee.Date(startDate) 
                  .format('yyyy-MM-dd')
                  .getInfo() ;

  Export.image.toDrive({
      image: composite,
      description: fileName,
      scale: 30,
      folder: 'landsat', 
      region: geometry
      });
}
\end{lstlisting}

\newpage

# Analysis

Because the analysis script assumes hundreds of imagery files and a large Excel file with 42 years of hourly weather observations have been saved locally, it cannot be run directly from this document but is provided here for transparency and reference.

## Remotely sensed imagery

\begin{lstlisting}[language=R]

# Load necessary packages
# Note that package terra is required but is not loaded
# (called directly to not create conflicts with dplyr verbs)
  pacman::p_load(tidyverse, sf, stars, foreach, doSNOW)

# Load location boundaries (data not provided)
  cgrec_gpkg = './CGREC_PBG_26914.gpkg'

  pastures <- st_read(cgrec_gpkg, 'Pastures') 
  patches <- st_read(cgrec_gpkg, 'PasturePatches') 
#
# Get veg data for unburned pastures
#
  NoFirePts <- st_read(cgrec_gpkg, 'SamplePoints') %>%
                filter(location == 'Refuge')
  # Map to directory with Sentinel-2 imagery
    imagery_dir = 'C:/Path/To/Sentinel'
    images <- list.files(imagery_dir)
  # Use parallel processing to chug imagery
  { 
    begin = Sys.time() 
    pacman::p_load()
    cores = parallel::detectCores()
    cl <- makeCluster(cores, methods = F, useXDR = F)
    registerDoSNOW(cl)
    NoFireIndices <-
      foreach(i=1:length(images), 
              .combine = 'bind_rows',
              .errorhandling = 'remove', 
              .packages=c('tidyverse', 'sf')) %dopar% {
        image = images[i]
        image_path = paste0(imagery_dir, '/', image)
        ras <- terra::rast(image_path)
        names(ras) <- c('nbr', 'ndvi') 
        ras <- ras[['ndvi']]
        float = (ras-10000)/10000
        terra::extract(float,
          NoFirePts %>%
            select(pasture, sample) %>%
            terra::vect() , 
          FUN = mean, 
          bind = TRUE)  %>%
        st_as_sf() %>%
          as_tibble() %>%
        mutate(ImageDate = substr(image, 1, 10)) %>%
        select(ImageDate, pasture, sample, ndvi) 
            }
    stopCluster(cl)
  Sys.time() - begin 
  }
#
# Get fuel greenness and dNBR for completed burns
#
  fires <- st_read(cgrec_gpkg, 'FirePerimeters') %>%
              mutate(Year = as.factor(Year)) %>%
              filter(status == 'Completed') %>%
              unite('fire', c(unit, Pasture, Patch), sep = "-") %>%
              select(fire, Year, Season, PreBurn, PostBurn )
  SamplePts <- 
    st_read(cgrec_gpkg, 'SamplePointsRegular')
{ 
  begin = Sys.time() 
  pacman::p_load(foreach, doSNOW)
  cores = parallel::detectCores()
  cl <- makeCluster(cores , methods = F, useXDR = F)
  registerDoSNOW(cl)
  BurnIndices <-
    foreach(i=1:length(fires$fire), 
            .combine = 'bind_rows',
            .errorhandling = 'remove', 
            .packages=c('tidyverse', 'sf')) %dopar% {
              # Get fire
                fire = slice(fires, i)
              # Get sample points
                pts <-
                  fire %>%
                    st_intersection(SamplePts)
              # Get dates
                pre_date = fire$PreBurn
                post_date = fire$PostBurn
              # Fetch & process multi-band rasters
                # pre image
                  pre_image = images[substr(images, 1, 10) == pre_date]
                  pre_path = paste0(imagery_dir, '/', pre_image)
                  pre_ras <- terra::rast(pre_path) %>%
                                terra::crop(terra::vect(fire))
                  names(pre_ras) <- c('nbr', 'ndvi') 
                  pre_ras <-  (pre_ras-10000)/10000
                # post-fire
                  post_image = images[substr(images, 1, 10) == post_date]
                  post_path = paste0(imagery_dir, '/', post_image)
                  post_ras <- terra::rast(post_path)[[1]] %>%
                                terra::crop(terra::vect(fire))
                  post_ras = (post_ras-10000)/10000
                # Calculate dNBR & replace in pre-fire raster
                  d_ras = pre_ras['nbr'] - post_ras
                  names(d_ras) <- 'dNBR'
                  pre_ras[[1]] <- d_ras
              # Sample rasters
                 terra::extract( pre_ras,
                          pts %>%
                            select(-PostBurn) %>%
                            terra::vect() , 
                          FUN = mean, 
                          bind = TRUE)  %>%
                  st_as_sf() %>%
                    as_tibble() %>%
                    select(-geometry) 
            }
  stopCluster(cl)
  Sys.time() - begin 
}
#
# Get historical Landsat data for unburned pastures
#
  imagery_dir = 'C:/Path/To/Landsat'
  images <- list.files(imagery_dir)
  { 
    begin = Sys.time() 
    pacman::p_load(foreach, doSNOW)
    cores = parallel::detectCores()
    cl <- makeCluster(cores, methods = F, useXDR = F)
    registerDoSNOW(cl)
    NoFireTrends <-
      foreach(i=1:length(images), 
              .combine = 'bind_rows',
              .errorhandling = 'remove', 
              .packages=c('tidyverse', 'sf')) %dopar% {
                image = images[i]
                image_path = paste0(imagery_dir, '/', image)
                ras <- terra::rast(image_path)
                ras <- ras[['ndvi']]
                 terra::extract(ras,
                               NoFirePts %>%
                                 select(pasture, sample) %>%
                                 st_transform(4326) %>%
                                 terra::vect() , 
                               FUN = mean, 
                               bind = TRUE)  %>%
                  st_as_sf() %>%
                  as_tibble() %>%
                  mutate(ImageDate = tools::file_path_sans_ext(image))  %>%
                  select(ImageDate, pasture, sample, ndvi) 
              }
    stopCluster(cl)
    Sys.time() - begin 
  }
\end{lstlisting}

## Weather data

This script wrangles the weather data downloaded from the [North Dakota Ag Weather Network's Streeter station](https://ndawn.ndsu.nodak.edu/station-info.html?station=48).

\begin{lstlisting}[language=R]

pacman::p_load(tidyverse, readxl)

seasons <- tibble(season = c('spring', 'summer'), 
                 start = c('04-20', '07-20'), 
                 end = c('06-10', '09-10') ) %>%
          mutate(start = as.Date(start, '%m-%d'), 
                 end = as.Date(end, '%m-%d'))

BurnDays <- 
  read_xlsx('./data/BurnData.xlsx', 'BurnDays') %>%
  filter(is.na(certainty)) %>%
    mutate(date = paste(year, date), 
           date = as.Date(date, '%Y %B %d')) %>%
    select(date) %>%
  distinct()

WxData <- lst() 
# Get daily data 
  DailyWx <- 
    read_xlsx('./data/CGREC_weather.xlsx', 'daily')
# Identify rainy days
  WxData$Rainfall <-
    DailyWx %>%
    select(Year, Month, Day, Rainfall) %>% 
    unite( c(Month, Day), col = 'day', sep = '-', remove = F) %>%
    mutate(day = as.Date(day, '%m-%d')  , 
           season = case_when(
             between(day, seasons$start[1], seasons$end[1]) ~'Spring', 
             between(day, seasons$start[2], seasons$end[2]) ~'Summer',
             TRUE ~ NA      
           ) ) %>% 
     unite( c(Year, Month, Day), col = 'date', sep = '-')

# Hourly data 
  WxData$HourlyWx <- 
    read_xlsx('./data/CGREC_weather.xlsx', 'hourly') %>%
    filter(between(Hour, 1000, 1700)) %>%
    unite( c(Year, Month, Day), col = 'date', sep = '-', remove = F) %>%
    unite( c(Month, Day), col = 'day', sep = '-', remove = F) %>%
    mutate(day = as.Date(day, '%m-%d'), 
           season = case_when(
             between(day, seasons$start[1], seasons$end[1]) ~'Spring', 
             between(day, seasons$start[2], seasons$end[2]) ~'Summer',
             TRUE ~ NA      
           ) ) %>%
    mutate(e = 6.11 * (10 ^ ( (7.5 * DewPoint)/ (237.3 + DewPoint) ) ), 
           es = 6.11 * (10 ^ ( (7.5 * AirTemp)/ (237.3 + AirTemp) ) ), 
           VPD = es - e) 
# Get burn day weather
  WxData$BurnDayWx <-
    HourlyWx  %>% 
    filter(date %in% BurnDays$date) %>%
    group_by(Year, date, season) %>%
    summarise_at(.vars = vars(c(DewPoint, RelHum, WindSpeed,VPD)),
                 .funs = c("mean")) %>%
    ungroup() %>%
    pivot_longer(names_to = 'variable', 
                 values_to = 'value', 
                 cols = DewPoint:VPD) %>%
    mutate(variable = recode(variable, 
                             'DewPoint' = 'Dew Point (°C)', 
                             'RelHum' = 'Relative Humidity (%)', 
                             'VPD' = 'Vapor Pressure Deficit', 
                             'WindSpeed' = 'Wind speed (m/s)') , 
           Year = as.character(Year)) 
\end{lstlisting}

### 2024-12-11 SpringSummerSeverity/paper/AllScriptTogetherSI.tex
draft SpringSummerSeverity

--- 2024-12-11 SpringSummerSeverity/paper/AllScriptTogetherSI.Rmd
@@
----
-title: Evaluating an attempt to restore summer fire in the Northern Great Plains
-author:
-- Devan Allen McGranahan and Jay P. Angerer
-subtitle:  'Supplementary Information'
-header-includes:
-  - \usepackage{amsmath}
-  - \renewcommand{\familydefault}{\sfdefault}
-  - \usepackage[colorlinks=true, citecolor=blue, urlcolor=blue,linkcolor=blue, pdfborder={0 0 0}]{hyperref}
-  - \urlstyle{same}
-  - \usepackage{xcolor} 
-  - \usepackage{booktabs}
-  - \raggedright
-  - \usepackage{listings}
-output:
-  pdf_document:
-    highlight: tango
-    keep_tex: yes
-    toc: false
----
+% Options for packages loaded elsewhere
+\PassOptionsToPackage{unicode}{hyperref}
+\PassOptionsToPackage{hyphens}{url}
+%
+\documentclass[
+]{article}
+\usepackage{amsmath,amssymb}
+\usepackage{iftex}
+\ifPDFTeX
+  \usepackage[T1]{fontenc}
+  \usepackage[utf8]{inputenc}
+  \usepackage{textcomp} % provide euro and other symbols
+\else % if luatex or xetex
+  \usepackage{unicode-math} % this also loads fontspec
+  \defaultfontfeatures{Scale=MatchLowercase}
+  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+\fi
+\usepackage{lmodern}
+\ifPDFTeX\else
+  % xetex/luatex font selection
+\fi
+% Use upquote if available, for straight quotes in verbatim environments
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+\IfFileExists{microtype.sty}{% use microtype if available
+  \usepackage[]{microtype}
+  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
+}{}
+\makeatletter
+\@ifundefined{KOMAClassName}{% if non-KOMA class
+  \IfFileExists{parskip.sty}{%
+    \usepackage{parskip}
+  }{% else
+    \setlength{\parindent}{0pt}
+    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
+}{% if KOMA class
+  \KOMAoptions{parskip=half}}
+\makeatother
+\usepackage{xcolor}
+\usepackage[margin=1in]{geometry}
+\usepackage{graphicx}
+\makeatletter
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
+\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
+\makeatother
+% Scale images if necessary, so that they will not overflow the page
+% margins by default, and it is still possible to overwrite the defaults
+% using explicit options in \includegraphics[width, height, ...]{}
+\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+% Set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\makeatother
+\setlength{\emergencystretch}{3em} % prevent overfull lines
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
+\usepackage{amsmath}
+\renewcommand{\familydefault}{\sfdefault}
+\usepackage[colorlinks=true, citecolor=blue, urlcolor=blue,linkcolor=blue, pdfborder={0 0 0}]{hyperref}
+\urlstyle{same}
+\usepackage{xcolor}
+\usepackage{booktabs}
+\raggedright
+\usepackage{listings}
+\ifLuaTeX
+  \usepackage{selnolig}  % disable illegal ligatures
+\fi
+\usepackage{bookmark}
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\urlstyle{same}
+\hypersetup{
+  pdftitle={Evaluating an attempt to restore summer fire in the Northern Great Plains},
+  pdfauthor={Devan Allen McGranahan and Jay P. Angerer},
+  hidelinks,
+  pdfcreator={LaTeX via pandoc}}
 
-This document provides script for fetching remotely-sensed imagery and processing results in the \textsf{R} statistical environment.
+\title{Evaluating an attempt to restore summer fire in the Northern
+Great Plains}
+\usepackage{etoolbox}
+\makeatletter
+\providecommand{\subtitle}[1]{% add subtitle to \maketitle
+  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
+}
+\makeatother
+\subtitle{Supplementary Information}
+\author{Devan Allen McGranahan and Jay P. Angerer}
+\date{}
 
-# Remote sensing
+\begin{document}
+\maketitle
 
-## Sentinel-2
+This document provides script for fetching remotely-sensed imagery and
+processing results in the \textsf{R} statistical environment.
 
-This \texttt{EvalScript} can be used in the [Copernicus browser](https://browser.dataspace.copernicus.eu/) as a Custom script to create a Custom visualization that can be exported as a 16-bit \texttt{.tiff} for raster analysis.
-The user must select the area of interest and imagery dates and is responsible for assessing cloudiness.
+\section{Remote sensing}\label{remote-sensing}
+
+\subsection{Sentinel-2}\label{sentinel-2}
+
+This \texttt{EvalScript} can be used in the
+\href{https://browser.dataspace.copernicus.eu/}{Copernicus browser} as a
+Custom script to create a Custom visualization that can be exported as a
+16-bit \texttt{.tiff} for raster analysis. The user must select the area
+of interest and imagery dates and is responsible for assessing
+cloudiness.
 
@@
 \clearpage
-## Landsat
 
-This script for [Google Earth Engine](https://code.earthengine.google.com/) exports a composited NDVI \texttt{.tiff} file to the user's Google Drive for the area of interest defined as \texttt{geometry} for an eight week period beginning with each date given in \texttt{listDate}.
-The script combines imagery from Landsat missions 5, 7, \& 8 and handles cloud masking.
+\subsection{Landsat}\label{landsat}
+
+This script for \href{https://code.earthengine.google.com/}{Google Earth
+Engine} exports a composited NDVI \texttt{.tiff} file to the user's
+Google Drive for the area of interest defined as \texttt{geometry} for
+an eight week period beginning with each date given in
+\texttt{listDate}. The script combines imagery from Landsat missions 5,
+7, \& 8 and handles cloud masking.
 
@@
 
-# Analysis
+\section{Analysis}\label{analysis}
 
-Because the analysis script assumes hundreds of imagery files and a large Excel file with 42 years of hourly weather observations have been saved locally, it cannot be run directly from this document but is provided here for transparency and reference.
+Because the analysis script assumes hundreds of imagery files and a
+large Excel file with 42 years of hourly weather observations have been
+saved locally, it cannot be run directly from this document but is
+provided here for transparency and reference.
 
-## Remotely sensed imagery
+\subsection{Remotely sensed imagery}
+
+The following \textsf{R} script processes the imagery produced by the scripts above after they have been saved to a drive that can be mapped to from the session as \texttt{imagery\_dir}.
 
@@
 
-## Weather data
+\newpage
+\subsection{Weather data}\label{weather-data}
 
-This script wrangles the weather data downloaded from the [North Dakota Ag Weather Network's Streeter station](https://ndawn.ndsu.nodak.edu/station-info.html?station=48).
+This script wrangles the weather data downloaded from the
+\href{https://ndawn.ndsu.nodak.edu/station-info.html?station=48}{North
+Dakota Ag Weather Network's Streeter station}.
 
@@
                  values_to = 'value', 
-                 cols = DewPoint:VPD) %>%
-    mutate(variable = recode(variable, 
-                             'DewPoint' = 'Dew Point (°C)', 
-                             'RelHum' = 'Relative Humidity (%)', 
-                             'VPD' = 'Vapor Pressure Deficit', 
-                             'WindSpeed' = 'Wind speed (m/s)') , 
-           Year = as.character(Year)) 
+                 cols = DewPoint:VPD) 
 \end{lstlisting}
+
+\end{document}

### 2024-12-11 SpringSummerSeverity/paper/CoverLetter/CoverLetter0.tex
draft SpringSummerSeverity

\documentclass[parskip=half, 
			   fontsize=12pt,
			   paper=a4]				
{scrartcl}
\input{SetupOptions} % contains scrletter

\newcommand{\journal}{\emph{Restoration Ecology}}

\vspace{-5em}

\begin{document}
	\begin{letter}{Editorial staff\\
		\journal }
\setlength{\parindent}{12pt}

\opening{Dear Sir or Madam,}

Please consider the attached submission for publication in \journal. 
We are excited to leverage a unique dataset to address a novel question about the restoration of pre-European fire regimes: How successful are attempts at summer burning?
Much research attention has been given to variation in fire effects across different seasons, but to our knowledge no one has yet directly addressed the feasability of burning during the growing season, along wth a comparison of burn outcomes (not fire effects) relative to conditions of the fire environment.

We also provide a historical context for the fuel and weather variables considered here, however the limited spatial resolution of the research station represents a narrow scope. 
Indeed, while the relatively high number of 40 fire events across four years provides plenty of food for thought, the present paper is best approached as a case study with a narrow scope in which we demonstrate how various data products can be used to assess patterns in fire activity and outcomes at broader scales.

We've provided suggested reviewers as part of the online submission.

We look forward to working with the editors and reviewers of \journal~to ensure our manuscript meets the standards of the journal.

\vspace{-3em} 
\closing{} %eg. Regards
%\cc{another dude}
\end{letter}

\end{document}

### 2024-12-11 Seasonality/script/SpatialWrangling.R

@@ -1,0 +1,33 @@
+pacman::p_load(tidyverse, sf)
+
+st_layers('./gis/GreatPlains.gpkg')
+gp <- read_sf('./gis/GreatPlains.gpkg', 'l3_state_boundaries') 
+
+padus %>%
+  ggplot() + theme_void() +
+  geom_sf() 
+
+# BLM's SMA 
+  st_layers("S:/DevanMcG/GIS/SpatialData/US/BLM/SMA_WM.gdb")
+  sma_us <- read_sf("S:/DevanMcG/GIS/SpatialData/US/BLM/SMA_WM.gdb", 'SurfaceManagementAgency')
+
+# USGS PADUS 4.0
+  st_layers("S:/DevanMcG/GIS/SpatialData/US/USGS/PADUS/PADUS4_0_Geodatabase.gdb")
+  padus <- read_sf("S:/DevanMcG/GIS/SpatialData/US/USGS/PADUS/PADUS4_0_Geodatabase.gdb", 
+                   'PADUS4_0Combined_Proclamation_Marine_Fee_Designation_Easement') %>%
+    st_transform(st_crs(gp)) %>%
+    st_intersection(gp) 
+
+
+sma_us %>%
+  ggplot() + theme_void() +
+  geom_sf() 
+sma_us %>%
+  st_transform(st_crs(gp)) %>%
+  
+  st_intersection(gp) 
+
+
+unique(sma_us$ADMIN_ST)
+
+unique(sma_us$ADMIN_AGENCY_CODE)

### 2024-12-11 script/ShapefileToGeoPackage.R

@@ -1,0 +1,50 @@
+pacman::p_load(tidyverse, sf)
+
+# Creating
+
+  # set a file path to where your spatial data live
+    fp = 'C:/Path/To/Spatial/Data'
+  
+  # import a shapefile
+    shp <-
+      paste0(fp, '/ShapefileName') %>%  # shapefile name
+        read_sf('Layer') # specific layer
+    
+  # convert layer to GeoPackage
+    shp %>%
+      st_write(paste0(fp, '/GeoPackageName.gpkg', 'LayerName')
+               
+# Loading
+  # specify the GeoPackage you want
+      gpkg = paste0(fp, '/GeoPackageName.gpkg')
+  # View available layers
+      st_layers(gpkg ) 
+  # load a layer
+      layer <- st_read(gpkg, 'LayerName') 
+      
+      
+shps <- list.files(fp)
+
+for(i in 1:length(shps)){
+  f = shps[i]
+  shp <-
+    paste0(fp, '/', f) %>%
+    st_write(paste0(fp, '/GeoPackageName.gpkg', f)
+  
+  
+}
+
+
+fp = 'C:/Path/To/Spatial/Data'
+shps <- list.files(fp, pattern = "\\.shp$") # should return unique list of shapefile names??
+
+for(i in 1:length(shps)){
+  f = shps[i]
+  lyrs = st_layers(f)
+  for(l in 1:length(lyers)){
+    lyr = lyrs[l]
+  paste0(fp, '/', f) %>%
+    read_sf(lyr) %>%
+    st_write(paste0(fp, '/GeoPackageName.gpkg', f))
+  }
+}

### 2024-12-12 Seasonality/script/SpatialWrangling.R
update perimeters

@@ -1,23 +1,53 @@
 pacman::p_load(tidyverse, sf)
 
-st_layers('./gis/GreatPlains.gpkg')
-gp <- read_sf('./gis/GreatPlains.gpkg', 'l3_state_boundaries') 
-
-padus %>%
+st_layers('./Seasonality/data/spatial/seasonality.gpkg')
+gp_l3 <- read_sf('./Seasonality/data/spatial/seasonality.gpkg', 'l3_gp_state_boundaries') 
+gp <- read_sf('./Seasonality/data/spatial/seasonality.gpkg', 'StudyRegion') 
+gp %>%
   ggplot() + theme_void() +
-  geom_sf() 
+  geom_sf()
 
 # BLM's SMA 
-  st_layers("S:/DevanMcG/GIS/SpatialData/US/BLM/SMA_WM.gdb")
-  sma_us <- read_sf("S:/DevanMcG/GIS/SpatialData/US/BLM/SMA_WM.gdb", 'SurfaceManagementAgency')
+  # Load as SpatVector
+    sma_gp <-     
+      terra::vect('S:/DevanMcG/GIS/SpatialData/US/BLM/SMA_WM.gdb', 
+                  'SurfaceManagementAgency') 
+  # Convert to sf
+    sma_sf <- sma_gp %>% st_as_sf() 
+  # Create gp object in SMA CRS
+    gp_3857 <- gp %>% 
+      st_transform(st_crs(sma_sf)) 
+  # crop SMA to GP 
+    sma_sf %<>% select(ADMIN_ST, ADMIN_AGENCY_CODE) %>% st_intersection(gp_3857)
+    
+    sma_st <- sma_sf %>%
+      filter(ADMIN_AGENCY_CODE == "ST") %>%
+      st_transform(st_crs(gp))
+    
+    sma_st %>%
+      ggplot() + theme_void() +
+      geom_sf() +
+      geom_sf(data = gp, fill = NA, color = 'blue')
+  
+  sma_ac <- 
+    sma_st  %>%
+      mutate(acres = st_area(.)*0.0002471054, 
+             acres = as.numeric(acres)) %>%
+      select(ADMIN_ST, acres)
 
 # USGS PADUS 4.0
   st_layers("S:/DevanMcG/GIS/SpatialData/US/USGS/PADUS/PADUS4_0_Geodatabase.gdb")
[…]

### 2025-04-21 SpringSummerSeverity/paper/ReviewResponses/ReviewResponse1.tex
Spring/summer seasonality R1 update

\documentclass[parskip=half, american]{scrartcl}
\usepackage{booktabs}
\usepackage[margin=2.5cm]{geometry}
\usepackage[dvipsnames]{xcolor}
\usepackage[colorlinks=true,
			citecolor=blue,
			urlcolor=blue,
			linkcolor=blue,
			pdfborder={0 0 0}]
	{hyperref}

\newcommand{\AR}[1]
	{\color{PineGreen}AR: #1\color{black} \par }

\newcommand{\journal}{\emph{Environmental Management}}

\title{\Large Evaluating an attempt to restore summer fire in the Northern Great Plains }
\subtitle{Response to first review}
\author{ }
\date{}

% \pagenumbering{gobble}
\raggedright

\begin{document}

\maketitle

\vspace{-2cm}

\AR{We appreciate the opportunity to revise this submission to \journal, and trust the editor will find these revisions address the important reviewer concerns. 
	Author responses (AR:) to each comment follow inline below, as colored text.\\
\vspace{1em}
Major points of revision include:
\begin{itemize}
\item Addition of statistical analysis for all graphed data and discussed relationships
\item Dropped references to burn ``success'' and emphasis on whether burns were completed as planned or not
\item Expanded discussion of land surface phenology and interactions with non-native cool-season grasses
\end{itemize} }

\section*{Reviewer 1}

I have read the manuscript ``Evaluating an attempt to restore summer fire in the Northern Great Plains'' aimed for publication in Environmental Management.

The study described four years of successful/unsuccessful prescribed burns on North Dakota grasslands. As a way to restore old traditions of a bimodal burn pattern of anthropogenic burns, greenness (in terms of satellite-retrieved NDVI) burn severity (differenced Normalized Burn Ratio) we studied during sprig and summer season burn series over the four years. The authors conclude that (1) spring burn dNBR decerases with increased NDVI, (2) summer burns are possible (50\% success rate in the study material), (3) summer burns can for some years be achieved with high burn severity and (4) 42 years of historic data indicate that conditions (weather and greenness) suitable for summer burns are more of the norm than an anomaly. Thus, late growing season burning could be included in prescribed fire programs (although not always achievable).

\AR{We appreciate the reviewer's attention to detail and accurate summary of the manuscript. }

My general positive thoughts on the manuscript are
1.	I really enjoyed reading it. The language is in many parts well-balanced, the text is short without unnecessary parts, and I found the results interesting.
2.	The study is well limited and the method sound.
3.	The topic is interesting and important.
4.	The study is based on an assessment of plenty field work.

\AR{ We greatly appreciate this feedback. }

On the other hand, although I praised the short and concise text, as interested in grassland fires, I feel some info on the experiments are missing:
5.	In addition to the NDVI, did you conduct and other measurement on curing degree? Either destructive sampling, visual assessment or levy-rod transects (as detailed in Anderson et al (2011) Int J Wildland Fire 20, 804–814).

\AR{In short, no. Fuels data for a subset of the spring fires were reported in the cited publication \emph{McGranahan et al 2023 Weather and fuel as modulators of grassland fire behavior in the northern Great Plains. Environmental Management 71:940–949}, but the present manuscript considers a broader suite of fires for which field data are not available.   }

6.	Were there any moisture content measurements of full fuel beds or the dead components during the burns?

\AR{In short, no; see response above. }

7.	Although I appreciate and respect that 48 burning attempts and 40 successful burns constitute a huge workload and plenty of data, the conclusion seems somewhat based on anecdotal evidence. This could however be covered if the authors dwell a bit on the generalisation to these specific two grass species, which are stated to be invasive. This is not clear from someone not familiar with North Dakota grasslands.

\AR{We agree with the reviewer's assessment. 
This is the main reason that we have submitted as a ``Methodology'' type manuscript, rather than an ``Original research paper'', which the reviewer might not have been aware of. 
Anyway, we recognize the limitations of extrapolating study results and offer the paper more as a case study on feasibility of burning in the summer, and effects of summer fires relative to spring burns. 
We have attempted to highlight this caveat throughout the revision.   }

However, as the topic is important and I feel the results should be published, I recommend acceptance subject to revisions. Below you will find some more detailed comments.

8.	Abstract: ``We found that burn severity declined with fuelbed greenness but was independent of burn season—summer burns could effect as high of severity as spring burns despite having greener fuelbeds''. The sentence makes sense but is a bit difficult at first reading and also when assessing the figures. Severity declines with greenness, but is independent on seasons. Summer burns have a high greenness but equally high severity. Thus, severity does not decline with greenness, or only within spring fires?

\AR{We have rewritten and restructured the first clause to accommodate the seasonal differences described in the second clause and depicted in the figures.  }

9.	Abstract: Next sentence. I would recommend adding a time period to the phenological change (such as ``four decades'' or so), as to not confuse it with changes over a single season.

\AR{Good point. We've added ``over 42 years''.  }

10.	Methods. Study location: In the introduction you discuss historical burn patterns but the grass species you describe is a non-native species. Could you dwell a bit on this, either here, in the introduction or the discussion. It should not be left uncommented.

\AR{ The effects of non-native grasses on fire behavior were specifically considered in the Discussion in the original submission, and this discussion remains in the revised version, as well.
Along with this and the other reviewer's suggestion, we have added material to the Discussion about drought susceptibility. }

11.	Methods. Study location: For someone not familiar with the local grasslands you are studying, could you provide us with a short description of these species in for this location? Examples include litter fuel load in spring, annual production, a characteristic season (growth peak, curing, onset of growth), characteristic height of the fuelbed (spring height after snow compression and typical height during the summer burns).

\AR{ While we appreciate the reviewer's interest, these details go beyond the scope of this paper and frankly beyond the scope of the paper already reporting specific components of the wildland fire environment.  }

12.	Methods. Study location: Photos would be helpful. If you don’t want to fill the manuscript with more figures, there is plenty of room in the supplementary information, which now only contains code.

\AR{ It is unclear what sort of photos the reviewer would find helpful. 
While we could fill the Supplementary Information with all sorts of cellphone photos from prescribed fire operations, their utility to the manuscript are unclear. }

13.	Methods. Data: Was there no post fire ground assessment of the data? Although dNBR is a good tool for burn severity over large areas did you sample e.g. remaining fuel after burns? This is especially interesting for the summer burns.

\AR{No ground assessments were conducted.  }

14.	Methods. Data. Line 136: (1000-1700) is this time. I would recommend 10:00 – 17:00 or 10.00 – 17.00 and the time zone this is evaluated in to avoid confusion.

\AR{Presentation of the burn period has been revised. 
Time zone is irrelevant in this context\textemdash in fact time zones exist such that times can be standardized within the diurnal period.  }

15.	Results: Please describe the aftermath of summer burns. For low severity, did only cured and semi-cured fuel burn while green was not consumed, or was the fuel bed consumed completely for patches were fire could spread but low severity implies a more mosaic post-burn fuel with more green areas and less consumed areas?

\AR{Such qualitative assessments were not made.  }

16.	Discussion: Please discuss why the burn severity is so high during the

\AR{ Unfortunately the reviewer's comment is abbreviated and remains too vague to infer what was meant. }

17.	Discussion: Lines 199-202: I think you should dwell a bit more on this. Can you assess the magnitude of these changes compared to the variables (greenness and weather) you studied here?

\AR{We appreciate the reviewer's focus on this important topic. 
It is unclear what additional dwelling might be done beyond the specific effects listed here. 
The cited papers, especially Yurkonis et al. 2019, substantiate this list\textemdash indeed, `assess the magnitude' as the reviewer says\textemdash  by using fire behavior models to show that early green-up reduces rate of spread. 
We feel that the point has been sufficiently delivered here based on the literature we are aware of.  }

18.	Results or Discussion: Please discuss also the effect of grazing. Was it uniform given your relatively large burn plots? Is one of the species more attractive to grazing then the other and does that influence the amount of curing?

\AR{We have no data on grazing $\times$ fire environment interactions. 
In general we have no ancedotal evidence that there was any particular effect of cattle being on these pastures\textemdash other than general reductions in potential standing biomass through offtake\textemdash nor do we have any data or observations related to the specific points posited by the reviewer.
It might be worth remembering that in most rangelands, grazing is the norm, not the exception.    }

19.	Figure 1: Summer pattern is described by only eight data points. If the 2017 data with NDVI~0.42 is ignored you’ll retrieve a strong burn severity dependence on NDVI. Could you assess the ground data of that point? Any differences in height, fuel load or pressure from the pastures?

\AR{ It is unclear what the reviewer is referring to\textemdash all summer NDVI values, 2017 \& 2018, are well above 0.42. 
If the reviewer meant to say \emph{dNBR} value above 0.42, we point out that the two 2017 dNBR values that exceed 0.4 represent the lowest and highest values of the NDVI range, so striking values above 0.42 would not improve the linear fit.  
Either way, unfortunately ground-level assessments of past fires is not possible. 
To our knowledge these pastures had been managed consistently prior to and during our study. }

20.	Figure 3. Caption: (Minor comment) the boxplots do not represent historical trends, do they? I would consider them being historical values rather than trends.

\AR{While we feel that comparing a latter time period to an earlier one does seem to count as a trend, we take the point and have revised to more accurately speak to the comparison of the endpoints.    }

Finally, after this list of potential improvements, I'd like to repeat that I enjoyed reading the manuscript.

\AR{We very much appreciate the reviewer's consideration, helpful comments, and support for publication. }

\section*{Reviewer 2}

The authors present an analysis describing the effects of fuel and weather conditions on successful or unsuccessful summer burns and place these burn seasons into historical context.  Their objective is to determine whether summer burning is a possible management option in addition to spring burning.  I have several questions about this analysis.
What defines a successful or unsuccessful burn? Is it a certain percentage of a patch burned? Was a patch unburned because the burn boss decided not to that season?  If that’s the case, why didn’t they decide to burn?

\AR{References to ``success'' have been replaced with specific language about completed vs. not completed burns. }

Is there a significant difference between spring and summer NDVI, cumulative rainfall, dew point, or any of the other variables? How different are the means and variability?  Can burn success be modeled as a function of the variables they describe?  What are the models shown in fig 1b, how well do they fit, and what does the shaded area show?

\AR{Extensive statistical tests have been added for all relationships. }

In addition, I have several line comments

63:  This is an interesting paragraph, but it seems out of place between the previous and next paragraphs.  Either moving it up to line 49 or into the study area part of the methods may fit better.

\AR{After consideration, we disagree with moving the paragraph. 
 	It ought not go before the previous paragraph (L 49 as the reviewer first suggested) because the focus is on the Northern Great Plains specifically, while the previous paragraph discusses the entire Great Plains and includes references to similar grassland systems worldwide. 
Thus moving it up would re-expand the narrowing scope of the Introduction. 
Moving it out of the Introduction entirely is not desirable, because in addition to framing our geographic focus\textemdash indeed, suitable for Methods material\textemdash this paragraph makes the socio-ecological case for this narrower geographic region still being of interest to an international audience because of the extreme discrepency between what is understood to be the pre-colonial fire regime and the one that has emerged from conventional management (and is thus so sought to be reversed by the topic of our study, restoring summer burning). 
This is a core conceptual aspect that merits inclusion in the Introduction as a selling point for the interesting components of our paper.  }

78.  What do you mean by `successful burn'?

\AR{We appreciate this opportunity to clarify the language in the paper. 
It was our intention that readers would understand our description in the first clause of the sentence\textemdash ``all spring burns were completed as planned''\textemdash to be synonymous with ``success'' in the second clause\textemdash i.e. \emph{not} completed as planned\textemdash but apparently this was not clear. 
We've revised to be more explicit, attempting generally to refrain from the word ``success''.   }

80.  So this objective really seems to hinge on the definition of success.  Are they successful because they were ignited and burned the specified area? Or unsuccessful because they were ignited and didn't burn?

\AR{The former. We've revised to simply speak of completing or not completing planned burns.  }

90 \& 96.  There are two strategies that involve spring burning – were both strategies reflected in the 32 planned spring burns, or are those just for the spring + summer burning?

\AR{All spring burns are included in the 32 spring burns.  }

92.  Livestock access is an unreported, possibly important variable.  Why wasn't it included, and if it’s not impactful, a line saying why it’s not is needed.

\AR{All pastures were grazed under exactly the same stocking regime. 
Without variability or comparison, there is little to be said quantitatively about any effects of grazing beyond the fact that it existed, which we have done. No relevant qualitative observations come to mind. It seems a slippery slope to start adding lines about why each thing that was not `impactful' was not considered; we've focused our discussion on elements that are relevant.
Again, as mentioned above, it is the norm for rangelands to be grazed, not the exception.  }

99.  How did the manner of ignition impact success/unsuccessful summer fires?

\AR{In short, it did not. Ignition patterns were selected and adapted to facilitate fire spread. }

136.  1000-1700 hours?

\AR{ Indeed, revised accordingly. }

146. How sensitive is the analysis to the selection of decade period?  If 1991-2000 were chosen, would the results change?  Are there any anomalous weather years that should be excluded from each decade? How were they summarized?

\AR{While there is a potential for different years slightly changing summary statistics, the nature of central tendency ensures the effects on the overall trends reported here would be minimal. 
We are unaware of `anomalous' weather years and are not sure what that would mean\textemdash it seems beyond the scope of this paper to define a metric for anomalous deviation from some baseline and create selection criteria to determine if a given year should be excluded from a 10-year mean. 
We are simply comparing the mean of a decent chunk of time at the beginning of the available weather period to a comparable chunk of time at the end. 
It seems excessive, especially for a case report from a single weather station, to create a bunch of rules for inclusion/exclusion when there is no apparent need.    }

152.  So a linear models is presented in figure 1b – what are the details of these models?  What are the slopes, r2,  etc?  This sort of information is important for the reader to make their own determination about whether a linear analysis is appropriate.  What about subsampling each fire and use a mixed effects model?

\AR{ Extensive statistical tests have been added for all relationships. }

156 -166.  These statements all need some sort of statistic or values with them.  What are the typical ranges? What is atypically high?  Are these statistically significant differences by some sort of comparative test?

\AR{Extensive statistical tests have been added for all relationships.  }

168.  This is something for the discussion rather than the results.

\AR{This has been removed.  }

177.  Again, if success is due to a burn manager deciding not to burn, then the analysis and its interpretation changes.  If it’s due to a percent unburned or incapable of burning, then this statement is much more impactful.

\AR{ We appeciate the reviewer's insight into helping us focus on the impactful elements of the story here. 
Indeed, the issue is ``incapable of burning.''
In addition to generally removing references to ``success'', we have added an anecdote that reinforces the barrier presented by environmental conditions in the summer\textemdash attempting to burn but being thwarted by high relative humidity. }

186.  What does `high end' mean? Top 90\%, 95\%?

\AR{This has been revised with reference to the calculated $z$ values. }

197.  this is an interesting point that I’d like to see expanded on with references.  What have other’s written about differences between expected changes between spring and summer climates?

\AR{This is a great suggestion, thank you. 
We have substantially revised this general section of the Discussion, incorporating both greater specificity of our data with the added statistical analyses, and several references on the topic of ``land surface phenology'', a new concept we found in the literature review prompted by this comment that does well to encapsulate the approach we employ here, and couch our discussion of fire seasonality within a broader ecological literature.  }

334. Is there a sample size for each point/fire? A table with the size/\# pixels for each fire and some additional descriptions (dates, ignition type, etc) in the supplemental would be helpful. What does the shaded region indicate, and how was it calculated? Adding the regression equation as an insert with r2 would help as well.

\AR{No\textemdash data for all burns are summarized to the experimental unit (8 or 16 ha patches). 
Revisions address comments about statistics.   }

\end{document}

### 2025-04-21 NorthDakotaFireSurvey/R/ResponsePlotting.Rmd
Spring/summer seasonality R1 update

---
title: "Rancher perceptions of Rx fire in North Dakota"
subtitle: "Survey responses"
date: "Version date: `r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    toc: TRUE
    toc_float: FALSE
    theme: journal
    highlight: tango
    code_folding: hide
    code_download: true
    self_contained: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, eval = TRUE, fig.path='../Writing/figures/', dev=c('png'))
# load packages
  pacman::p_load(tidyverse, magrittr, sf, grid, gridExtra, cowplot)
  pacman::p_load_gh('devanmcg/wesanderson')
  ER_pal = wes_palette("Zissou1")[c(1, 4, 5, 2)]
# load pre-wrangled data
  load('./Data/responses.Rdata')
```

# Synthesis and potential recommentations

* Despite not relying on scientific products themselves, respondents do expect administrators of grazinglands to use the best available science
* Respondents generally trust agencies and NGOs, but not necessarily with fire
* Respondents believe their community sees little value in Rx fire and do not want its use expanded
* Initiatives related to Rx fire are likely to be best recieved when they come from Pheasants Forever and Extension
* To increase palatability of Rx fire on public grazinglands, ranchers will likely need to come to appreciate burning as a component of good stewardship, not just a tool for specific management objectives

# Results

## Perceptions of, and trust in, various organizations

Take-aways regarding general use and trust:

* Respondents generally trust most resources
* County Extension resources are the most used
* BLM resources are the least used, consistent with BLM footprint in ND

Take-aways regarding fire, specifically:

* Overall trust is considerably lower with fire
* USFS is most associated with Rx fire, but trusted least
* Pheasants Forever is associated with Rx fire and most trusted with it
* Extension has a high degree of trust but low association

```{r agency_responses, fig.height=12, fig.width=10, fig.cap='Perceptions of various federal (US) and state (ND) agencies and non-governmental organizations regarding (A) use and trust in provided resources, in general; and (B) association and trust with respect to prescribed fire, specifically. TNC = The Nature Conservancy, NRCS = USDA Natural Resource Conservation Service; BLM = US Dept. of Interior Bureau of Land Management; ARS = USDA Agricultural Research Service.'}
# Prepare agency response 
  agency_responses <- 
    responses %>%
      filter(QuestionKey != 'Agency_Resources', 
             str_starts(QuestionKey, "Agency") ) %>%
      mutate(QuestionKey = str_remove(QuestionKey, "Agency_"), 
             QuestionKey = str_replace_all(QuestionKey,  "([[:upper:]])"," \\1"), 
             QuestionKey = str_trim(QuestionKey)) %>% 
      separate(QuestionKey, into = c('category', 'topic'), sep = ' ') %>%
      mutate(agency = case_when(
                        OptionKey == 'NDGF' ~      'ND Game & Fish', 
                        OptionKey == 'PF' ~        'Pheasants Forever', 
                        OptionKey == 'USFWS' ~     'US Fish & Wildlife', 
                        OptionKey == 'audubon' ~   'Audubon Society', 
                        OptionKey == 'extension' ~ 'County Extension', 
                        OptionKey == 'USFS' ~      'US Forest Service', 
                        TRUE ~ str_to_upper(OptionKey) ))
# Resources in general 
  agency_general <- 
    agency_responses %>%
      filter(category != "Fire") %>%
      mutate(Response = recode(topic, 
                               "Trust" = "Trust", 
                               "Use" = "Use"), 
             Response = fct_rev(Response)) %>%
      ggplot() + theme_bw(16) + 
      geom_vline(xintercept = 0, lty = 2) + 
      geom_errorbarh(aes(xmin = min, 
                         xmax = max, 
                         y = agency, 
                         color = Response), 
                     lwd = 0.9,
                     position = position_dodge(width = 0.6)) +
      geom_point(aes(x = mid, y = agency, 
                     fill = Response, 
                     shape = Response), 
                 size = 4, 
                 stroke = 1,
                 position = position_dodge(width = 0.6)) + 
     coord_cartesian(xlim = c(-1.15, 1.9)) + 
      labs(title = 'A. "In general, I personally use/trust resources from this organization"', 
           x = 'Agreement Index') + 
      scale_shape_manual(values = c(21, 24)) + 
      scale_color_manual(values = wes_palette("FantasticFox1")[c(3,5)]) + 
      scale_fill_manual(values = wes_palette("FantasticFox1")[c(3,5)]) + 
      theme(axis.title.y = element_blank(), 
            axis.text.y = element_text(color = 'black', size = 14), 
            plot.title.position = "plot", 
            legend.position = 'top', 
            legend.title=element_blank() )
# WRT fire, specifically
  agency_fire <- 
    agency_responses %>%
      filter(category == "Fire") %>%
      mutate(Response = recode(topic, 
                            "Trust" = "Trust regarding Rx fire", 
                            "Use" = "Associate with Rx fire")) %>%
      ggplot() + theme_bw(16) + 
      geom_vline(xintercept = 0, lty = 2) + 
        geom_errorbarh(aes(xmin = min, 
                           xmax = max, 
                           y = agency, 
                           color = Response), 
                       lwd = 0.9,
                       height = 0.5,
                       position = position_dodge(width = 0.6)) +
        geom_point(aes(x = mid, y = agency, 
                       fill = Response, 
                       shape = Response), 
                   size = 4, 
                   stroke = 1,
                   position = position_dodge(width = 0.6)) + 
       coord_cartesian(xlim = c(-1.15, 1.9)) + 
      labs(title = 'B. "I associate/trust this organization with Rx fire activity or information"', 
           x = 'Agreement Index') + 
       scale_shape_manual(values = c(21, 24)) + 
      scale_color_manual(values = wes_palette("FantasticFox1")[c(3,5)]) + 
      scale_fill_manual(values = wes_palette("FantasticFox1")[c(3,5)]) + 
        theme(axis.title.y = element_blank(), 
              axis.text.y = element_text(color = 'black', size = 14), 
              plot.title.position = "plot", 
              legend.position = 'top', 
              legend.title=element_blank() )
plot_grid(agency_general, 
          agency_fire, 
           # labels = c('A', 'B'),
          ncol = 1)
```

## Contributions to management decisionmaking

Take-aways:

* Respondents mostly value open rangeland 
* Respondents have little concern for managing wildfire risk or invasive species
* Respondents are mostly influenced by a sense of stewardship and passed-down knowledge
* Little direct impact of neighbors and science on decisionmaking

```{r decisionmaking, fig.height=8, fig.width=10, fig.cap='Degree to which respondents agreed that various management considerations (A) were important in their decisions, and (B) influenced their decisionmaking.'}
importance <-
    responses %>%
      filter(str_starts(QuestionKey, "Importance") ) %>%
      mutate(response = case_when(
              OptionKey == 'restore_prairie' ~      'Restore prairie/grassland', 
              OptionKey == 'reduce_wildfire' ~        'Reduce wildfire risk', 
              OptionKey == 'protect_wildlife' ~     'Protect wildlife habitat', 
              OptionKey == 'heterogeneity' ~   'Increase plant diversity', 
              OptionKey == 'control_woody' ~ 'Woody plant control', 
              OptionKey == 'control_invasives' ~      'Invasive species control' )) %>%
  ggplot() + theme_bw(16) + 
  geom_vline(xintercept = 0, lty = 2) + 
  geom_errorbarh(aes(xmin = min, 
                     xmax = max, 
                     y = fct_reorder(response, mid)), 
                 color = wes_palette("FantasticFox1")[3],
                  height = 0.5,
                 lwd = 0.9 ) +
  geom_point(aes(x = mid, y = response ),
             fill = wes_palette("FantasticFox1")[3], 
             pch = 21,
             size = 5, 
             stroke = 1 ) + 
  # coord_cartesian(xlim = c(-1.15, 1.9)) + 
  labs(title = 'A. "This environmental concern is important to me"', 
       x = 'Agreement Index') + 
  theme(axis.title.y = element_blank(), 
        axis.text.y = element_text(color = 'black', size = 14), 
        plot.title.position = "plot" )
influence <- 
    responses %>%
    filter(str_starts(QuestionKey, "Influence") ) %>%
    mutate(response = case_when(
              OptionKey == 'research' ~      'Scientific research/\nUniversity publications', 
              OptionKey == 'neighbors' ~     'How my neighbors\nmanage their land', 
              OptionKey == 'good' ~          'Being a good steward', 
              OptionKey == 'gen' ~   'What previous generations\ntaught me', 
              OptionKey == 'advice' ~ 'Professional advice' )) %>%
    ggplot() + theme_bw(16) + 
    geom_vline(xintercept = 0, lty = 2) + 
    geom_errorbarh(aes(xmin = min, 
                       xmax = max, 
                       y = fct_reorder(response, mid)), 
                   color = wes_palette("FantasticFox1")[5],
                   height = 0.5,
                   lwd = 0.9 ) +
    geom_point(aes(x = mid, y = response ),
               fill = wes_palette("FantasticFox1")[5], 
               pch = 21,
               size = 5, 
               stroke = 1 ) + 
    # coord_cartesian(xlim = c(-1.15, 1.9)) + 
    labs(title = 'B. "This consideration affects my management decisionmaking"', 
         x = 'Agreement Index') + 
    theme(axis.title.y = element_blank(), 
          axis.text.y = element_text(color = 'black', size = 14), 
          plot.title.position = "plot" )
plot_grid(importance, 
          influence, 
           # labels = c('A', 'B'),
          ncol = 1)
```

## Attitudes about fire

Take-aways

* Respondents generally believe their community has negative perceptions of Rx fire, except for wildlife
* Respondents have a conservative attitude towards Rx fire use and show a disinclination to use it themselves

```{r attitudes, fig.height=10, fig.width=10, fig.cap='Degree to which respondents agreed with statements about (A) community perceptions of prescribed fire effects, and (B) personal attitudes about prescribed fire use.'}
# Attitudes and beliefs about Rx fire
beliefs <-
  responses %>%
    filter(QuestionKey == "CommunityFireBeliefs")  %>%
    mutate(OptionKey = str_remove(OptionKey, 'burn_'), 
           response = case_when(
              OptionKey == 'wildlife' ~ 'Negative impacts to wildlife', 
              OptionKey == 'soil_erosion' ~   'Results in soil erosion', 
              OptionKey == 'smoke' ~    'Produces harmful smoke', 
              OptionKey == 'safety' ~   'Poses risk to human safety', 
              OptionKey == 'property_risk' ~   "Poses risk to nearby property", 
              OptionKey == 'forage' ~      'Results in loss of forage' )) %>%
    ggplot() + theme_bw(16) + 
    geom_vline(xintercept = 0, lty = 2) + 
    geom_errorbarh(aes(xmin = min, 
                       xmax = max, 
                       y = fct_reorder(response, desc(SurveyOrder))) , 
                   color = wes_palette("FantasticFox1")[3], 
                   lwd = 0.9 ) +
    geom_point(aes(x = mid, y = response ),
               fill = wes_palette("FantasticFox1")[3], 
               pch = 21,
               size = 4, 
               stroke = 1 ) + 
    # coord_cartesian(xlim = c(-1.15, 1.9)) + 
    labs(title = 'A. "Members of my community generally believe the following about Rx fire"', 
         x = 'Agreement Index') + 
    theme(axis.title.y = element_blank(), 
          axis.text.y = element_text(color = 'black', size = 14), 
          plot.title.position = "plot" )
attitudes <-
  responses %>%
    filter(str_starts(QuestionKey, "Attitudes") ) %>%
    mutate(OptionKey = str_remove(OptionKey, 'burn_'), 
           response = case_when(
             OptionKey == 'beneficial' ~ 'Rx fire is a beneficial\ntool for managing rangelands', 
             OptionKey == 'frequent' ~   'People should be able to\nburn as frequently as they\n want for management', 
             OptionKey == 'limited' ~    'Burning can be used in a\nlimited set of instances', 
             OptionKey == 'man_plan' ~   'My management plan includes\nusing prescribed fire', 
             OptionKey == 'prepared' ~   "I'm prepared to use Rx fire\nif I choose to", 
             OptionKey == 'pros' ~      'Fire should only be used\nby professionals', 
             OptionKey == 'shouldnt' ~   'Burning is a poor management\noption that shouldn’t be used', 
             OptionKey == 'training' ~ 'Burning is a good option\nfor anyone with training')) %>%
    ggplot() + theme_bw(16) + 
    geom_vline(xintercept = 0, lty = 2) + 
    geom_errorbarh(aes(xmin = min, 
                       xmax = max, 
                       y = fct_reorder(response, desc(SurveyOrder))) ,
                   height = 0.5,
                   color = wes_palette("FantasticFox1")[5], 
                   lwd = 0.9 ) +
    geom_point(aes(x = mid, y = response ),
               fill = wes_palette("FantasticFox1")[5], 
               pch = 21,
               size = 4, 
               stroke = 1 ) + 
    # coord_cartesian(xlim = c(-1.15, 1.9)) + 
    labs(title = 'B. "I personally believe the following"', 
         x = 'Agreement Index') + 
    theme(axis.title.y = element_blank(), 
          axis.text.y = element_text(color = 'black', size = 14), 
          plot.title.position = "plot" )
plot_grid(beliefs, 
          attitudes, 
          rel_heights = c(0.75, 1.5), 
          ncol = 1)
```

## Public grazing land management and personal recreation

Take-aways

* Respondents expect administrators to use good science
* Respondents do not want to see Rx fire expanded on public grazingland
* Respondents really don't use public grazlinglands for recreation

```{r public_lands, fig.height=5, fig.width=10, fig.cap='Respondent attitudes towards public grazingland administration and prescribed fire use.'}
responses %>%
  filter(QuestionKey == "PublicGrazingBeliefs", 
         OptionKey != 'public_fire')  %>%
  mutate(response = case_when(
           OptionKey == 'increase_public_fire' ~ 'There should be more Rx fire\non public grazingland',            OptionKey == 'amount_public_fire' ~'There is an appropriate amount of\nRx fire on public grazingland', 
           OptionKey == 'admin_users' ~ 'Administrators should manage to meet\nuser needs or preferences', 
           OptionKey == 'admin_science' ~ "Administrators should manage land\naccording to the best available science", 
           OptionKey == 'admin_good' ~ 'Public grazingland administrators\n do a good job managing the land' )) %>%
  ggplot() + theme_bw(16) + 
  geom_vline(xintercept = 0, lty = 2) + 
  geom_errorbarh(aes(xmin = min, 
                     xmax = max, 
                     y = fct_reorder(response, desc(SurveyOrder))) , 
                 color = wes_palette("FantasticFox1")[3], 
                 height = 0.5,
                 lwd = 0.9 ) +
  geom_point(aes(x = mid, y = response ),
             fill = wes_palette("FantasticFox1")[3], 
             pch = 21,
             size = 4, 
             stroke = 1 ) + 
  # coord_cartesian(xlim = c(-1.15, 1.9)) + 
  labs(title = '"I personally believe the following about public grazingland management"', 
       x = 'Agreement Index') + 
  theme(axis.title.y = element_blank(), 
        axis.text.y = element_text(color = 'black', size = 14), 
        plot.title.position = "plot" )
```

```{r recreation, fig.height=5, fig.width=10, fig.cap="Respondents' use of public grazingland for various recreation activies."}
responses %>%
  filter(QuestionKey == 'Recreation' ) %>%
  ggplot() + theme_bw(16) + 
  geom_vline(xintercept = 0, lty = 2) + 
  geom_errorbarh(aes(xmin = min, 
                     xmax = max, 
                     y = fct_reorder(SurveyText, mid) ) , 
                 color = wes_palette("FantasticFox1")[3], 
                 height = 0.5,
                 lwd = 0.9 ) +
  geom_point(aes(x = mid, y = SurveyText ),
             fill = wes_palette("FantasticFox1")[3], 
             pch = 21,
             size = 4, 
             stroke = 1 ) + 
  # coord_cartesian(xlim = c(-1.15, 1.9)) + 
  labs(title = '"Myself and/or my family use public grazingland for these recreation activities"', 
       x = 'Agreement Index') + 
  theme(axis.title.y = element_blank(), 
        axis.text.y = element_text(color = 'black', size = 14), 
        plot.title.position = "plot" )
```

### 2025-04-21 SpringSummerSeverity/paper/CoverLetter/CoverLetterEM0.tex
Spring/summer seasonality R1 update

--- 2024-12-11 SpringSummerSeverity/paper/CoverLetter/CoverLetter0.tex
@@
 \documentclass[parskip=half, 
-			   fontsize=12pt,
+			   fontsize=10pt,
 			   paper=a4]				
@@
 
-\newcommand{\journal}{\emph{Restoration Ecology}}
+\newcommand{\journal}{\emph{Environmental Management}}
 
@@
 
-We've provided suggested reviewers as part of the online submission.
+Please note that we have prepared and submitted this as a \textbf{Methodology} paper, for two basic reasons: Firstly, our emphasis is on the feasibility of conducting the treatments, not evaluating their ecological outcomes with reference to experimental hypotheses; and secondly, this is a case study focused on a single research station.
+Although 40+ fires is a decent sample size, we are reticent to extrapolate the conclusions beyond the local context beyond the caveat of the case study. 
+As such, we have refrained from elaborate statistical analysis that would likely be irrelevant in a case study. 
+Instead, we have focused on providing this food for thought for managers.
 

### 2025-04-21 SpringSummerSeverity/paper/CoverLetter/CoverLetterEM1.tex
Spring/summer seasonality R1 update

\documentclass[parskip=half, 
			   fontsize=10pt,
			   paper=a4]				
{scrartcl}
\input{SetupOptions} % contains scrletter

\newcommand{\journal}{\emph{Environmental Management}}

\vspace{-5em}

\begin{document}
	\begin{letter}{Editorial staff\\
		\journal }
\setlength{\parindent}{12pt}

\opening{Dear Sir or Madam,}

Please find attached our revised manuscript for further consideration for publication in \journal. 
We appreciate the opportunity to consider the thoughtful comments of both reviewers. 
The editor will note that we have responded strongly to the suggestions for statistical analysis and the greater precision in discussing results these analyses provide, as well as better definitions of burn objectives, and an expanded discussion of global patterns in phenology as they interact with local factors.

Two points from the reviewers have not been incorporated: Reviewer 1's comments about field data from the fires cannot be satisfied as these data do not exist, and comments from both reviewers about the role of grazing were not incorporated into the paper as we do not have anything substantial to say about a factor we did not observe to have any impact on the variables reported here. 
Certainly, pages can be written about the potential effects of grazing on a fuelbed but without any way to connect grazing to the data we present here, the whole topic strikes us as tangential.

Please recall that, from our initial submission, we have prepared this as a \textbf{Methodology} paper, for two basic reasons: Firstly, our emphasis is on the feasibility of conducting the treatments, not evaluating their ecological outcomes with reference to experimental hypotheses; and secondly, this is a case study focused on a single research station. 
As a reviewer pointed out, although 40+ fires is a decent sample size, we are reticent to extrapolate the conclusions beyond the local context beyond the caveat of the case study.

We look forward to working with the editors and reviewers of \journal~to ensure our manuscript meets the standards of the journal.

\vspace{-3em} 
\closing{} %eg. Regards
%\cc{another dude}
\end{letter}

\end{document}

### 2025-04-21 SpringSummerSeverity/paper/FigureGenerator.Rmd
draft SpringSummerSeverity
Spring/summer seasonality R1 update

@@
 title: Evaluating an attempt to restore summer fire in the Northern Great Plains
-author: DA McGranahan & JP Angerer??
+author: DA McGranahan & JP Angerer
 output: 
@@
   load('../data/WxData.Rdata')
+  # stats results
+  load('../data/z_tab.Rdata')
+  load('../data/comp_tab.Rdata')
 # some necessary things
+  sys_yr <- 
+    Sys.Date() %>%
+      as.POSIXct(., '%Y-%M-%d') %>%
+      year() 
   seasons <- 
     tibble(season = c('Spring', 'Summer'), 
-           start = c(as.Date('2024-04-25'), as.Date('2024-08-07')), 
-           end = c(as.Date('2024-05-25'), as.Date('2024-08-27'))) 
-```
-
-```{r}
+           start = c(as.Date(paste0(sys_yr, '-04-25')), as.Date(paste0(sys_yr, '-08-07'))), 
+           end = c(as.Date(paste0(sys_yr, '-05-25')), as.Date(paste0(sys_yr, '-08-27'))))
 LandsatSums <-
@@
                               year %in% c('2018', '2019') ~ "Not completed", 
-                              TRUE ~ "Completed") )%>% 
+                              TRUE ~ "Completed") ) %>% 
   ggplot(aes(x = season)) + theme_bw(14) + 
@@
 
+```{r ndvi_precip_z}
+z_tab$ndvi_prcp %>%
+    pander::pander("Study period deviations from historical NDVI and YTD precipitation across burn seasons.") 
+```
+
 ## Weather data
@@
 
+```{r weather_season_z}
+z_tab$wxd %>%
+    pander::pander("Study period deviations from historical fire weather across burn seasons.") 
+```
+
+## Historical patterns
+
+```{r }
+StudySums <- 
+  LandsatSums %>%
+    filter(year %in% c(2017:2020)) %>%
+    group_by(year, season, index) %>%
+    summarize(Mean = mean(Value),
+              SEM = sd(Value)/sqrt(n()), 
+              .groups = 'drop') %>%
+    mutate(status = case_when(
+      season == 'Summer' & year %in% c('2018', '2019') ~ "Not completed", 
+      TRUE ~ "Completed"), 
+      index = toupper(index), 
+      index = fct_rev(index) )
+
+HistoricalSpectralData <- 
+  bind_rows(
+  LandsatSums %>%
+  filter(year %in% c(1990:1999))%>% 
+  mutate(range = '1990-1999', 
+         index = toupper(index), 
+         index = fct_rev(index)), 
+  LandsatSums %>%
+    filter(year %in% c(2012:2022))%>% 
+    mutate(range = '2012-2022', 
+           index = toupper(index), 
+           index = fct_rev(index)), 
+  LandsatSums %>%
+    mutate(range = '1990-2022', 
+           index = toupper(index), 
+           index = fct_rev(index)) )
+```
+
+```{r historical_phenology, eval = TRUE, fig.width = 8, fig.height =11.5}
+WxData$HourlyWx %>%
+  filter(! date %in% RainyDays$date, # remove rainy days with no chance of burning
+         ! is.na(season)) %>%
+  group_by(Year, date, season) %>%
+  summarise_at(.vars = vars(c(DewPoint, RelHum, WindSpeed,VPD)),
+               .funs = c("mean")) %>%
+  ungroup() %>%
+  pivot_longer(names_to = 'variable', 
+               values_to = 'value', 
+               cols = DewPoint:VPD) %>%
+  group_by(Year, season, variable) %>%
+  summarise(Value = mean(value), 
+            .groups = 'drop') %>%
+  mutate(Year = as.character(Year), 
+         #season = fct_rev(season), 
+         period = case_when(
+           Year %in% c(1990:1999) ~  '1990-1999', 
+           Year %in% c(2012:2022) ~  '2012-2022', 
+           TRUE ~ NA
+         )) %>%
+  filter(!is.na(period)) %>%
+  bind_rows(WxData$Rainfall %>% 
+              filter(month(date) %in% c(04:09)) %>%
+              separate(date, into = c('Year', 'Month', 'Day'),sep = '-') %>%
+              group_by(Year) %>%
+              mutate(CumPrecip = cumsum(Rainfall), 
+                     Month = recode(Month, '05' ="May", '08'='Aug'), 
+                     season = paste0(Month, ' ', Day)) %>%
+              ungroup() %>%
+              filter(season %in% c('May 15', 'Aug 01' )) %>%
+              mutate(season = ifelse(season == 'May 15', 'Spring', 'Summer'), 
+                     period = case_when(
+                       Year %in% c(1990:1999) ~  '1990-1999', 
+                       Year %in% c(2012:2022) ~  '2012-2022', 
+                       TRUE ~ NA
+                     ) )  %>%
+              filter(!is.na(period)) %>%
+              select(Year, season, CumPrecip, period) %>%
+              pivot_longer(names_to = 'variable', 
+                           values_to = 'Value', 
+                           cols = CumPrecip))  %>%
+  bind_rows(HistoricalSpectralData  %>% 
+              select(-pasture) %>%
+              filter(index == 'NDVI', 
+                     range != '1990-2022') %>%
+              rename(Year = year, period = range, variable = index)
+            ) %>%
+  mutate( variable = factor(variable, levels = c('NDVI', 'CumPrecip', 'DewPoint', 
+                                                 'RelHum', 'VPD', 'WindSpeed')), 
+          variable = recode(variable, 
+                           'NDVI' = "Fuelbed\ngreenness\n(NDVI)", 
+                           'CumPrecip' = 'Cumulative\nrainfall\n(mm)',
+                           'DewPoint' = 'Dew\nPoint\n(°C)', 
+                           'RelHum' = 'Relative\nHumidity\n(%)', 
+                           'VPD' = 'Vapor\nPressure\nDeficit', 
+                           'WindSpeed' = 'Wind\nspeed\n(m/s)'), 
+          period = recode(period, '1990-1999' = '1990-99', '2012-2022' = '2012-22')) %>%
+  ggplot() + theme_bw(16) + 
+  geom_hline(data = . %>%
+               filter(as.numeric(Year) %in% c(2017:2020)) %>%
+               mutate(status = case_when(
+                 season == 'Summer' & 
+                   Year %in% c('2018', '2019') ~ "Not completed", 
+                 TRUE ~ "Completed") ) %>%
+               group_by(variable, Year, season, status) %>%
+               slice(1) , 
+             aes(yintercept = Value,
+                 color = Year, lty = status), 
+             lwd = 1 ) + 
+  geom_boxplot(aes(x=period, 
+                   y = Value),
+               outliers = F, 
+               staplewidth = 0.25) +
+  labs(x = 'Historical period', y = ' ') + 
+  facet_grid(variable~season, scales = "free_y")  +
+  scale_linetype_manual("Burn status", values = c(1,2)) +
+  scale_color_manual("Study year", values = FFgrad) +
+  theme(axis.text.x = element_text(color = 'black'), 
+        panel.grid.major.x = element_blank() , 
+        strip.text.y = element_text(angle = 0), 
+        strip.background = element_rect(fill = '#A1D99B')) 
+```
+
+```{r historical_phenology_z}
+z_tab$hist %>%
+    pander::pander("Study period deviations from two historical periods of fuel conditions and fire weather across burn seasons.") 
+```
+
 # Supplemenary Info
@@
   WxData$Rainfall %>% 
+  mutate(day = day + years(1)) %>%
     filter(month(date) %in% c(04:09)) %>%
@@
 
-## Historical trends
-
 ```{r eval = FALSE}
@@
         panel.grid.major.x = element_blank() )
-```
-
-```{r }
-StudySums <- 
-  LandsatSums %>%
-    filter(year %in% c(2017:2020)) %>%
-    group_by(year, season, index) %>%
-    summarize(Mean = mean(Value),
-              SEM = sd(Value)/sqrt(n()), 
-              .groups = 'drop') %>%
-    mutate(status = case_when(
-      season == 'Summer' & year %in% c('2018', '2019') ~ "Not completed", 
-      TRUE ~ "Completed"), 
-      index = toupper(index), 
-      index = fct_rev(index) )
-
-HistoricalSpectralData <- 
-  bind_rows(
-  LandsatSums %>%
-  filter(year %in% c(1990:1999))%>% 
-  mutate(range = '1990-1999', 
-         index = toupper(index), 
-         index = fct_rev(index)), 
-  LandsatSums %>%
-    filter(year %in% c(2012:2022))%>% 
-    mutate(range = '2012-2022', 
-           index = toupper(index), 
-           index = fct_rev(index)), 
-  LandsatSums %>%
-    mutate(range = '1990-2022', 
-           index = toupper(index), 
-           index = fct_rev(index)) )
 ```
@@
 
-```{r historical_phenology, eval = TRUE, fig.width = 8, fig.height =11.5}
-WxData$HourlyWx %>%
-  filter(! date %in% RainyDays$date, # remove rainy days with no chance of burning
-         ! is.na(season)) %>%
-  group_by(Year, date, season) %>%
-  summarise_at(.vars = vars(c(DewPoint, RelHum, WindSpeed,VPD)),
-               .funs = c("mean")) %>%
-  ungroup() %>%
-  pivot_longer(names_to = 'variable', 
-               values_to = 'value', 
-               cols = DewPoint:VPD) %>%
-  group_by(Year, season, variable) %>%
-  summarise(Value = mean(value), 
-            .groups = 'drop') %>%
-  mutate(Year = as.character(Year), 
-         #season = fct_rev(season), 
-         period = case_when(
-           Year %in% c(1990:1999) ~  '1990-1999', 
-           Year %in% c(2012:2022) ~  '2012-2022', 
-           TRUE ~ NA
-         )) %>%
-  filter(!is.na(period)) %>%
-  bind_rows(WxData$Rainfall %>% 
-              filter(month(date) %in% c(04:09)) %>%
-              separate(date, into = c('Year', 'Month', 'Day'),sep = '-') %>%
-              group_by(Year) %>%
-              mutate(CumPrecip = cumsum(Rainfall), 
-                     Month = recode(Month, '05' ="May", '08'='Aug'), 
-                     season = paste0(Month, ' ', Day)) %>%
-              ungroup() %>%
-              filter(season %in% c('May 15', 'Aug 01' )) %>%
-              mutate(season = ifelse(season == 'May 15', 'Spring', 'Summer'), 
-                     period = case_when(
-                       Year %in% c(1990:1999) ~  '1990-1999', 
-                       Year %in% c(2012:2022) ~  '2012-2022', 
-                       TRUE ~ NA
-                     ) )  %>%
-              filter(!is.na(period)) %>%
-              select(Year, season, CumPrecip, period) %>%
-              pivot_longer(names_to = 'variable', 
-                           values_to = 'Value', 
-                           cols = CumPrecip))  %>%
-  bind_rows(HistoricalSpectralData  %>% 
-              select(-pasture) %>%
-              filter(index == 'NDVI', 
-                     range != '1990-2022') %>%
-              rename(Year = year, period = range, variable = index)
-            ) %>%
-  mutate( variable = factor(variable, levels = c('NDVI', 'CumPrecip', 'DewPoint', 
-                                                 'RelHum', 'VPD', 'WindSpeed')), 
-          variable = recode(variable, 
-                           'NDVI' = "Fuelbed\ngreenness\n(NDVI)", 
-                           'CumPrecip' = 'Cumulative\nrainfall\n(mm)',
-                           'DewPoint' = 'Dew\nPoint\n(°C)', 
-                           'RelHum' = 'Relative\nHumidity\n(%)', 
-                           'VPD' = 'Vapor\nPressure\nDeficit', 
-                           'WindSpeed' = 'Wind\nspeed\n(m/s)'), 
-          period = recode(period, '1990-1999' = '1990-99', '2012-2022' = '2012-22')) %>%
-  ggplot() + theme_bw(16) + 
-  geom_hline(data = . %>%
-               filter(as.numeric(Year) %in% c(2017:2020)) %>%
-               mutate(status = case_when(
-                 season == 'Summer' & 
-                   Year %in% c('2018', '2019') ~ "Not completed", 
-                 TRUE ~ "Completed") ) %>%
-               group_by(variable, Year, season, status) %>%
-               slice(1) , 
-             aes(yintercept = Value,
-                 color = Year, lty = status), 
-             lwd = 1 ) + 
-  geom_boxplot(aes(x=period, 
-                   y = Value),
-               outliers = F, 
-               staplewidth = 0.25) +
-  labs(x = 'Historical period', y = ' ') + 
-  facet_grid(variable~season, scales = "free_y")  +
-  scale_linetype_manual("Burn status", values = c(1,2)) +
-  scale_color_manual("Study year", values = FFgrad) +
-  theme(axis.text.x = element_text(color = 'black'), 
-        panel.grid.major.x = element_blank() , 
-        strip.text.y = element_text(angle = 0), 
-        strip.background = element_rect(fill = '#A1D99B')) 
-```
-
 ```{r eval = FALSE}

### 2025-04-21 SpringSummerSeverity/paper/SummerFireRestorationEM1.Rmd
Spring/summer seasonality R1 update

--- 2024-11-25 SpringSummerSeverity/paper/SummerFireRestoration.Rmd
@@
 title: Evaluating an attempt to restore summer fire in the Northern Great Plains
+#subtitle: 'Running head: Summer burning in the Northern Plains'
 author:
@@
     keep_md: yes
-csl: restoration-ecology.csl
+csl: environmental-management.csl
 bibliography: 
@@
                       echo=FALSE, eval=TRUE)
-pacman::p_load(tidyverse, knitr, kableExtra)
+pacman::p_load(tidyverse, knitr, kableExtra, pander)
 # Some caption tools 
@@
   f_ref = function(x) { str_extract(f_nums(x), "[^:]*")   }
+
+load('../data/z_tab.Rdata')
+load('../data/comp_tab.Rdata')
 ```
+
+**Author contributions:** Conception, analysis, writing: DAM; Programming and editing: JPA
 
@@
 In many North American grasslands, late growing season burns re-introduce fire to periods most prone to lightning-driven fire prior to wildfire suppression policies. 
-We report here on restoring summer fire in central North Dakota, USA, from a research project in which summer fire was only successfully achieved in two out of four years. 
+We report here on restoring summer fire in central North Dakota, USA, from a research project in which summer burns were only completed in two out of four years for which summer burns were planned. 
 We use remotely-sensed imagery and local weather data to assess whether fuel or weather conditions limited burning in the summer, and to compare fire environmental conditions and subsequent burn severity across prescribed burns conducted in the spring and summer.
 Finally, we review historical data to determine if conditions have changed in either the spring or summer burn seasons over 42 years. 
-We found that burn severity declined with fuelbed greenness but was independent of burn season---summer burns could effect as high of severity as spring burns despite having greener fuelbeds.
-What little phenological change seems to have occurred at the study location---slightly greener fuelbeds and slightly lower relative humidity in spring---likely offset each other to some degree. 
-Overall, we found little evidence that successful summer burns were anomalies, and conclude that it is reasonable for managers to incorporate late growing season fire into prescribed fire programs with the caveat that some summers will simply be too wet and/or too green to burn.
+Although burn severity generally declined with fuelbed greenness in the spring, summer burns could effect as high of severity as spring burns despite having greener fuelbeds.
+What little phenological change seems to have occurred at the study location over 42 years---slightly greener fuelbeds and slightly lower relative humidity in spring---likely offset each other to some degree.
+Overall, we found little evidence that being able to complete summer burns was anomalous, and conclude that it is reasonable for managers to incorporate late growing season fire into prescribed fire programs with the caveat that some summers will simply be too wet and/or too green to burn.
 
 **Keywords:** Burn severity; Phenology; Seasonality of fire; Pyrodiversity
+
+**Statements and Declarations** The authors declare no competing interests.
 
@@
 
-In this paper we report on mixed success in conducting summer prescribed fire in the Northern Great Plains. 
+In this paper we report on mixed results in conducting summer prescribed fire in the Northern Great Plains. 
 As researchers attempted to restore spatially-patchy prescribed fire to grazed mixed-grass prairie, two patterns of fire were introduced: one in which an entire one-quarter patch was burned each spring, and another in which half of the patch---one-eighth of the pasture---was burned in the spring, followed by the second half-patch scheduled for a late summer burn, in an attempt to provide a late-season flush of high-quality forage [e.g., @allred2011; @nichols2021] and determine whether a greater number of patches would increase spatial heterogeneity  [e.g., @mcgranahan2018b].
-While all spring burns were completed as planned, summer fires were only successful in two out of four years. 
-Here we present seasonal and long-term remotely-sensed burn severity and fuel data alongside weather information in considering whether successful summer burns are likely to be the exception or the norm based on historical patterns.
+While all spring burns were completed as planned, summer fires were only completed in two out of four years. 
+Here we present seasonal and long-term remotely-sensed burn severity and fuel data alongside weather information in considering whether being able to conduct summer burns was exceptional or more likely to be the norm based on historical patterns.
 
@@
 
-A total of 40 prescribed fires were conducted over four years, 2017--2020. 
+Forty prescribed fires were conducted over four years, 2017--2020. 
 All 32 planned spring burns were completed, between 25 April and 25 May. 
@@
 We used remotely-sensed data from two separate sources. 
-For the study period alone, 2017--2020, we used the European Space Agency's Sentinel-2 mission, which provides hyperspectral imagery at 10 $\times$ 10 m resolution at 5-10 day intervals. 
-For historical context, 1990--2022, we used Landsat resources from the US Geological Survey, which provides hyperspectral imagery at 30 $\times$ 30 m resolution at 7-14 day intervals; Sentinel-2 only began in 2016, while the first Landsat mission launched in the mid 1980s.
+For the study period alone, 2017--2020, we used the European Space Agency's Sentinel-2 mission, multispectral imagery at 10 $\times$ 10 m resolution in the visible and near-infrared region and 20 $\times$ 20 m in the red-edge and shortwave infrared regions.
+Data are collected and produced at 5-10 day intervals. 
+For historical context, 1990--2022, we used Landsat resources from the US Geological Survey (Landsat 5, 7, and 8 Level 2, Collection 2, Tier 1 Surface Reflectance), which provide multispectral imagery at 30 $\times$ 30 m resolution at 7-14 day intervals; Sentinel-2 only began in 2016, while the first Landsat 5 mission launched in the mid 1980s.
 
-Remotely-sensed data consisted of two indices: the Normalized Burn Ratio (NBR), which is used to estimate burn severity, and Normalized Difference Vegetation Index (NDVI), used to estimate greenness of vegetation.
+Remotely-sensed data consisted of two indices: the Normalized Burn Ratio (NBR; @key2006), which is used to estimate burn severity, and Normalized Difference Vegetation Index (NDVI; @rousejr1974), used to estimate greenness of vegetation.
 Sentinel-2 imagery was retrieved from the online Copernicus Browser using a custom EvalScript to return values for each index as a separate band in a geo-referenced raster file; each scene was inspected to ensure cloud-free imagery over pasture locations beginning with the first snow-free scene through to the final snow-free scene each year.
-To calculate burn severity, we subtracted the first post-fire image from the most-recent pre-fire image for each burn unit, generating a burn severity index known as differenced Normalized Burn Ratio (dNBR).
+To calculate burn severity, we subtracted the first post-fire image from the most-recent pre-fire image for each burn unit, generating a burn severity index known as differenced Normalized Burn Ratio (dNBR; @key2006).
 Estimates of fuel greenness were made with the NDVI from the same most-recent pre-burn image.
 
-Landsat imagery was processed with Google Earth Engine using a custom script to merge missions over time, mask clouds, and calculate NDVI for the burn seasons defined above, 1990--2022. 
+Landsat imagery was processed with Google Earth Engine [@gorelick2017] using a custom script to merge missions over time, mask clouds, and calculate NDVI for the burn seasons defined above, 1990--2022.
 See Supplemental Information for scripts used for both Copernicus and Google Earth Engine.
@@
 Weather data---which included air temperature, relative humidity, dew point, windspeed, and vapor pressure deficit---were retrieved from the North Dakota Agricultural Weather Network's Streeter station [@ndawn2023].
-To characterize day-of fire weather for successful burn days, we calculated mean values for operational burn periods (1000--1700) from hourly data for days recorded as having at least one completed fire.
+To characterize day-of fire weather for burn days, we calculated mean values for operational burn periods (1000--1700 hrs) from hourly data for days recorded as having at least one completed fire. 
 To characterize seasonal fire weather, we used daily precipitation totals to eliminate days with more than 25 mm precipitation---days no fire would be conducted under any other circumstance---and calculated mean values for the operational period for each spring (25 April--25 May) and summer (7--27 August) burn season.
-Seasonal values were determined for both the study period (spring and summer burn seasons, 2017--2020) and the historical range (1 April- 31 October, 1990--2023).
+Seasonal values were determined for both the study period (spring and summer burn seasons, 2017--2020) and the historical range (1 April--31 October, 1990--2023). 
+Historical trends in both Landsat NDVI and fire weather are available in Supplemental Information.
 
@@
 
+## Statistical analysis
+
+Difference in NDVI between spring and summer for the four study years was tested via analysis of deviance ($\chi^2$ test), in which with a linear mixed-effect model was compared against an intercept-only, null model, each fit with the `lmer` function in the *lme4* package for **R** [@bates2015].
+Linear relationships between dNBR and NDVI were first tested for a difference between spring and summer trends using analysis of deviance ($\chi^2$ test), then within each season ($t$ test), using generalized mixed-effect regression models fit with a Gamma distribution using `glmer` from *lme4*.
+Pasture (block) and year were fit as random effects.
+
+To compare mean values for each year of the study period to historical ranges, we calculated $z$ scores as a measure of deviation from the mean [@schober2021] by subtracting the historical mean from the seasonal value, and dividing by the historical standard deviation.
+*Anomalies* were defined as values more than one standard deviation above or below the historical mean; *statistical significance* at $\alpha$ = 0.05 occurs when $z > |2|$.
+
+Potential temporal changes in fuel greenness and fire weather variables from early in the historical dataset (1990--1999) and recent values (2012-2022) were tested as differences in means using ordinary least squares regression ($t$ test) at $\alpha$ = 0.05.
+
 # Results
 
-Grassland fuelbeds were less green during each of the spring fire seasons (Sentinel-2 NDVI 0.3-0.4) than in the summer (NDVI > 0.5), with the two seasons in which summer burns could not be completed being the most green (NDVI > 0.6; `r f_ref("sentinel")`A).
-Burn severity ranged widely within and across seasons and years (`r f_ref("sentinel")`B). 
-Among the 32 spring burns, there was a general decline in burn severity with increasing fuelbed greenness. 
-The low number of successful summer burns precluded a linear assessment, but despite having categorically higher greenness, summer burning effected the same range of variability in burn severity as in the spring (`r f_ref("sentinel")`B).
+Grassland fuelbeds were less green during each of the spring fire seasons (Sentinel-2 NDVI 0.3-0.4) than in the summer (NDVI > 0.5; $\chi^2$ = 134, $P <$ 0.001), with the two seasons in which summer burns could not be completed being the most green (NDVI > 0.6; `r f_ref("sentinel")`A).
+Burn severity ranged widely within and across seasons and years (`r f_ref("sentinel")`B); despite having categorically higher greenness (`r f_ref("sentinel")`A), summer burning effected the same range of variability in burn severity as in the spring.
+The relationship between dNBR and NDVI varied between seasons ($\chi^2$ = 8.4, $P <$ 0.01). 
+Among summer burns, there was no linear relationship between dNBR and NDVI ($t$ = -0.97, $P >$ 0.05).
+Among spring burns, there was a general decline in burn severity with increasing fuelbed greenness ($t$ = -4.7, $P <$ 0.001).
 
-Fuelbed greenness during successful burn seasons was close to the typical range over the 42-year historical period (`r f_ref("GreenRain")`A), while cumulative rainfall was at or below the typical historical range (`r f_ref("GreenRain")`B).
-Conversely, for years in which summer burns were not completed, both summer fuelbed greenness and accumulated rainfall by 1 August were atypically high relative to their historical ranges (`r f_ref("GreenRain")`).
+Fuelbed greenness, as measured by Landsat NDVI, during the four spring burn seasons was within or slightly above typical values for the 42-year historical period (`r f_ref("GreenRain")`A; `r t_ref("GreenRain_z")`).
+Summer seasons with completed burns were very near the historical mean NDVI (`r t_ref("GreenRain_z")`), while summer seasons without completed burns had the highest NDVI anomalies (2018, $z$ = 0.76; 2019, $z$ = 1.52).
+In terms of cumulative rainfall, all four spring seasons were at or below historical values (`r f_ref("GreenRain")`B; `r t_ref("GreenRain_z")`). 
+Years in which summer burns were not completed had the highest anomalies in cumulative precipitation (2018, $z$ = 1.6; 2019, $z$ = 0.8). 
+Conversely, cumulative precipitation in years in which summer burns were completed was at least 1 standard deviation below the historical mean (2017, $z$ = -1.04; 2020, $z$ = -1.3).
 <!-- Note that spring NDVI derived from Landsat (`r f_ref("GreenRain")`A) was slightly higher than the same data from Sentinel-2 (`r f_ref("sentinel")`A); the reason for the difference is unclear but might relate to the lower temporal resolution of Landsat imagery skewing the spring average with later, greener values and fewer early, snow-free images.
@@
 
-Observed fire weather varied widely within and across the four study years, and generally spanned the entire range of historical values, although wind speeds on the day of prescribed burns never exceeded the historical median in either season, and only one burn day exceeded the historical median relative humidity (`r f_ref("weather")`).
-Throughout the 42-year dataset, summer burn seasons generally had higher moisture, greater vapor pressure deficient, and lower wind speed than spring burn seasons.
+Day-of-burn fire weather varied widely within and across the four study years, and generally spanned the entire range of historical values (`r f_ref("weather")`). 
+However, wind speeds on the day of prescribed burns never exceeded the historical median in either season, and only one burn day exceeded the historical median relative humidity (`r f_ref("weather")`).
+Seasonal averages in fire weather for each of the four study years also ranged widely relative to historical means. 
+In summer, especially, there were no patterns in the magnitude and signs of fire weather anomalies that distinguished years in which summer burns were completed, and those in which they were not (`r t_ref("weather_z")`).
 
-Few substantial changes in the fire environment appear to have occurred since the 1990s in either spring or summer burn seasons (`r f_ref("historical")`). 
-The two most notable changes---increased spring fuelbed greenness and lower spring dewpoint and relative humidity---might cancel each other out, with drier air potentially mitigating the impact of higher-moisture fine fuels.
+Few substantial changes in the fire environment appear to have occurred since the 1990s in either spring or summer burn seasons (`r f_ref("historical")`, `r t_ref("comp_stats")`). 
+The only statistically-significant change was an increase in spring fuelbed greenness ($t$ = 3.8, $P$ < 0.001).
 In two instances---spring windspeed and summer rainfall---the more apparent temporal trend was for greater variability around the typical range, rather than substantial shifts in typical values.
@@
 
-We used remotely-sensed imagery and weather data to inform whether, in the four years summer burns were attempted in a research project in the Northern Great Plains, the two successful years, or the two unsuccessful years, were more likely outcomes based on typical fuel and weather conditions.
-Comparing seasonal values to historical ranges suggests that unsuccessful burn seasons were the anomalies. 
-As such, successfully completing prescribed fire in the late growing season in the Northern Great Plains is a reasonable management objective. 
+We used remotely-sensed imagery and weather data to assess whether, in the four years summer burns were planned in a research project in the Northern Great Plains, the fuel and weather conditions of the two years in which burns were completed, or conditions of the two years in which burns were not completed, were more likely to represent normal conditions based on historical patterns.
+Comparing seasonal values to historical ranges suggested that not being able to complete summer burns were the anomalies. 
+As such, completing prescribed fire in the late growing season in the Northern Great Plains is a reasonable management objective. 
 However, prairie fuelbeds might simply be too green to carry fire in some years, likely those with above-normal rainfall over the course of the growing season.
 
-The data provide little indication that fire weather posed a barrier to successful summer burning. 
-Overall, burn season fire weather was highly variable across study years, and burn day weather spanned the gamut (the exception was wind speed, which was below average in all study years). 
-For example, for the four years summer burns were planned, one successful year and one unsuccessful year were at the high end of typical relative humidity, and the other successful and unsuccessful years were at the low end of typical relative humidity.
+At a seasonal level, these data provide little indication that fire weather posed a barrier to conducting summer fire, and instead pointed to fuelbed greenness.
+This pattern complements regional wildfire trends, which show an increase in wildfire frequency with lower summer precipitation and higher temperatures and drought indices [@clabo2018]. 
+Overall, burn season fire weather in our data was highly variable across study years, and burn day weather spanned the gamut (the exception was wind speed, which was below average in all study years).
+Take, for example, relative humidity: for the four years summer burns were planned, two years were above average ($z$ scores 0.9 and 0.5), and two years were below average ($z$ scores -0.6 and -0.9), but each set included a year in which summer burns were completed and a year in which burns were not completed.
 In a study of these and similar grasslands that included a subset of the spring fires reported here, higher wind speed and greater vapor pressure deficit contributed to higher rates of spread, a measure of fire intensity [@mcgranahan2023a].
@@
 
-<!-- Experience does suggest fire weather conditions can limit specific prescribed fire operations.  -->
-<!-- For example, a burn was planned around a favorable weather forecast on August 13, 2018, and a full crew was on-site that morning.  -->
-<!-- But relative humidity did not fall as predicted in the spot weather forecast, remaining above 50\% until 1500 (the burn was called off by 1400 after repeated test fires failed to spread).  -->
-<!-- Certainly, actual weather defies forecasts in all seasons.  -->
-<!-- But in the context of summer prescribed fire, when fuels are more green and windspeed generally lower (`r f_ref("historical")`), stubborn relative humidity is more of a barrier in the absence of mitigating factors.
--->
+However, experience does suggest fire weather conditions can limit specific prescribed fire operations.
+For example, a burn was planned around a favorable weather forecast on 13 August 2018, and a full crew was on-site that morning. 
+But relative humidity did not fall as predicted in the spot weather forecast, remaining above 50\% until 1500 hrs (the burn was called off by 1400 hrs after repeated test fires failed to spread).
+Certainly, actual weather defies forecasts in all seasons.
+But in the context of summer prescribed fire, when fuels are more green and windspeed potentially lower (`r f_ref("historical")`), stubborn relative humidity is more of a barrier in the absence of mitigating factors.
 
-Although trends towards lower relative humidity might mitigate trends towards greener fuelbeds in the spring, cumulative seasonal rainfall, fuelbed greenness, and relative humidity likely all increase together in the summer, when plants are photosynthetically active.
-Conversely, our historical comparisons suggest the spring burn season is perhaps most susceptible to global changes, although complex interactions between drivers of global change introduce substantial uncertainty.
+Consistent with a wide body of research on land surface phenology, the single statistically-significant change in the fire environment over the 42 year period at our study location was increased spring fuelbed greenness (`r t_ref("comp_stats")`).
+Earlier spring green-up over recent decades has been documented worldwide in northern temperate ecosystems generally [@xu2018; @ren2024], and in temperate US rangelands, specifically [@zimmer2022]. 
+While earlier spring green-up in grasslands has been broadly attributed to increased spring precipitation and higher temperatures [@post2022], managers in the Northern Great Plains must also deal with non-native cool-season grasses that are particularly responsive to earlier onset of the growing season [@gasch2020].
+Conversely, observed trends toward lower spring dewpoint and relative humidity (`r f_ref("historical")`), although not statistically-significantly different at this single weather station, suggests drier air could potentially mitigate spring fuelbed greenness.
+
+While these historical comparisons suggest the spring burn season is perhaps most susceptible to global changes, complex interactions between drivers of global change introduce substantial uncertainty.
 In addition to continental-scale increases in air temperature and variability in precipitation [@kukal2016; @pendergrass2017], fire managers in the Northern Great Plains must also deal with local changes.
-For example, native grasslands of the Northern Plains---including the pastures considered here---have been invaded with exotic, cool-season grasses that green up early, change fuel structure, and reduce rate of spread [@gasch2020; @yurkonis2019].
+For example, native grasslands in the region---including the rangeland considered here---have been invaded with exotic, cool-season grasses that green up early, change fuel structure, and reduce rate of spread [@yurkonis2019].
+Even if trends towards lower relative humidity might mitigate greener fuelbeds in the spring, cumulative seasonal rainfall, fuelbed greenness, and relative humidity likely all increase together in the summer, when plants are photosynthetically active, creating potential barriers to summer burns.
+Conversely, the non-native cool-season grasses of the Northern Great Plains, especially *Poa pratensis*, are widely known to be susceptible to summer drought [@etter1951; @meyer1989; @liu2008];  dormancy due to water stress reduces fine fuel moisture and potentially facilitates fire spread.
+Given these bimodalities, we suggest that the greater the presence of non-native cool-season grasses in fuelbeds of the Northern Great Plains, the more extreme managers might expect the potential magnitude of fuel and fire responses to year-to-year variability in precipitation.
 
@@
 
-Research into opportunities to conduct prescribed fire might consider phenological changes in the fire environment, rather than conventional attempts to assess changes in the availability of "weather windows"---the set of conditions under which prescribed fire might possibly be conducted [e.g., @weir2011].
+Research into opportunities to conduct prescribed fire might apply the "land surface phenology" approach to the fire environment [@hanes2014; @zimmer2022], rather than conventional attempts to assess changes in the availability of "weather windows"---the set of conditions under which prescribed fire might possibly be conducted [e.g., @weir2011].
 While some research has used the concept of the weather window to identify seasonal shifts in  potentially favorable burn conditions---e.g., @yurkonis2019 showed a decrease in early spring burn days and an increase in late summer burn days for central North Dakota---the focus on prescriptions generally masks interactions between mitigating and exacerbating factors, fails to account for fuelbed dynamics, and is agnostic to antecedent weather and fuel conditioning.
@@
 
+```{r tab_caps}
+gr_tab_cap <- t_nums(name = 'GreenRain_z', 
+                       caption = '\\label{GreenRain_z}Seasonal deviations from historical averages in Normalized Differenced Vegetation Index (NDVI) and year-to-date precipitation for spring and summer burn seasons, as presented in Fig.
+2.
+Anomalies determined by $z$ scores (parentheses) and categorized as above, below, or within 1 standard deviation of the historical mean.
+2018 and 2019 were the years in which no summer burns were conducted; all spring burns were conducted in all years.
+')
+wx_tab_cap <- t_nums(name = 'weather_z', 
+                       caption = '\\label{weather_z}Seasonal deviations from historical averages in four fire weather variables for spring and summer burn seasons, as presented in Fig.
+3.
+Anomalies determined by $z$ scores (parentheses) and categorized as above, below, or within 1 standard deviation of the historical mean.
+2018 and 2019 were the years in which no summer burns were conducted; all spring burns were conducted in all years ')
+hist_tab_cap <- t_nums(name = 'historical_z', 
+                       caption = '\\label{historical_z}Seasonal deviations in fuel conditions and fire weather for spring and summer burn seasons, relative to two decades (1990-1999, 2012-2022) as presented in Fig.
+4.
+Anomalies determined by $z$ scores (parentheses) and categorized as above, below, or within 1 standard deviation of the mean for the given historical range.
+2018 and 2019 were the years in which no summer burns were conducted; all spring burns were conducted in all years')
+comp_tab_cap <- t_nums(name = 'comp_stats', 
+                       caption = '\\label{comp_stats}Results of statistical models comparing, within each season, early period (1990-1999) to late period (2012-2022) values for fuelbed greenness and five fire weather variables in the historical dataset (Fig.
+4).')
+```
+
+```{r GreenRain_z, eval = TRUE, tab.cap=gr_tab_cap}
+ z_tab$ndvi_prcp %>%
+  kable() 
+```
+
+\newpage
+
+```{r weather_z, eval = TRUE, tab.cap=wx_tab_cap}
+ z_tab$wxd %>%
+  kable() 
+```
+
+\newpage
+
+```{r comp_stats, eval = TRUE, tab.cap=comp_tab_cap}
+ comp_tab %>%
+  kable() 
+```
+
+\newpage
+
+```{r historical_z, eval = FALSE, tab.cap=hist_tab_cap}
+ z_tab$hist %>%
+  kable() 
+```
+
+\newpage
+
 ```{r sentinel, fig.cap = sent_cap}
@@
                    caption = "\\label{sentinel}Sentinel imagery summarized for the study years.
-A: Normalized Difference Vegetation Index (NDVI) from four adjacent unburned pastures avoids any confounding effects of prescribed burns.
+A: Normalized Difference Vegetation Index (NDVI), gathered from four adjacent unburned pastures to avoid confounding effects of prescribed burns, was greater in summer burn seasons than spring burn seasons ($\\chi^2$ = 134, $P <$ 0.001).
 X denotes seasons in which burns were not completed.
 B: Burn severity (differenced Normalized Burn Ratio, dNBR) plotted against fuelbed greenness (NDVI) retrieved from the pre-burn Sentinel-2 image used for pre-fire NBR.
-Points represent mean for individual prescribed fires with associated standard error; circles and solid trendline indicate spring burns while broken trendline and triangles denote summer burns.")
+Points represent mean for individual prescribed fires with associated standard error: circles indicate spring burns while triangles denote summer burns.
+The statistical relationship between dNBR and NDVI differs between seasons in generalized mixed-effect models ($\\chi^2$ = 8.4, $P$ < 0.01).
+dNBR declined as NDVI increased in spring burns ($t$ = -4.7, $P$ < 0.001; solid trendline, plotted as ordinary least-squares regression with 95% confidence interval, $R^2$ = 0.22); summer burns have no significant relationship (broken trendline).
+")
 include_graphics('ndvi_gg-1.png')
@@
 hist_cap <- f_nums(name = "historical", 
-                   caption = "\\label{historical}Temporal trends in the phenology of the fire environment.
+                   caption = "\\label{historical}Temporal comparison of phenological changes in the fire environment over 42 years.
 Boxplots represent data from the first decade of the dataset used here (1990-1999) to the final decade (2012-2022), by burn season.
-Horizontal lines represent average conditions for the four study years (2017-2020).")
+Horizontal lines represent average conditions for the four study years (2017-2020).
+Table 3 reports statistical tests of early vs.
+late periods for each season by each variable; Supplemental Information Table 1 reports z scores for each study year against historical values.")
 include_graphics('historical_phenology-1.png')

### 2025-04-21 SpringSummerSeverity/paper/SummerFireRestorationRE0.Rmd
Spring/summer seasonality R1 update

--- 2024-11-25 SpringSummerSeverity/paper/SummerFireRestoration.Rmd
@@
 title: Evaluating an attempt to restore summer fire in the Northern Great Plains
+subtitle: 'Running head: Summer burning in the Northern Plains'
 author:
@@
 
+**Author contributions:** Conception, analysis, writing: DAM; Programming and editing: JPA
+
+\newpage 
 # Abstract
@@
 **Keywords:** Burn severity; Phenology; Seasonality of fire; Pyrodiversity
+
+# Implications for practice
+
+* Successful Rx fire in the summer is likely the norm, not the exception. 
+* At least at this specific research station, there is little evidence that fuel and weather conditions have changed over 42 years except for slightly greener spring fuels and lower relative humidity, which might cancel each other out.
 
@@
 To characterize seasonal fire weather, we used daily precipitation totals to eliminate days with more than 25 mm precipitation---days no fire would be conducted under any other circumstance---and calculated mean values for the operational period for each spring (25 April--25 May) and summer (7--27 August) burn season.
-Seasonal values were determined for both the study period (spring and summer burn seasons, 2017--2020) and the historical range (1 April- 31 October, 1990--2023).
+Seasonal values were determined for both the study period (spring and summer burn seasons, 2017--2020) and the historical range (1 April--31 October, 1990--2023).
 
@@
 
+The authors declare no conflict of interest.
+
 \newpage

### 2025-04-21 SpringSummerSeverity/paper/SupplementalInformation.Rmd
Spring/summer seasonality R1 update

--- 2024-12-11 SpringSummerSeverity/paper/AllScriptTogetherSI.Rmd
@@
 
+```{r setup, echo=FALSE, warning=FALSE, message = FALSE, results='hide'}
+knitr::opts_chunk$set(message = FALSE, warning=FALSE, 
+                      echo=FALSE, eval=TRUE)
+pacman::p_load(tidyverse, xtable, wesanderson)
+  BuGnPal = c( "#67A9CF" ,"#3690C0" ,"#02818A", "#016450")
+  FFgrad = wes_palette("FantasticFox1")[c(3,2,4,5)]
+  load('../data/NoFireIndices.Rdata')
+  # load('../data/BurnIndices.Rdata')
+  # load('../data/NoFireTrends.Rdata')
+  load('../data/WxData.Rdata')
+# some necessary things
+  sys_yr <- 
+    Sys.Date() %>%
+      as.POSIXct(., '%Y-%M-%d') %>%
+      year() 
+  seasons <- 
+    tibble(season = c('Spring', 'Summer'), 
+           start = c(as.Date(paste0(sys_yr, '-04-25')), as.Date(paste0(sys_yr, '-08-07'))), 
+           end = c(as.Date(paste0(sys_yr, '-05-25')), as.Date(paste0(sys_yr, '-08-27'))))
+```
+
+# Additional data and results
+
+## Seasonal trends
+
+```{r ndvi_trend, eval = TRUE, fig.cap="Seasonal trends in Landsat-derived NDVI for the study period.
+Green shaded areas span the first and last dates of completed burns in the spring and summer.", fig.height = 4, fig.width = 8}
+NoFireIndices %>%
+  group_by(ImageDate, pasture) %>%
+    summarise(Mean = mean(ndvi), 
+              SEM = sd(ndvi)/sqrt(n()), 
+              .groups = 'drop') %>%
+  mutate(ImageDate = as.Date(ImageDate, '%Y-%m-%d'), 
+         Year = lubridate::year(ImageDate), 
+         Year = as.factor(Year), 
+         SampleDate = format(ImageDate, '%m-%d'), 
+         SampleDate = as.Date(SampleDate, '%m-%d') ) %>%
+  filter(Year != 2016) %>%
+ggplot(aes(x = SampleDate, 
+           color = Year, 
+           fill = Year, 
+           group = interaction(pasture, Year))) + theme_bw(14) +
+  geom_rect(data= seasons, inherit.aes = FALSE,
+              aes(xmin=start, xmax=end, ymin=-Inf, ymax=+Inf), 
+              fill='#A1D99B', alpha=0.5) +
+  geom_smooth(aes(y = Mean, group = Year),
+              se = T,
+              level = 0.99) +
+  geom_errorbar(aes(ymin = Mean - SEM, 
+                    ymax = Mean + SEM),
+                alpha = 0.5) +
+  geom_line(aes(y = Mean),
+             alpha = 0.5) +
+  labs(x = 'Image date', y = 'NDVI') + 
+    scale_fill_manual(values = FFgrad) + 
+  scale_color_manual(values = FFgrad) + 
+  scale_x_date( date_breaks = 'month', 
+                date_labels = '%b-%d', 
+                minor_breaks = NULL) 
+```
+
+```{r precip_trend, eval = TRUE, fig.cap="Historical precipitation data for the entire 42-year period.
+Thick black line shows smoothed 42-yr trend, gray lines individual years, and colored lines study years.
+Green shaded areas span the first and last dates of completed burns in the spring and summer.", fig.height = 4, fig.width = 8}
+  WxData$Rainfall %>% 
+  mutate(day = day + years(1)) %>%
+    filter(month(date) %in% c(04:09)) %>%
+    separate(date, into = c('Year', 'Month', 'Day'),sep = '-') %>%
+    group_by(Year) %>%
+    mutate(CumPrecip = cumsum(Rainfall)) %>%
+    ggplot(aes(x = day)) + theme_bw(14) +
+    geom_rect(data= seasons, inherit.aes = FALSE,
+              aes(xmin=start, xmax=end, ymin=-Inf, ymax=+Inf),
+              fill='#A1D99B', alpha=0.5) +
+    geom_line(aes(y = CumPrecip, group = Year),
+              color = 'grey40', 
+              alpha = 0.5) +
+    geom_smooth(aes(y = CumPrecip),
+                se = F,
+                color = 'black',
+                size = 1,
+                level = 0.99) +
+    geom_line(data = . %>%
+                filter(as.numeric(Year) %in% c(2017:2020)) , 
+              aes(y = CumPrecip, 
+                  color = Year,
+                  group = Year),
+              size = 2) +
+    labs(x = 'Day of year', y = 'Cumulative rainfall (mm)') + 
+    scale_fill_manual(values = FFgrad) + 
+    scale_color_manual(values = FFgrad) + 
+    scale_x_date( date_breaks = 'month', 
+                  date_labels = '%b-%d', 
+                  minor_breaks = NULL)
+```
+
+\clearpage
+
+## $z$ scores for historical data by burn season
+
+```{r historical_z, eval = TRUE, results='asis'}
+load('../data/z_tab.Rdata')
+ z_tab$hist %>%
+  xtable(caption = 'Seasonal deviations in fuel conditions and fire weather for spring and summer burn seasons, relative to two decades (1990-1999, 2012-2022) as presented in Fig.
+4.
+Anomalies determined by $z$ scores (parentheses) and categorized as above, below, or within 1 standard deviation of the mean for the given historical range.
+2018 and 2019 were the years in which no summer burns were conducted; all spring burns were conducted in all years') %>%
+   print(include.rownames = F, comment = F)
+```
+
+\newpage
+
+# Script
+
 This document provides script for fetching remotely-sensed imagery and processing results in the \textsf{R} statistical environment.
 
-# Remote sensing
+## Remote sensing
 
-## Sentinel-2
+### Sentinel-2
 
@@
 \clearpage
-## Landsat
+### Landsat
 
@@
 
-# Analysis
+## Analysis
 
@@
 
-## Remotely sensed imagery
+### Remotely sensed imagery
 
@@
 
-## Weather data
+### Weather data
 
@@
 
-seasons <- tibble(season = c('spring', 'summer'), 
-                 start = c('04-20', '07-20'), 
-                 end = c('06-10', '09-10') ) %>%
-          mutate(start = as.Date(start, '%m-%d'), 
-                 end = as.Date(end, '%m-%d'))
+  sys_yr <- 
+    Sys.Date() %>%
+      as.POSIXct(., '%Y-%M-%d') %>%
+      year() 
+  seasons <- 
+    tibble(season = c('Spring', 'Summer'), 
+           start = c(as.Date(paste0(sys_yr, '-04-20')), 
+                     as.Date(paste0(sys_yr, '-07-20'))), 
+           end = c(as.Date(paste0(sys_yr, '06-10')), 
+                   as.Date(paste0(sys_yr, '09-10'))))
 
@@
              between(day, seasons$start[2], seasons$end[2]) ~'Summer',
-             TRUE ~ NA      
-           ) ) %>% 
+             TRUE ~ NA  ) ) %>% 
      unite( c(Year, Month, Day), col = 'date', sep = '-')
@@
              between(day, seasons$start[2], seasons$end[2]) ~'Summer',
-             TRUE ~ NA      
-           ) ) %>%
+             TRUE ~ NA  ) ) %>%
     mutate(e = 6.11 * (10 ^ ( (7.5 * DewPoint)/ (237.3 + DewPoint) ) ), 
@@
                  values_to = 'value', 
+                 cols = DewPoint:VPD) 
+\end{lstlisting}
+
+### Statistical analysis
+
+\textsf{R} script for statistical analysis is organized by the manuscript figure for which the tests support.
+
+**Figure 1**
+
+\begin{lstlisting}[language=R]
+
+# A: Test NDVI in study years across seasons 
+  # Wrangle NDVI data from adjacent unburned pastures
+    ndvi_dat <-
+      NoFireIndices %>%
+        group_by(ImageDate, pasture) %>%
+        summarise(NDVI = median(ndvi) , 
+                  .groups = 'drop') %>%
+        mutate(ImageDate = as.Date(ImageDate, '%Y-%m-%d'), 
+               Year = lubridate::year(ImageDate), 
+               Year = as.factor(Year), 
+               SampleDate = format(ImageDate, '%m-%d'), 
+               SampleDate = as.Date(SampleDate, '%m-%d'), 
+               season = case_when(
+                 between(SampleDate, seasons$start[1], seasons$end[1]) ~ 'Spring',
+                 between(SampleDate, seasons$start[2], seasons$end[2]) ~ 'Summer', 
+                 TRUE ~ NA ) ) %>%
+        filter(!is.na(season)) %>%
+      select(pasture, Year, season, NDVI)
+  # Fit mixed-effect models
+    m0 <- lme4::lmer(NDVI ~ 1 + (1|Year), REML=FALSE, data = ndvi_dat) 
+    m1 <- lme4::lmer(NDVI ~ season + (1|Year), REML=FALSE, data = ndvi_dat)
+  # Compare models for significance of season term
+    anova(m0, m1)
+
+# B: Test for difference between seasons
+  # Wrangle severity and NDVI data from burned pastures
+    reg_dat <-
+      BurnIndices  %>%
+        group_by(Year, Season, fire) %>%
+        summarize(dNBR = mean(dNBR), 
+                  NDVI = mean(ndvi), 
+                  .groups = 'drop') %>%
+        separate(fire, into=c('block','pasture','patch'), sep='-')
+  # Fit models
+    m0 <- lme4::glmer(dNBR ~ 1 + (1|block) , 
+                  family = Gamma(link = "identity"), 
+                  data = reg_dat)
+    m1 <- lme4::glmer(dNBR ~ NDVI + (1|block), 
+                  family = Gamma(link = "identity"), 
+                  data = reg_dat)
+    m2 <- lme4::glmer(dNBR ~ NDVI + Season + (1|block) , 
+                      family = Gamma(link = "identity"), 
+                      data = reg_dat)
+  # Compare models
+    anova(m0, m1, m2)
+
+\end{lstlisting}
+
+**Figure 2**
+
+Calculate $z$ scores.
+
+\begin{lstlisting}[language=R]
+# NDVI x burn season 
+  ndvi_z <-
+    LandsatSums %>%
+    filter(year %in% c(2017:2020), 
+           index == 'ndvi') %>%
+    group_by(year, season) %>%
+    summarize(Mean = mean(Value), 
+              .groups = 'drop') %>%
+    left_join(by = 'season', 
+              LandsatSums %>%
+                filter(index == 'ndvi') %>%
+                group_by(season) %>%
+                summarize(HistMean = mean(Value, na.rm = T),
+                          HistSD = sd(Value, na.rm = T),
+                          .groups = 'drop') ) %>%
+    mutate(z = (Mean - HistMean)/HistSD, 
+           interp = case_when(
+             z < -1 ~ "Below", 
+             z > 1 ~ "Above", 
+             TRUE ~ 'Within' ) )
+# YTD precip x burn season 
+  prcp <- 
+    WxData$Rainfall %>% 
+    filter(month(date) %in% c(04:09)) %>%
+    separate(date, into = c('Year', 'Month', 'Day'),sep = '-') %>%
+    group_by(Year) %>%
+    mutate(CumPrecip = cumsum(Rainfall), 
+           Month = recode(Month, '05' ="May", '08'='Aug'), 
+           Month = paste0(Month, ' ', Day)) %>%
+    ungroup() %>%
+    filter(Month %in% c('May 15', 'Aug 01' )) %>%
+    mutate(season = recode(Month, 'May 15' = 'Spring', 'Aug 01' = 'Summer' )) %>%
+    select(Year, season, CumPrecip)
+  prcp_z <-
+    prcp %>%
+    filter(as.numeric(Year) %in% c(2017:2020)) %>%
+    group_by(Year, season) %>%
+    summarize(Mean = mean(CumPrecip), 
+              .groups = 'drop') %>%
+    left_join(by = 'season', 
+              prcp %>%
+                group_by(season) %>%
+                summarize(HistMean = mean(CumPrecip, na.rm = T),
+                          HistSD = sd(CumPrecip, na.rm = T),
+                          .groups = 'drop') ) %>%
+    mutate(z = (Mean - HistMean)/HistSD, 
+           interp = case_when(
+             z < -1 ~ "Below", 
+             z > 1 ~ "Above", 
+             TRUE ~ 'Within' ))
+
+\end{lstlisting}
+
+**Figure 3**
+
+Compare 1990-1999 period to 2012-2022 for fuel and fire weather.
+
+\begin{lstlisting}[language=R]
+
+# test historical fuel and fire weather across time periods 
+    comp_tab <-
+      hist_wx %>%
+        group_by(season, variable) %>%
+        group_modify(~ broom::tidy(lm(Value ~ period, data = .x))) %>%
+        ungroup() %>%
+        filter(term != '(Intercept)') %>%
+        mutate(P = case_when(
+          p.value > 0.05 ~ 'P > 0.05', 
+          between(p.value, 0.00099, 0.01) ~ 'P < 0.01', 
+          p.value < 0.001 ~ 'P < 0.001', 
+          TRUE ~ paste0('P = ', round(p.value, 2)) ), 
+          result = paste0('t = ', signif(statistic, 2), ", ", P)) %>%
+        select(variable, season, result) %>%
+        pivot_wider(names_from = season, 
+                    values_from = result) 
+\end{lstlisting}
+
+Calculate $z$ scores.
+
+\begin{lstlisting}[language=R]
+
+ hist_wx <-
+    WxData$HourlyWx %>%
+      filter(! date %in% RainyDays$date, 
+             ! is.na(season)) %>%
+      group_by(Year, date, season) %>%
+      summarise_at(.vars = vars(c(DewPoint, RelHum, WindSpeed,VPD)),
+                   .funs = c("mean")) %>%
+      ungroup() %>%
+      pivot_longer(names_to = 'variable', 
+                   values_to = 'value', 
                    cols = DewPoint:VPD) %>%
-    mutate(variable = recode(variable, 
-                             'DewPoint' = 'Dew Point (°C)', 
-                             'RelHum' = 'Relative Humidity (%)', 
-                             'VPD' = 'Vapor Pressure Deficit', 
-                             'WindSpeed' = 'Wind speed (m/s)') , 
-           Year = as.character(Year)) 
+      group_by(Year, season, variable) %>%
+      summarise(Value = mean(value), 
+                .groups = 'drop') %>%
+      mutate(Year = as.character(Year), 
+             #season = fct_rev(season), 
+             period = case_when(
+               Year %in% c(1990:1999) ~  '1990-1999', 
+               Year %in% c(2012:2022) ~  '2012-2022', 
+               TRUE ~ NA)) %>%
+      filter(!is.na(period)) %>%
+      bind_rows(WxData$Rainfall %>% 
+                  filter(month(date) %in% c(04:09)) %>%
+                  separate(date, into = c('Year', 'Month', 'Day'),sep = '-') %>%
+                  group_by(Year) %>%
+                  mutate(CumPrecip = cumsum(Rainfall), 
+                         Month = recode(Month, '05' ="May", '08'='Aug'), 
+                         season = paste0(Month, ' ', Day)) %>%
+                  ungroup() %>%
+                  filter(season %in% c('May 15', 'Aug 01' )) %>%
+                  mutate(season = ifelse(season == 'May 15', 'Spring', 'Summer'), 
+                         period = case_when(
+                           Year %in% c(1990:1999) ~  '1990-1999', 
+                           Year %in% c(2012:2022) ~  '2012-2022', 
+                           TRUE ~ NA ) )  %>%
+                  filter(!is.na(period)) %>%
+                  select(Year, season, CumPrecip, period) %>%
+                  pivot_longer(names_to = 'variable', 
+                               values_to = 'Value', 
+                               cols = CumPrecip))  %>%
+      bind_rows(HistoricalSpectralData  %>% 
+                  select(-pasture) %>%
+                  filter(index == 'NDVI', 
+                         range != '1990-2022') %>%
+                  rename(Year = year, period = range, variable = index) ) 
+    hist_wx %>%
+      filter(as.numeric(Year) %in% c(2017:2020)) %>%
+      group_by(Year, season, variable) %>%
+      slice(1) %>%
+      summarize(Mean = mean(Value), 
+                .groups = 'drop') %>%
+      left_join(by = c('season', 'variable'), 
+                relationship = "many-to-many",
+                hist_wx %>%
+                  group_by(period, season, variable) %>%
+                  summarize(HistMean = mean(Value, na.rm = T),
+                            HistSD = sd(Value, na.rm = T),
+                            .groups = 'drop') ) %>%
+      mutate(z = (Mean - HistMean)/HistSD, 
+             interp = case_when(
+               z < -1 ~ "Below", 
+               z > 1 ~ "Above", 
+               TRUE ~ 'Within') )
+
 \end{lstlisting}

### 2025-04-21 SpringSummerSeverity/paper/SupplementalInformation.tex
Spring/summer seasonality R1 update

--- 2025-04-21 SpringSummerSeverity/paper/SupplementalInformation.Rmd
@@
----
-title: Evaluating an attempt to restore summer fire in the Northern Great Plains
-author:
-- Devan Allen McGranahan and Jay P. Angerer
-subtitle:  'Supplementary Information'
-header-includes:
-  - \usepackage{amsmath}
-  - \renewcommand{\familydefault}{\sfdefault}
-  - \usepackage[colorlinks=true, citecolor=blue, urlcolor=blue,linkcolor=blue, pdfborder={0 0 0}]{hyperref}
-  - \urlstyle{same}
-  - \usepackage{xcolor} 
-  - \usepackage{booktabs}
-  - \raggedright
-  - \usepackage{listings}
-output:
-  pdf_document:
-    highlight: tango
-    keep_tex: yes
-    toc: false
----
+% Options for packages loaded elsewhere
+\PassOptionsToPackage{unicode}{hyperref}
+\PassOptionsToPackage{hyphens}{url}
+%
+\documentclass[
+]{article}
+\usepackage{amsmath,amssymb}
+\usepackage{iftex}
+\ifPDFTeX
+  \usepackage[T1]{fontenc}
+  \usepackage[utf8]{inputenc}
+  \usepackage{textcomp} % provide euro and other symbols
+\else % if luatex or xetex
+  \usepackage{unicode-math} % this also loads fontspec
+  \defaultfontfeatures{Scale=MatchLowercase}
+  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+\fi
+\usepackage{lmodern}
+\ifPDFTeX\else
+  % xetex/luatex font selection
+\fi
+% Use upquote if available, for straight quotes in verbatim environments
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+\IfFileExists{microtype.sty}{% use microtype if available
+  \usepackage[]{microtype}
+  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
+}{}
+\makeatletter
+\@ifundefined{KOMAClassName}{% if non-KOMA class
+  \IfFileExists{parskip.sty}{%
+    \usepackage{parskip}
+  }{% else
+    \setlength{\parindent}{0pt}
+    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
+}{% if KOMA class
+  \KOMAoptions{parskip=half}}
+\makeatother
+\usepackage{xcolor}
+\usepackage[margin=1in]{geometry}
+\usepackage{graphicx}
+\makeatletter
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
+\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
+\makeatother
+% Scale images if necessary, so that they will not overflow the page
+% margins by default, and it is still possible to overwrite the defaults
+% using explicit options in \includegraphics[width, height, ...]{}
+\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+% Set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\makeatother
+\setlength{\emergencystretch}{3em} % prevent overfull lines
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
+\usepackage{amsmath}
+\renewcommand{\familydefault}{\sfdefault}
+\usepackage[colorlinks=true, citecolor=blue, urlcolor=blue,linkcolor=blue, pdfborder={0 0 0}]{hyperref}
+\urlstyle{same}
+\usepackage{xcolor}
+\usepackage{booktabs}
+\raggedright
+\usepackage{listings}
+\ifLuaTeX
+  \usepackage{selnolig}  % disable illegal ligatures
+\fi
+\usepackage{bookmark}
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\urlstyle{same}
+\hypersetup{
+  pdftitle={Evaluating an attempt to restore summer fire in the Northern Great Plains},
+  pdfauthor={Devan Allen McGranahan and Jay P. Angerer},
+  hidelinks,
+  pdfcreator={LaTeX via pandoc}}
 
-```{r setup, echo=FALSE, warning=FALSE, message = FALSE, results='hide'}
-knitr::opts_chunk$set(message = FALSE, warning=FALSE, 
-                      echo=FALSE, eval=TRUE)
-pacman::p_load(tidyverse, xtable, wesanderson)
-  BuGnPal = c( "#67A9CF" ,"#3690C0" ,"#02818A", "#016450")
-  FFgrad = wes_palette("FantasticFox1")[c(3,2,4,5)]
-  load('../data/NoFireIndices.Rdata')
-  # load('../data/BurnIndices.Rdata')
-  # load('../data/NoFireTrends.Rdata')
-  load('../data/WxData.Rdata')
-# some necessary things
-  sys_yr <- 
-    Sys.Date() %>%
-      as.POSIXct(., '%Y-%M-%d') %>%
-      year() 
-  seasons <- 
-    tibble(season = c('Spring', 'Summer'), 
-           start = c(as.Date(paste0(sys_yr, '-04-25')), as.Date(paste0(sys_yr, '-08-07'))), 
-           end = c(as.Date(paste0(sys_yr, '-05-25')), as.Date(paste0(sys_yr, '-08-27'))))
-```
+\title{Evaluating an attempt to restore summer fire in the Northern
+Great Plains}
+\usepackage{etoolbox}
+\makeatletter
+\providecommand{\subtitle}[1]{% add subtitle to \maketitle
+  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
+}
+\makeatother
+\subtitle{Supplementary Information}
+\author{Devan Allen McGranahan and Jay P. Angerer}
+\date{}
 
-# Additional data and results
+\begin{document}
+\maketitle
 
-## Seasonal trends
+\section{Additional data and results}\label{additional-data-and-results}
 
-```{r ndvi_trend, eval = TRUE, fig.cap="Seasonal trends in Landsat-derived NDVI for the study period.
-Green shaded areas span the first and last dates of completed burns in the spring and summer.", fig.height = 4, fig.width = 8}
-NoFireIndices %>%
-  group_by(ImageDate, pasture) %>%
-    summarise(Mean = mean(ndvi), 
-              SEM = sd(ndvi)/sqrt(n()), 
-              .groups = 'drop') %>%
-  mutate(ImageDate = as.Date(ImageDate, '%Y-%m-%d'), 
-         Year = lubridate::year(ImageDate), 
-         Year = as.factor(Year), 
-         SampleDate = format(ImageDate, '%m-%d'), 
-         SampleDate = as.Date(SampleDate, '%m-%d') ) %>%
-  filter(Year != 2016) %>%
-ggplot(aes(x = SampleDate, 
-           color = Year, 
-           fill = Year, 
-           group = interaction(pasture, Year))) + theme_bw(14) +
-  geom_rect(data= seasons, inherit.aes = FALSE,
-              aes(xmin=start, xmax=end, ymin=-Inf, ymax=+Inf), 
-              fill='#A1D99B', alpha=0.5) +
-  geom_smooth(aes(y = Mean, group = Year),
-              se = T,
-              level = 0.99) +
-  geom_errorbar(aes(ymin = Mean - SEM, 
-                    ymax = Mean + SEM),
-                alpha = 0.5) +
-  geom_line(aes(y = Mean),
-             alpha = 0.5) +
-  labs(x = 'Image date', y = 'NDVI') + 
-    scale_fill_manual(values = FFgrad) + 
-  scale_color_manual(values = FFgrad) + 
-  scale_x_date( date_breaks = 'month', 
-                date_labels = '%b-%d', 
-                minor_breaks = NULL) 
-```
+\subsection{Seasonal trends}\label{seasonal-trends}
 
-```{r precip_trend, eval = TRUE, fig.cap="Historical precipitation data for the entire 42-year period.
-Thick black line shows smoothed 42-yr trend, gray lines individual years, and colored lines study years.
-Green shaded areas span the first and last dates of completed burns in the spring and summer.", fig.height = 4, fig.width = 8}
-  WxData$Rainfall %>% 
-  mutate(day = day + years(1)) %>%
-    filter(month(date) %in% c(04:09)) %>%
-    separate(date, into = c('Year', 'Month', 'Day'),sep = '-') %>%
-    group_by(Year) %>%
-    mutate(CumPrecip = cumsum(Rainfall)) %>%
-    ggplot(aes(x = day)) + theme_bw(14) +
-    geom_rect(data= seasons, inherit.aes = FALSE,
-              aes(xmin=start, xmax=end, ymin=-Inf, ymax=+Inf),
-              fill='#A1D99B', alpha=0.5) +
-    geom_line(aes(y = CumPrecip, group = Year),
-              color = 'grey40', 
-              alpha = 0.5) +
-    geom_smooth(aes(y = CumPrecip),
-                se = F,
-                color = 'black',
-                size = 1,
-                level = 0.99) +
-    geom_line(data = . %>%
-                filter(as.numeric(Year) %in% c(2017:2020)) , 
-              aes(y = CumPrecip, 
-                  color = Year,
-                  group = Year),
-              size = 2) +
-    labs(x = 'Day of year', y = 'Cumulative rainfall (mm)') + 
-    scale_fill_manual(values = FFgrad) + 
-    scale_color_manual(values = FFgrad) + 
-    scale_x_date( date_breaks = 'month', 
-                  date_labels = '%b-%d', 
-                  minor_breaks = NULL)
-```
+\begin{figure}
+\centering
+\includegraphics{SupplementalInformation_files/figure-latex/ndvi_trend-1.pdf}
+\caption{Seasonal trends in Landsat-derived NDVI for the study period.
+Green shaded areas span the first and last dates of completed burns in
+the spring and summer.}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics{SupplementalInformation_files/figure-latex/precip_trend-1.pdf}
+\caption{Historical precipitation data for the entire 42-year period.
+Thick black line shows smoothed 42-yr trend, gray lines individual
+years, and colored lines study years. Green shaded areas span the first
+and last dates of completed burns in the spring and summer.}
+\end{figure}
 
@@
 
-## $z$ scores for historical data by burn season
+\subsection{\texorpdfstring{\(z\) scores for historical data by burn
+season}{z scores for historical data by burn season}}\label{z-scores-for-historical-data-by-burn-season}
 
-```{r historical_z, eval = TRUE, results='asis'}
-load('../data/z_tab.Rdata')
- z_tab$hist %>%
-  xtable(caption = 'Seasonal deviations in fuel conditions and fire weather for spring and summer burn seasons, relative to two decades (1990-1999, 2012-2022) as presented in Fig.
+\begin{table}[ht]
+\centering
+\begin{tabular}{llllll}
+  \hline
+Variable & Year & Spring 1990-99 & Spring 2012-22 & Summer 1990-99 & Summer 2012-22 \\ 
+  \hline
+Fuelbed Greenness (NDVI) & 2017 & Above (1.3) & Within (0.96) & Within (0.34) & Within (-0.05) \\ 
+    & 2018 & Within (0.15) & Within (-0.87) & Within (0.92) & Within (0.64) \\ 
+    & 2019 & Within (-0.54) & Below (-1.96) & Above (1.67) & Above (1.51) \\ 
+    & 2020 & Above (1) & Within (0.48) & Within (0.37) & Within (-0.01) \\ 
+  Cumulative rainfall & 2017 & Within (-0.63) & Within (-0.89) & Within (-0.89) & Within (-0.91) \\ 
+    & 2018 & Within (-0.76) & Below (-1.08) & Above (1.87) & Above (1.38) \\ 
+    & 2019 & Within (-0.27) & Within (-0.37) & Above (1.04) & Within (0.69) \\ 
+    & 2020 & Within (0.02) & Within (0.06) & Below (-1.16) & Below (-1.13) \\ 
+  Dew Point & 2017 & Below (-1.21) & Within (-0.98) & Below (-1.15) & Below (-1.07) \\ 
+    & 2018 & Within (-0.05) & Within (0.63) & Within (-0.27) & Within (-0.07) \\ 
+    & 2019 & Below (-1.01) & Within (-0.7) & Within (-0.21) & Within (0) \\ 
+    & 2020 & Within (0.41) & Above (1.28) & Within (0.9) & Above (1.27) \\ 
+  Relative Humidity & 2017 & Within (-0.85) & Within (-0.46) & Within (-0.56) & Within (-0.37) \\ 
+    & 2018 & Within (-0.99) & Within (-0.67) & Within (-0.82) & Within (-0.65) \\ 
+    & 2019 & Within (-0.48) & Within (0.12) & Within (0.67) & Above (1.01) \\ 
+    & 2020 & Within (0.37) & Above (1.43) & Within (0.32) & Within (0.62) \\ 
+  Vapor Pressure Deficit & 2017 & Within (0.48) & Within (0.26) & Within (-0.17) & Within (-0.32) \\ 
+    & 2018 & Above (1.26) & Above (1.3) & Within (0.91) & Within (0.82) \\ 
+    & 2019 & Within (-0.38) & Within (-0.9) & Below (-1.15) & Below (-1.36) \\ 
+    & 2020 & Within (-0.62) & Below (-1.21) & Within (-0.13) & Within (-0.28) \\ 
+  Wind speed & 2017 & Within (-0.69) & Within (-0.19) & Below (-1.69) & Within (-0.91) \\ 
+    & 2018 & Within (-0.97) & Within (-0.42) & Below (-2.06) & Below (-1.31) \\ 
+    & 2019 & Below (-1.54) & Within (-0.89) & Below (-1.3) & Within (-0.5) \\ 
+    & 2020 & Below (-1.24) & Within (-0.65) & Within (-0.74) & Within (0.1) \\ 
+   \hline
+\end{tabular}
+\caption{Seasonal deviations in fuel conditions and fire weather for spring and summer burn seasons, relative to two decades (1990-1999, 2012-2022) as presented in Fig.
 4.
 Anomalies determined by $z$ scores (parentheses) and categorized as above, below, or within 1 standard deviation of the mean for the given historical range.
-2018 and 2019 were the years in which no summer burns were conducted; all spring burns were conducted in all years') %>%
-   print(include.rownames = F, comment = F)
-```
+2018 and 2019 were the years in which no summer burns were conducted; all spring burns were conducted in all years}
+\end{table}
 
@@
 
-# Script
+\section{Script}\label{script}
 
-This document provides script for fetching remotely-sensed imagery and processing results in the \textsf{R} statistical environment.
+This document provides script for fetching remotely-sensed imagery and
+processing results in the \textsf{R} statistical environment.
 
-## Remote sensing
+\subsection{Remote sensing}\label{remote-sensing}
 
-### Sentinel-2
+\subsubsection{Sentinel-2}\label{sentinel-2}
 
-This \texttt{EvalScript} can be used in the [Copernicus browser](https://browser.dataspace.copernicus.eu/) as a Custom script to create a Custom visualization that can be exported as a 16-bit \texttt{.tiff} for raster analysis.
-The user must select the area of interest and imagery dates and is responsible for assessing cloudiness.
+This \texttt{EvalScript} can be used in the
+\href{https://browser.dataspace.copernicus.eu/}{Copernicus browser} as a
+Custom script to create a Custom visualization that can be exported as a
+16-bit \texttt{.tiff} for raster analysis. The user must select the area
+of interest and imagery dates and is responsible for assessing
+cloudiness.
 
@@
 \clearpage
-### Landsat
 
-This script for [Google Earth Engine](https://code.earthengine.google.com/) exports a composited NDVI \texttt{.tiff} file to the user's Google Drive for the area of interest defined as \texttt{geometry} for an eight week period beginning with each date given in \texttt{listDate}.
-The script combines imagery from Landsat missions 5, 7, \& 8 and handles cloud masking.
+\subsubsection{Landsat}\label{landsat}
+
+This script for \href{https://code.earthengine.google.com/}{Google Earth
+Engine} exports a composited NDVI \texttt{.tiff} file to the user's
+Google Drive for the area of interest defined as \texttt{geometry} for
+an eight week period beginning with each date given in
+\texttt{listDate}. The script combines imagery from Landsat missions 5,
+7, \& 8 and handles cloud masking.
 
@@
 
-## Analysis
+\subsection{Analysis}\label{analysis}
 
-Because the analysis script assumes hundreds of imagery files and a large Excel file with 42 years of hourly weather observations have been saved locally, it cannot be run directly from this document but is provided here for transparency and reference.
+Because the analysis script assumes hundreds of imagery files and a
+large Excel file with 42 years of hourly weather observations have been
+saved locally, it cannot be run directly from this document but is
+provided here for transparency and reference.
 
-### Remotely sensed imagery
+\subsubsection{Remotely sensed imagery}\label{remotely-sensed-imagery}
 
@@
 
-### Weather data
+\subsubsection{Weather data}\label{weather-data}
 
-This script wrangles the weather data downloaded from the [North Dakota Ag Weather Network's Streeter station](https://ndawn.ndsu.nodak.edu/station-info.html?station=48).
+This script wrangles the weather data downloaded from the
+\href{https://ndawn.ndsu.nodak.edu/station-info.html?station=48}{North
+Dakota Ag Weather Network's Streeter station}.
 
@@
 
-### Statistical analysis
+\subsubsection{Statistical analysis}\label{statistical-analysis}
 
-\textsf{R} script for statistical analysis is organized by the manuscript figure for which the tests support.
+\textsf{R} script for statistical analysis is organized by the
+manuscript figure for which the tests support.
 
-**Figure 1**
+\textbf{Figure 1}
 
@@
 
-**Figure 2**
+\textbf{Figure 2}
 
-Calculate $z$ scores.
+Calculate \(z\) scores.
 
@@
 
-**Figure 3**
+\textbf{Figure 3}
 
@@
 
-Calculate $z$ scores.
+Calculate \(z\) scores.
 
@@
 \end{lstlisting}
+
+\end{document}

### 2025-04-21 SpringSummerSeverity/writeup.Rmd
Spring/summer seasonality R1 update

--- 2024-10-29 SpringSummerSeverity/writeup.Rmd
@@
 ---
-title: Spring vs. summer fire regimes in central North Dakota
-author: DA McGranahan & JP Angerer??
+title: Energy infrastructure as built environment in rural rangelands
+author: DA McGranahan
 output: 
@@
                       echo=FALSE, eval=TRUE)
-pacman::p_load(tidyverse)
+pacman::p_load(tidyverse, cowplot, patchwork, wesanderson)
+  BuGnPal = c( "#67A9CF" ,"#3690C0" ,"#02818A", "#016450")
+  FFgrad = wes_palette("FantasticFox1")[c(3,2,4,5)]
 # load data 
-  load('./data/NoFireIndices.Rdata')
-  load('./data/BurnIndices.Rdata')
-# some necessary things
-  seasons <- 
-    tibble(season = c('Spring', 'Summer'), 
-           start = c(as.Date('2024-04-25'), as.Date('2024-08-07')), 
-           end = c(as.Date('2024-05-25'), as.Date('2024-08-27'))) 
+  load('./data/NoPlainsEnergy.Rdata')
 ```
 
-# In a nutshell
+```{r region_map}
+footprint <- 
+    bind_rows(
+      NoPlainsEnergy$TurbineFootprint %>%
+        select( ) %>%
+        mutate(type = 'Wind farm'), 
+      NoPlainsEnergy$WellFootprint %>%
+         select( ) %>%
+        mutate(type = 'Petroleum basin') ) %>% 
+    st_simplify(dTolerance = 500)
 
-## Background
+np_us <-
+  ggplot() + theme_void() + 
+  geom_sf(data = us, fill = 'grey90', color = NA) +
+  geom_sf(data =  np %>% st_union( ),
+          fill = 'lightblue', color = NA)  + 
+  geom_sf(data = us, 
+          size = 0.15, 
+          fill = NA, color = 'grey60') 
+np_footprints <-
+  ggplot() + theme_void() +
+  geom_sf(data = np, fill = NA) +
+  geom_sf(data = footprint, aes(fill = type, color = type)) +
+  scale_color_manual("Energy type", values = wes_palette('Zissou1')[c(2,5)]) +
+  scale_fill_manual("Energy type", values = wes_palette('Zissou1')[c(1,5)])
 
-* From 2017-2020, a project was conducted to restore spatially-patchy Rx fire in Central ND
-* Two burning regimes:
-  - 4 x 40-acre, full-patch burns each spring 
-  - 4 x 20-acre, half-patch burns, first half in spring, second half in summer
-* Mixed success in restoring fire: 
-  - All spring burns were completed in each year 
-  - Summer burns only completed in 2/4 years *but not for lack of trying*
+v1<-viewport(width = 1, height = 1, x = 0.5, y = 0.5) #  main map
+v2<-viewport(width = 0.3, height = 0.3, x = 0.7, y = 0.2) # inset map
+print(np_footprints,vp=v1) 
+print(np_us,vp=v2)
 
-## Questions
-
-* Are there biophysical limitations on summer fires? 
-  - *e.g., maybe it can just be too green and/or too wet and/or too humid to burn*
-* Can trends in vegetation and weather explain why summer burns were successful in some years and not in others? 
-  - *If so, maybe one can predict whether a summer burn is worth trying or not, or predicting summer wildfire risk*
-* Potential implications for intentions behind growing season burns vs. pre-European summer fire regimes?
-  - *Folks recognize pre-European fire regimes included considerable summer fire, and talk about greater variability in burn severity with growing-season fire as a biodiversity-friendly reason to burnin the summer.
-But what if pre-European summer burns were actually quite intense and burned severely because they actually only occurred under drought conditions?*
-
-## Objectives
-
-* Explore patterns in greenness and humidity to determine effects on fire effectiveness
-* Identify anomalies and ranges of conditions conducive to prescribed fire
-
-# Remotely-sensed data
-
-Pulled burn severity (NBR) and several vegetation indices from Sentinel-2 imagery for 2017-2020 growing seasons. 
-Data sampled from regular gridded points within burn units
-
-Vegetation indices:
-
-* NDVI (Normalized Difference Vegetation Index)---Measure of vegetation greenness
-* NDMI (Normalized Difference Moisture Index)---Measure of moisture content of vegetation
-* MSI (Moisture Stress Index)---Another measure of moisture content sometimes used in fire risk analysis. Higher = drier
-
-## Vegetation data
-
-These data are from four *unburned* pastures in an adjacent continuous grazing treatment to remove any confounding effects of prescribed burns.
-
-### Seasonal trends
-
-Summer fires were successful in 2017 and 2020. 
-Vegetation in both years was less green and generally more dry.
-
-```{r veg_trends, fig.cap="Blue shaded areas span the first and last dates of completed burns in the spring and summer.", fig.height = 8, fig.width = 10}
-NoFireIndices %>%
-  pivot_longer(names_to = 'index', 
-               values_to = 'value', 
-               cols = c(ndvi:ndmi)) %>%
-  group_by(ImageDate, pasture, index) %>%
-    summarise(Mean = mean(value), 
-              SEM = sd(value)/sqrt(n()), 
-              .groups = 'drop') %>%
-  mutate(ImageDate = as.Date(ImageDate, '%Y-%m-%d'), 
-         Year = lubridate::year(ImageDate), 
-         Year = as.factor(Year), 
-         SampleDate = format(ImageDate, '%m-%d'), 
-         SampleDate = as.Date(SampleDate, '%m-%d'), 
-         index = toupper(index), 
-          index = fct_rev(index)) %>%
-  filter(Year != 2016) %>%
-ggplot(aes(x = SampleDate, 
-           color = Year, 
-           fill = Year, 
-           group = interaction(pasture, Year))) + theme_bw(14) +
-  geom_rect(data= seasons, inherit.aes = FALSE,
-              aes(xmin=start, xmax=end, ymin=-Inf, ymax=+Inf), 
-              fill='lightblue', alpha=0.5) +
-  geom_smooth(aes(y = Mean, group = Year),
-              se = T,
-              level = 0.99) +
-  geom_errorbar(aes(ymin = Mean - SEM, 
-                    ymax = Mean + SEM),
-                alpha = 0.5) +
-  geom_line(aes(y = Mean),
-             alpha = 0.5) +
-  labs(x = 'Image date', y = 'Index value') + 
-  facet_wrap(~index, ncol=1, scale = 'free_y') +
-  scale_x_date( date_breaks = 'month', 
-                date_labels = '%b-%d', 
-                minor_breaks = NULL) 
 ```
-
-### Seasons summarized
-
-* Greenness:
-  - NDVI during all spring burns generally pretty low
-  - Successful summer burn years higher greenness than any spring...
-  - ... still slightly but distinctly lower than unsuccessful years
-
-* Moisture content: 
-  - High variability in spring seasons across years
-  - Successful summer seasons drier than in unsuccessful years
-
-```{r veg_by_season, fig.cap="Data summarized by year from shaded blue ranges above.", fig.height = 4, fig.width = 10}
-# Get average values for burn seasons 
-  NoFireIndices %>%
-    pivot_longer(names_to = 'index', 
-                 values_to = 'value', 
-                 cols = c(ndvi:ndmi)) %>%
-    group_by(ImageDate, pasture, index) %>%
-    summarise(Median = median(value) , 
-              .groups = 'drop') %>%
-    mutate(ImageDate = as.Date(ImageDate, '%Y-%m-%d'), 
-           Year = lubridate::year(ImageDate), 
-           Year = as.factor(Year), 
-           SampleDate = format(ImageDate, '%m-%d'), 
-           SampleDate = as.Date(SampleDate, '%m-%d'), 
-           season = case_when(
-             between(SampleDate, seasons$start[1], seasons$end[1]) ~ 'Spring',
-             between(SampleDate, seasons$start[2], seasons$end[2]) ~ 'Summer', 
-             TRUE ~ NA ) ) %>%
-    filter(!is.na(season)) %>%
-    group_by(season, Year, index) %>%
-    summarize(SeasonMean = mean(Median), 
-              SeasonSEM = sd(Median)/sqrt(n()), 
-              .groups = 'drop') %>%
-    mutate(status = case_when(
-      season == 'Summer' & Year %in% c('2018', '2019') ~ "Not completed", 
-      TRUE ~ "Completed"), 
-          index = toupper(index), 
-          index = fct_rev(index)) %>%
-    ggplot(aes(x = season, y = SeasonMean, color = Year)) + theme_bw(14) +
-    geom_errorbar(aes(ymin = SeasonMean - SeasonSEM, 
-                      ymax = SeasonMean + SeasonSEM), 
-                  width = 0.25, lwd = 0.95, 
-                  position = position_dodge(width = 0.33), 
-                  key_glyph = "point")  +
-    geom_point(aes(pch = interaction(status, Year)), 
-               position = position_dodge(width = 0.33), 
-                size = 4, stroke = 2, show.legend = F) +
-    labs(x = 'Burn season', 
-         y = 'Index value', 
-         caption = 'X = Year/season in which burns were not completed') + 
-    scale_shape_manual(values = c(16, 16, 4, 16, 4, 16)) + 
-    facet_wrap(~index, ncol = 3, scale = "free") +
-    theme(axis.text.x = element_text(color = 'black')) +
-    guides(color=guide_legend(override.aes = list(pch=15, size =4))) 
-```
-
-## Burn severity
-
-Overall it appears greener fuels consistently result in less-complete burns (although 2018 was weird).
-
-Trends between fuel moisture content and severity were less consistent. 
-Again 2018 behaved strangely---fuel in 4 burn units in 1 block of pastures was extremely dry but still didn't burn with particularly high severity; less-dry fuel in the other block tended to burn slightly more severely.
-The dry block was burned a couple weeks earlier than the other block.
-
-```{r sev_veg, fig.cap="Vegetation indices pulled from closest pre-burn image, same as pre-burn NBR.", fig.height = 3.8, fig.width = 10}
-BurnIndices  %>%
-  pivot_longer(names_to = 'name', 
-               values_to = 'index', 
-               cols = c(ndvi, msi, ndmi)) %>%
-  group_by(Year, Season, fire, name) %>%
-  summarise_at(.vars = vars(dNBR, index),
-               .funs = c(Mean="mean", 
-                         SEM = function(x) {sd(x)/sqrt(n()) } )) %>%
-  ungroup() %>%
-  mutate(name = toupper(name), 
-         name = fct_rev(name)) %>% 
-  filter(Season == "Spring") %>%
-  ggplot(aes(x = index_Mean, y = dNBR_Mean, color = Year)) + theme_bw(14) +
-  geom_smooth(aes(x = index_Mean, y = dNBR_Mean), 
-              method = 'lm', 
-              color = 'black',
-              se = T, alpha = 0.2) +
-  geom_errorbar(aes(ymin = dNBR_Mean - dNBR_SEM, 
-                    ymax = dNBR_Mean + dNBR_SEM)) +
-  geom_errorbarh(aes(xmin = index_Mean - index_SEM, 
-                     xmax = index_Mean + index_SEM)) +
-  geom_point() + 
-  labs(x = 'Pre-burn index value', 
-       y = 'Burn severity (dNBR)') + 
-  geom_smooth(method = 'lm', se = F) +
-  facet_wrap(~name, scales = 'free')
-```
-
-```{r, eval = FALSE}
-full_join(by = 'fire', 
-fires %>%
-  as_tibble() %>%
-  filter(Year == '2018') %>%
-  select(fire, PreBurn),
-
-BurnIndices  %>%
-  pivot_longer(names_to = 'name', 
-               values_to = 'index', 
-               cols = c(ndvi, msi, ndmi)) %>%
-  group_by(Year, Season, fire, name) %>%
-  summarise_at(.vars = vars(dNBR, index),
-               .funs = c(Mean="mean", 
-                         SEM = function(x) {sd(x)/sqrt(n()) } )) %>%
-  ungroup() %>%
-  filter(Year == '2018') %>%
-  select(fire, name, dNBR_Mean, index_Mean) %>%
-  pivot_wider(names_from = name, 
-              values_from = index_Mean) ) %>%
-  rename(dNBR = dNBR_Mean) %>%
-  arrange(desc(msi)) 
-```
-
-# Weather data
-
-*Pending*

### 2025-04-21 NorthDakotaFireSurvey/R/PlottingSandbox.R

@@ -1,0 +1,285 @@
+{
+  pacman::p_load(tidyverse, readxl)
+pacman::p_load_gh('devanmcg/wesanderson')
+}
+
+
+keys <- read_xlsx('./Data/DigitizedData.xlsx', 'OptionKeys' )
+
+responses <- 
+  read_xlsx('./Data/DigitizedData.xlsx', 'AgreementIndices' ) %>%
+    separate(Coordinates, into = c('X', 'Y'), sep = ', ') %>%
+    mutate(across(c(X, Y), ~ as.numeric(.)), 
+           Y = round(Y, 0) ) %>%
+    pivot_wider(names_from = Endpoint, 
+                values_from = X, 
+                id_cols = -Y) %>%
+    rowwise() %>% 
+    mutate(mid = mean(c(min, max))) %>%
+    ungroup() %>% 
+    full_join(by = c('QuestionKey', 'OptionKey'), 
+              keys) 
+
+# save(responses, file = './Data/responses.Rdata')
+
+#
+# Public grazing land management and personal recreation
+#
+responses %>%
+  filter(QuestionKey == "PublicGrazingBeliefs", 
+         OptionKey != 'public_fire')  %>%
+  mutate(response = case_when(
+           OptionKey == 'increase_public_fire' ~ 'There should be more Rx fire\non public grazing lands', 
+           OptionKey == 'amount_public_fire' ~   'There is an appropriate amount of\nRx fire on public grazing lands', 
+           OptionKey == 'admin_users' ~          'Administrators should manage to meet\nuser needs or preferences', 
+           OptionKey == 'admin_science' ~        "Administrators should manage land\naccording to the best available science", 
+           OptionKey == 'admin_good' ~           'Public grazing land administrators\n do a good job managing the land' )) %>%
+  ggplot() + theme_bw(16) + 
+  geom_vline(xintercept = 0, lty = 2) + 
+  geom_errorbarh(aes(xmin = min, 
+                     xmax = max, 
+                     y = fct_reorder(response, desc(SurveyOrder))) , 
+                 color = wes_palette("FantasticFox1")[3], 
+                 height = 0.5,
+                 lwd = 0.9 ) +
+  geom_point(aes(x = mid, y = response ),
+             fill = wes_palette("FantasticFox1")[3], 
+             pch = 21,
+             size = 4, 
+             stroke = 1 ) + 
+  # coord_cartesian(xlim = c(-1.15, 1.9)) + 
[…]

### 2025-04-21 NorthDakotaFireSurvey/data/NationalGrasslandSummaries.R

@@ -1,0 +1,27 @@
+pacman::p_load(tidyverse, magrittr, sf)
+
+gp <- read_sf('./gis/GreatPlains.gpkg', 'outline')
+
+ng <- read_sf("S:/DevanMcG/GIS/SpatialData/US/USFS/NationalGrasslands/S_USA.NationalGrassland.shp") %>% 
+        st_transform(st_crs(gp)) %>% 
+        st_intersection(gp)
+allots <- read_sf('S:/DevanMcG/GIS/SpatialData/US/USFS/RangeAllotments/S_USA.Allotment.shp') %>% 
+        st_transform(st_crs(gp)) %>%
+        select(ALLOTMENT_, ALLOTMENT1, ALLOTMEN_1) %>%
+        st_intersection(gp)
+
+ng %>%
+  mutate(area = st_area(.), 
+         ha = as.numeric(area) * 0.0001) %>%
+  summarize(TotalArea = sum(ha) ) 
+
+ng %>%
+  filter(GRASSLANDN  %in% c('Cedar River National Grassland', 'Grand River National Grassland', 
+                            'Sheyenne National Grassland', 'Little Missouri National Grassland')) %>%
+  st_intersection(allots)
+  mutate(area = st_area(.), 
+         ha = as.numeric(area) * 0.0001) %>%
+  summarize(TotalArea = sum(ha) ) 
+  
+
+

### 2025-04-21 Seasonality/script/SpatialWrangling.R

@@ -39,27 +39,26 @@ gp <- read_sf('S:/DevanMcG/Projects/FuelWeatherSeasonality/seasonality.gpkg', 'S
  
  
  # Rangeland 
-   reeves_rast <- terra::rast('S:/DevanMcG/GIS/SpatialData/US/USFS/ReevesRangelands/InteriorWestRangelands.tif')
-    gp_sp <- state_polygons %>%
-               st_union() %>%
-               st_transform(5070) %>%
-               as_Spatial()
-   gp_reeves_rast <-
-     reeves_rast %>%
-       terra::crop(gp_sp ) 
-   
-   terra::plot(gp_reeves_rast)
-   
-   terra::writeRaster(gp_reeves_rast, 'S:/DevanMcG/Projects/FuelWeatherSeasonality/StudyRegionReeves.tif')
-   
-   gp_reeves_r <- raster::raster(gp_reeves_rast)
-   
-   reeves_SPDF <- as(gp_reeves_r, "SpatialPointsDataFrame")
-   reeves_SPDF %>%
-     st_as_sf() %>%
-     st_intersection(
-       st_transform(
-         state_sections, raster::crs(c_wui_SPDF)), 
-       .) 
-
+ reeves_rast <- terra::rast('S:/DevanMcG/GIS/SpatialData/US/USFS/ReevesRangelands/InteriorWestRangelands.tif')
+ gp_sp <- state_polygons %>%
+   st_union() %>%
+   st_transform(5070) %>%
+   as_Spatial()
+ gp_reeves_rast <-
+   reeves_rast %>%
+   terra::crop(gp_sp ) 
+ 
+ terra::plot(gp_reeves_rast)
+ 
+ terra::writeRaster(gp_reeves_rast, 'S:/DevanMcG/Projects/FuelWeatherSeasonality/StudyRegionReeves.tif')
+ 
+ gp_reeves_r <- raster::raster(gp_reeves_rast)
+ 
+ reeves_SPDF <- as(gp_reeves_r, "SpatialPointsDataFrame")
+ reeves_SPDF %>%
+   st_as_sf() %>%
+   st_intersection(
+     st_transform(
+       state_sections, raster::crs(c_wui_SPDF)), 
+     .) 
     

### 2025-04-21 SpringSummerSeverity/script/stats.R

@@ -1,0 +1,402 @@
+pacman::p_load(tidyverse) 
+
+# The Z-score for a data point is calculated by subtracting the historical mean from the data point and 
+# dividing by the historical standard deviation.
+# 1 or greater (or -1 or less) suggests the data point is significantly outside the historical mean
+
+
+# load data 
+load('./data/NoFireIndices.Rdata')
+load('./data/BurnIndices.Rdata')
+load('./data/NoFireTrends.Rdata')
+load('./data/WxData.Rdata')
+
+seasons <- 
+  tibble(season = c('Spring', 'Summer'), 
+         start = c(as.Date('2024-04-25'), as.Date('2024-08-07')), 
+         end = c(as.Date('2024-05-25'), as.Date('2024-08-27'))) 
+
+LandsatSums <-
+  NoFireTrends  %>%
+  separate(ImageDate, into=c('year', 'month', 'day'), sep = '-') %>%
+  mutate(season = ifelse(month=='04', 'Spring', 'Summer')) %>%
+  pivot_longer(names_to = 'index', 
+               values_to = 'value', 
+               cols = c(ndvi:ndmi)) %>%
+  group_by(year, season, pasture, index) %>%
+  summarize(Value = mean(value), 
+            .groups = 'drop') 
+
+#
+# linear regression
+#
+
+# Check distribution
+
+MASS::fitdistr(reg_dat$dNBR, "normal")
+MASS::fitdistr(reg_dat$dNBR, "Gamma")
+reg_dat <-
+  BurnIndices  %>%
+    group_by(Year, Season, fire) %>%
+    summarize(dNBR = mean(dNBR), 
+              NDVI = mean(ndvi), 
+              .groups = 'drop') %>%
+    separate(fire, into = c('block', 'pasture', 'patch'), sep = '-')
+
+ggplot(reg_dat, aes(x=dNBR)) + theme_bw(16) + 
+  geom_histogram(aes(y=after_stat(density)),      
+                 binwidth=0.05,
+                 colour="black",
+                 fill="lightgreen") + 
+  geom_density(alpha=.5, fill="lightblue")  + 
+  coord_cartesian(xlim = c(0,0.5)) + 
+  stat_function(data= tibble(dNBR=seq(0,0.5, 0.01) ), 
+                fun = dnorm, 
+                args=list(mean = 0.208,
+                          sd = 0.106),
+                colour="blue", 
+                size=1.1)  + 
[…]

### 2025-09-02 NorthDakotaBurnSeverity/paper/AllScriptTogether.Rmd
ND severity paper first draft

---
title: "Remote sensing in rangeland fire ecology: Comparing imagery to measured fire behavior, and burn severity across prescribed burns and wildfires"
subtitle:  'Supplementary Information:  script for \textsf{R} and Copernicus browser'
author: "DA McGranahan"
header-includes:
  - \usepackage{amsmath}
  - \renewcommand{\familydefault}{\sfdefault}
  - \usepackage[colorlinks=true, citecolor=blue, urlcolor=blue,linkcolor=blue, pdfborder={0 0 0}]{hyperref}
  - \urlstyle{same}
  - \usepackage{xcolor} 
  - \usepackage{booktabs}
  - \usepackage{listings}
  - \raggedright
output:
  pdf_document:
    #highlight: tango
    keep_tex: yes
    toc: true
---

```{r setup, echo=FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, warning= FALSE, message= FALSE)
```

This document is organized around the principal steps used to wrangle and analyze data in this project, and summarizes the \textsf{R} script used to perform each step. 
Script is provided for readers to understand the means by which data were sourced, assembled, and analyzed to better contextualize the results and their interpretations in the main manuscript. 
While in some cases it might serve as a resource for solutions to the data wrangling and analysis problems encountered in this project, is not presented as a linear, start-to-finish set of reproducible code. 
Some steps rely on data manipulated externally in QGIS.

# Setup

```{r packages}
# Load necessary packages via pacman utility
  pacman::p_load(tidyverse, magrittr, readxl, # data wrangling
                 foreach, doSNOW,             # parallel processing
                 sf, terra,                   # Spatial data
                 tigris,                      # Access US census data
                 climateR,                    # Fetch TerraClim data
                 lme4, emmeans       )        # statistical analysis
```

## Study region

```{r}
# bounding box for focal area of study
  region_box <- 
    tibble(feature = 'region', 
           Easting = c(-900000, -215000, -215000, -900000), 
           Northing = c(2470000, 2950000, 2470000, 2950000) ) %>%
    st_as_sf(coords = c("Easting", "Northing"), crs = 102039) %>%
    group_by(feature) %>%
    summarize(geometry = st_combine(geometry)) %>%
    st_cast("POLYGON") %>%
    st_bbox() %>%
    st_as_sfc(st_bbox(.))
# EPA Level III ecoregions
# (https://www.epa.gov/eco-research/level-iii-and-iv-ecoregions-continental-united-states)
  ngp <- 
    read_sf(local_dir, # downloaded to local directory
            "us_eco_l3_state_boundaries") %>%
      filter(NA_L3NAME %in% c('Northwestern Glaciated Plains', 
                              'Northwestern Great Plains')) %>%
      select(NA_L3NAME, STATE_NAME) %>%
      rename(L3 = NA_L3NAME, state = STATE_NAME) %>%
     st_crop(region_box)
# Get counties for the region
  ngp_counties <- 
    tigris::counties(state = c('MT', "ND", 'SD'), cb = T) %>%
    st_transform(st_crs(ngp)) %>% 
    st_intersection(ngp)
```

## Data aquisition

## Regional precipitation and anomalies

```{r}
# Create grid for region 
  ngp_grid <- 
    ngp %>%
      st_make_grid(cellsize = c(10000, 10000)) %>%
      st_intersection(ngp)  %>% 
      st_as_sf() %>%
      rowid_to_column('cell')
# PReciptation data for study period
{
  # set up for parallel processing
  cores= parallel::detectCores()
  cl <- makeCluster(cores) 
  registerDoSNOW(cl )

  begin = Sys.time()
  ngp_ppt <- 
    foreach(i=1:length(ngp_grid$cell), 
            .combine = bind_rows, 
            .errorhandling = 'remove', 
            .inorder = FALSE, 
            .packages = c('tidyverse', 'sf', 'climateR')) %dopar% {
              getTerraClim(
                AOI = ngp_grid %>%
                      slice(i) %>%
                       st_centroid() , 
                varname = 'ppt',
                startDate = '2017-01-01', 
                endDate = '2024-12-31'  ) %>% 
                as_tibble() %>%
                separate(date, into = c('year', 'month', 'day'), sep = "-") %>%
                 group_by(year) %>%
                summarize(ppt = sum(ppt_total, na.rm = TRUE)) %>%
                summarize(ppt = mean(ppt, na.rm = TRUE)) %>%
                add_column(cell = i)
            }
  stopCluster(cl)
  Sys.time() - begin 
}

  ppt_dat <- lst() 
  ppt_grd <- 
    ngp_grid  %>%
      merge(by = 'cell', 
                ngp_ppt) 
  ppt_dat$ppt_grd <- ppt_grd
# Find precip for counties with Research Extension Centers
  ppt_dat$REC_PPT <-
  ppt_grd %>%
    st_intersection(ngp_counties %>%
                      filter(STUSPS == 'ND', 
                             NAME %in% c('Stutsman', 'Kidder', 'Adams')) %>%
                      select(L3) ) %>%
  filter(st_geometry_type(., by_geometry = TRUE) %in% c('MULTIPOLYGON', 'POLYGON'))  %>%
    as_tibble() %>%
    group_by(L3) %>%
    summarize(PPT_Mean = mean(ppt), 
              PPT_SD = sd(ppt)) 
# Assign anomalies to grid
  SDs = 3 # define anomaly by standard deviation
  ppt_dat$anomaly <- 
    ppt_grd %>%
      st_intersection(ngp, .) %>%
      filter(st_geometry_type(., by_geometry=TRUE) %in% c('MULTIPOLYGON','POLYGON')) %>%
      mutate(anomaly = case_when(
        L3 == "Northwestern Glaciated Plains" ~ ppt - 447,  
        L3 == "Northwestern Great Plains" ~ ppt - 431 ), 
        anom_cat = case_when(
          L3 == "Northwestern Glaciated Plains" &
            abs(anomaly) > 12 * SDs ~ 'beyond',
          L3 == "Northwestern Glaciated Plains" & 
            abs(anomaly) < 12 * SDs ~ 'within',
          L3 == "Northwestern Great Plains" & 
            abs(anomaly) > 19 * SDs ~ 'beyond',
          L3 == "Northwestern Great Plains" & 
            abs(anomaly) < 19 * SDs ~ 'within') )

```

## Wildfire perimeters

```{r}
# Get fire perimeters for the study region by L3
  # https://data-nifc.opendata.arcgis.com/datasets/
  # nifc::interagencyfireperimeterhistory-all-years-view/
region_perims2 <-
  read_sf(local_dir, # downloaded to local directory
          "InterAgencyFirePerimeterHistory_All_Years_View") %>%
    select(FIRE_YEAR, INCIDENT, FEATURE_CA) %>%
    st_transform(st_crs(l3)) %>%
    st_make_valid(.) %>%
    st_intersection(l3) %>%
    rename(FireYear = FIRE_YEAR, FireName = INCIDENT, FireType = FEATURE_CA)

region_perims <- 
  region_perims2  %>%
  mutate(FireType = case_when(
          str_sub(FireType, 1,4)=='Wild' ~ 'Wildfire', 
          TRUE ~ 'RxFire' )) %>%
  filter(FireType == 'Wildfire', 
         between(as.numeric(FireYear), 2017, 2024), 
         state != "Wyoming") %>%
  mutate(FireName = case_when(
              FireName == 'nd-crr-fy22-wf-knutsen' ~ 'Knutsen', 
              TRUE ~ FireName )) %>%
  mutate(FireCode = str_remove_all(FireName, ' Fire'), 
         FireCode = gsub(pattern     = "[^a-z,A-Z,0-9]", 
                         replacement = "", 
                         FireCode), 
         FireCode = paste0(FireCode, '_', FireYear), 
         Ha = st_area(.), 
         Ha = as.numeric(Ha) * 0.0001) %>%
  group_by(FireCode, FireYear, FireName, FireType, L3, state) %>%
  summarize(AreaHa = sum(Ha), 
            .groups ='drop')  %>%
  group_by(FireCode) %>%
  arrange(desc(AreaHa)) %>%
  slice(1) %>% # in case fire crosses ecoregion/state lines 
  ungroup() 
# Identify wildfires within the precipitation anomaly
  win_fires <-
    region_perims %>%
      st_centroid() %>%
      st_intersection(ppt_dat$anomaly) %>%
      filter(anom_cat == 'within', 
             AreaHa > 10) %>%
    as_tibble() %>%
    select(FireCode, L3, state, anomaly)

  precip_perims <-
    region_perims %>%
    filter(FireCode %in% win_fires$FireCode)
# Get rangeland classifications 
  # Locally-saved USFS Extent of US Rangelands raster product 
    # https://apps.fs.usda.gov/arcx/rest/services/
    # RDW_LandscapeAndWildlife/Extent_of_US_Rangelands/MapServer
  rr_tr <- terra::rast(paste0(local_dir, "/USrangelands.tif")) %>%
            terra::crop(st_transform(region_perims, 5070))
# Find the proportion rangeland for each fire
  range_fires <- tibble() 
  for(i in 1:length(unique(precip_perims$FireCode))){      
    fire = unique(precip_perims$FireCode)[i]
    precip_perims %>%
      filter(FireCode == fire )%>% 
      st_transform(st_crs(rr_tr)) %>%
      terra::extract(rr_tr, .) %>%
      group_by(LABEL) %>%
      summarize(pixels = n() ) %>%
      mutate(prop = pixels / sum(pixels)) %>%
      add_column(FireCode = fire, .before = 1) %>%
      filter(LABEL == 'Rangeland') %>%
      bind_rows(range_fires) -> range_fires }
  range_fires %<>%
    mutate(RangeHa = (pixels * 900)* 0.0001) %>%
    select(-LABEL, -pixels)  %>%
    rename(PropRange = prop) 
# create comparison wildfires object 
  gp_perims <- 
    bind_rows(
    # NW Glaciated Plains
      precip_perims %>% 
        right_join(by = 'FireCode', 
                   range_fires) %>% 
        filter(L3 == 'Northwestern Glaciated Plains') %>%
        arrange(desc(PropRange)) %>%
        filter(PropRange > 0.2), 
    # NW Great Plains 
      precip_perims %>% 
        right_join(by = 'FireCode', 
                   range_fires) %>% 
        merge(by ='FireCode', 
              win_fires %>% select(FireCode, anomaly)) %>% 
        filter(L3 == 'Northwestern Great Plains' &
               RangeHa > 10) %>%
        arrange((abs(anomaly)), desc(PropRange)) %>%
        slice(1:20) ) %>%
     mutate(type = 'Wildfire', 
                zone = case_when(
                  FireCode %in% c('Swather_2021','Hwy31101STWest_2022', 'CB00121_2021', 
                                  'CoalSeamWest_2021', '1806HunkpapaCreek_2021') ~
                    'east', 
                  L3 == 'Northwestern Great Plains' ~ 'west',
                  L3 == 'Northwestern Glaciated Plains' ~ 'east'))
```

# Remotely-sensed data

## Retrieving imagery

This script was used to create and export rectangular Area of Interest (AOI) \texttt{.kml} files for each wildfire for input into Copernicus browser.

```{r}
  for(i in 1:length(unique(gp_perims$FireCode))) {
    wd = "./AOIs"
    fire = unique(gp_perims$FireCode)[i] 
    gp_perims %>%
      filter(FireCode == fire) %>%
      st_transform(4326) %>%
      st_bbox() %>%
      st_as_sfc() %>% 
      st_write(., paste0(wd, '/', fire, '.kml'), append = FALSE )}
```

This EvalScript for the Copernicus browser exports Normalized Burn Ratio (NBR) and Normalized Differenced Vegetation Index (NDVI) for the active AOI.

\begin{lstlisting}[language=Java]

//VERSION=3
function setup() {
  return {
    input: ["B04", "B08", "B12"],
    output: { bands: 2, sampleType: "UINT16" }
  };
}

function evaluatePixel(sample) {
  let nbr = index(sample.B08, sample.B12);
  let ndvi = index(sample.B08, sample.B04);
  // apply offset for UINT16 
  return [10000 * nbr + 10000, 
          10000 * ndvi + 10000]; 
}
\end{lstlisting}

## Processing imagery

```{r}
  # load perimeters after editing in QGIS
    wf_sf <- read_sf('./SpatialData.gpkg', 'GreatPlainsModifiedWF')

    wf_dir = 'Folder/With/Sentinel/Imagery'

    wf_images <- 
      tibble(file = list.files(wf_dir, pattern = "\\.tiff$", ignore.case = TRUE) ) %>% 
      mutate(info = str_remove(file, '.tiff')) %>%
      separate(info, into = c('FireCode', 'period', 'ImageDate'), sep = '_') %>%
      mutate(FireCode = str_replace(FireCode, '-', '_'))

    wildfires <- wf_sf %>% 
                filter(FireCode %in% unique(wf_images$FireCode) ) %>%
                st_transform(albersEAC) %>%
                mutate(AreaHa = st_area(.), 
                       AreaHa = as.numeric(AreaHa) * 0.0001) 
  # Sample raster imagery
   WfBurnIndices <- tibble() 
    for(i in 1:length(wildfires$FireCode)) {
      # Get fire
        fire <-  wildfires %>% 
                  filter(FireCode == unique(wildfires$FireCode)[i]) %>%
                  st_transform(st_crs(r_rr))
      # create gridded sample points
        buff <- fire  %>% st_buffer(-20)
        # cell size factor scaled to total area
          cs = case_when(
            fire$AreaHa < 10 ~ 25, 
            between(fire$AreaHa, 10, 50) ~ 36,
            between(fire$AreaHa, 50, 100) ~ 50, 
            between(fire$AreaHa, 100, 200) ~ 100, 
            between(fire$AreaHa, 200, 300) ~ 125, 
            between(fire$AreaHa, 300, 500) ~ 150, 
            between(fire$AreaHa, 500, 700) ~ 200, 
            between(fire$AreaHa, 700, 3000) ~ 300, 
            between(fire$AreaHa, 3000, 5000) ~ 500, 
            between(fire$AreaHa, 5000, 7000) ~ 700,
            fire$AreaHa > 7000 ~ 1000 )
        pts <- buff %>%
          st_make_grid(cellsize = cs, 
                       square = FALSE) %>% 
          st_centroid() %>%
          st_intersection(buff) %>%
          st_as_sf() %>%
          rowid_to_column('ID')
       # Filter sample points to rangeland cells
          r_pts <-  
            terra::extract(rr_tr, 
                          terra::vect(pts), 
                          df = TRUE) %>%
            filter(LABEL %in% c('Rangeland', 
                                'Transitional Rangeland', 
                                'Afforested CO') ) 
          pts %<>% filter(ID %in% r_pts$ID) %>%
                    st_transform(4326)
    # dNBR & NDVI
      # Fetch & process multi-band rasters
      # pre image
        pre_image = filter(wf_images, FireCode == fire$FireCode, , 
                           period == 'B')$file
        pre_path = paste0(wf_dir, '/', pre_image)
        pre_ras <- terra::rast(pre_path) %>%
                    terra::crop(terra::vect(fire %>% st_transform(4326)))
        pre_ras <-  (pre_ras-10000)/10000
      # post-fire
        post_image = filter(wf_images, FireCode == fire$FireCode, , 
                            period == 'A')$file
        post_path = paste0(wf_dir, '/', post_image)
        post_ras <- terra::rast(post_path) %>%
                    terra::crop(terra::vect(fire %>% st_transform(4326)))
        post_ras = (post_ras-10000)/10000
      # Calculate dNBR & replace in pre-fire raster
        d_ras = pre_ras[[1]]- post_ras[[1]]
        pre_ras[[1]] <- d_ras
        names(pre_ras) <- c('dNBR', 'ndvi')
      # Sample rasters
        terra::extract(pre_ras[[1:2]],
                       pts %>%
                         terra::vect(), 
                       fun = mean, 
                       method = 'bilinear',
                       bind = TRUE)  %>% 
        as_tibble() %>%
     add_column(FireCode=fire$FireCode, .before = 1 ) %>%
     bind_rows(., WfBurnIndices) -> WfBurnIndices
    }
```

# Statistical analysis

## Space-based vs. field-based data

```{r}
PlotBurnIndices <- readxl::read_xlsx('SeverityComparisonData.xlsx', 
                                     'PlotBurnIndices')
# Testing drivers of burn severity
  # Canopy temperature
    mc_0 <- lme4::lmer(dNBR ~ 1 + (1|location/burn), 
                       REML = FALSE, data = PlotBurnIndices)
    mc_1 <- lme4::lmer(dNBR ~ MaxC + (1|location/burn), 
                       REML = FALSE, data = PlotBurnIndices)
    BurnModels$Sev_v_MaxC <- anova(mc_0, mc_1)
  # Soil surface temperature
    sc_0 <- lme4::lmer(dNBR ~ 1 + (1|location/burn), REML = FALSE, 
                       data = filter(PlotBurnIndices, !is.na(SoilMaxC) ) )
    sc_1 <- lme4::lmer(dNBR ~ SoilMaxC + (1|location/burn), REML = FALSE, 
                       data = filter(PlotBurnIndices, !is.na(SoilMaxC)) )
    BurnModels$Sev_v_SoilMaxC <- anova(sc_0, sc_1)
  # Rate of spread
    rs_0 <- lme4::lmer(dNBR ~ 1 + (1|location/burn), REML = FALSE,  
                        data = filter(PlotBurnIndices, ! is.na(ros)) )
    rs_1 <- lme4::lmer(dNBR ~ ros + (1|location/burn), REML = FALSE,  
                        data = filter(PlotBurnIndices, ! is.na(ros)) )
    BurnModels$Sev_v_ROS <- anova(rs_0, rs_1)
```

## Wildfires vs. Rx burns

```{r}
reg_dat <- readxl::read_xlsx('SeverityComparisonData.xlsx', 
                                     'BurnSeverityData') %>%
            rename(dNBR = dNBR_Mean) %>%
            select(zone, type, dNBR)

    sev_0 <- lm(dNBR ~ 1, data = reg_dat)
    sev_int <- lm(dNBR ~ type*zone, data = reg_dat)

    anova(sev_0, sev_int)

    BurnModels$type_comp <- emmeans::emmeans(sev_int, ~  type | zone)

    emmeans::joint_tests(BurnModels$type_comp, by = 'zone') 
```

### 2025-09-02 NorthDakotaBurnSeverity/paper/AllScriptTogether.tex
ND severity paper first draft

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{amsmath}
\renewcommand{\familydefault}{\sfdefault}
\usepackage[colorlinks=true, citecolor=blue, urlcolor=blue,linkcolor=blue, pdfborder={0 0 0}]{hyperref}
\urlstyle{same}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{listings}
\raggedright
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Remote sensing in rangeland fire ecology: Comparing imagery to measured fire behavior, and burn severity across prescribed burns and wildfires},
  pdfauthor={DA McGranahan},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Remote sensing in rangeland fire ecology: Comparing imagery to
measured fire behavior, and burn severity across prescribed burns and
wildfires}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{Supplementary Information: script for \textsf{R} and
Copernicus browser}
\author{DA McGranahan}
\date{}

\begin{document}
\maketitle

{
\setcounter{tocdepth}{2}
\tableofcontents
}
This document is organized around the principal steps used to wrangle
and analyze data in this project, and summarizes the \textsf{R} script
used to perform each step. Script is provided for readers to understand
the means by which data were sourced, assembled, and analyzed to better
contextualize the results and their interpretations in the main
manuscript. While in some cases it might serve as a resource for
solutions to the data wrangling and analysis problems encountered in
this project, is not presented as a linear, start-to-finish set of
reproducible code. Some steps rely on data manipulated externally in
QGIS.

\section{Setup}\label{setup}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Load necessary packages via pacman utility}
\NormalTok{  pacman}\SpecialCharTok{::}\FunctionTok{p\_load}\NormalTok{(tidyverse, magrittr, readxl, }\CommentTok{\# data wrangling}
\NormalTok{                 foreach, doSNOW,             }\CommentTok{\# parallel processing}
\NormalTok{                 sf, terra,                   }\CommentTok{\# Spatial data}
\NormalTok{                 tigris,                      }\CommentTok{\# Access US census data}
\NormalTok{                 climateR,                    }\CommentTok{\# Fetch TerraClim data}
\NormalTok{                 lme4, emmeans       )        }\CommentTok{\# statistical analysis}
\end{Highlighting}
\end{Shaded}

\subsection{Study region}\label{study-region}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# bounding box for focal area of study}
\NormalTok{  region\_box }\OtherTok{\textless{}{-}} 
    \FunctionTok{tibble}\NormalTok{(}\AttributeTok{feature =} \StringTok{\textquotesingle{}region\textquotesingle{}}\NormalTok{, }
           \AttributeTok{Easting =} \FunctionTok{c}\NormalTok{(}\SpecialCharTok{{-}}\DecValTok{900000}\NormalTok{, }\SpecialCharTok{{-}}\DecValTok{215000}\NormalTok{, }\SpecialCharTok{{-}}\DecValTok{215000}\NormalTok{, }\SpecialCharTok{{-}}\DecValTok{900000}\NormalTok{), }
           \AttributeTok{Northing =} \FunctionTok{c}\NormalTok{(}\DecValTok{2470000}\NormalTok{, }\DecValTok{2950000}\NormalTok{, }\DecValTok{2470000}\NormalTok{, }\DecValTok{2950000}\NormalTok{) ) }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{st\_as\_sf}\NormalTok{(}\AttributeTok{coords =} \FunctionTok{c}\NormalTok{(}\StringTok{"Easting"}\NormalTok{, }\StringTok{"Northing"}\NormalTok{), }\AttributeTok{crs =} \DecValTok{102039}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{group\_by}\NormalTok{(feature) }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{summarize}\NormalTok{(}\AttributeTok{geometry =} \FunctionTok{st\_combine}\NormalTok{(geometry)) }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{st\_cast}\NormalTok{(}\StringTok{"POLYGON"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{st\_bbox}\NormalTok{() }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{st\_as\_sfc}\NormalTok{(}\FunctionTok{st\_bbox}\NormalTok{(.))}
\CommentTok{\# EPA Level III ecoregions}
\CommentTok{\# (https://www.epa.gov/eco{-}research/level{-}iii{-}and{-}iv{-}ecoregions{-}continental{-}united{-}states)}
\NormalTok{  ngp }\OtherTok{\textless{}{-}} 
    \FunctionTok{read\_sf}\NormalTok{(local\_dir, }\CommentTok{\# downloaded to local directory}
            \StringTok{"us\_eco\_l3\_state\_boundaries"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
      \FunctionTok{filter}\NormalTok{(NA\_L3NAME }\SpecialCharTok{\%in\%} \FunctionTok{c}\NormalTok{(}\StringTok{\textquotesingle{}Northwestern Glaciated Plains\textquotesingle{}}\NormalTok{, }
                              \StringTok{\textquotesingle{}Northwestern Great Plains\textquotesingle{}}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%}
      \FunctionTok{select}\NormalTok{(NA\_L3NAME, STATE\_NAME) }\SpecialCharTok{\%\textgreater{}\%}
      \FunctionTok{rename}\NormalTok{(}\AttributeTok{L3 =}\NormalTok{ NA\_L3NAME, }\AttributeTok{state =}\NormalTok{ STATE\_NAME) }\SpecialCharTok{\%\textgreater{}\%}
     \FunctionTok{st\_crop}\NormalTok{(region\_box)}
\CommentTok{\# Get counties for the region}
\NormalTok{  ngp\_counties }\OtherTok{\textless{}{-}} 
\NormalTok{    tigris}\SpecialCharTok{::}\FunctionTok{counties}\NormalTok{(}\AttributeTok{state =} \FunctionTok{c}\NormalTok{(}\StringTok{\textquotesingle{}MT\textquotesingle{}}\NormalTok{, }\StringTok{"ND"}\NormalTok{, }\StringTok{\textquotesingle{}SD\textquotesingle{}}\NormalTok{), }\AttributeTok{cb =}\NormalTok{ T) }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{st\_transform}\NormalTok{(}\FunctionTok{st\_crs}\NormalTok{(ngp)) }\SpecialCharTok{\%\textgreater{}\%} 
    \FunctionTok{st\_intersection}\NormalTok{(ngp)}
\end{Highlighting}
\end{Shaded}

\subsection{Data aquisition}\label{data-aquisition}

\subsection{Regional precipitation and
anomalies}\label{regional-precipitation-and-anomalies}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Create grid for region }
\NormalTok{  ngp\_grid }\OtherTok{\textless{}{-}} 
\NormalTok{    ngp }\SpecialCharTok{\%\textgreater{}\%}
      \FunctionTok{st\_make\_grid}\NormalTok{(}\AttributeTok{cellsize =} \FunctionTok{c}\NormalTok{(}\DecValTok{10000}\NormalTok{, }\DecValTok{10000}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%}
      \FunctionTok{st\_intersection}\NormalTok{(ngp)  }\SpecialCharTok{\%\textgreater{}\%} 
      \FunctionTok{st\_as\_sf}\NormalTok{() }\SpecialCharTok{\%\textgreater{}\%}
      \FunctionTok{rowid\_to\_column}\NormalTok{(}\StringTok{\textquotesingle{}cell\textquotesingle{}}\NormalTok{)}
\CommentTok{\# PReciptation data for study period}
\NormalTok{\{}
  \CommentTok{\# set up for parallel processing}
\NormalTok{  cores}\OtherTok{=}\NormalTok{ parallel}\SpecialCharTok{::}\FunctionTok{detectCores}\NormalTok{()}
\NormalTok{  cl }\OtherTok{\textless{}{-}} \FunctionTok{makeCluster}\NormalTok{(cores) }
  \FunctionTok{registerDoSNOW}\NormalTok{(cl )}

\NormalTok{  begin }\OtherTok{=} \FunctionTok{Sys.time}\NormalTok{()}
\NormalTok{  ngp\_ppt }\OtherTok{\textless{}{-}} 
    \FunctionTok{foreach}\NormalTok{(}\AttributeTok{i=}\DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(ngp\_grid}\SpecialCharTok{$}\NormalTok{cell), }
            \AttributeTok{.combine =}\NormalTok{ bind\_rows, }
            \AttributeTok{.errorhandling =} \StringTok{\textquotesingle{}remove\textquotesingle{}}\NormalTok{, }
            \AttributeTok{.inorder =} \ConstantTok{FALSE}\NormalTok{, }
            \AttributeTok{.packages =} \FunctionTok{c}\NormalTok{(}\StringTok{\textquotesingle{}tidyverse\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}sf\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}climateR\textquotesingle{}}\NormalTok{)) }\SpecialCharTok{\%dopar\%}\NormalTok{ \{}
              \FunctionTok{getTerraClim}\NormalTok{(}
                \AttributeTok{AOI =}\NormalTok{ ngp\_grid }\SpecialCharTok{\%\textgreater{}\%}
                      \FunctionTok{slice}\NormalTok{(i) }\SpecialCharTok{\%\textgreater{}\%}
                       \FunctionTok{st\_centroid}\NormalTok{() , }
                \AttributeTok{varname =} \StringTok{\textquotesingle{}ppt\textquotesingle{}}\NormalTok{,}
                \AttributeTok{startDate =} \StringTok{\textquotesingle{}2017{-}01{-}01\textquotesingle{}}\NormalTok{, }
                \AttributeTok{endDate =} \StringTok{\textquotesingle{}2024{-}12{-}31\textquotesingle{}}\NormalTok{  ) }\SpecialCharTok{\%\textgreater{}\%} 
                \FunctionTok{as\_tibble}\NormalTok{() }\SpecialCharTok{\%\textgreater{}\%}
                \FunctionTok{separate}\NormalTok{(date, }\AttributeTok{into =} \FunctionTok{c}\NormalTok{(}\StringTok{\textquotesingle{}year\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}month\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}day\textquotesingle{}}\NormalTok{), }\AttributeTok{sep =} \StringTok{"{-}"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
                 \FunctionTok{group\_by}\NormalTok{(year) }\SpecialCharTok{\%\textgreater{}\%}
                \FunctionTok{summarize}\NormalTok{(}\AttributeTok{ppt =} \FunctionTok{sum}\NormalTok{(ppt\_total, }\AttributeTok{na.rm =} \ConstantTok{TRUE}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%}
                \FunctionTok{summarize}\NormalTok{(}\AttributeTok{ppt =} \FunctionTok{mean}\NormalTok{(ppt, }\AttributeTok{na.rm =} \ConstantTok{TRUE}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%}
                \FunctionTok{add\_column}\NormalTok{(}\AttributeTok{cell =}\NormalTok{ i)}
\NormalTok{            \}}
  \FunctionTok{stopCluster}\NormalTok{(cl)}
  \FunctionTok{Sys.time}\NormalTok{() }\SpecialCharTok{{-}}\NormalTok{ begin }
\NormalTok{\}}

\NormalTok{  ppt\_dat }\OtherTok{\textless{}{-}} \FunctionTok{lst}\NormalTok{() }
\NormalTok{  ppt\_grd }\OtherTok{\textless{}{-}} 
\NormalTok{    ngp\_grid  }\SpecialCharTok{\%\textgreater{}\%}
      \FunctionTok{merge}\NormalTok{(}\AttributeTok{by =} \StringTok{\textquotesingle{}cell\textquotesingle{}}\NormalTok{, }
\NormalTok{                ngp\_ppt) }
\NormalTok{  ppt\_dat}\SpecialCharTok{$}\NormalTok{ppt\_grd }\OtherTok{\textless{}{-}}\NormalTok{ ppt\_grd}
\CommentTok{\# Find precip for counties with Research Extension Centers}
\NormalTok{  ppt\_dat}\SpecialCharTok{$}\NormalTok{REC\_PPT }\OtherTok{\textless{}{-}}
\NormalTok{  ppt\_grd }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{st\_intersection}\NormalTok{(ngp\_counties }\SpecialCharTok{\%\textgreater{}\%}
                      \FunctionTok{filter}\NormalTok{(STUSPS }\SpecialCharTok{==} \StringTok{\textquotesingle{}ND\textquotesingle{}}\NormalTok{, }
\NormalTok{                             NAME }\SpecialCharTok{\%in\%} \FunctionTok{c}\NormalTok{(}\StringTok{\textquotesingle{}Stutsman\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}Kidder\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}Adams\textquotesingle{}}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%}
                      \FunctionTok{select}\NormalTok{(L3) ) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{filter}\NormalTok{(}\FunctionTok{st\_geometry\_type}\NormalTok{(., }\AttributeTok{by\_geometry =} \ConstantTok{TRUE}\NormalTok{) }\SpecialCharTok{\%in\%} \FunctionTok{c}\NormalTok{(}\StringTok{\textquotesingle{}MULTIPOLYGON\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}POLYGON\textquotesingle{}}\NormalTok{))  }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{as\_tibble}\NormalTok{() }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{group\_by}\NormalTok{(L3) }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{summarize}\NormalTok{(}\AttributeTok{PPT\_Mean =} \FunctionTok{mean}\NormalTok{(ppt), }
              \AttributeTok{PPT\_SD =} \FunctionTok{sd}\NormalTok{(ppt)) }
\CommentTok{\# Assign anomalies to grid}
\NormalTok{  SDs }\OtherTok{=} \DecValTok{3} \CommentTok{\# define anomaly by standard deviation}
\NormalTok{  ppt\_dat}\SpecialCharTok{$}\NormalTok{anomaly }\OtherTok{\textless{}{-}} 
\NormalTok{    ppt\_grd }\SpecialCharTok{\%\textgreater{}\%}
      \FunctionTok{st\_intersection}\NormalTok{(ngp, .) }\SpecialCharTok{\%\textgreater{}\%}
      \FunctionTok{filter}\NormalTok{(}\FunctionTok{st\_geometry\_type}\NormalTok{(., }\AttributeTok{by\_geometry=}\ConstantTok{TRUE}\NormalTok{) }\SpecialCharTok{\%in\%} \FunctionTok{c}\NormalTok{(}\StringTok{\textquotesingle{}MULTIPOLYGON\textquotesingle{}}\NormalTok{,}\StringTok{\textquotesingle{}POLYGON\textquotesingle{}}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%}
      \FunctionTok{mutate}\NormalTok{(}\AttributeTok{anomaly =} \FunctionTok{case\_when}\NormalTok{(}
\NormalTok{        L3 }\SpecialCharTok{==} \StringTok{"Northwestern Glaciated Plains"} \SpecialCharTok{\textasciitilde{}}\NormalTok{ ppt }\SpecialCharTok{{-}} \DecValTok{447}\NormalTok{,  }
\NormalTok{        L3 }\SpecialCharTok{==} \StringTok{"Northwestern Great Plains"} \SpecialCharTok{\textasciitilde{}}\NormalTok{ ppt }\SpecialCharTok{{-}} \DecValTok{431}\NormalTok{ ), }
        \AttributeTok{anom\_cat =} \FunctionTok{case\_when}\NormalTok{(}
\NormalTok{          L3 }\SpecialCharTok{==} \StringTok{"Northwestern Glaciated Plains"} \SpecialCharTok{\&}
            \FunctionTok{abs}\NormalTok{(anomaly) }\SpecialCharTok{\textgreater{}} \DecValTok{12} \SpecialCharTok{*}\NormalTok{ SDs }\SpecialCharTok{\textasciitilde{}} \StringTok{\textquotesingle{}beyond\textquotesingle{}}\NormalTok{,}
\NormalTok{          L3 }\SpecialCharTok{==} \StringTok{"Northwestern Glaciated Plains"} \SpecialCharTok{\&} 
            \FunctionTok{abs}\NormalTok{(anomaly) }\SpecialCharTok{\textless{}} \DecValTok{12} \SpecialCharTok{*}\NormalTok{ SDs }\SpecialCharTok{\textasciitilde{}} \StringTok{\textquotesingle{}within\textquotesingle{}}\NormalTok{,}
\NormalTok{          L3 }\SpecialCharTok{==} \StringTok{"Northwestern Great Plains"} \SpecialCharTok{\&} 
            \FunctionTok{abs}\NormalTok{(anomaly) }\SpecialCharTok{\textgreater{}} \DecValTok{19} \SpecialCharTok{*}\NormalTok{ SDs }\SpecialCharTok{\textasciitilde{}} \StringTok{\textquotesingle{}beyond\textquotesingle{}}\NormalTok{,}
\NormalTok{          L3 }\SpecialCharTok{==} \StringTok{"Northwestern Great Plains"} \SpecialCharTok{\&} 
            \FunctionTok{abs}\NormalTok{(anomaly) }\SpecialCharTok{\textless{}} \DecValTok{19} \SpecialCharTok{*}\NormalTok{ SDs }\SpecialCharTok{\textasciitilde{}} \StringTok{\textquotesingle{}within\textquotesingle{}}\NormalTok{) ) }
\end{Highlighting}
\end{Shaded}

\subsection{Wildfire perimeters}\label{wildfire-perimeters}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Get fire perimeters for the study region by L3}
  \CommentTok{\# https://data{-}nifc.opendata.arcgis.com/datasets/}
  \CommentTok{\# nifc::interagencyfireperimeterhistory{-}all{-}years{-}view/}
\NormalTok{region\_perims2 }\OtherTok{\textless{}{-}}
  \FunctionTok{read\_sf}\NormalTok{(local\_dir, }\CommentTok{\# downloaded to local directory}
          \StringTok{"InterAgencyFirePerimeterHistory\_All\_Years\_View"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{select}\NormalTok{(FIRE\_YEAR, INCIDENT, FEATURE\_CA) }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{st\_transform}\NormalTok{(}\FunctionTok{st\_crs}\NormalTok{(l3)) }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{st\_make\_valid}\NormalTok{(.) }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{st\_intersection}\NormalTok{(l3) }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{rename}\NormalTok{(}\AttributeTok{FireYear =}\NormalTok{ FIRE\_YEAR, }\AttributeTok{FireName =}\NormalTok{ INCIDENT, }\AttributeTok{FireType =}\NormalTok{ FEATURE\_CA)}

\NormalTok{region\_perims }\OtherTok{\textless{}{-}} 
\NormalTok{  region\_perims2  }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{FireType =} \FunctionTok{case\_when}\NormalTok{(}
          \FunctionTok{str\_sub}\NormalTok{(FireType, }\DecValTok{1}\NormalTok{,}\DecValTok{4}\NormalTok{)}\SpecialCharTok{==}\StringTok{\textquotesingle{}Wild\textquotesingle{}} \SpecialCharTok{\textasciitilde{}} \StringTok{\textquotesingle{}Wildfire\textquotesingle{}}\NormalTok{, }
          \ConstantTok{TRUE} \SpecialCharTok{\textasciitilde{}} \StringTok{\textquotesingle{}RxFire\textquotesingle{}}\NormalTok{ )) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{filter}\NormalTok{(FireType }\SpecialCharTok{==} \StringTok{\textquotesingle{}Wildfire\textquotesingle{}}\NormalTok{, }
         \FunctionTok{between}\NormalTok{(}\FunctionTok{as.numeric}\NormalTok{(FireYear), }\DecValTok{2017}\NormalTok{, }\DecValTok{2024}\NormalTok{), }
\NormalTok{         state }\SpecialCharTok{!=} \StringTok{"Wyoming"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{FireName =} \FunctionTok{case\_when}\NormalTok{(}
\NormalTok{              FireName }\SpecialCharTok{==} \StringTok{\textquotesingle{}nd{-}crr{-}fy22{-}wf{-}knutsen\textquotesingle{}} \SpecialCharTok{\textasciitilde{}} \StringTok{\textquotesingle{}Knutsen\textquotesingle{}}\NormalTok{, }
              \ConstantTok{TRUE} \SpecialCharTok{\textasciitilde{}}\NormalTok{ FireName )) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{FireCode =} \FunctionTok{str\_remove\_all}\NormalTok{(FireName, }\StringTok{\textquotesingle{} Fire\textquotesingle{}}\NormalTok{), }
         \AttributeTok{FireCode =} \FunctionTok{gsub}\NormalTok{(}\AttributeTok{pattern     =} \StringTok{"[\^{}a{-}z,A{-}Z,0{-}9]"}\NormalTok{, }
                         \AttributeTok{replacement =} \StringTok{""}\NormalTok{, }
\NormalTok{                         FireCode), }
         \AttributeTok{FireCode =} \FunctionTok{paste0}\NormalTok{(FireCode, }\StringTok{\textquotesingle{}\_\textquotesingle{}}\NormalTok{, FireYear), }
         \AttributeTok{Ha =} \FunctionTok{st\_area}\NormalTok{(.), }
         \AttributeTok{Ha =} \FunctionTok{as.numeric}\NormalTok{(Ha) }\SpecialCharTok{*} \FloatTok{0.0001}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{group\_by}\NormalTok{(FireCode, FireYear, FireName, FireType, L3, state) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{summarize}\NormalTok{(}\AttributeTok{AreaHa =} \FunctionTok{sum}\NormalTok{(Ha), }
            \AttributeTok{.groups =}\StringTok{\textquotesingle{}drop\textquotesingle{}}\NormalTok{)  }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{group\_by}\NormalTok{(FireCode) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{arrange}\NormalTok{(}\FunctionTok{desc}\NormalTok{(AreaHa)) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{slice}\NormalTok{(}\DecValTok{1}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%} \CommentTok{\# in case fire crosses ecoregion/state lines }
  \FunctionTok{ungroup}\NormalTok{() }
\CommentTok{\# Identify wildfires within the precipitation anomaly}
\NormalTok{  win\_fires }\OtherTok{\textless{}{-}}
\NormalTok{    region\_perims }\SpecialCharTok{\%\textgreater{}\%}
      \FunctionTok{st\_centroid}\NormalTok{() }\SpecialCharTok{\%\textgreater{}\%}
      \FunctionTok{st\_intersection}\NormalTok{(ppt\_dat}\SpecialCharTok{$}\NormalTok{anomaly) }\SpecialCharTok{\%\textgreater{}\%}
      \FunctionTok{filter}\NormalTok{(anom\_cat }\SpecialCharTok{==} \StringTok{\textquotesingle{}within\textquotesingle{}}\NormalTok{, }
\NormalTok{             AreaHa }\SpecialCharTok{\textgreater{}} \DecValTok{10}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{as\_tibble}\NormalTok{() }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{select}\NormalTok{(FireCode, L3, state, anomaly)}

\NormalTok{  precip\_perims }\OtherTok{\textless{}{-}}
\NormalTok{    region\_perims }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{filter}\NormalTok{(FireCode }\SpecialCharTok{\%in\%}\NormalTok{ win\_fires}\SpecialCharTok{$}\NormalTok{FireCode)}
\CommentTok{\# Get rangeland classifications }
  \CommentTok{\# Locally{-}saved USFS Extent of US Rangelands raster product }
    \CommentTok{\# https://apps.fs.usda.gov/arcx/rest/services/}
    \CommentTok{\# RDW\_LandscapeAndWildlife/Extent\_of\_US\_Rangelands/MapServer}
\NormalTok{  rr\_tr }\OtherTok{\textless{}{-}}\NormalTok{ terra}\SpecialCharTok{::}\FunctionTok{rast}\NormalTok{(}\FunctionTok{paste0}\NormalTok{(local\_dir, }\StringTok{"/USrangelands.tif"}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%}
\NormalTok{            terra}\SpecialCharTok{::}\FunctionTok{crop}\NormalTok{(}\FunctionTok{st\_transform}\NormalTok{(region\_perims, }\DecValTok{5070}\NormalTok{))}
\CommentTok{\# Find the proportion rangeland for each fire}
\NormalTok{  range\_fires }\OtherTok{\textless{}{-}} \FunctionTok{tibble}\NormalTok{() }
  \ControlFlowTok{for}\NormalTok{(i }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(}\FunctionTok{unique}\NormalTok{(precip\_perims}\SpecialCharTok{$}\NormalTok{FireCode)))\{      }
\NormalTok{    fire }\OtherTok{=} \FunctionTok{unique}\NormalTok{(precip\_perims}\SpecialCharTok{$}\NormalTok{FireCode)[i]}
\NormalTok{    precip\_perims }\SpecialCharTok{\%\textgreater{}\%}
      \FunctionTok{filter}\NormalTok{(FireCode }\SpecialCharTok{==}\NormalTok{ fire )}\SpecialCharTok{\%\textgreater{}\%} 
      \FunctionTok{st\_transform}\NormalTok{(}\FunctionTok{st\_crs}\NormalTok{(rr\_tr)) }\SpecialCharTok{\%\textgreater{}\%}
\NormalTok{      terra}\SpecialCharTok{::}\FunctionTok{extract}\NormalTok{(rr\_tr, .) }\SpecialCharTok{\%\textgreater{}\%}
      \FunctionTok{group\_by}\NormalTok{(LABEL) }\SpecialCharTok{\%\textgreater{}\%}
      \FunctionTok{summarize}\NormalTok{(}\AttributeTok{pixels =} \FunctionTok{n}\NormalTok{() ) }\SpecialCharTok{\%\textgreater{}\%}
      \FunctionTok{mutate}\NormalTok{(}\AttributeTok{prop =}\NormalTok{ pixels }\SpecialCharTok{/} \FunctionTok{sum}\NormalTok{(pixels)) }\SpecialCharTok{\%\textgreater{}\%}
      \FunctionTok{add\_column}\NormalTok{(}\AttributeTok{FireCode =}\NormalTok{ fire, }\AttributeTok{.before =} \DecValTok{1}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
      \FunctionTok{filter}\NormalTok{(LABEL }\SpecialCharTok{==} \StringTok{\textquotesingle{}Rangeland\textquotesingle{}}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
      \FunctionTok{bind\_rows}\NormalTok{(range\_fires) }\OtherTok{{-}\textgreater{}}\NormalTok{ range\_fires \}}
\NormalTok{  range\_fires }\SpecialCharTok{\%\textless{}\textgreater{}\%}
    \FunctionTok{mutate}\NormalTok{(}\AttributeTok{RangeHa =}\NormalTok{ (pixels }\SpecialCharTok{*} \DecValTok{900}\NormalTok{)}\SpecialCharTok{*} \FloatTok{0.0001}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{select}\NormalTok{(}\SpecialCharTok{{-}}\NormalTok{LABEL, }\SpecialCharTok{{-}}\NormalTok{pixels)  }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{rename}\NormalTok{(}\AttributeTok{PropRange =}\NormalTok{ prop) }
\CommentTok{\# create comparison wildfires object }
\NormalTok{  gp\_perims }\OtherTok{\textless{}{-}} 
    \FunctionTok{bind\_rows}\NormalTok{(}
    \CommentTok{\# NW Glaciated Plains}
\NormalTok{      precip\_perims }\SpecialCharTok{\%\textgreater{}\%} 
        \FunctionTok{right\_join}\NormalTok{(}\AttributeTok{by =} \StringTok{\textquotesingle{}FireCode\textquotesingle{}}\NormalTok{, }
\NormalTok{                   range\_fires) }\SpecialCharTok{\%\textgreater{}\%} 
        \FunctionTok{filter}\NormalTok{(L3 }\SpecialCharTok{==} \StringTok{\textquotesingle{}Northwestern Glaciated Plains\textquotesingle{}}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
        \FunctionTok{arrange}\NormalTok{(}\FunctionTok{desc}\NormalTok{(PropRange)) }\SpecialCharTok{\%\textgreater{}\%}
        \FunctionTok{filter}\NormalTok{(PropRange }\SpecialCharTok{\textgreater{}} \FloatTok{0.2}\NormalTok{), }
    \CommentTok{\# NW Great Plains }
\NormalTok{      precip\_perims }\SpecialCharTok{\%\textgreater{}\%} 
        \FunctionTok{right\_join}\NormalTok{(}\AttributeTok{by =} \StringTok{\textquotesingle{}FireCode\textquotesingle{}}\NormalTok{, }
\NormalTok{                   range\_fires) }\SpecialCharTok{\%\textgreater{}\%} 
        \FunctionTok{merge}\NormalTok{(}\AttributeTok{by =}\StringTok{\textquotesingle{}FireCode\textquotesingle{}}\NormalTok{, }
\NormalTok{              win\_fires }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{select}\NormalTok{(FireCode, anomaly)) }\SpecialCharTok{\%\textgreater{}\%} 
        \FunctionTok{filter}\NormalTok{(L3 }\SpecialCharTok{==} \StringTok{\textquotesingle{}Northwestern Great Plains\textquotesingle{}} \SpecialCharTok{\&}
\NormalTok{               RangeHa }\SpecialCharTok{\textgreater{}} \DecValTok{10}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
        \FunctionTok{arrange}\NormalTok{((}\FunctionTok{abs}\NormalTok{(anomaly)), }\FunctionTok{desc}\NormalTok{(PropRange)) }\SpecialCharTok{\%\textgreater{}\%}
        \FunctionTok{slice}\NormalTok{(}\DecValTok{1}\SpecialCharTok{:}\DecValTok{20}\NormalTok{) ) }\SpecialCharTok{\%\textgreater{}\%}
     \FunctionTok{mutate}\NormalTok{(}\AttributeTok{type =} \StringTok{\textquotesingle{}Wildfire\textquotesingle{}}\NormalTok{, }
                \AttributeTok{zone =} \FunctionTok{case\_when}\NormalTok{(}
\NormalTok{                  FireCode }\SpecialCharTok{\%in\%} \FunctionTok{c}\NormalTok{(}\StringTok{\textquotesingle{}Swather\_2021\textquotesingle{}}\NormalTok{,}\StringTok{\textquotesingle{}Hwy31101STWest\_2022\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}CB00121\_2021\textquotesingle{}}\NormalTok{, }
                                  \StringTok{\textquotesingle{}CoalSeamWest\_2021\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}1806HunkpapaCreek\_2021\textquotesingle{}}\NormalTok{) }\SpecialCharTok{\textasciitilde{}}
                    \StringTok{\textquotesingle{}east\textquotesingle{}}\NormalTok{, }
\NormalTok{                  L3 }\SpecialCharTok{==} \StringTok{\textquotesingle{}Northwestern Great Plains\textquotesingle{}} \SpecialCharTok{\textasciitilde{}} \StringTok{\textquotesingle{}west\textquotesingle{}}\NormalTok{,}
\NormalTok{                  L3 }\SpecialCharTok{==} \StringTok{\textquotesingle{}Northwestern Glaciated Plains\textquotesingle{}} \SpecialCharTok{\textasciitilde{}} \StringTok{\textquotesingle{}east\textquotesingle{}}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\section{Remotely-sensed data}\label{remotely-sensed-data}

\subsection{Retrieving imagery}\label{retrieving-imagery}

This script was used to create and export rectangular Area of Interest
(AOI) \texttt{.kml} files for each wildfire for input into Copernicus
browser.

\begin{Shaded}
\begin{Highlighting}[]
  \ControlFlowTok{for}\NormalTok{(i }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(}\FunctionTok{unique}\NormalTok{(gp\_perims}\SpecialCharTok{$}\NormalTok{FireCode))) \{}
\NormalTok{    wd }\OtherTok{=} \StringTok{"./AOIs"}
\NormalTok{    fire }\OtherTok{=} \FunctionTok{unique}\NormalTok{(gp\_perims}\SpecialCharTok{$}\NormalTok{FireCode)[i] }
\NormalTok{    gp\_perims }\SpecialCharTok{\%\textgreater{}\%}
      \FunctionTok{filter}\NormalTok{(FireCode }\SpecialCharTok{==}\NormalTok{ fire) }\SpecialCharTok{\%\textgreater{}\%}
      \FunctionTok{st\_transform}\NormalTok{(}\DecValTok{4326}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
      \FunctionTok{st\_bbox}\NormalTok{() }\SpecialCharTok{\%\textgreater{}\%}
      \FunctionTok{st\_as\_sfc}\NormalTok{() }\SpecialCharTok{\%\textgreater{}\%} 
      \FunctionTok{st\_write}\NormalTok{(., }\FunctionTok{paste0}\NormalTok{(wd, }\StringTok{\textquotesingle{}/\textquotesingle{}}\NormalTok{, fire, }\StringTok{\textquotesingle{}.kml\textquotesingle{}}\NormalTok{), }\AttributeTok{append =} \ConstantTok{FALSE}\NormalTok{ )\}}
\end{Highlighting}
\end{Shaded}

This EvalScript for the Copernicus browser exports Normalized Burn Ratio
(NBR) and Normalized Differenced Vegetation Index (NDVI) for the active
AOI.

\begin{lstlisting}[language=Java]

//VERSION=3
function setup() {
  return {
    input: ["B04", "B08", "B12"],
    output: { bands: 2, sampleType: "UINT16" }
  };
}

function evaluatePixel(sample) {
  let nbr = index(sample.B08, sample.B12);
  let ndvi = index(sample.B08, sample.B04);
  // apply offset for UINT16 
  return [10000 * nbr + 10000, 
          10000 * ndvi + 10000]; 
}
\end{lstlisting}

\subsection{Processing imagery}\label{processing-imagery}

\begin{Shaded}
\begin{Highlighting}[]
  \CommentTok{\# load perimeters after editing in QGIS}
\NormalTok{    wf\_sf }\OtherTok{\textless{}{-}} \FunctionTok{read\_sf}\NormalTok{(}\StringTok{\textquotesingle{}./SpatialData.gpkg\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}GreatPlainsModifiedWF\textquotesingle{}}\NormalTok{)}

\NormalTok{    wf\_dir }\OtherTok{=} \StringTok{\textquotesingle{}Folder/With/Sentinel/Imagery\textquotesingle{}}

\NormalTok{    wf\_images }\OtherTok{\textless{}{-}} 
      \FunctionTok{tibble}\NormalTok{(}\AttributeTok{file =} \FunctionTok{list.files}\NormalTok{(wf\_dir, }\AttributeTok{pattern =} \StringTok{"}\SpecialCharTok{\textbackslash{}\textbackslash{}}\StringTok{.tiff$"}\NormalTok{, }\AttributeTok{ignore.case =} \ConstantTok{TRUE}\NormalTok{) ) }\SpecialCharTok{\%\textgreater{}\%} 
      \FunctionTok{mutate}\NormalTok{(}\AttributeTok{info =} \FunctionTok{str\_remove}\NormalTok{(file, }\StringTok{\textquotesingle{}.tiff\textquotesingle{}}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%}
      \FunctionTok{separate}\NormalTok{(info, }\AttributeTok{into =} \FunctionTok{c}\NormalTok{(}\StringTok{\textquotesingle{}FireCode\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}period\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}ImageDate\textquotesingle{}}\NormalTok{), }\AttributeTok{sep =} \StringTok{\textquotesingle{}\_\textquotesingle{}}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
      \FunctionTok{mutate}\NormalTok{(}\AttributeTok{FireCode =} \FunctionTok{str\_replace}\NormalTok{(FireCode, }\StringTok{\textquotesingle{}{-}\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}\_\textquotesingle{}}\NormalTok{))}

\NormalTok{    wildfires }\OtherTok{\textless{}{-}}\NormalTok{ wf\_sf }\SpecialCharTok{\%\textgreater{}\%} 
                \FunctionTok{filter}\NormalTok{(FireCode }\SpecialCharTok{\%in\%} \FunctionTok{unique}\NormalTok{(wf\_images}\SpecialCharTok{$}\NormalTok{FireCode) ) }\SpecialCharTok{\%\textgreater{}\%}
                \FunctionTok{st\_transform}\NormalTok{(albersEAC) }\SpecialCharTok{\%\textgreater{}\%}
                \FunctionTok{mutate}\NormalTok{(}\AttributeTok{AreaHa =} \FunctionTok{st\_area}\NormalTok{(.), }
                       \AttributeTok{AreaHa =} \FunctionTok{as.numeric}\NormalTok{(AreaHa) }\SpecialCharTok{*} \FloatTok{0.0001}\NormalTok{) }
  \CommentTok{\# Sample raster imagery}
\NormalTok{   WfBurnIndices }\OtherTok{\textless{}{-}} \FunctionTok{tibble}\NormalTok{() }
    \ControlFlowTok{for}\NormalTok{(i }\ControlFlowTok{in} \DecValTok{1}\SpecialCharTok{:}\FunctionTok{length}\NormalTok{(wildfires}\SpecialCharTok{$}\NormalTok{FireCode)) \{}
      \CommentTok{\# Get fire}
\NormalTok{        fire }\OtherTok{\textless{}{-}}\NormalTok{  wildfires }\SpecialCharTok{\%\textgreater{}\%} 
                  \FunctionTok{filter}\NormalTok{(FireCode }\SpecialCharTok{==} \FunctionTok{unique}\NormalTok{(wildfires}\SpecialCharTok{$}\NormalTok{FireCode)[i]) }\SpecialCharTok{\%\textgreater{}\%}
                  \FunctionTok{st\_transform}\NormalTok{(}\FunctionTok{st\_crs}\NormalTok{(r\_rr))}
      \CommentTok{\# create gridded sample points}
\NormalTok{        buff }\OtherTok{\textless{}{-}}\NormalTok{ fire  }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{st\_buffer}\NormalTok{(}\SpecialCharTok{{-}}\DecValTok{20}\NormalTok{)}
        \CommentTok{\# cell size factor scaled to total area}
\NormalTok{          cs }\OtherTok{=} \FunctionTok{case\_when}\NormalTok{(}
\NormalTok{            fire}\SpecialCharTok{$}\NormalTok{AreaHa }\SpecialCharTok{\textless{}} \DecValTok{10} \SpecialCharTok{\textasciitilde{}} \DecValTok{25}\NormalTok{, }
            \FunctionTok{between}\NormalTok{(fire}\SpecialCharTok{$}\NormalTok{AreaHa, }\DecValTok{10}\NormalTok{, }\DecValTok{50}\NormalTok{) }\SpecialCharTok{\textasciitilde{}} \DecValTok{36}\NormalTok{,}
            \FunctionTok{between}\NormalTok{(fire}\SpecialCharTok{$}\NormalTok{AreaHa, }\DecValTok{50}\NormalTok{, }\DecValTok{100}\NormalTok{) }\SpecialCharTok{\textasciitilde{}} \DecValTok{50}\NormalTok{, }
            \FunctionTok{between}\NormalTok{(fire}\SpecialCharTok{$}\NormalTok{AreaHa, }\DecValTok{100}\NormalTok{, }\DecValTok{200}\NormalTok{) }\SpecialCharTok{\textasciitilde{}} \DecValTok{100}\NormalTok{, }
            \FunctionTok{between}\NormalTok{(fire}\SpecialCharTok{$}\NormalTok{AreaHa, }\DecValTok{200}\NormalTok{, }\DecValTok{300}\NormalTok{) }\SpecialCharTok{\textasciitilde{}} \DecValTok{125}\NormalTok{, }
            \FunctionTok{between}\NormalTok{(fire}\SpecialCharTok{$}\NormalTok{AreaHa, }\DecValTok{300}\NormalTok{, }\DecValTok{500}\NormalTok{) }\SpecialCharTok{\textasciitilde{}} \DecValTok{150}\NormalTok{, }
            \FunctionTok{between}\NormalTok{(fire}\SpecialCharTok{$}\NormalTok{AreaHa, }\DecValTok{500}\NormalTok{, }\DecValTok{700}\NormalTok{) }\SpecialCharTok{\textasciitilde{}} \DecValTok{200}\NormalTok{, }
            \FunctionTok{between}\NormalTok{(fire}\SpecialCharTok{$}\NormalTok{AreaHa, }\DecValTok{700}\NormalTok{, }\DecValTok{3000}\NormalTok{) }\SpecialCharTok{\textasciitilde{}} \DecValTok{300}\NormalTok{, }
            \FunctionTok{between}\NormalTok{(fire}\SpecialCharTok{$}\NormalTok{AreaHa, }\DecValTok{3000}\NormalTok{, }\DecValTok{5000}\NormalTok{) }\SpecialCharTok{\textasciitilde{}} \DecValTok{500}\NormalTok{, }
            \FunctionTok{between}\NormalTok{(fire}\SpecialCharTok{$}\NormalTok{AreaHa, }\DecValTok{5000}\NormalTok{, }\DecValTok{7000}\NormalTok{) }\SpecialCharTok{\textasciitilde{}} \DecValTok{700}\NormalTok{,}
\NormalTok{            fire}\SpecialCharTok{$}\NormalTok{AreaHa }\SpecialCharTok{\textgreater{}} \DecValTok{7000} \SpecialCharTok{\textasciitilde{}} \DecValTok{1000}\NormalTok{ )}
\NormalTok{        pts }\OtherTok{\textless{}{-}}\NormalTok{ buff }\SpecialCharTok{\%\textgreater{}\%}
          \FunctionTok{st\_make\_grid}\NormalTok{(}\AttributeTok{cellsize =}\NormalTok{ cs, }
                       \AttributeTok{square =} \ConstantTok{FALSE}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%} 
          \FunctionTok{st\_centroid}\NormalTok{() }\SpecialCharTok{\%\textgreater{}\%}
          \FunctionTok{st\_intersection}\NormalTok{(buff) }\SpecialCharTok{\%\textgreater{}\%}
          \FunctionTok{st\_as\_sf}\NormalTok{() }\SpecialCharTok{\%\textgreater{}\%}
          \FunctionTok{rowid\_to\_column}\NormalTok{(}\StringTok{\textquotesingle{}ID\textquotesingle{}}\NormalTok{)}
       \CommentTok{\# Filter sample points to rangeland cells}
\NormalTok{          r\_pts }\OtherTok{\textless{}{-}}  
\NormalTok{            terra}\SpecialCharTok{::}\FunctionTok{extract}\NormalTok{(rr\_tr, }
\NormalTok{                          terra}\SpecialCharTok{::}\FunctionTok{vect}\NormalTok{(pts), }
                          \AttributeTok{df =} \ConstantTok{TRUE}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
            \FunctionTok{filter}\NormalTok{(LABEL }\SpecialCharTok{\%in\%} \FunctionTok{c}\NormalTok{(}\StringTok{\textquotesingle{}Rangeland\textquotesingle{}}\NormalTok{, }
                                \StringTok{\textquotesingle{}Transitional Rangeland\textquotesingle{}}\NormalTok{, }
                                \StringTok{\textquotesingle{}Afforested CO\textquotesingle{}}\NormalTok{) ) }
\NormalTok{          pts }\SpecialCharTok{\%\textless{}\textgreater{}\%} \FunctionTok{filter}\NormalTok{(ID }\SpecialCharTok{\%in\%}\NormalTok{ r\_pts}\SpecialCharTok{$}\NormalTok{ID) }\SpecialCharTok{\%\textgreater{}\%}
                    \FunctionTok{st\_transform}\NormalTok{(}\DecValTok{4326}\NormalTok{)}
    \CommentTok{\# dNBR \& NDVI}
      \CommentTok{\# Fetch \& process multi{-}band rasters}
      \CommentTok{\# pre image}
\NormalTok{        pre\_image }\OtherTok{=} \FunctionTok{filter}\NormalTok{(wf\_images, FireCode }\SpecialCharTok{==}\NormalTok{ fire}\SpecialCharTok{$}\NormalTok{FireCode, , }
\NormalTok{                           period }\SpecialCharTok{==} \StringTok{\textquotesingle{}B\textquotesingle{}}\NormalTok{)}\SpecialCharTok{$}\NormalTok{file}
\NormalTok{        pre\_path }\OtherTok{=} \FunctionTok{paste0}\NormalTok{(wf\_dir, }\StringTok{\textquotesingle{}/\textquotesingle{}}\NormalTok{, pre\_image)}
\NormalTok{        pre\_ras }\OtherTok{\textless{}{-}}\NormalTok{ terra}\SpecialCharTok{::}\FunctionTok{rast}\NormalTok{(pre\_path) }\SpecialCharTok{\%\textgreater{}\%}
\NormalTok{                    terra}\SpecialCharTok{::}\FunctionTok{crop}\NormalTok{(terra}\SpecialCharTok{::}\FunctionTok{vect}\NormalTok{(fire }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{st\_transform}\NormalTok{(}\DecValTok{4326}\NormalTok{)))}
\NormalTok{        pre\_ras }\OtherTok{\textless{}{-}}\NormalTok{  (pre\_ras}\DecValTok{{-}10000}\NormalTok{)}\SpecialCharTok{/}\DecValTok{10000}
      \CommentTok{\# post{-}fire}
\NormalTok{        post\_image }\OtherTok{=} \FunctionTok{filter}\NormalTok{(wf\_images, FireCode }\SpecialCharTok{==}\NormalTok{ fire}\SpecialCharTok{$}\NormalTok{FireCode, , }
\NormalTok{                            period }\SpecialCharTok{==} \StringTok{\textquotesingle{}A\textquotesingle{}}\NormalTok{)}\SpecialCharTok{$}\NormalTok{file}
\NormalTok{        post\_path }\OtherTok{=} \FunctionTok{paste0}\NormalTok{(wf\_dir, }\StringTok{\textquotesingle{}/\textquotesingle{}}\NormalTok{, post\_image)}
\NormalTok{        post\_ras }\OtherTok{\textless{}{-}}\NormalTok{ terra}\SpecialCharTok{::}\FunctionTok{rast}\NormalTok{(post\_path) }\SpecialCharTok{\%\textgreater{}\%}
\NormalTok{                    terra}\SpecialCharTok{::}\FunctionTok{crop}\NormalTok{(terra}\SpecialCharTok{::}\FunctionTok{vect}\NormalTok{(fire }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{st\_transform}\NormalTok{(}\DecValTok{4326}\NormalTok{)))}
\NormalTok{        post\_ras }\OtherTok{=}\NormalTok{ (post\_ras}\DecValTok{{-}10000}\NormalTok{)}\SpecialCharTok{/}\DecValTok{10000}
      \CommentTok{\# Calculate dNBR \& replace in pre{-}fire raster}
\NormalTok{        d\_ras }\OtherTok{=}\NormalTok{ pre\_ras[[}\DecValTok{1}\NormalTok{]]}\SpecialCharTok{{-}}\NormalTok{ post\_ras[[}\DecValTok{1}\NormalTok{]]}
\NormalTok{        pre\_ras[[}\DecValTok{1}\NormalTok{]] }\OtherTok{\textless{}{-}}\NormalTok{ d\_ras}
        \FunctionTok{names}\NormalTok{(pre\_ras) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{\textquotesingle{}dNBR\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}ndvi\textquotesingle{}}\NormalTok{)}
      \CommentTok{\# Sample rasters}
\NormalTok{        terra}\SpecialCharTok{::}\FunctionTok{extract}\NormalTok{(pre\_ras[[}\DecValTok{1}\SpecialCharTok{:}\DecValTok{2}\NormalTok{]],}
\NormalTok{                       pts }\SpecialCharTok{\%\textgreater{}\%}
\NormalTok{                         terra}\SpecialCharTok{::}\FunctionTok{vect}\NormalTok{(), }
                       \AttributeTok{fun =}\NormalTok{ mean, }
                       \AttributeTok{method =} \StringTok{\textquotesingle{}bilinear\textquotesingle{}}\NormalTok{,}
                       \AttributeTok{bind =} \ConstantTok{TRUE}\NormalTok{)  }\SpecialCharTok{\%\textgreater{}\%} 
        \FunctionTok{as\_tibble}\NormalTok{() }\SpecialCharTok{\%\textgreater{}\%}
     \FunctionTok{add\_column}\NormalTok{(}\AttributeTok{FireCode=}\NormalTok{fire}\SpecialCharTok{$}\NormalTok{FireCode, }\AttributeTok{.before =} \DecValTok{1}\NormalTok{ ) }\SpecialCharTok{\%\textgreater{}\%}
     \FunctionTok{bind\_rows}\NormalTok{(., WfBurnIndices) }\OtherTok{{-}\textgreater{}}\NormalTok{ WfBurnIndices}
\NormalTok{    \}}
\end{Highlighting}
\end{Shaded}

\section{Statistical analysis}\label{statistical-analysis}

\subsection{Space-based vs.~field-based
data}\label{space-based-vs.-field-based-data}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{PlotBurnIndices }\OtherTok{\textless{}{-}}\NormalTok{ readxl}\SpecialCharTok{::}\FunctionTok{read\_xlsx}\NormalTok{(}\StringTok{\textquotesingle{}SeverityComparisonData.xlsx\textquotesingle{}}\NormalTok{, }
                                     \StringTok{\textquotesingle{}PlotBurnIndices\textquotesingle{}}\NormalTok{)}
\CommentTok{\# Testing drivers of burn severity}
  \CommentTok{\# Canopy temperature}
\NormalTok{    mc\_0 }\OtherTok{\textless{}{-}}\NormalTok{ lme4}\SpecialCharTok{::}\FunctionTok{lmer}\NormalTok{(dNBR }\SpecialCharTok{\textasciitilde{}} \DecValTok{1} \SpecialCharTok{+}\NormalTok{ (}\DecValTok{1}\SpecialCharTok{|}\NormalTok{location}\SpecialCharTok{/}\NormalTok{burn), }
                       \AttributeTok{REML =} \ConstantTok{FALSE}\NormalTok{, }\AttributeTok{data =}\NormalTok{ PlotBurnIndices)}
\NormalTok{    mc\_1 }\OtherTok{\textless{}{-}}\NormalTok{ lme4}\SpecialCharTok{::}\FunctionTok{lmer}\NormalTok{(dNBR }\SpecialCharTok{\textasciitilde{}}\NormalTok{ MaxC }\SpecialCharTok{+}\NormalTok{ (}\DecValTok{1}\SpecialCharTok{|}\NormalTok{location}\SpecialCharTok{/}\NormalTok{burn), }
                       \AttributeTok{REML =} \ConstantTok{FALSE}\NormalTok{, }\AttributeTok{data =}\NormalTok{ PlotBurnIndices)}
\NormalTok{    BurnModels}\SpecialCharTok{$}\NormalTok{Sev\_v\_MaxC }\OtherTok{\textless{}{-}} \FunctionTok{anova}\NormalTok{(mc\_0, mc\_1)}
  \CommentTok{\# Soil surface temperature}
\NormalTok{    sc\_0 }\OtherTok{\textless{}{-}}\NormalTok{ lme4}\SpecialCharTok{::}\FunctionTok{lmer}\NormalTok{(dNBR }\SpecialCharTok{\textasciitilde{}} \DecValTok{1} \SpecialCharTok{+}\NormalTok{ (}\DecValTok{1}\SpecialCharTok{|}\NormalTok{location}\SpecialCharTok{/}\NormalTok{burn), }\AttributeTok{REML =} \ConstantTok{FALSE}\NormalTok{, }
                       \AttributeTok{data =} \FunctionTok{filter}\NormalTok{(PlotBurnIndices, }\SpecialCharTok{!}\FunctionTok{is.na}\NormalTok{(SoilMaxC) ) )}
\NormalTok{    sc\_1 }\OtherTok{\textless{}{-}}\NormalTok{ lme4}\SpecialCharTok{::}\FunctionTok{lmer}\NormalTok{(dNBR }\SpecialCharTok{\textasciitilde{}}\NormalTok{ SoilMaxC }\SpecialCharTok{+}\NormalTok{ (}\DecValTok{1}\SpecialCharTok{|}\NormalTok{location}\SpecialCharTok{/}\NormalTok{burn), }\AttributeTok{REML =} \ConstantTok{FALSE}\NormalTok{, }
                       \AttributeTok{data =} \FunctionTok{filter}\NormalTok{(PlotBurnIndices, }\SpecialCharTok{!}\FunctionTok{is.na}\NormalTok{(SoilMaxC)) )}
\NormalTok{    BurnModels}\SpecialCharTok{$}\NormalTok{Sev\_v\_SoilMaxC }\OtherTok{\textless{}{-}} \FunctionTok{anova}\NormalTok{(sc\_0, sc\_1)}
  \CommentTok{\# Rate of spread}
\NormalTok{    rs\_0 }\OtherTok{\textless{}{-}}\NormalTok{ lme4}\SpecialCharTok{::}\FunctionTok{lmer}\NormalTok{(dNBR }\SpecialCharTok{\textasciitilde{}} \DecValTok{1} \SpecialCharTok{+}\NormalTok{ (}\DecValTok{1}\SpecialCharTok{|}\NormalTok{location}\SpecialCharTok{/}\NormalTok{burn), }\AttributeTok{REML =} \ConstantTok{FALSE}\NormalTok{,  }
                        \AttributeTok{data =} \FunctionTok{filter}\NormalTok{(PlotBurnIndices, }\SpecialCharTok{!} \FunctionTok{is.na}\NormalTok{(ros)) )}
\NormalTok{    rs\_1 }\OtherTok{\textless{}{-}}\NormalTok{ lme4}\SpecialCharTok{::}\FunctionTok{lmer}\NormalTok{(dNBR }\SpecialCharTok{\textasciitilde{}}\NormalTok{ ros }\SpecialCharTok{+}\NormalTok{ (}\DecValTok{1}\SpecialCharTok{|}\NormalTok{location}\SpecialCharTok{/}\NormalTok{burn), }\AttributeTok{REML =} \ConstantTok{FALSE}\NormalTok{,  }
                        \AttributeTok{data =} \FunctionTok{filter}\NormalTok{(PlotBurnIndices, }\SpecialCharTok{!} \FunctionTok{is.na}\NormalTok{(ros)) )}
\NormalTok{    BurnModels}\SpecialCharTok{$}\NormalTok{Sev\_v\_ROS }\OtherTok{\textless{}{-}} \FunctionTok{anova}\NormalTok{(rs\_0, rs\_1)}
\end{Highlighting}
\end{Shaded}

\subsection{Wildfires vs.~Rx burns}\label{wildfires-vs.-rx-burns}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{reg\_dat }\OtherTok{\textless{}{-}}\NormalTok{ readxl}\SpecialCharTok{::}\FunctionTok{read\_xlsx}\NormalTok{(}\StringTok{\textquotesingle{}SeverityComparisonData.xlsx\textquotesingle{}}\NormalTok{, }
                                     \StringTok{\textquotesingle{}BurnSeverityData\textquotesingle{}}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
            \FunctionTok{rename}\NormalTok{(}\AttributeTok{dNBR =}\NormalTok{ dNBR\_Mean) }\SpecialCharTok{\%\textgreater{}\%}
            \FunctionTok{select}\NormalTok{(zone, type, dNBR)}

\NormalTok{    sev\_0 }\OtherTok{\textless{}{-}} \FunctionTok{lm}\NormalTok{(dNBR }\SpecialCharTok{\textasciitilde{}} \DecValTok{1}\NormalTok{, }\AttributeTok{data =}\NormalTok{ reg\_dat)}
\NormalTok{    sev\_int }\OtherTok{\textless{}{-}} \FunctionTok{lm}\NormalTok{(dNBR }\SpecialCharTok{\textasciitilde{}}\NormalTok{ type}\SpecialCharTok{*}\NormalTok{zone, }\AttributeTok{data =}\NormalTok{ reg\_dat)}

    \FunctionTok{anova}\NormalTok{(sev\_0, sev\_int)}

\NormalTok{    BurnModels}\SpecialCharTok{$}\NormalTok{type\_comp }\OtherTok{\textless{}{-}}\NormalTok{ emmeans}\SpecialCharTok{::}\FunctionTok{emmeans}\NormalTok{(sev\_int, }\SpecialCharTok{\textasciitilde{}}\NormalTok{  type }\SpecialCharTok{|}\NormalTok{ zone) }

\NormalTok{    emmeans}\SpecialCharTok{::}\FunctionTok{joint\_tests}\NormalTok{(BurnModels}\SpecialCharTok{$}\NormalTok{type\_comp, }\AttributeTok{by =} \StringTok{\textquotesingle{}zone\textquotesingle{}}\NormalTok{) }
\end{Highlighting}
\end{Shaded}

\end{document}

### 2025-09-02 NorthDakotaBurnSeverity/paper/FigureGenerator_SeverityFireBehavior.Rmd
ND severity paper first draft

---
title: 'Remote sensing in rangeland fire ecology: Comparing imagery to measured fire behavior, and burn severity across prescribed burns and wildfires'
subtitle:  'Figures and results'
author: "DAM"
date: "Last compiled: `r Sys.Date()`"
output:
  html_document:
    theme: united
    highlight: tango
    toc: true
    toc_float: true
---

```{r setup, echo=FALSE, message = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning= FALSE,
                      message= FALSE, eval = TRUE, 
                      fig.path = "./", dev = c('png', 'pdf')) 
pacman::p_load(tidyverse, sf, ggpattern, pander, readxl, grid, gridExtra)
pacman::p_load_gh('devanmcg/wesanderson')
```

```{r load_data, eval = TRUE}
# Mapping 
load('../../albersEAC.Rdata')
 rx_sf <- read_sf( '../data/OriginalRxBurnPerims.gpkg', 'RxLocations')
 wf_sf <- read_sf('../data/SeverityComparison.gpkg', 'GreatPlainsModifiedWF')
 ngp <- read_sf('../data/MapStuff.gpkg', 'ngp')
 ngp_counties <- read_sf('../data/MapStuff.gpkg', 'ngp_counties')
 usa <- read_sf('../data/MapStuff.gpkg', 'usa')
 gp <- read_sf('../data/MapStuff.gpkg', 'gp')
 NGP <- read_sf('../data/MapStuff.gpkg', 'NGP_states') # L3 for MT, ND, SD
 NGP_fires <- read_sf('../data/MapStuff.gpkg', 'NGP_fires')
 load('../data/ppt_dat.Rdata')

# Data 
  load('../data/BurnIndices.Rdata')
  load('../data/BurnModels.Rdata')
  load('../data/BurnSeverityData.Rdata')
# Aesthetics
  blues <- c(wes_palette('Zissou1')[2], wes_palette('Zissou1')[1])
```

# Overview

## Objectives

* Determine the relationship between *in agris* fire behavior measurements and remotely-sensed burn severity
* Compare burn severity across planned (Rx fire) and unplanned (wildfire) events
* Assess utility of fuelbed NDVI as predictor of fire behavior and burn severity

## Hypotheses

1. Positive linear relationship between *in agris* fire behavior and RS burn severity (higher intensity removes more aboveground plant material)
2. No difference in burn severity among wildfires and prescribed fires (Rx fires demonstrate a wide range of burn severity that likely overlaps that of wildfires)
3. Negative relationship between burn severity and NDVI (intensity declines with fine fuel moisture, indicated by high NDVI)

# Maps

```{r study_map}
 locations <- 
   bind_rows( 
    wf_sf %>% 
      st_transform(albersEAC) %>% 
      st_centroid() %>%
      mutate(type = 'Wildfire',
             L3 = case_when(
               FireCode %in% c('Swather_2021','Hwy31101STWest_2022', 'CB00121_2021', 
                               'CoalSeamWest_2021', '1806HunkpapaCreek_2021') ~
                 'Northwestern Glaciated Plains', 
               TRUE ~ L3), 
      zone = ifelse(L3 == "Northwestern Great Plains", 'Western', 'Eastern'), 
      zone = fct_rev(zone)) %>%
      select(type, zone), 
    rx_sf %>%
      mutate(type = 'Prescribed fire', 
             zone = ifelse(location == "cgrec", 'Eastern', 'Western')) %>%
      select(type, zone) )

 states <-  
   ngp %>%
   mutate(Area = st_area(.)) %>%
   group_by(state) %>%
   summarize(Area = sum(Area))

reg_gg <- 
  ppt_dat$ppt_grd %>%
    st_intersection(ngp, .) %>%
    filter(st_geometry_type(., by_geometry = TRUE) %in% c('MULTIPOLYGON', 'POLYGON'))  %>%
    mutate(anomaly = case_when(
      L3 == "Northwestern Glaciated Plains" ~ ppt - 447,  
      L3 == "Northwestern Great Plains" ~ ppt - 431 )) %>%
    ggplot() + theme_void() +
    geom_sf(aes(fill = anomaly), color = 'NA') +
    geom_sf_pattern(data =  ppt_dat$anom_map %>% filter(anom_cat == 'beyond'), 
                    aes(pattern = anom_cat), 
                    pattern_color = NA,
                    pattern_fill = "grey80",
                    pattern_angle = 45,
                    pattern_alpha = 0.5,
                    pattern_density = 0.5,
                    pattern_spacing = 0.025,
                    pattern_key_scale_factor = 1,, 
                    show.legend = F) +
    geom_sf(data = ngp_counties %>% 
  st_intersection(states %>% st_transform(st_crs(ngp_counties))), 
            fill = NA, color = 'grey70') + 
    geom_sf(data = states, fill = NA, color = 'black') + 
    geom_sf_text(data = states %>%
                   st_centroid(), 
                 aes(label = state),
                 fontface = 'bold',
                  color = 'black') + 
    geom_sf_label(data = locations %>% filter(type != 'Wildfire'), 
            aes(color = zone),
            label = expression(R[X]), 
            label.padding = unit(0.1, "lines"),
            label.r = unit(0.5, "lines"),
            label.size = 0.15,
            size = 5) +
    geom_sf(data = locations %>% filter(type == 'Wildfire'), 
            aes(shape = zone, color = zone), 
             stroke = 2, size = 3) +
    # geom_sf(data = locations, aes(shape = type, color = type), 
    #         fill = 'lightgreen', 
    #         stroke = 2, size = 3) +
    scale_fill_gradient2("Precipitation\nanomaly (mm)", 
                         low = wes_palette("Zissou1")[4],
                         mid = 'white',
                         high = wes_palette("Zissou1")[1], 
                         na.value = "grey70") +
    scale_pattern_manual(values = c(beyond = "circle", within = "none")) + 
    scale_shape_manual('Zone', values = c(4,3)) + 
    scale_color_manual('Fire type', values = c('darkgreen', 'darkred'), guide = 'none')

us_gg <- 
  ggplot() + theme_void() + 
 # geom_sf(data = usa %>% st_buffer(10000), color = NA, fill = 'white')  + 
  geom_sf(data = usa, color = 'white', fill = 'grey60')  + 
  geom_sf(data = gp, fill = 'white', 
          color =  'black', 
          alpha = 0.5, size = 2) +
  geom_sf(data = ngp, 
          fill =  wes_palette('Zissou1')[5],  
          color = NA, 
          show.legend = F) +
  scale_fill_manual('EPA Level III\necoregion', 
                    values = c("#74A089","#E1BD6D"))

grid.newpage() 
main <- viewport(width = 1, height = 0.95, x = 0.5, y = 0.5) 
inset <- viewport(width = 0.3, height = 0.4, x = 0.725, y = 0.7)
print(reg_gg, vp = main)
print(us_gg, vp = inset)  
```

```{r region_map}
reg_gg <-
ppt_dat$ngp_ppt %>%
  mutate(NA_L3NAME = str_replace(NA_L3NAME, ' ', '\n')) %>%
ggplot() + theme_void() +
  geom_sf(aes(fill = NA_L3NAME, alpha = ppt), 
          color = 'NA') +
  geom_sf(data = NGP, 
          fill = NA) + 
   geom_sf_label(data = locations %>% 
                  filter(type != 'Wildfire'), 
                label = expression(R[X]), 
                label.padding = unit(0.1, "lines"),
                label.r = unit(0.5, "lines"),
                label.size = 0.15,
                size = 4) +
  geom_sf(data = NGP_fires, 
          pch = 1, alpha = 0.5) + 
  geom_sf_label(data = NGP %>%
                  group_by(STATE_NAME) %>%
                  summarize(area = sum(Shape_Area)) , 
                aes(label = state.abb[match(STATE_NAME,state.name)]), 
                fontface='bold') + 
  scale_fill_manual('EPA Level III\necoregion', 
                    values = c("#74A089","#E1BD6D")) +
  scale_alpha_continuous("Precipitation\nmean annual (mm)", 
                         range = c(0.1, 1)) + 
  theme(legend.position = 'right', 
        legend.key.height = unit(1, 'cm')) + 
  guides(alpha = guide_legend(order = 2), fill = guide_legend(order = 1) )

us_gg <- 
  ggplot() + theme_void() + 
  geom_sf(data = usa %>% st_buffer(10000), color = NA, fill = 'white')  + 
  geom_sf(data = usa, color = 'white', fill = 'grey60')  + 
  geom_sf(data = gp, fill = 'white', 
          color =  'black', 
          alpha = 0.5, size = 2) +
  geom_sf(data = NGP, 
          fill =  wes_palette('Zissou1')[5],  
          color = NA, 
          show.legend = F) +
  scale_fill_manual('EPA Level III\necoregion', 
                    values = c("#74A089","#E1BD6D"))

grid.newpage() 
main <- viewport(width = 1, height = 0.95, x = 0.5, y = 0.5) 
inset <- viewport(width = 0.3, height = 0.4, x = 0.25, y = 0.25)
print(reg_gg, vp = main)
print(us_gg, vp = inset)
```

```{r}
NGP_fires %>% 
  as_tibble() %>%
  group_by(STATE_NAME, NA_L3NAME) %>% 
  summarize(Fires = n()) %>%
  pivot_wider(names_from = NA_L3NAME, 
              values_from = Fires) %>%
   mutate(Total = `Northwestern Great Plains` + `Northwestern Glaciated Plains`) %>%
  #xtable::xtable() 
  pander::pander("Fire incidents within each LIII ecoregion by state.")
```

# Fire behavior and burn severity

## Burn severity as response

```{r dNBR_v_FB, fig.width = 8, fig.height = 3, fig.cap = " "}
  BurnIndices %>%
  filter(dNBR > 0) %>%
    mutate(ros = ifelse(ros >= 13, NA, ros)) %>%
    pivot_longer(names_to = 'behavior', 
                 values_to = 'value', 
                 cols = c(MaxC:ros)) %>%
    mutate(behavior = fct_relevel(behavior, c('MaxC', 'SoilMaxC', 'ros')), 
           Behavior = recode(behavior, 
                        ros = 'Spread rate (m/min)', 
                        SoilMaxC = 'Soil surface temp (°C)',
                        MaxC = 'Flame temp (°C)') )   %>%

    ggplot(aes(x = value, y = dNBR)) + theme_bw(16) +
    geom_smooth(data = . %>%
                  filter(behavior != 'SoilMaxC'), 
                method = 'lm', 
                color = blues[1],
                fill = blues[2] ) +
    geom_smooth(data = . %>%
                  filter(behavior == 'SoilMaxC'), 
                method = 'lm', 
                lty = 2, se = F, 
                color = blues[1]) +
    geom_point(pch = 21, size =2, stroke = 1,
               color = 'grey20', fill=blues[1]  ) +
  labs(x = '', y = 'Burn severity (ΔNBR)') + 
    facet_wrap(~Behavior, scales = 'free_x') +
  theme(strip.text = element_text(face = 'bold'), 
        plot.margin=margin(c(3,10,-10,3), 'em'))
```

```{r }
      bind_rows(
      BurnModels$Sev_v_MaxC[2,c(6,8)] %>% 
        tibble() %>%
        add_column(Predictor = 'Flame temperature', .before = 1), 
      BurnModels$Sev_v_SoilMaxC[2,c(6,8)] %>% 
        tibble() %>%
        add_column(Predictor = 'Soil surface temperature', .before = 1), 
      BurnModels$Sev_v_ROS[2,c(6,8)] %>% 
        tibble() %>%
        add_column(Predictor = 'Rate of spread', .before = 1) ) %>%
        mutate(Chisq = signif(Chisq, 3), 
               Chisq = round(Chisq, 1)) %>%
        arrange(desc(Chisq)) %>%
      rename('χ$^2$' = Chisq, 
            P = `Pr(>Chisq)` ) %>%
      mutate(P = case_when(
              between(P,0.0009, 0.01) ~ '< 0.01',
              P <= 0.0009 ~ '< 0.001', 
              P >= 0.05 ~ '> 0.05', 
              TRUE ~ as.character(round(P, 2) ) ))  %>%
        pander::pander( )
```

## Fuelbed greenness as the predictor

```{r FB_v_ndvi, fig.cap = " "}
  BurnIndices %>%
   filter(dNBR > 0) %>%
    mutate(ros = ifelse(ros >= 13, NA, ros)) %>%
    pivot_longer(names_to = 'response', 
                 values_to = 'value', 
                 cols = c(MaxC:dNBR)) %>%
    mutate(response = fct_relevel(response, c('MaxC', 'SoilMaxC', 'ros', 'dNBR')), 
           Response = recode(response, 
                        ros = 'Spread rate (m/min)', 
                        SoilMaxC = 'Soil surface temp (°C)',
                        MaxC = 'Flame temp (°C)', 
                        dNBR = 'Burn severity (ΔNBR)') )   %>%

    ggplot(aes(x = ndvi, y = value)) + theme_bw(16) +
    geom_smooth(data = . %>%
                  filter(response != 'SoilMaxC'), 
                method = 'lm', 
                color = blues[1],
                fill = blues[2] ) +
    geom_smooth(data = . %>%
                  filter(response == 'SoilMaxC'), 
                method = 'lm', 
                lty = 2, se = F, 
                color = blues[1]) +
    geom_point(pch = 21, size =2, stroke = 1,
               color = 'grey20', fill=blues[1]  ) +
  labs(x = 'Fuelbed greenness (NDVI)', y = '' ) + 
    facet_wrap(~Response, scales = 'free_y') +
  theme(strip.text = element_text(face = 'bold'), 
        plot.margin=margin(c(2,2,2,-10), 'mm'))
```

```{r}
      bind_rows(
        BurnModels$MaxC_v_NDVI[2,c(6,8)] %>% 
          tibble() %>%
          add_column(Response = 'Flame temperature', .before = 1), 
        BurnModels$SoilMaxC_v_NDVI[2,c(6,8)] %>% 
          tibble() %>%
          add_column(Response = 'Soil surface temperature', .before = 1), 
        BurnModels$ROS_v_NDVI[2,c(6,8)] %>% 
          tibble() %>%
          add_column(Response = 'Rate of spread', .before = 1), 
        BurnModels$dNBR_v_NDVI[2,c(6,8)] %>% 
          tibble() %>%
          add_column(Response = 'Burn severity', .before = 1)  ) %>%
        mutate(Chisq = signif(Chisq, 3), 
               Chisq = round(Chisq, 1)) %>%
        arrange(desc(Chisq)) %>%
        rename('χ$^2$' = Chisq, 
               P = `Pr(>Chisq)` ) %>%
        mutate(P = case_when(
          between(P,0.0009, 0.01) ~ '< 0.01',
          P <= 0.0009 ~ '< 0.001', 
          P >= 0.05 ~ '> 0.05', 
          TRUE ~ as.character(round(P, 2) ) ))  %>%
        pander::pander( )
```

# Wildfires vs. Rx burns

```{r}
BurnSeverityData %>%
  group_by(zone, type) %>%
  summarize(fires = n() ) %>%
  pivot_wider(names_from = zone, 
              values_from = fires) %>%
  mutate(Total = east + west) %>%
  pander::pander("Incidents used in this study.")
```

## Burn severity across zones and fire types

```{r WF_v_Rx, fig.width = 8, fig.height = 4, fig.cap = " "}
    BurnSeverityData  %>% 
      mutate(zone = fct_rev(zone), 
             zone = paste0(str_to_title(zone), 'ern zone'),
             type = fct_relevel(type, c("Wildfire",'Rx' )), 
             type = recode(type, 'Rx'= "Prescribed\nburn")) %>%
      ggplot(aes(x= type, y = dNBR_Mean)) + theme_bw(16) +
       geom_boxplot(varwidth = TRUE, 
                    fill = 'lightblue', 
                    staplewidth = 0.25) +
       geom_point(data = . %>%
                   group_by(type, zone) %>%
                   summarize(Mean = mean(dNBR_Mean)), 
                 aes(y = Mean), 
                 pch = 24, size = 4, 
                 color = 'white', fill = 'darkblue') + 
      labs(x = 'Fire type', 
           y = 'Burn severity (ΔNBR)') +           
      facet_wrap(~zone, scales = 'free_x') +
      theme(panel.grid.major.x = element_blank(), 
            axis.text.x = element_text(color = 'black'))
```

```{r wf_rx_int}
  emmeans::joint_tests(BurnModels$type_comp) %>%
    tibble() %>%
    rename(`F` = F.ratio, P = p.value) %>%
    mutate(`F` = signif(F, 2),
           P = case_when(
             between(P,0.0009, 0.01) ~ '< 0.01',
             P <= 0.0009 ~ '< 0.001', 
             P >= 0.05 ~ '> 0.05', 
             TRUE ~ as.character(round(P, 2) ) ) )  %>%
  pander::pander('emmeans test indicates significant type x zone interaction.')
```

```{r wf_rx_comp}
  emmeans::joint_tests(BurnModels$type_comp, by = 'zone')  %>%
    tibble() %>%
    rename(`F` = F.ratio, P = p.value) %>%
    mutate(`F` = signif(F, 2),
           P = case_when(
             between(P,0.0009, 0.01) ~ '< 0.01',
             P <= 0.0009 ~ '< 0.001', 
             P >= 0.05 ~ '> 0.05', 
             TRUE ~ as.character(round(P, 2) ) ) )  %>%
  pander::pander('emmeans test on contrasts within the interactive linear model indicates significant difference between burn severity in wildfires vs prescribed burns in the eastern zone, but not in the western zone.')
```

### 2025-09-02 NorthDakotaBurnSeverity/paper/RemoteSensingRangeFireEcology.tex
ND severity paper first draft

\documentclass[fire,article,submit,oneauthor,pdftex]{Definitions/mdpi}

% MDPI internal commands
\firstpage{1} 
\makeatletter 
\setcounter{page}{\@firstpage} 
\makeatother
\pubvolume{1}
\issuenum{1}
\articlenumber{0}
\pubyear{2022}
\copyrightyear{2022}
\externaleditor{Academic Editor: Firstname Lastname} 
\datereceived{} 
\dateaccepted{} 
\datepublished{} 
\hreflink{https://doi.org/} % If needed use \linebreak

% Additional packages
\usepackage{longtable,booktabs}
\usepackage[utf8]{inputenc}

\usepackage{lmodern}

%=================================================================
% Full title of the paper (Capitalized)
\Title{Remote sensing in rangeland fire ecology: Comparing imagery to measured fire behavior, and burn severity across prescribed burns and wildfires}

% MDPI internal command: Title for citation in the left column
\TitleCitation{Remote sensing in rangeland fire ecology}

\newcommand{\orcidauthorA}{0000-0002-3763-7641}

% Authors, for the paper (add full first names)
\Author{Devan Allen McGranahan $^{1,\dagger,}$\orcidA{} }

\firstnote{The US Department of Agriculture is an equal opportunity lender, provider, and employer. }

% Authors, for metadata in PDF
\AuthorCitation{McGranahan, D.A.}

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address{%
$^{1}$ \quad USDA Agricultural Research Service, Livestock \& Range Research Laboratory, Miles City, MT 59301 USA; Devan.McGranahan@usda.gov}

% Contact information of the corresponding author
\corres{Correspondence: \href{mailto:Devan.McGranahan@usda.gov}{\nolinkurl{Devan.McGranahan@usda.gov}}}

% Abstract (Do not use inserted blank lines, i.e. \\)
\abstract{Wildland fire scientists have made substantial advances in measuring fire behavior, but properly collecting data is often beyond the capacity of prescribed fire managers and by definition all but impossible for wildfire events. 
	While a method for the immediate assessment of burn severity has been developed around multispectral imagery from space-based Earth observation systems, there has been little comparison of these post-hoc metrics to actual fire behavior.
	Meanwhile, the application of research results from experimental prescribed burns to rangeland affected by wildfire can be impeded by a lack of understanding of how immediate burn severity  differs between wildfires and prescribed burns, especially in rangelands. 
	Overall, much of what is known about wildland fire behavior, severity, and effects comes from forests, whereas rangelands are characterized by having lower fuel loads comprised of fine vegetation that promotes high rates of spread and brief residence time. 
	This paper provides rangeland-specific information on the relationships between direct field-based fire behavior measurements and a space-based index of burn severity (differenced Normalized Burn Ratio, $\Delta$NBR, from Sentinel-2 imagery), and use those data to compare burn severity across 54 prescribed burns in North Dakota, USA, and 29 nearby wildfires in the US Northern Great Plains.
 	In prescribed burns, remotely-sensed burn severity increased with rate of spread and flame temperature 15 cm above the ground, but had no statistically-significant relationship with soil surface temperature. 
 	In the semi-arid western zone of the Northern Great Plains, wildfires and prescribed burns had similar, low-moderate severity; wildfires in the eastern zone tended to be moderately-high severity and thus greater than the low severity of the experimental prescribed burns. 
	 By describing meaningful gradients in surface fire behavior in rangelands with $\Delta$NBR, even those without the capacity to measure fire behavior in the field can monitor prescribed fire effectiveness and incorporate burn severity in adaptive management plans. 
	Understanding the relationship between burn severity across wildfires and prescribed burns is a critical step in applying knowledge gained from research on prescribed fires to areas impacted by wildfire. 
	Resistance to prescribed burning might be overcome by increasing livestock managers' experience with post-fire forage resources through grazing areas burned in unintentional wildfires, but current practice and policy dissuades or outright prevents ranchers from doing so.
	Future research ought to connect burn severity with ecosystem recovery metrics to ensure post-fire grazing does not impair rangeland sustainability. }

% Keywords
\keyword{Differenced Normalized Burn Ratio (dNBR), Rangeland fire ecology and management, Rate of spread, Thermocouple dataloggers, US Northern Great Plains, Wildland fire behavior }

\begin{document}

\section{Introduction}

Given the ubiquity of fire around the Earth and throughout its history, it is of little wonder that fire has had a primary role in human understandings of nature, shaping both the content and methods of natural philosophy and scientific investigations. 
From the four natural elements of Empedocles and Aristotle to Faraday's \emph{The Chemical History of a Candle}, the demystification of fire is emblematic of how Western civilization advanced from mythological to empirical ways of knowing \cite{pyne2016}. 
However, despite the efforts of reductionism, the scientific method, and controlled experimentation, a complete understanding of \emph{wildland fire}\textemdash combustion in its natural habitat, openly burning through vegetation\textemdash has remained elusive. 
While the essential contributions of each component in the fire triangle\textemdash heat, oxygen, and fuel\textemdash are broadly understood and can be tightly controlled in closed industrial environments, the degree of nuance in their dynamics in nature led Sullivan \cite[][p. 133]{sullivan2017} to describe wildland fires as ``a complex combination of highly chaotic chemical reactions and physical processes that continually and freely propagate through spatially-variable biomass fuels across variable terrain influenced by spatially and temporally varying atmospheric conditions.''

Wildland fire scientists use the term \emph{fire behavior} to describe energy release by the combustion of vegetation in terms of how fast fire moves, how much heat is released, and how long burning lasts \cite{mcgranahan2021a, finney2021}. 
Some of these are immediately observable\textemdash one can measure rate of spread simply by timing the passage of a flame front between points of a known distance apart; the length of a flame is related to energy release \cite{rothermel1980}. 
But each of these metrics require direct observation of the flame front, which is often precluded by smoke, lack of access, or the general danger to personnel posed by fire moving freely about. 
And while post-burn observations such as scorch height on trees can be used to infer intensity after the fact \cite{rothermel1980}, such methods have limited value in the rangeland ecosystems that cover approximately 54\% of the Earth's terrestrial surface and are characterized by having few if any trees \cite{ilri2021}.

\subsection{Direct measurements of fire behavior}

Several methods directly measure fire behavior beyond human line of sight. 
As a prominent property of fire is heat, wildland fire scientists have long tried to measure fire behavior as temperature.
For at least 75 years, measuring temperature in wildland fires has predominantly relied upon \emph{thermocouples}\textemdash pairs of two wires, composed of different metals, each of which transmit slightly different electric signals when their shared junction is heated; at the other end of the cable, differences in electrical properties across the two wires can be detected and converted to temperature \cite{vaartaja1949, engle1989,mcgranahan2020a, pavlasek2015, shannon2003}.

Despite their ubiquity in wildland fire research, thermocouple systems have substantial limitations. 
Firstly, the physical limitations: \emph{e.g.}, response lags as the thermocouple material itself gains and loses heat from the material with which it is in contact, and  underestimations due to the fact the junction is losing heat to the ambient environment as well as absorbing it \cite{walker1968,lemaire2017}.  These errors are greatest when the temperature discrepancy between the junction and the media is greatest \cite{blevins1999}, which is exactly when flame fronts arrive at the sensor.
Finally, the junction retains heat even after the flame front passes and continues to register elevated temperatures; this post-fontal period is often erroneously interpreted as ``residual heating'' even when the media the junction is in contact with has returned to ambient temperature \cite{mcgranahan2020a}.

\begin{figure}[t]
	\centering
	\includegraphics[width=1\columnwidth]{FeatherFlame.pdf}
	\caption{The FeatherFlame is an example of a thermocouple-based system designed to measure two-dimensional rate of spread, shown here deployed before a prescribed fire (left) and after the fire had passed (right). 
		The 1 m equilateral triangle holds three K-type thermocouples 15 cm above the soil surface. 
		Using the timestamps associated with flame temperatures recorded by the Arduino-based datalogger allows calculation of rate of spread through the array. 
		A fourth thermocouple is placed under litter at the soil surface. 
		The cylindrical galvanized cap protects the datalogger from heat, allowing \emph{in agris} fire behavior measurements.
	The system on the left is shown deployed with optional passive flame height sensors following \cite{finney1992}.   }
	\label{Fig:FeatherFlame} % Fig.~\ref{Fig:FeatherFlame}
\end{figure}

Secondly, even accurately measured temperature is a poor metric of fire behavior. 
Sensor placement has an outsized impact on temperature readings: the temperature of a single flame varies with distance from its base, and the vertical distribution of fuel can effect differences in the heating profile as great as whether temperatures increase or decrease with height away from the ground. 
Furthermore, recent work has called into question the decades-old concept that organisms have specific temperature mortality thresholds \cite{smith2016, pingree2019}; better is to quantify the total exposure of tissue to heat energy over time \cite{wright1970, dickinson2004, bova2005}.
Thus, Alexander \cite[][p. 350]{alexander1982} concluded, ``Measuring temperature merely introduces a cumbersome, secondary step in the study of fire effects when direct linkage to one or more fire-behavior characteristics would be more profitable,'' and Bova \& Dickinson \cite{bova2008} urge fire ecologists to move ``beyond fire temperatures.''

For many wildland fire scientists, rate of spread has become the most reliable metric of fire behavior\textemdash conceptually simple; relatively easy to observe, measure, and report; and a step towards calculating intensity, as an input in the Byram equation \cite{byram1959}. 
Rate of spread isn't perfect\textemdash it is affected by each component of the fire triangle but not singularly attributable to any, meaning several combinations of variables in the wildland fire environment can produce the same rate of spread \cite{finney2021}. 
But rate of spread is a common output in fire behavior prediction models \cite{perry1998, cruz2018}, particularly of head fires spreading with the wind. 
While such one-dimensional rate of spread is theoretically relatively easy to measure, especially during prescribed burns\textemdash stakes can be pre-positioned and a stopwatch used to time flame front travel between them\textemdash in reality, these measurements are challenged first by needing to ensure stakes are positioned perpendicular to the flame front, and second by needing to be within visual contact of the fuelbed ahead of the advancing fire.

Alternatively, measuring fire spread through a two-dimensional array of sensors has two advantages: it can be calculated regardless of the direction of spread, and does not need a human present \cite{finney2021}.
Thermocouple dataloggers can be used, but with the twist that the information of most value is the timestamp recorded when the flame front hits each sensor in the array: trigonometry can be applied to these arrival times to calculate the average speed at which the flame front moved through the array \cite[][Fig.~\ref{Fig:FeatherFlame}]{simard1982, simard1984, clements2019, mcgranahan2021, mcgranahan2023}.

\subsection{Remote sensing in fire ecology}

Beyond field-based measurements, space-based earth observation systems are increasingly being applied in wildland fire science \cite{szpakowski2019}. 
Fire ecologists have made particular use of the Normalized Burn Ratio (NBR), a multispectral product derived from Near-Infrared and Shortwave Infrared wavelengths, which have specific responses to reflectance changes caused by burning \cite{key2006a}. 
Changes effected by a specific fire event can be isolated by subtracting, or \emph{differencing}, post-fire NBR values from NBR values of the same point prior to the fire, creating a metric called Differenced Normalized Burn Ratio ($\Delta$NBR; Fig.~\ref{Fig:NBRexample}).

\begin{figure}[t]
	\centering
	%\includegraphics[width=1\columnwidth, trim={30 30 340 15},clip]	{example_wide.png} % left bottom right top
	\includegraphics[width=0.72\columnwidth, trim={10 0 0 0},clip]{nbr_gg.png}~
	\includegraphics[width=0.28\columnwidth, trim={150 0 150 0},clip]{dnbr_gg.png}

	\caption{An example of a wildland fire from multipsectral imagery; 	see Fig.~\ref{Fig:TrueColor} for true color imagery, including a capture of the fire while it was burning.
		The longest axis of this fire extended 12 km (7.5 miles).
				\emph{Left:} The Normalized Burn Ratio for Before and After scenes from a fire in North Dakota, derived from Sentinel-2 Near-Infrared and Shortwave Infrared sensors.
		\emph{Right:} Burn severity is calculated as the difference between the Before and After NBR (Differenced NBR, or $\Delta$NBR), a measure of how much aboveground material the fire removed via combustion.
 }
	\label{Fig:NBRexample} % Fig.~\ref{Fig:NBRexample}
\end{figure}

Fire ecologists use $\Delta$NBR to estimate burn severity, defined in the context of interpreting remotely-sensed products as ``the degree of environmental change caused by fire \cite[][p. LA-6]{key2006a}.''
Temporal scale is a very important factor in the definition and interpretation of a $\Delta$NBR value. 
Generally speaking, burn severity assessments can be divided into \emph{extended} and \emph{immediate} \cite{key2006a,veraverbeke2010}.
The Monitoring Trends in Burn Severity program uses both imagery from immediately after a fire, to assess first-order effects on soil and vegetation consumption, and from the next growing season, to identify overall mortality to tree canopies in forests \cite{eidenshink2007a}.
The temporal distinction is particularly important in grasslands, where fire might remove all aboveground plant biomass down to mineral soil\textemdash by most definitions, high severity in an immediate assessment\textemdash but herbaceous plants have typically regrown by the end of the next growing season, which an extended assessment would classify as low severity \cite{lentile2006}.

\subsection{Fire management in the 21\textsuperscript{st} century}

European colonization altered fire regimes around the world through top-down, command-and-control management that viewed wildland fire as destructive\textemdash and ironically, might have exacerbated the issue. 
As in many European and post-European nations, the United States had tight control over wildland fire by the middle of the 20\textsuperscript{th} century by suppressing unintentional ignitions and excluding intentional ignitions, especially cultural burning \cite{mcgranahan2021a, donovan2007}. 
But several ecosystems have been diagnosed with fire deficits in which the decades-long elimination of fire has had negative ecological effects and led to a buildup of vegetation biomass that drives high-intensity fires that are often uncharacteristically severe and difficult to control \cite{parks2015,kolden2019}. 
Management interventions often focus on fuel reduction through mechanical thinning or prescribed burning; conducted under less extreme conditions than wildfires typically occur, these ``good'' fires consume hazardous fuels without effecting high severity ot the rest of the ecosystem, and have shown varying degrees of effectiveness in reducing the severity of subsequent wildfires \cite{fernandes2003, brodie2024}.

While the fire deficit affects most ecosystems in the United States, the prevalence of the ``good fire/bad fire'' duality has its origins in forests, with little critical examination of how or even whether it applies to rangelands. 
The colonial vanguard described a substantial amount of fire throughout the US Great Plains \cite{courtwright2007}, including the Northern Great Plains, specifically \cite{higgins1984,umbanhowar1996}. 
But colonization forced a transition from Indigenous fire use and native ungulates to a commercial ranching industry that relied upon grass being fed to cattle, not destroyed by fires \cite{courtwright2007}. 
Although this ethos is present today, especially in North Dakota \cite{boland2025}, a growing body of work demonstrates the benefits of prescribed fire for livestock production \cite{wanchuk2024, spiess2020, wanchuk2024a}. 
Even without active prescribed fire programs, a substantial portion of the Northwestern Great Plains already has a high incidence of fire\textemdash the ecoregion has one of the highest densities of wildfires in the US \cite{mcgranahan2024}. 
But there is a paucity of information on whether burn severity differs between prescribed burns and wildfires in the Northern Plains.

\subsection{Objectives and hypotheses}

The intent of this paper is two-fold: to provide rangeland-specific information on the relationships between direct field-based fire behavior measurements and space-based measurements of burn severity, and use those data to compare burn severity across prescribed burns and wildfires. 
Connecting remotely-sensed data to direct measurements helps calibrate assessments and monitoring programs that are unable to collect field-based fire behavior data but want insight into whether variation in remotely-sensed products has a meaningful relationship with ecological processes on the ground. 
Understanding how burn severity differs between prescribed burns and wildfires informs the transferability of ecological data on fire effects from burning experiments to broader landscapes affected by wildfire.

The specific hypotheses tested here include: 
\begin{enumerate} 
	\item A positive linear relationship between field-based fire behavior and space-based burn severity metrics, as higher-intensity fires remove more aboveground plant material and expose more of the mineral surface to satellites.
	\item No difference in burn severity among wildfires and prescribed burns, as prescribed burns have been shown to demonstrate a wide range of burn severity that likely overlaps with that of wildfires \cite[e.g., ][]{mcgranahan2025}. 
\end{enumerate}

\section{Methods}

\subsection{Study area}

The primary geographical scope of this study is the Northern Great Plains of the United States, specifically two experimental research locations in south-central and southwestern North Dakota (Fig.~\ref{Fig:StudyMap}). 
The south-central research location, referred to hereafter as the eastern location and in the eastern zone, is the North Dakota State University Central Grasslands Research Extension Center near Streeter, ND, located within the Northwestern Glaciated Plains in a relatively humid, mixed-grass prairie pothole landscape known as the Missouri Coteau.

The southwestern research location, referred to hereafter as the western location and in the western zone, is focused around the North Dakota State University Hettinger Research Extension Center in Hettinger, ND, located within the semi-arid mixed-grass prairie of the Northwestern Great Plains. 
Research activities at the western location were conducted by North Dakota State University personnel on two blocks of Center-managed rangeland pastures within 5 km of the Center and another privately-owned ranch approximately 20 km to the east of Hettinger.

\begin{figure}[H]
	\centering
	\includegraphics[width=1\columnwidth]{study_map-1.pdf}
	\caption{Location of rangeland research areas with experimental prescribed burns ($R_{X}$) and comparison wildfires ($\times$, $+$) in the Northern Great Plains. 
		Precipitation anomaly refers to the difference from mean annual precipitation at each of the western and eastern research locations for the study period (2017-2024); grey shaded areas have precipitation anomalies that exceed $\pm$ 3 standard deviations of mean annual precipitation and thus wildfires from these areas were excluded.
		Eastern and western zones align loosely with EPA Level III ecoregions NW Glaciated Plains and NW Great Plains, respectively (Fig.~\ref{AppFig:RegionMap}), with the exception of five wildfires near the Missouri River in south-central North Dakota being assigned to the eastern zone, which has substantially fewer wildfires than the western zone for comparison. 
		Inset: Study region within broader US Great Plains and continental US. }
	\label{Fig:StudyMap} %Fig.~\ref{Fig:StudyMap}
\end{figure}

\subsection{Data sources}

This study reports on both prescribed burns conducted as part of rangeland ecology and management research, and wildfires that spread from unintentional ignitions in the region.
Data in this study are derived from two sources: fire behavior data collected \emph{in agris}\textemdash from field-based sensors recording temperature as prescribed fires occurred\textemdash and remotely-sensed products from space-based Earth observation satellites. 
Fire behavior data collected \emph{in agris} consist of flame front rate of spread, flame temperature, and soil surface temperature. 
Remotely-sensed data consist of immediate burn severity (differenced Normalized Burn Ratio, or $\Delta$NBR). 
Each of these data types and products are described below, and script for their processing and analysis in the \textsf{R} statistical environment \cite{rcoreteam2024} is freely available from Ag Data Commons \cite{mcgranahan2025b}.

\subsubsection{Location of fire incidents}

\textbf{Prescribed burns.\textemdash}54 individual experimental burn units were treated with prescribed broadcast burning between 2017 and 2020 by North Dakota State University personnel associated with the Hettinger (22) and Central Grasslands (32) Research Extension Centers. 
Burns were conducted as investigations into rangeland ecosystem responses to prescribed burning; results of this research and more extensive descriptions of experimental design have been published for both Central Grasslands \cite{duquette2022, wanchuk2024, mcgranahan2025a} and Hettinger \cite{spiess2024, spiess2025} Research Extension Centers. 
All burns at Central Grasslands and 15 at Hettinger were part of patch-burn grazing research in which 64 ha pastures were divided into four, 16 ha patches, one of which was burned annually. 
Burns at Central Grasslands were conducted in the spring (April\textendash May) and in the fall (October) at Hettinger using adaptive ignition strategies to maximize burned area within each unit, including ring, strip, and spiral ignitions. 
Seven additional fall burns at Hettinger were conducted on private ranch, with areas ranging from 5\textendash 12 ha \cite{mcgranahan2022}.

\textbf{Wildfires.\textemdash}To compare burn severity effected by experimental prescribed burns to that produced by wildfires, a set of comparison wildfires were identified from the National Interagency Fire Center (NIFC) Interagency Fire Perimeter History\textemdash All Years geospatial database \cite{nifc2024}. 
Incidents in the database within the study region were filtered to wildfires of at least 10 ha that occurred between 2017\textendash 2024. 
As North Dakota had the fewest wildfire incidents in the region (Table~\ref{AppTab:StatesFires}), and very few occurred near the Research Extension Centers (Fig.~\ref{AppFig:RegionMap}), two additional steps were taken to provide sufficient sample size by extending beyond North Dakota while maximizing biophysical comparability with the prescribed burn locations:

Firstly, because wildfires spread across crop fields and pastures as well as through rangeland, the area within each wildfire perimeter identified as rangeland in the United States Forest Service (USFS) Extent of Coterminous US Rangelands raster layer \cite{usdaforestservicenodate} was summed, and incidents $<$20\% rangeland were excluded.

Secondly, because the study area spans a wide precipitation gradient (Fig.~\ref{AppFig:RegionMap}) that effects substantial differences in fuel load and plant community composition, comparison wildfires were limited to those within 3 standard deviations of mean annual precipitation at the Research Extension Center in each zone (Fig.~\ref{Fig:StudyMap}).
Mean and standard deviation in annual precipitation for the period 2017\textendash 2024 were calculated from daily weather data collected by the respective North Dakota Agricultural Weather Network stations at each Center \cite{ndawn2025, ndawn2025a}. 
Precipitation anomalies across the study region were determined by creating a 10 $\times$ 10 km resolution grid and retrieving the mean annual precipitation, 2017\textendash 2024, for each cell centroid from the TerraClimate dataset \cite{abatzoglou2018}, accessed with the \emph{climateR} package \cite{johnson2021} for \textsf{R}. 
Centroids from wildfire perimeters were spatially joined with the nearest precipitation anomaly value in the grid, and incidents where the mean annual precipitation differed  by more than 3 standard deviations from that of the Research Extension Center in that zone were excluded.
Ten incidents were included as comparison wildfires in the eastern zone, and 19 in the western zone.

Both prescribed burn and wildfire features were overlaid as polygons onto true color land surface imagery from the United States Geological Survey's National Map \cite{usgs_nm} to identify and exclude non-vegetated areas, such as ponds, areas of bare soil, and especially for wildfire perimeters, elements of the built environment such as farmsteads and petroleum well pads. 
As the NIFC perimeters often reflect general outlines of suppression efforts rather than boundaries of actual wildfire burned area, all perimeters were inspected visually over true color post-fire imagery and vertices edited for spatial accuracy.

\subsubsection{Fire behavior data}

Field measurements of fire behavior were collected from 26 of the prescribed burns described above \cite{mcgranahan2022,mcgranahan2023}.
Briefly, fire behavior data were collected with K-type thermocouples (Omega Engineering, Inc., Norwalk, Connecticut, USA) attached to electronic data loggers using two different systems:

Firstly, most prescribed burns (23) were sampled with the FeatherFlame system \cite{mcgranahan2021}, which consists of three thermocouples arranged in a 1m equilateral triangle 15 cm above the soil surface, and one thermocouple placed at the soil surface, all read by a single Arduino-based microcontroller writing temperature data at 1.5 Hz to a removable data storage card.
When arranged as such, the FeatherFlame system provides information on rate of spread, flame temperature at 15 cm, and soil surface temperature \cite{mcgranahan2022}. 
Nine FeatherFlame units were deployed in each of these 23 burns.

Secondly, fire behavior at three prescribed burns was measured with single-channel HOBO\textregistered~ U12 thermocouple dataloggers (Onset Computer Corporation, Bourne, Massachusetts, USA) logging at 1.0 Hz. 
At three sub-sample points within each burn, two loggers were deployed, one with a thermocouple placed at the soil surface and another at 15 cm above the soil surface. 
Rate of spread information was not available from these three instances of data from single-channel loggers.

\subsubsection{Remotely-sensed data}

Burn severity\textemdash a measure of how much aboveground material was removed by wildland fire combustion\textemdash was calculated as the difference in Normalized Burn Ratio ($\Delta$NBR) from Normalized Burn Ratio (NBR) values in two images, the most-recent cloud-free image taken just before the burn, and the first cloud-free image taken just after the burn \cite[e.g.,][]{llorens2021}.
The NBR index is derived from the Near-Infrared and Shortwave Infrared bands \cite{key2006a}, and the subtraction of the NBR value in the post-burn image from the pre-burn NBR value at the same pixel provide $\Delta$NBR.

NBR data were retrieved from the European Space Agency (ESA) Sentinel-2 mission's L2A collection, which provides multispectral imagery at 10 m $\times$ 10 m resolution from 2017 to present. 
All data were downloaded as geo-referenced raster files using the freely-available ESA Copernicus Browser, cropped to the bounding box of each visually-verified perimeter, generated with a custom EvalScript \cite{mcgranahan2025}, following visual inspection of each area of interest in true color images to ensure quality, cloud-free imagery.

For the subset of prescribed burns with fire behavior measurements, the coordinates recorded when each datalogger was deployed were used to extract remotely-sensed data from the pre-burn and post-burn multispectral Sentinel-2 rasters using the \emph{terra} package for \textsf{R} \cite{hijmans2022}, to allow spatially-explicit tests of the relationship between \emph{in agris} measurements and remotely-sensed data at the finest scale available. 
For the broader comparison of prescribed burns and wildfires, each perimeter was reduced with an internal 20 m buffer (at least two raster cells in Sentinel-2 imagery) to minimize edge effects, and $\pm$100 gridded sample points were distributed within each perimeter. 
For wildfires, sampled pixels were limited to those identified as rangeland in the USFS product \cite{usdaforestservicenodate}. 
Remotely-sensed data were again extracted from the locally-downloaded multispectral rasters using \emph{terra}.

\subsection{Statistical analysis}

Linear regression models were used to compare remotely-sensed burn severity against \emph{in agris} fire behavior data from prescribed burns, and to compare burn severity across prescribed burns and wildfires. 
In models comparing the 26 prescribed burns only, plot-level data from each experimental location were combined and fit with location and burn event as a nested random effect in linear mixed-effect models using  the \texttt{lmer} function in the \emph{lme4} package for \textsf{R} \cite{bates2015}. 
Statistical significance was determined with a $\chi^{2}$ test of deviance against null, intercept-only models.

Models comparing burn severity across the 54 prescribed burns and 29 wildfires were fit with ordinary least squares regression using the \texttt{lm} function. 
These data were summarized as mean values for each unique fire event to eliminate pseudoreplication. 
Visual evidence suggested a likely statistical interaction between fire type and the two zones of the study area, which was confirmed by a significant interaction term in a global model fitting burn severity against fire type $\times$ zone.
Post-hoc comparisons of prescribed fires and wildfires within each zone were conducted with the \emph{emmeans} package \cite{lenth2018} by applying the \texttt{joint\_tests} function to the full model to produce an ANOVA table of test statistics for each zone.

\section{Results}

Remotely-sensed burn severity\textemdash as measured by differenced Normalized Burn Ratio ($\Delta$NBR)\textemdash increased with two directly-measured fire behavior responses (Fig.~\ref{Fig:SevBehav})\textemdash flame temperature ($\chi^{2}=$ 28.2, $P<$ 0.001) and rate of spread ($\chi^{2}=$ 4.3, $P=$ 0.04)\textemdash but burn severity had no statistically-significant relationship with soil surface temperature ($\chi^{2=}$ 2.0, $P>$  0.05).

%Likewise, three fire response variables declined as fuelbed greenness (Normalized Differenced Vegetation Index, NDVI) increased (Fig.~\ref{Fig:BehavNDVI})\textemdash flame temperature ($\chi^{2}=$ 6.7, $P<$ 0.01), rate of spread ($\chi^{2}=$ 20, $P<$ 0.001), and burn severity ($\chi^{2}=$ 6.9, $P<$ 0.01)\textemdash but again there was no statistically-significant relationship with soil surface temperature ($\chi^{2}=$ 2.9, $P>$ 0.05).

In the comparison of burn severity ($\Delta$NBR) across prescribed burns and wildfires in the two zones of the study region ( Fig.~\ref{Fig:WF_Rx}), there was a statistically-significant interaction between fire type and zone ($F_{1,79}=$ 10, $P<$ 0.01). 
In post-hoc pairwise contrasts, burn severity did not differ between prescribed burns and wildfires in the semi-arid western zone ($F_{1,79}=$ 1.5, $P>$ 0.05), but burn severity was statistically-significantly greater in wildfires that prescribed burns in the relatively higher-rainfall eastern zone ($F_{1,79}=$ 28, $P<$ 0.001).

\begin{figure}[t]
	\centering
	\includegraphics[width=1\columnwidth]{dNBR_v_FB-1.png}
	\caption{Burn severity\textemdash as measured by differenced Normalized Burn Ratio ($\Delta$NBR)\textemdash plotted against three fire behavior metrics measured during 26 prescribed burns in south-central and southwestern North Dakota.
	Solid trendlines with 95\% confidence intervals denote statistically-significant linear relationships in linear mixed-effect regression models.   }
	\label{Fig:SevBehav} % Fig.~\ref{Fig:SevBehav}
\end{figure}

%\begin{figure}[t]
%	\centering
%	\includegraphics[width=1\columnwidth]{FB_v_ndvi-1.png}
%	\caption{Three field-sampled fire behavior metrics and a remotely-sensed measure of burn severity (differenced Normalized Burn Ratio, $\Delta$NBR) plotted against fuelbed greenness, measured as Normalized Difference Vegetation Index (NDVI) from the pre-burn Sentinel-2 image.
%		Solid trendlines with 95\% confidence intervals denote statistically-significant linear relationships in linear mixed-effect regression models.   }
%	\label{Fig:BehavNDVI} % Fig.~\ref{Fig:BehavNDVI}
%\end{figure}

\begin{figure}[t]
	\centering
	\includegraphics[width=1\columnwidth]{WF_v_Rx-1.png}
	\caption{Burn severity\textemdash as measured by differenced Normalized Burn Ratio ($\Delta$NBR)\textemdash across 29 wildfires and 54 prescribed burns in the two zones of the study region (see map, Fig.~\ref{Fig:StudyMap}). 
	Triangles represent group means. 
	Width of boxes scale with group sample size.   }
	\label{Fig:WF_Rx} % Fig.~\ref{Fig:WF_Rx}
\end{figure}

\section{Discussion}

Using the US Northern Great Plains as a case study, this paper makes two primary comparisons that can support wildland fire scientists and managers in rangeland ecosystems worldwide. 
Firstly, three field-based fire behavior measurements (rate of spread, flame temperature, and soil surface temperature) from prescribed burns are compared to remotely-sensed burn severity (differenced Normalized Burn Ratio, $\Delta$NBR), and secondly, burn severity is compared between prescribed burns and wildfires. 
These comparisons provide partial support for each of two stated hypotheses: there is a positive linear relationship between remotely-sensed burn severity and measured flame temperature and rate of spread (but not soil surface temperature), and prescribed burn severity is not different from that effected by wildfires in the semi-arid western zone of the region (but wildfire severity is greater in the eastern zone).

These results increase the range of options that provide insight into the variability of fire behavior and immediate fire effects in rangelands, facilitating better application of research from prescribed burns to areas affected by wildfire. 
Below are some considerations for both conventionally-collected and remotely-sensed data, and knowledge gaps that will require additional research to enhance the utility of these resources in wildland fire science and management applications.

\subsection{Remote sensing as complement, or even surrogate, to field-based measurements}

Most apparently, managers unable to deploy fire behavior sensors during prescribed burns appear to be able to use freely-available remotely-sensed data on burn severity to describe meaningful gradients in fire behavior, and even those able to deploy sensors can use remotely-sensed data to describe fire behavior at  spatial scales beyond sample points. 
As such, remotely-sensed burn severity can enhance \emph{wildland fire science literacy}, the ability of scientists and managers to express and understand characteristics of the fire environment that give context for fire effects \cite{mcgranahan2018, mcgranahan2022a}. 
More specifically, rangeland fire ecologists can include summaries of burn severity alongside, or in lieu of, actual fire behavior measurements in their reports to help contextualize fire effects within a dose-response framework instead of a simple burned/unburned dichotomy \cite{smith2016a}.

Although using time-temperature data from thermocouple dataloggers to calculate two-dimensional rate of spread is considered a more robust measure of wildland fire behavior than flame temperature alone, rate of spread is still an imperfect metric.
On one hand, rate of spread is a description of the wildland fire system, and does quantify specific physical processes within the fire environment \cite{finney2021}. 
Thus, some fire scientists measure responses such as total energy flux using radiometers to detect infrared radiation \cite{butler2004, kremens2012, obrien2018}, but such systems can be logistically challenging and expensive \cite{butler2010}.

On the other hand, rate of spread is the initial output from several standard mathematical fire behavior prediction models \cite{rothermel1972,noble1980}, and is thus a useful connection between predicting and evaluating fire behavior. 
Rate of spread is also an input for equations that calculate fire intensity, the value of which ``lies not so much in the exact description of energy, but rather in the provision of numerical data for comparing fires \cite[][p. 354]{alexander1982}''. 
Knowing that burn severity scales with rate of spread adds a monitoring option for managers to use fire behavior prediction models as part of planning burns that meet specific objectives\textemdash and deciding under what conditions to conduct those burns effectively and safely.

A particular concern with space-based multispectral products is a potentially coarser resolution than many processes within the fire environment and plant community play out. 
The inherent heterogeneity of wildland fuelbeds introduces variability in fire behavior that is critical to the survival of many species in fire-dependent ecosystems, such as how gaps between the vegetation that serves as wildland fuels reduce microsite fire intensity and thus severity to propagules \cite{daibes2018, atchley2021}. 
Analyses of post-fire patch dynamics suggest grain sizes\textemdash the finest scale of spatial independence\textemdash ranging from 0.5\textendash 10 m \cite{lamont1993, gimeno-garcia2004, pereira2013, gongalsky2016, moore2021}. 
But Sentinel-2 products are comprised of an average value for each 10 $\times$ 10 m pixel, and in the case of indices including NBR, one of the bands used in the calculations is collected at  20 $\times$ 20 m resolution. 
Thus, the pixel-scale averages in remotely-sensed severity data might obscure finer-scale instances of high severity that are ecologically meaningful to the individual organisms that comprise the broader-scale community. 
Research from forests suggest this might be more more of a concern for prescribed fires: Extreme fire weather conditions, more typical of wildfires, can override stand-level heterogeneity to effect more uniform severity \cite{romme2006, mcfarland2025}. 
In either case, remotely-sensed severity metrics should be compared to ecosystem-specific research on direct and indirect fire effects at the scale of relevant biological processes to determine how well severity indices correlate with rangeland fire effects, in addition to fire behavior.

These data corroborate other evidence that the biotic and abiotic factors that typically have linear relationships with surface fire behavior metrics (\emph{e.g.}, rate of spread, flame properties) appear to have little correlation with soil heating. 
A subset of the prescribed burns included here were subject to a previous analysis of how fuel and fire weather conditions affected measured fire behavior, which reported that rate of spread increased with wind speed and flame temperature increased with fuel load and declined with fuel moisture; however, no fuel or weather factor had a statistically-significant effect on soil surface temperature \cite{mcgranahan2023}.
Likewise, this study showed no linear relationship between burn severity and soil surface temperature (Fig.~\ref{Fig:SevBehav}).
Similar patterns\textemdash or rather, lack thereof\textemdash between variables that drive rate of spread not correlating with soil surface temperature have been reported elsewhere \cite{gimeno-garcia2004}.

The consistent lack of connections between the fire environment, soil heating, and burn severity suggest that at least for rangeland fuelbeds in which the primary carrier of fire is fine fuels (graminoids and herbaceous vegetation), a burned/unburned dichotomy might in fact be sufficient to explain first-order (direct) fire effects to soil. 
While it is certainly true that nutrient pools and microbial abundance are directly affected by soil heating, soil properties in rangelands recover within days or weeks of even relatively severe fires \cite{mcgranahan2022,docherty2012}; fungal communities are especially resilient to the low soil heating in grass-carried fires \cite{egidi2016}. 
The primary driver of belowground tissue damage is duration of heating; as such, fast-moving fires through grass-dominated fuelbeds effect the lowest amount of soil heating due to the shorter residence time of fine fuel combustion \cite{neary1999a, robichaud2000}. 
And because soil is a good insulator, even when soil surface temperatures in a subset of the prescribed burns here ranged from 200\textendash 400\textsuperscript{$\circ$}C, temperatures were consistently at or below 100\textsuperscript{$\circ$}C at depths of just 2 cm \cite{mcgranahan2022}.

Thus, even intense fires through grass-dominated fuels might not exceed a threshold of soil heating to immediately effect anything beyond a general burned vs. unburned response pattern. 
However, second-order (indirect) effects to soil microbes might depend more on surface fire behavior\textemdash as modulated by abiotic and biotic conditions, and correlated with burn severity\textemdash because these organisms can be affected by post-fire soil conditions and plant community dynamics \cite{neary1999a}.

\subsection{Transferability of research from prescribed burns to wildfire-affected rangeland}

Although research on prescribed fire effects on rangeland plant communities has been conducted in the US Great Plains for over 100 years \cite{hensel1923, aldous1934, hulbert1969}, conceptions of wildfires and knowledge of their effects have largely been derived from forests \cite{crist2023}.
Due to the high potential fuel loading of forests (especially those under decades of fire deficit), there is a general concern that 'natural' and prescribed burns in forests are ecologically different \cite{williams1994}.  
But little scientific information substantiates such a difference in rangelands.

As most rangeland ecosystems in the US are used for commercial livestock grazing, there is a general concern among managers throughout the western US about the impacts of livestock grazing on rangeland assumed to have been negatively affected by a wildfire. 
For example, the general guidance for the US Department of Interior Bureau of Land Management is to have their permittees defer, delay, or avoid altogether livestock grazing for up to two seasons on a grazing allotment burned by a wildfire \cite{blm2007}. 
While \citet{kluth2024} attribute the two-year convention to \citet{blaisdell1982}, it was more explicitly stated earlier, by \citet[][p. 9]{wright1979}: ``Prescribed fire can be a useful too in many big sagebrush communities if the fires are carefully planned and livestock do not graze the burn for two growing seasons.''
(Ironically, both sources were cautioning on the return of livestock to range recently treated with prescribed burns, a practice both \emph{recommended be used} to maintain herbaceous plant cover and reduce woody plants; the prescribed fire predicate has since been detached and the 2-year grazing exclusion adopted into pyro-skeptical wildfire management.)

It remains an open question whether rangeland management recommendations from the arid Great Basin region apply to semi-arid rangeland in the Northern Great Plains. 
Studying wildfire is difficult: by its very nature it occurs unpredictably, precluding the pre-treatment data collection that constitutes a robust sampling design, as simple burned-unburned comparisons are limited in their capacity to control for environmental and temporal variability \cite{zaitsev2016}. 
What research does exist on post-wildfire vegetation dynamics in the region suggests the caution over damage to plant communities is unjustified: productivity in the season after two wildfires in the Northwestern Great Plains was resilient to both wildfire and grazing \cite{gates2017, kral-obrien2020, williams2022}. 
Meanwhile, benefits of prescribed fire to the rangeland ecosystem and livestock grazing recent burns have been widely documented, especially from the prescribed fire experiments included in this study \cite{powell2018, spiess2024, mcgranahan2022, wanchuk2024, wanchuk2024a}.
What remains is an understanding of whether data from experimentally-controlled prescribed burns are relevant to wildfire incidents.

The data presented here suggest that across both wildfires and prescribed burns, burn severity in the Northern Great Plains is generally low to moderate, especially in the western zone; wildfires in the more humid eastern zone tend to be moderately high (Fig.~\ref{Fig:WF_Rx}).
Based on the lack of difference in burn severity between wildfires and prescribed burns in semi-arid rangeland of the Northwestern Great Plains, it is likely that research results from prescribed burns in the region are applicable to ecologically-similar areas affected by wildfire. 
Although less certain, there is possibly a useful degree of transferability between prescribed burn research and wildfires in the eastern zone as well, because despite the statistically-significant difference in burn severity, even the incidents with the most severe burns fall below the 0.66 $\Delta$NBR threshold for high burn severity \cite{key2006a}.

Consistently low to moderate burn severity aside, a number of biotic and abiotic differences between most wildfires and prescribed burns ought to be resolved before the ecological effects of these fires can be considered consistent, as well. 
Many of these differences fall under the general umbrella of seasonality, which affects both the conditions of the fire environment and the biological status of organisms impacted by heating, \emph{e.g.} seasonal phenology and dormancy considerations. 
A common denominator of weather conditions conducive to burning relate to moisture: dry air from cold fronts, dry surface conditions from periods with little to no precipitation, and low relative humidity \cite{flannigan2001}. 
Additional factors include fuel load and wind speed at the time of an ignition further modulate energy release \cite{cheney1995, reid2010, kidnie2015, gomes2020}.

Prescribed burn operations are typically reserved for seasons more conducive to manageable fire behavior, while wildfires occur whenever fuels and weather align with ignitions. 
In the eastern zone, prescribed burns reported here were conducted entirely in April and May, whereas the wildfires in the  eastern zone included here occurred fairly consistently between January and August (Table~\ref{AppTab:FireMonths}). 
Prescribed burns in the western zone were restricted to October, whereas all wildfires included here occurred January-September, and mostly in July and August (Table~\ref{AppTab:FireMonths}). 
And then within seasons, prescribed burns are often only conducted under moderate weather conditions, which likely accounts for substantially lower burn severity in experimental prescribed fires in the eastern zone: for instance, prescribed burns at the Central Grasslands Research Extension Center were only conducted on days with wind speeds below the 42-yr median \cite{mcgranahan2025}. 
Future research should measure post-fire effects on soils and plants on wildfires, as well, to determine that seasonal differences do not limit ecosystem recovery beyond that observed after prescribed burns despite consistent low to moderate severities in immediate assessments based on remotely-sensed data.

\subsection{Conclusion}

These data suggest wildland fire professionals can describe meaningful gradients in surface fire behavior in rangelands within an immediate assessment framework using differenced Normalized Burn Ratio ($\Delta$NBR) from remotely-sensed imagery as a metric of burn severity. 
Thus, even those without the capacity to measure fire behavior in the field can monitor prescribed fire effectiveness and incorporate burn severity in adaptive management plans. 
Understanding the relationship between burn severity across wildfires and prescribed burns is a critical step in applying knowledge gained from research on prescribed fires to areas impacted by wildfire. 
Resistance to prescribed burning might be overcome by increasing livestock managers' experience with post-fire forage resources through grazing areas burned in unintentional wildfires, but current practice and policy dissuades or outright prevents ranchers from doing so.
Despite the know improvements to forage, it is essential that future research study the association between burn severity with ecosystem recovery metrics to ensure post-fire grazing does not impair the sustainability of rangeland ecosystems.

\vspace{6pt}

\acknowledgments{The prescribed fires reported here depended on efforts of several North Dakota State University faculty and graduate students, especially Kevin Sedivec and Ben Geaumont for location management and equipment; and Megan Wanchuk, Jonathan Spiess, and Megan Zopfi (University of North Dakota) for their efforts in fire behavior data collection. 
Jay Angerer provided expertise for many aspects of remotely-sensed data management. }

\authorcontributions{DA McGranahan did everything related to this paper. }

\conflictsofinterest{The author declares no conflict of interest.}

\dataavailability{Data for formal analysis, modified perimeters of included wildfires, and script are freely available on Ag Data Commons \cite{mcgranahan2025b}. }

\funding{No specific funding was acquired for this study.}

\appendixtitles{yes} 
\appendixstart
\appendix
 \section{Additional information}

\begin{figure}[H]
	\centering
	\includegraphics[width=1\columnwidth]{SatelliteFireSpread.png}
	\caption{True-color images of a fire in North Dakota, which a satellite from the Sentinel-2 mission caught shortly after the fire started\textemdash a tall smoke plume coming off the wind-driven head fire is clearly visible on August 21, 2022.
	NBR and $\Delta$NBR in Fig.~\ref{Fig:NBRexample}.  }
	\label{Fig:TrueColor} % Fig.~\ref{Fig:TrueColor}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=1\columnwidth]{region_map-1.png}
	\caption{Portions of Montana, North Dakota, and South Dakota encompassed by the primary EPA Level III ecoregions of the Northern Great Plains\textemdash Northwestern Glaciated Plains and Northwestern Great Plains\textemdash within the broader US Great Plains (inset, red area). 
	Ecoregions shaded by mean total annual precipitation for the complete years in the study period (2017\textendash 2024). 
	$R_{x}$ badges indicate locations of prescribed burn experiments. 
	Open circles denote centroids of wildfire events in the NIFC Interagency Wildfire Perimeter database, 2017-2023; total counts presented in Table~\ref{AppTab:StatesFires}. }
	\label{AppFig:RegionMap} % Fig.~\ref{AppFig:RegionMap}
\end{figure}

\begin{specialtable}[H] 
\caption{Summarized totals of wildfire incidents presented in Fig.~\ref{AppFig:RegionMap}.  }
\label{AppTab:StatesFires}
\begin{tabular}{p{3cm}p{4cm}p{3cm}p{1cm}}
\toprule
\textbf{State} & \textbf{NW Glaciated Plains} & \textbf{NW Great Plains} & \textbf{Total} \\ 
\hline
 Montana (MT) & 153 & 998 & 1151 \\ 
South Dakota (SD) &   9 & 339 & 348 \\ 
North Dakota (ND) &  21 & 150 & 171 \\
\hline
\bottomrule
\end{tabular}
\end{specialtable}

\begin{specialtable}[H] 
	\caption{Monthy distribution of the wildfire subset used in this study, by zone, as determined by midpoint of pre- and post-fire image dates.  }
	\label{AppTab:FireMonths}
\begin{tabular}{lccccccccc}
	\hline
	Zone & Jan & Feb & Mar & Apr & May & Jun & Jul & Aug & Sep \\ 
	\hline
	East &   1 &  1 &   1  &   2 &  2 & 0   & 2 &  1 & 0  \\ 
	West &   1 &  0 &   2 &   3 &  0  & 0   & 5 &  6 &  2 \\ 
	\bottomrule
	\end{tabular}
\end{specialtable}

\end{paracol}

\reftitle{References}
\externalbibliography{yes}
\bibliography{RemoteSensingFireBehavior.bib}

\end{document}

### 2025-09-02 NorthDakotaBurnSeverity/writeup/SeverityFireBehavior.Rmd
ND severity paper first draft

--- 2025-09-02 NorthDakotaBurnSeverity/paper/FigureGenerator_SeverityFireBehavior.Rmd
@@
 title: 'Remote sensing in rangeland fire ecology: Comparing imagery to measured fire behavior, and burn severity across prescribed burns and wildfires'
-subtitle:  'Figures and results'
+subtitle:  'Data and analyses'
 author: "DAM"
@@
 knitr::opts_chunk$set(echo = FALSE, warning= FALSE,
-                      message= FALSE, eval = TRUE, 
-                      fig.path = "./", dev = c('png', 'pdf')) 
+                      message= FALSE, eval = TRUE)
+                      #fig.path = "./, dev = c('png', 'pdf')) 
 pacman::p_load(tidyverse, sf, ggpattern, pander, readxl, grid, gridExtra)
@@
 
+```{r region_map}
+ggplot() + theme_void() +
+  geom_sf(data = NGP, 
+          aes(fill = NA_L3NAME), 
+          alpha = 0.6) +
+  geom_sf(data = NGP_fires, 
+          pch = 1, alpha = 0.5) + 
+  geom_sf_label(data = NGP %>%
+                 group_by(STATE_NAME) %>%
+                 summarize(area = sum(Shape_Area)) , 
+               aes(label = state.abb[match(STATE_NAME,state.name)])) + 
+  scale_fill_manual('L3 ecoregion', 
+                    values = c("#74A089","#E1BD6D")) +
+  theme(legend.position = 'inside', 
+        legend.position.inside = c(0.3, 0.23))
+```
+
 ```{r study_map}
@@
 
-reg_gg <- 
   ppt_dat$ppt_grd %>%
@@
                     show.legend = F) +
-    geom_sf(data = ngp_counties %>% 
-  st_intersection(states %>% st_transform(st_crs(ngp_counties))), 
-            fill = NA, color = 'grey70') + 
+    geom_sf(data = ngp_counties, fill = NA, color = 'grey70') + 
     geom_sf(data = states, fill = NA, color = 'black') + 
@@
     scale_color_manual('Fire type', values = c('darkgreen', 'darkred'), guide = 'none')
-
-us_gg <- 
-  ggplot() + theme_void() + 
- # geom_sf(data = usa %>% st_buffer(10000), color = NA, fill = 'white')  + 
-  geom_sf(data = usa, color = 'white', fill = 'grey60')  + 
-  geom_sf(data = gp, fill = 'white', 
-          color =  'black', 
-          alpha = 0.5, size = 2) +
-  geom_sf(data = ngp, 
-          fill =  wes_palette('Zissou1')[5],  
-          color = NA, 
-          show.legend = F) +
-  scale_fill_manual('EPA Level III\necoregion', 
-                    values = c("#74A089","#E1BD6D"))
-
-grid.newpage() 
-main <- viewport(width = 1, height = 0.95, x = 0.5, y = 0.5) 
-inset <- viewport(width = 0.3, height = 0.4, x = 0.725, y = 0.7)
-print(reg_gg, vp = main)
-print(us_gg, vp = inset)  
-```
-
-```{r region_map}
-reg_gg <-
-ppt_dat$ngp_ppt %>%
-  mutate(NA_L3NAME = str_replace(NA_L3NAME, ' ', '\n')) %>%
-ggplot() + theme_void() +
-  geom_sf(aes(fill = NA_L3NAME, alpha = ppt), 
-          color = 'NA') +
-  geom_sf(data = NGP, 
-          fill = NA) + 
-   geom_sf_label(data = locations %>% 
-                  filter(type != 'Wildfire'), 
-                label = expression(R[X]), 
-                label.padding = unit(0.1, "lines"),
-                label.r = unit(0.5, "lines"),
-                label.size = 0.15,
-                size = 4) +
-  geom_sf(data = NGP_fires, 
-          pch = 1, alpha = 0.5) + 
-  geom_sf_label(data = NGP %>%
-                  group_by(STATE_NAME) %>%
-                  summarize(area = sum(Shape_Area)) , 
-                aes(label = state.abb[match(STATE_NAME,state.name)]), 
-                fontface='bold') + 
-  scale_fill_manual('EPA Level III\necoregion', 
-                    values = c("#74A089","#E1BD6D")) +
-  scale_alpha_continuous("Precipitation\nmean annual (mm)", 
-                         range = c(0.1, 1)) + 
-  theme(legend.position = 'right', 
-        legend.key.height = unit(1, 'cm')) + 
-  guides(alpha = guide_legend(order = 2), fill = guide_legend(order = 1) )
-
-us_gg <- 
-  ggplot() + theme_void() + 
-  geom_sf(data = usa %>% st_buffer(10000), color = NA, fill = 'white')  + 
-  geom_sf(data = usa, color = 'white', fill = 'grey60')  + 
-  geom_sf(data = gp, fill = 'white', 
-          color =  'black', 
-          alpha = 0.5, size = 2) +
-  geom_sf(data = NGP, 
-          fill =  wes_palette('Zissou1')[5],  
-          color = NA, 
-          show.legend = F) +
-  scale_fill_manual('EPA Level III\necoregion', 
-                    values = c("#74A089","#E1BD6D"))
-
-grid.newpage() 
-main <- viewport(width = 1, height = 0.95, x = 0.5, y = 0.5) 
-inset <- viewport(width = 0.3, height = 0.4, x = 0.25, y = 0.25)
-print(reg_gg, vp = main)
-print(us_gg, vp = inset)
-```
-
-```{r}
-NGP_fires %>% 
-  as_tibble() %>%
-  group_by(STATE_NAME, NA_L3NAME) %>% 
-  summarize(Fires = n()) %>%
-  pivot_wider(names_from = NA_L3NAME, 
-              values_from = Fires) %>%
-   mutate(Total = `Northwestern Great Plains` + `Northwestern Glaciated Plains`) %>%
-  #xtable::xtable() 
-  pander::pander("Fire incidents within each LIII ecoregion by state.")
 ```
@@
 
-```{r}
-BurnSeverityData %>%
-  group_by(zone, type) %>%
-  summarize(fires = n() ) %>%
-  pivot_wider(names_from = zone, 
-              values_from = fires) %>%
-  mutate(Total = east + west) %>%
-  pander::pander("Incidents used in this study.")
-```
-
-## Burn severity across zones and fire types
-
 ```{r WF_v_Rx, fig.width = 8, fig.height = 4, fig.cap = " "}
     BurnSeverityData  %>% 
-      mutate(zone = fct_rev(zone), 
-             zone = paste0(str_to_title(zone), 'ern zone'),
+      mutate(zone = ifelse(L3 == "Northwestern Great Plains", 
+                           'Western zone', 'Eastern zone'), 
+             zone = fct_rev(zone), 
              type = fct_relevel(type, c("Wildfire",'Rx' )), 

### 2025-09-02 NorthDakotaBurnSeverity/script/ChuggingImagery.R

@@ -1,0 +1,466 @@
+pacman::p_load(tidyverse, magrittr, sf, stars)
+
+load('../albersEAC.Rdata')
+
+##
+##  T H E R M O C O U P L E   C O M P A R I S O N S
+##
+
+# This will not run now, 
+# but the output remains valid. 
+# object 'BurnIndices' has plot-level data for 26 Rx burns 
+# renamed as-is to 'PlotBurnIndices' 8/8/2025 to avoid conflict with 
+# broader-scale data from CGREC from previous project with same name :(
+
+  # Get fuel greenness and dNBR for completed burns
+  fires <- st_read('./data/ThermocouplePlacements.gpkg', 'AllBurnData') 
+  
+  dates <- read_csv('./data/BurnDates.csv') %>%
+            mutate(across(c(BurnDate:PostBurn), ~ as.Date(.x, '%m/%d/%Y')))
+  
+  # locations of imagery on S
+  {
+    cgrec_imagery = 'S:/DevanMcG/Projects/SpringSummerSeverity/sentinel/MultiBand'
+    hrec_imagery = 'S:/DevanMcG/Projects/NorthDakotaBurnSeverity/sentinel/HREC'
+    day_imagery = 'S:/DevanMcG/Projects/NorthDakotaBurnSeverity/sentinel/dayton'
+    }
+
+    BurnIndices <- tibble() 
+    for(i in 1:length(unique(fires$burn))) {
+                # Get fire & extract spatial, temporal info
+                  fire = fires %>% filter(burn == unique(fires$burn)[i]) 
+                  BurnDay = unique(fire$date)
+                  loc = unique(fire$location)
+                # Get dates
+                  pre_date = dates %>% filter(BurnDate == BurnDay) %>% select(PreBurn)
+                  post_date = dates %>% filter(BurnDate == BurnDay) %>% select(PostBurn)
+                # Map to location's imagery
+                  imagery_dir <- case_when(
+                      loc == 'CGREC' ~ cgrec_imagery, 
+                      loc == 'HREC' ~ hrec_imagery,
+                      loc == 'Dayton' ~ day_imagery
+                    )
+                # Fetch images in location's folder
+                  images <- list.files(imagery_dir, pattern = "\\.tiff$", ignore.case = TRUE)
+                # Set location CRS to match rasters
+                 fire %<>% st_transform(case_when(
+                                        loc == 'CGREC' ~ 32614, 
[…]

### 2025-09-02 NorthDakotaBurnSeverity/script/DeterminePrecipAnomalies.R

@@ -1,0 +1,212 @@
+pacman::p_load(tidyverse, magrittr, climateR, foreach, doSNOW)
+
+# study area 
+  ngp <- read_sf('./data/SeverityComparison.gpkg', 'StudyRegion')
+# Census tracts
+  ngp_counties <- tigris::counties(state = c('MT', "ND", 'SD'), cb = T)
+  
+  ngp_counties%<>% st_transform(st_crs(ngp)) %>% st_intersection(ngp)
+ngpFIPS <-
+  read_sf('S:/DevanMcG/GIS/SpatialData/US/census/tracts2020', 
+          'ConterminousTracts2020') %>%
+  select(-state) %>%
+  st_transform(st_crs(ngp)) %>%
+  st_intersection(ngp)  
+
+ngp_grid <- 
+ngp %>%
+  st_make_grid(cellsize = c(10000, 10000)) %>%
+  st_intersection(ngp)  %>% 
+  st_as_sf() %>%
+  rowid_to_column('cell')
+ggplot() + geom_sf( )  + 
+  geom_sf(data = ngp, fill = NA, color = 'blue')
+
+setdiff(ngpFIPS$fips, ngpFIPS2$fips)
+
+length(unique(ngpFIPS2$fips))
+
+
+{
+  cores= parallel::detectCores()
+  cl <- makeCluster(cores) 
+  registerDoSNOW(cl )
+  
+  begin = Sys.time()
+  ngp_ppt <- 
+    foreach(i=1:length(ngp_grid$cell), 
+            .combine = bind_rows, 
+            .errorhandling = 'remove', 
+            .inorder = FALSE, 
+            .packages = c('tidyverse', 'sf', 'climateR')) %dopar% {
+              getTerraClim(
+                AOI = ngp_grid %>%
+                      slice(i) %>%
+                       st_centroid() , 
+                varname = 'ppt',
+                startDate = '2017-01-01', 
+                endDate = '2024-12-31'  ) %>% 
+                as_tibble() %>%
+                separate(date, into = c('year', 'month', 'day'), sep = "-") %>%
+                 group_by(year) %>%
+                summarize(ppt = sum(ppt_total, na.rm = TRUE)) %>%
+                summarize(ppt = mean(ppt, na.rm = TRUE)) %>%
+                add_column(cell = i)
+            }
+  stopCluster(cl)
+  Sys.time() - begin 
+}
+
[…]

### 2025-09-02 NorthDakotaBurnSeverity/script/DeterminePrecipNGP.R

@@ -1,0 +1,60 @@
+pacman::p_load(tidyverse, magrittr, climateR, foreach, doSNOW)
+
+# study area 
+NGP <- read_sf('./data/MapStuff.gpkg', 'NGP_states')
+
+NGP_grid <- 
+NGP %>%
+  st_make_grid(cellsize = c(20000, 20000)) %>%
+  st_intersection(NGP)  %>% 
+  st_as_sf() %>%
+  rowid_to_column('cell') 
+ggplot() + geom_sf( )  + 
+  geom_sf(data = NGP, fill = NA, color = 'blue')
+
+
+{
+  cores= parallel::detectCores()
+  cl <- makeCluster(cores) 
+  registerDoSNOW(cl )
+  
+  begin = Sys.time()
+  NGP_ppt <- 
+    foreach(i=1:length(NGP_grid$cell), 
+            .combine = bind_rows, 
+            .errorhandling = 'remove', 
+            .inorder = FALSE, 
+            .packages = c('tidyverse', 'sf', 'climateR')) %dopar% {
+              getTerraClim(
+                AOI = NGP_grid %>%
+                      slice(i) %>%
+                       st_centroid() , 
+                varname = 'ppt',
+                startDate = '2017-01-01', 
+                endDate = '2024-12-31'  ) %>% 
+                as_tibble() %>%
+                separate(date, into = c('year', 'month', 'day'), sep = "-") %>%
+                 group_by(year) %>%
+                summarize(ppt = sum(ppt_total, na.rm = TRUE)) %>%
+                summarize(ppt = mean(ppt, na.rm = TRUE)) %>%
+                add_column(cell = i)
+            }
+  stopCluster(cl)
+  Sys.time() - begin 
+  }
+
+ppt_dat$ngp_ppt <-  
+NGP_grid %>%
+  st_intersection(NGP %>% 
+                    select(NA_L3NAME)) %>%
+  mutate(Area = st_area(.)) %>%
+  group_by(cell, NA_L3NAME) %>%
+  summarize(Area = sum(Area)) %>%
+  arrange(desc(Area)) %>%
+  slice(1) %>%
+  select(-Area) %>%
+  merge(by = 'cell', 
+        NGP_ppt) 
+  
+
+# save(ppt_dat, file = './data/ppt_dat.Rdata')

### 2025-09-02 NorthDakotaBurnSeverity/script/GetTCHeatingCurves.R

@@ -1,0 +1,146 @@
+pacman::p_load(tidyverse, magrittr, FSA)
+
+
+  FilePath = "./A_cana/Data/ThermocoupleData/FA21/"
+  MinTemp = 15
+  ExcludeColumns = c("logger", "timestamp")
+
+# list.dirs(FilePath)
+tc_files <- list.files(FilePath, pattern = "\\.TXT$")
+
+f = 6 # edit manually to go through the files one by one
+l = tools::file_path_sans_ext(tc_files[f])
+
+{
+  rd1 <- 
+    paste0(FilePath, tc_files[f]) %>% 
+    read_csv(col_names = FALSE, show_col_types = FALSE) %>% 
+    mutate(X1 = l)
+
+  # Prepare data for pivoting
+  tcs <- dim(rd1)[2] - length(ExcludeColumns)
+  colnames(rd1) <- c(ExcludeColumns, seq(1, tcs, 1))
+  # Pivot logger data into long format
+  rd <-
+    rd1 %>%
+    pivot_longer(values_to = 'degC', 
+                 names_to = 'TC', 
+                 -all_of(ExcludeColumns) ) %>%
+    group_by(TC) %>%
+    mutate(obs = seq(1, n(), 1)) %>%
+    ungroup() 
+}
+  # check with plots
+    rd %>%
+    group_by(TC) %>%
+     slice( which(row_number() %% 10 == 1)) %>%
+    ggplot() + theme_bw(14) + 
+    geom_line(aes(x = obs, 
+                  y = degC)) + 
+    labs(title = paste0("Logger ", unique(rd$logger))) + 
+    facet_wrap(~TC, scales = "free_y")
+    
+    #HeatingCurves <- tibble() 
+{   
+    # Prompt channel selection 
+    channels <- readline(prompt = message("Select channels to import (see graphics device) by entering numbers of desired channels separated by commas, or...\n Enter: select all channels\n 0: None (and skip to next file)\n Q: quit (then select Cancel to avoid closing R altogether)")) # \nTo *exclude* certain channels, precede channel number with negative sign (-)."))
+    
+    if(channels == "Q") { 
+      quit(save = 'ask') 
+    } else {
+      if(channels == "") { 
+        selections <- unique(rd$TC) 
+      } else {
+        selections <- ifelse(channels %in% c('0', 'O'), 'None', channels) 
+        selections <- unlist(strsplit(selections, ",")) %>% 
+          trimws()  
+      } 
+    }
+
[…]

### 2025-09-02 NorthDakotaBurnSeverity/script/GetWildfiresByPrecipAnomaly.R

@@ -1,0 +1,210 @@
+
+# study area 
+  ngp <- read_sf('./data/SeverityComparison.gpkg', 'StudyRegion')
+  ngp_counties <- tigris::counties(state = c('MT', "ND", 'SD'), cb = T)
+
+# How many fires by state? 
+  NGP <- read_sf('S:/DevanMcG/GIS/SpatialData/US/EPAecoregions/L3/us_eco_l3.shp') %>%
+    filter(NA_L3NAME %in% c("Northwestern Great Plains", 'Northwestern Glaciated Plains')) %>%
+   st_intersection(  ngp_counties %>%
+        group_by(STATE_NAME) %>%
+        summarize(area = sum(ALAND)) %>%
+        st_transform(st_crs(NGP))) 
+  
+  st_write(NGP, './data/MapStuff.gpkg', 'NGP_states')
+  
+  read_sf('./data/MapStuff.gpkg', 'NGP') %>%
+    ggplot() + geom_sf() 
+  cities <- tigris::urban_areas(filter_by= st_transform(NGP, st_crs(ngp_counties)), 
+                                 'ftp', cb = T)
+  
+  
+  AllFires <-  read_sf("S:/DevanMcG/GIS/SpatialData/US/NIFC", 
+                       "InterAgencyFirePerimeterHistory_All_Years_View") %>%
+    select(FIRE_YEAR, INCIDENT, FEATURE_CA) %>%
+    st_transform(st_crs(NGP)) %>%
+    st_centroid() %>%
+    # st_make_valid(.) %>%
+    st_intersection(NGP) %>%
+    rename(FireYear = FIRE_YEAR, FireName = INCIDENT, FireType = FEATURE_CA ) 
+  
+  AllFires %>%
+  mutate(FireType = case_when(
+    str_sub(FireType, 1,4)=='Wild' ~ 'Wildfire', 
+    TRUE ~ 'RxFire' )) %>%
+    filter(FireType == 'Wildfire', 
+           between(as.numeric(FireYear), 2017, 2024), 
+           STATE_NAME != "Wyoming") %>%
+    # st_write('./data/MapStuff.gpkg', 'NGP_fires')
+    as_tibble() %>%
+    group_by(STATE_NAME,  NA_L3NAME) %>%
+    summarize(fires = n(), 
+              .groups = 'drop') %>%
+    pivot_wider(names_from =  NA_L3NAME, 
+                values_from = fires) %>%
[…]

### 2025-09-02 NorthDakotaBurnSeverity/script/InitialDataWrangling.R

@@ -1,0 +1,159 @@
+pacman::p_load(tidyverse, lubridate, magrittr, readxl,  sf)
+
+# Combine burn dates with non-imputed fire behavior data 
+  # CGREC
+    BurnData_CGREC  <- 
+    left_join(by = 'FireCode', 
+      # Medium-rare version of the fire behavior data with burn dates
+        read_csv("S:/DevanMcG/Projects/SpatialFireBehavior/data/fromMZ/CompiledData2.csv") %>%
+        filter(location == "CGREC") %>%
+        mutate(date = as.Date(date, format = "%m/%d/%Y"),
+               L = str_remove(location, "REC"), 
+               B = str_sub(block, 1,3), 
+               Ps = str_replace(pasture, "[.]", ""), 
+               Ps = str_sub(Ps, 1,2), 
+               patch = str_replace(patch, "[.]", ""),
+               sp = str_sub(patch, 3,3),
+               sp = ifelse(sp == '', NA, sp), 
+               patch = str_sub(patch, 1,2),
+               y = format(date, "%y")) %>% 
+        unite("FireCode", c(L,B,Ps,patch,sp,y), sep=".") %>%
+        mutate(FireCode = str_to_upper(FireCode) ) %>%
+        select(FireCode, date) %>%
+        group_by(FireCode) %>%
+        slice(1) %>%
+        ungroup() %>%
+        arrange(date) %>%
+        mutate(burn = paste0('CGREC_', row_number()) ) , 
+      # Medium-done fire behavior data prior to imputation
+        read_csv("S:/DevanMcG/Projects/SpatialFireBehavior/data/AnalysisDataKY.csv") %>%
+        filter(location == "CG") %>%
+        mutate(patch = str_replace(patch, "[.]", ""),
+               sp = str_sub(patch, 3,3),
+               sp = ifelse(sp == '', NA, sp),
+               patch = str_sub(patch, 1,2)) %>%
+        unite("FireCode", c(location,block,pasture,patch,sp,year), sep=".") %>%
+        select(FireCode, plot, array, MaxC, SoilMaxC, ros) ) %>%
+          group_by(FireCode, burn, date, plot) %>%
+          summarise(across(c(MaxC:ros), ~mean(.x, na.rm = TRUE)), .groups = 'drop' )
+    
+    cgrec_bd <- 
+      BurnData_CGREC %>%
[…]

### 2025-09-02 NorthDakotaBurnSeverity/script/LinearRegressions.R

@@ -1,0 +1,210 @@
+pacman::p_load(tidyverse)
+
+
+load('./data/PlotBurnIndices.Rdata') # plot-level data for 26 Rx burns
+
+length(unique(filter(PlotBurnIndices, ! is.na(ros))$burn) ) 
+  
+
+BurnModels <- lst() 
+
+# Plot-level data 
+
+# Testing drivers of burn severity
+  # Canopy temperature
+    mc_0 <- lme4::lmer(dNBR ~ 1 + (1|location/burn), REML = FALSE, data = PlotBurnIndices)
+    mc_1 <- lme4::lmer(dNBR ~ MaxC + (1|location/burn), REML = FALSE, data = PlotBurnIndices)
+  
+    BurnModels$Sev_v_MaxC <- anova(mc_0, mc_1)
+
+  # Soil surface temperature
+    sc_0 <- lme4::lmer(dNBR ~ 1 + (1|location/burn), REML = FALSE, 
+                       data = filter(PlotBurnIndices, !is.na(SoilMaxC) ) )
+    sc_1 <- lme4::lmer(dNBR ~ SoilMaxC + (1|location/burn), REML = FALSE, 
+                       data = filter(PlotBurnIndices, !is.na(SoilMaxC)) )
+    
+    BurnModels$Sev_v_SoilMaxC <- anova(sc_0, sc_1)
+
+  # Rate of spread
+    rs_0 <- lme4::lmer(dNBR ~ 1 + (1|location/burn), REML = FALSE,  
+                        data = filter(PlotBurnIndices, ! is.na(ros)) )
+    rs_1 <- lme4::lmer(dNBR ~ ros + (1|location/burn), REML = FALSE,  
+                        data = filter(PlotBurnIndices, ! is.na(ros)) )
+    
+    BurnModels$Sev_v_ROS <- anova(rs_0, rs_1)
+    
+# Testing responses to fuelbed greenness
+    # Canopy temperature
+      mc_0 <- lme4::lmer(MaxC ~ 1 + (1|location/burn), REML = FALSE, data = PlotBurnIndices)
+      mc_1 <- lme4::lmer(MaxC ~ ndvi + (1|location/burn), REML = FALSE, data = PlotBurnIndices)
+      
+      BurnModels$MaxC_v_NDVI <- anova(mc_0, mc_1)
+    
+    # Soil surface temperature
+      sc_0 <- lme4::glmer(SoilMaxC ~ 1 + (1|location/burn), 
[…]

### 2025-09-02 NorthDakotaBurnSeverity/script/PaperOddsAndEnds.R

@@ -1,0 +1,241 @@
+pacman::p_load(tidyverse, ggpattern, magrittr, sf, stars, wesanderson)
+load('../albersEAC.Rdata')
+
+# Create objects for data sharing 
+  # Space-based vs Rx comparison
+  load('./data/PlotBurnIndices.Rdata')
+  PlotBurnIndices %>%
+    select(location, burn, plot:dNBR)  %>%
+    mutate(across(c(MaxC, SoilMaxC), ~ round(., 1)),
+           ros = round(ros, 2),
+           dNBR = round(dNBR, 3)) %>%
+    write_csv('./data/PlotBurnIndices.csv')
+
+  # Wildfire vs Rx comparison
+    load('./data/BurnSeverityData.Rdata')
+    BurnSeverityData %>%
+      select(type:dNBR_Mean) %>%
+      mutate(dNBR_Mean = round(dNBR_Mean, 3)) %>%
+      write_csv('./data/BurnSeverityData.csv')
+
+  # spatial data 
+    read_sf('./data/SeverityComparison.gpkg', 'StudyRegion') %>%
+      st_write('./data/SpatialData.gpkg', 'StudyRegion')
+    read_sf('./data/SeverityComparison.gpkg', 'GreatPlainsModifiedWF') %>%
+      select(-STUSPS, -FireType, -NAME) %>%
+      st_write('./data/SpatialData.gpkg', 'GreatPlainsModifiedWF', append = F)
+    
+    st_layers('./data/SpatialData.gpkg')
+    
+    read_sf('./data/SeverityComparison.gpkg', 'StudyRegion') %>%
+      st_union() %>%
+      st_as_sf() %>%
+      st_simplify(dTolerance = 1000) %>% 
+      geojsonsf::sf_geojson( simplify = TRUE,
+                             digits = 3) %>%
+      write.table("clipboard", sep="\t", row.names=FALSE, col.names=FALSE)
+# Get burn dates 
+# Regional comparison wildfires 
+
+st_layers('./data/SeverityComparison.gpkg')
+
+wf_sf <- read_sf('./data/SeverityComparison.gpkg', 'GreatPlainsModifiedWF') %>%
+  mutate(zone = case_when(
+    FireCode %in% c('Swather_2021','Hwy31101STWest_2022', 'CB00121_2021', 
+                    'CoalSeamWest_2021', '1806HunkpapaCreek_2021') ~
+      'east', 
[…]

### 2025-09-02 NorthDakotaBurnSeverity/script/PlottingBurnIndices.R

@@ -1,0 +1,192 @@
+pacman::p_load(tidyverse)
+
+
+load('./data/BurnIndices.Rdata')
+
+# Burn severity as the response 
+  BurnIndices %>%
+    mutate(ros = ifelse(ros >= 13, NA, ros)) %>%
+    pivot_longer(names_to = 'behavior', 
+                 values_to = 'value', 
+                 cols = c(MaxC:ros)) %>%
+    ggplot(aes(x = value, y = dNBR)) + theme_bw() +
+    geom_smooth(data = . %>%
+                  filter(behavior != 'SoilMaxC'), 
+                method = 'lm', color = 'black') +
+    geom_smooth(data = . %>%
+                  filter(behavior == 'SoilMaxC'), 
+                method = 'lm', lty = 2,
+                se = F, color = 'black') +
+    geom_point() +
+    facet_wrap(~behavior, scales = 'free_x')
+  
+# NDVI as the predictor 
+  BurnIndices %>%
+    mutate(ros = ifelse(ros >= 13, NA, ros)) %>%
+    pivot_longer(names_to = 'response', 
+                 values_to = 'value', 
+                 cols = c(MaxC:dNBR)) %>%
+    ggplot(aes(x = ndvi, y = value, color = location)) + theme_bw() +
+    geom_smooth(data = . %>%
+                  filter(response != 'SoilMaxC'), 
+                method = 'lm', color = 'black') +
+   #  geom_smooth(method = 'lm', se = FALSE) +
+    geom_point() +
+    facet_wrap(~response, scales = 'free_y')
+
+# Histograms 
+
+  # Maximum canopy temp
+    MASS::fitdistr(BurnIndices$MaxC, 'normal')
+    MASS::fitdistr(BurnIndices$MaxC, 'Gamma')
+    
+    BurnIndices %>%
+    ggplot(aes(x=MaxC)) + theme_bw(16) + 
+      geom_histogram(aes(y=after_stat(density)),      
+                     binwidth=20,
+                     colour="black",
+                     fill="lightgreen") +
+      coord_cartesian(xlim = c(0,500)) + 
+      geom_density(alpha=.5, fill="lightblue")  +
+      stat_function(data= BurnIndices , 
+                    fun = dnorm, 
+                    args=list(mean = 261.1,
+                              sd = 103.9),
+                    colour="blue", 
+                    size=1.1)  +
+      stat_function(data= BurnIndices , 
+                    fun = dgamma, 
[…]

### 2025-09-02 NorthDakotaBurnSeverity/script/SeverityCalculations.R

@@ -1,0 +1,80 @@
+pacman::p_load(tidyverse, sf, wesanderson) 
+
+# Sentinel imagery
+  sev_dir = 'S:/DevanMcG/FireScience/Sentinel/SeverityComparison'
+  images = list.files(sev_dir, include.dirs = FALSE)
+  
+  ImgTab <- 
+    images %>%
+      as_tibble() %>%
+      separate(value, into = c('incident', 'period', 'date', 'script'), sep = '_') %>%
+      rowid_to_column('image')
+# Fire perimeters and sample points 
+  perims <- 'C:/Users/devan.mcgranahan/USDA/Rangeland responses to fire - Fire x herbivory/Lemonade Fire/talks/data' %>%
+    paste0(., '/SeverityComparison.gpkg') %>%
+    read_sf('perimeters')
+  points <- 'C:/Users/devan.mcgranahan/USDA/Rangeland responses to fire - Fire x herbivory/Lemonade Fire/talks/data' %>%
+    paste0(., '/SeverityComparison.gpkg') %>%
+    read_sf('SamplePoints')
+
+SeveritySamples <- tibble() 
+
+for(i in 1:length(unique(ImgTab$incident))){
+  inc = unique(ImgTab$incident)[i]
+  # Get pre-burn image
+    b_num <- ImgTab %>% filter(incident == inc, period == 'B') %>% select(image)
+    b_fn = images[as.vector(b_num$image)]
+    b_ras <- paste0(sev_dir, '/', b_fn) %>% terra::rast() 
+    names(b_ras) <- c('nbr', 'ndvi') 
+    b_fl = (b_ras-10000)/10000
+  # Get post-burn image
+    a_num <- ImgTab %>% filter(incident == inc, period == 'A') %>% select(image)
+    a_fn = images[as.vector(a_num$image)]
+    a_ras <- paste0(sev_dir, '/', a_fn) %>% terra::rast() 
+    names(a_ras) <- c('nbr', 'ndvi') 
+    a_ras <- a_ras[['nbr']]
+    a_fl = (a_ras-10000)/10000
+  # differencing NBR
+    dNBR = b_fl[['nbr']] - a_fl
+    names(dNBR) <- 'dNBR'
+    #terra::plot(dNBR)
+  # Subsetting vectors 
+    perim <- filter(perims, incident == inc)
+    pts <- st_intersection(perim, points)
+  # Sample rasters & export data as tibble
+    full_join(by = c('incident','FireYear','FireType','id'), 
+      # get dNBR for sample points
[…]

### 2025-09-02 NorthDakotaBurnSeverity/script/SeverityComparisons.R

@@ -1,0 +1,116 @@
+FFgrad = wes_palette("FantasticFox1")[c(3,2,4,5)]
+
+BurnSeverityData <- tibble() 
+# Load data 
+  # HREC & Dayton
+   load('./data/RxBurnIndices.Rdata') # 22 Rx burns in sw ND
+ # CGREC data from previous paper
+   # loading this previously conflicted with the plot-level BurnIndices but it has been renamed
+   load("C:/Users/devan.mcgranahan/GitHubProjects/GreatPlainsFire/SpringSummerSeverity/data/BurnIndices.Rdata")
+# Wildfires 
+   load('./data/WfBurnIndices.Rdata') # wildfire burn indices
+   wf_sf <- read_sf('./data/SeverityComparison.gpkg', 'GreatPlainsModifiedWF') # Perimeter data
+
+BurnSeverityData <- 
+  bind_rows(
+    # CGREC
+      BurnIndices  %>%
+        filter(Season == 'Spring') %>%
+        rename(FireCode = fire) %>%
+        group_by(FireCode) %>%
+        summarise_at(.vars = vars(dNBR, ndvi),
+                     .funs = c(Mean="mean", 
+                               SEM = function(x) {sd(x)/sqrt(n()) } )) %>%
+        ungroup() %>%
+        mutate(type = 'Rx', 
+               zone = 'east') %>%
+        rowid_to_column() %>%
+        mutate(burn = paste0('CGREC_', rowid)) %>%
+        select(type, zone, FireCode:ndvi_SEM),  
+    # HREC & Dayton
+      RxBurnIndices  %>%
+      rename(FireCode = burn) %>%
+        filter(complete.cases(.), 
+               dNBR > 0) %>%
+        group_by(location, FireCode) %>%
+        summarise_at(.vars = vars(dNBR, ndvi),
+                     .funs = c(Mean="mean", 
+                               SEM = function(x) {sd(x)/sqrt(n()) } )) %>%
+        ungroup() %>%
+        mutate(type = 'Rx', 
+               zone = 'west') %>%
+         select(type, zone, FireCode:ndvi_SEM) , 
+    # Wildfires
+      wf_sf %>%
+        as_tibble() %>%
+        select(FireCode, L3) %>%
+        left_join(by = 'FireCode', 
+          WfBurnIndices %>%
+            group_by(FireCode) %>%
+            summarize(across(c(dNBR, ndvi), 
+                             list('Mean' = mean, 
[…]

### 2025-09-02 NorthDakotaBurnSeverity/script/SpatialWrangling.R

@@ -1,0 +1,385 @@
+pacman::p_load(tidyverse, sf)
+
+load('../albersEAC.Rdata')
+
+# Rx burn locations for mapping 
+  gpkg = './data/OriginalRxBurnPerims.gpkg'
+
+  purrr::map(1:length(st_layers(gpkg)$name),
+             function(i) {
+               lyr = st_layers(gpkg)$name[i]
+               read_sf(gpkg, lyr) %>%
+                 st_transform(albersEAC) %>%
+                 st_union() %>%
+                 st_centroid() %>%
+                 st_as_sf() %>%
+                 mutate(location = lyr) 
+             } ) %>%
+    do.call(rbind, .) %>%
+    st_write(gpkg, 'RxLocations')
+    
+ # put some map stuff together
+  ngp <- read_sf('../data/SeverityComparison.gpkg', 'StudyRegion')
+  ngp %>% 
+    st_write('./data/MapStuff.gpkg', 'ngp')
+    
+  ngp_counties <- tigris::counties(state = c('MT', "ND", 'SD'), cb = T)  
+  
+  ngp_counties %>%
+    st_transform(st_crs(ngp)) %>% 
+    st_intersection(ngp %>% 
+                      st_union()) %>% 
+    st_write('./data/MapStuff.gpkg', 'ngp_counties')
+  load("C:/Users/devan.mcgranahan/USDA/Rangeland responses to fire - Fire x herbivory/Lemonade Fire/talks/data/gp.Rdata")
+   load("C:/Users/devan.mcgranahan/USDA/Rangeland responses to fire - Fire x herbivory/Lemonade Fire/talks/data/usa.Rdata")
+
+   st_write(usa, './data/MapStuff.gpkg', 'usa')
+   st_write(gp, './data/MapStuff.gpkg', 'gp')
+# Fetching original thermocouple sampling locations
+  # Hettinger
+    h_dr = 'D:/NDSU/GoogleDrive/QGIS/HREC/sampling/FireBehavior'
+    read_sf(h_dr, 'FireTowerPlacements') %>%
+      st_transform(albersEAC) %>%
+      select(PastureLoc, PastureRep, Livestock, Patch, SubPlot, FireTower) %>%
+      st_write('./data/ThermocouplePlacements.gpkg', 'HREC', append = FALSE)
+  # Dayton Tract
[…]

### 2025-09-02 NorthDakotaBurnSeverity/script/mapping.R

@@ -1,0 +1,164 @@
+pacman::p_load(tidyverse, sf)
+
+# Data loading 
+load('../albersEAC.Rdata')
+rx_sf <- read_sf( './data/OriginalRxBurnPerims.gpkg', 'RxLocations')
+wf_sf <- read_sf('./data/SeverityComparison.gpkg', 'GreatPlainsModifiedWF')
+ngp <- read_sf('./data/MapStuff.gpkg', 'ngp')
+ngp_counties <- read_sf('./data/MapStuff.gpkg', 'ngp_counties')
+usa <- read_sf('./data/MapStuff.gpkg', 'usa')
+gp <- read_sf('./data/MapStuff.gpkg', 'gp')
+NGP <- read_sf('./data/MapStuff.gpkg', 'NGP_states') # L3 for MT, ND, SD
+NGP_fires <- read_sf('./data/MapStuff.gpkg', 'NGP_fires')
+load('./data/ppt_dat.Rdata')
+
+# Region map 
+reg_gg <-
+ppt_dat$ngp_ppt %>%
+  mutate(NA_L3NAME = str_replace(NA_L3NAME, ' ', '\n')) %>%
+ggplot() + theme_void() +
+  geom_sf(aes(fill = NA_L3NAME, alpha = ppt), 
+          color = 'NA') +
+  geom_sf(data = NGP, 
+          fill = NA) + 
+  geom_sf_label(data = locations %>% 
+                  filter(type != 'Wildfire'), 
+                label = expression(R[X]), 
+                label.padding = unit(0.1, "lines"),
+                label.r = unit(0.5, "lines"),
+                label.size = 0.15,
+                size = 4) +
+  geom_sf(data = NGP_fires, 
+          pch = 1, alpha = 0.5) + 
+  geom_sf_label(data = NGP %>%
+                  group_by(STATE_NAME) %>%
+                  summarize(area = sum(Shape_Area)) , 
+                aes(label = state.abb[match(STATE_NAME,state.name)])) + 
+  scale_fill_manual('EPA Level III\necoregion', 
+                    values = c("#74A089","#E1BD6D")) +
+  scale_alpha_continuous("Precipitation\nmean annual (mm)", 
+                         range = c(0.1, 1)) + 
+  theme(legend.position = 'right', 
+        legend.key.height = unit(1, 'cm')) + 
+  guides(alpha = guide_legend(order = 2), fill = guide_legend(order = 1) ) 
[…]
